### 2016-07-11 manuscript/greenwell (Kristi-HP's conflicted copy 2016-07-06).tex
added manuscript

\title{partial: An R Package for Creating Partial Dependence Plots}
\author{by Author One, Author Two and Author Three}

\maketitle

\abstract{
An abstract of less than 150 words.
}

\section{Outline}

\begin{itemize}
  \item Other packages of interest: 
  \begin{description}
    \item[\CRANpkg{plotmo}] allows for up to two variables and fixes other variables at there median values (or the first level for factors). Less accurate than partial dependence, but flexible and fast! For additive models (i.e., no interactions), these plots are identical in shape to partial dependence plots.
    \item[\CRANpkg{ICEbox}] can also calculate derivatives, but only allows for one variable at a time (i.e., no multivariate displays) -- however, color can be used to diplay information about an additional predictor.
    \item[\CRANpkg{car}] contains many functions for constructing partial-residual and marginal-model plots. The \CRANpkg{effects} is also of interest. However, these packages are orientated towards parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{partial} are mainly for nonparametric and black box models (though, they can be used for simple parametric models as well).
  \end{description}
\end{itemize}

\section{Introduction}

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" variables is identified it is often necessary to assess the relationship between them and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs). PDPs help visualize the relationship between a subset of the predictors and the response while accounting for the average effect of the other independent variables. They are particularly effective with black box models like support vector machines or random forest.

%computing and plotting \dfn{partial dependence functions}. Partial dependence plots help visualize the relationship between a subset of the predictors and the response while accounting for the average effect of the other independent variables. They are particularly effective with \dfn{black box} models (e.g., a support vector machine or random forest).

Many techniques exist for visualizing the relationship between a predictor and the response based on a fitted model. For example, the output from complex parametric models are typically augmented with partial residual plots (see J. FOX for details). Most existing techniques arose in applied regression and typically only apply to parametric models. \dfn{Partial dependence functions} \citep{friedman-2000-greedy}, on the other hand, offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.

Targeted implementations of Friedman's partial dependence plots are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others. While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space. Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., neural networks). However, to our knowledge, there is no general package for obtaining partial dependence plots for other types of model fits (e.g., a \dfn{conditional inference forest} as implemented by the \code{cforest} function in the \CRANpkg{party} package). The \CRANpkg{partial} package tries to close this gap by offering a general framework that can be applied to several types of fitted models -- though, package \CRANpkg{plotmo} has come a long way in this direction..

Partial dependence plots are very useful but can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}. To overcome this issue \citet{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots -- available in the \CRANpkg{ICEbox} package.

There are two main functions available in the \pkg{partial} package (PKG REF):
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predctor vaulues. By default, \code{partial} returns a data frame with an dditional class: \code{"partial"}. Otherwise a \code{"lattice"} object (i.e., a plot) is returned. For more advanced plotting, the \code{plotPartial} functions will take a \code{"partial"} object and display a customizable lattice plot.

There are two ways to install the \pkg{partial} package. Obtaining the latest stable release from CRAN:
\begin{example}
install.packages("partial")
\end{example}
or installing the development version from GitHub:
\begin{example}
devtools::install_github("bgreenwell/partial")
\end{example}

\section{Partial dependence plots}

Let $\boldsymbol{x}$ represent the set of predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$ (these are the values returned by \code{predict} in R). If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_{c} = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
\begin{equation}
\label{eqn:pdf}
\bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
\end{equation}
where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample. This can be quite computationally intensive since Equation~\eqref{eqn:pdf} involves a pass over the training records for each set of combinations of $\boldsymbol{z}_s$. Fortunately, this problem is \dfn{embarrassingly parallel}.

The basic algorithm for computing a partial dependence function is as follows:
\begin{enumerate}
  \item Fit a model
  \item Suppose $x_i$ is the predictor set of interest. For each unique value of $x_i$, do the following:
  \begin{enumerate}
    \item Make a copy of the original training data and replace $x_i$ with the unique value -- everything else remains th same.
    \item Compute the predicted values using the modified copy of the training data.
    \item Average the predictions over each of the observations.
  \end{enumerate}
  \item Plot the average prediction against $x_i$.
  \item Repeat for each predictor of interest.
\end{enumerate}

\section{Partial dependence plots in R}

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # load the data
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}

Next, we fit a traditional random forest (as implemented in the \CRANpkg{randomForest} package) to the entire data set. 
\begin{example}
# Load required packages
library(partial)
library(randomForest)

# Fit a random forest
set.seed(101)  # for reproducibility
fit.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
print(fit.rf)  # check model results
varImpPlot(fit.rf, main = "")  # variable importance plot

# Call:
#   randomForest(formula = cmedv ~ ., data = boston, importance = TRUE) 
# Type of random forest: regression
# Number of trees: 500
# No. of variables tried at each split: 5
# 
# Mean of squared residuals: 9.114248
# % Var explained: 89.17
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes. The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, individually and together.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{boston_rf_vimp}
  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
  \label{fig:plotmo_vs_partial}
\end{figure}

\subsection{Univariate partial dependence}

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for displaying partial dependence of the response on a single predictor. For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(fit.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot (implemented in \pkg{lattice}) can be achieved using the \code{partial} function and setting \code{plot = TRUE} (the default is \code{FALSE}):
\begin{example}
partial(fit.rf, pred.var = "lstat", plot = TRUE)
\end{example}
\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat}.}
  \label{fig:pd_lstat}
\end{figure}

By default, \code{partial} returns a data frame:
\begin{example}
head(partial(fit.rf, pred.var = "lstat"))

#    lstat        y
# 1 1.7300 31.34696
# 2 2.4548 31.34564
# 3 3.1796 31.28286
# 4 3.9044 30.82356
# 5 4.6292 29.12636
# 6 5.3540 26.89093
\end{example}

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for other types of model fits (e.g., a \dfn{conditional inference forest} as implemented in the \CRANpkg{party} package), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package (more on this later). For example, to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} we could use
\begin{example}
partial(fit.rf, pred.var = c("lstat", "rm"), plot = TRUE)
\end{example}

Finer plotting control can be achieved using the accompanying \code{plotPartial} function. \code{plotPartial} takes a \code{"partial"} object and provides a convenient \CRANpkg{lattice} display that can be easily tailored. For example, the code chunk
\begin{example}
pd.lstat.rm <- partial(fit.rf, pred.var = c("lstat", "rm"))
p1 <- plotPartial(pd.lstat.rm)
p2 <- plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE, 
                  colorkey = TRUE, screen = list(z = -20, x = -60))
print(p1, position = c(0, 0, 0.5, 1), more = TRUE)
print(p2, position = c(0.5, 0, 1, 1))
\end{example}
produces the three-dimensional displays in Figure~\ref{fig:pd_lstat_rm}.
\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm}.}
  \label{fig:pd_lstat_rm}
\end{figure}

Unfortunately, the \code{cforest} implementation contains no such function. In this case we can use \code{partial}:
\begin{example}
crf.lstat <- partial_1d(fit.crf, x.name = "lstat", n = 51)
crf.rm <- partial_1d(fit.crf, x.name = "rm", n = 51)
par(mfrow = c(1, 2))
plot(crf.lstat, type = "l")
plot(crf.rm, type = "l")
\end{example}

In some cases it is difficult for \code{partial} to tell whether the response is continuous (regression) or categorical (classification). For these cases, we added the option \code{super.type} which accepts a character string specifying either \code{"regression"} or \code{"classification"} -- this distinction is important because for classification problems, PDPs use something similar to a \dfn{logit} scale for response estimation.

By default \code{partial} returns a data frame. If called with \code{plot = TRUE}, however, a lattice plot is displayed instead
\begin{example}
partial_1d(fit.crf, x.name = "lstat", n = 51, plot = TRUE)
\end{example}

\subsection{Multivariate partial dependence}

Bivariate (and higher order) partial dependence can also be handled

\begin{example}
# Bivariate partial dependence
rf.rm.lstat <- partial_2d(fit.rf, x1.name = "rm", x2.name = "lstat", 
                          n1 = 51, n2 = 51, .progress = "text")
wireframe(y ~ rm * lstat, data = rf.rm.lstat, zlab = "cmedv", drape = TRUE, 
          colorkey = TRUE, screen = list(z = 110, x = -60))
\end{example}

The variables are plotted in the order they are given; that is, the first variable is plotted along the $x$-axis, the second on the $y$-axis, and the rest becom panel variables. Hence, it is best to use numeric variables for the first two, and factors for the remaining. The following snippet of code plots the partial dependence of \code{cmedv} on \code{rm}, \code{lstat}, and \code{chas}, but since \code{chas} is a binary variable (whether or not the census tract bounds river), a separate panel is displayed for each category.
\begin{example}
# Trivariate partial dependence
rf.rm.lstat.ptratio <- partial_3d(fit.rf, x1.name = "rm", x2.name = "lstat", 
                                  x3.name = "chas", n1 = 51, n2 = 51, n2 = 51, 
                                  .progress = "text")
wireframe(y ~ rm * lstat, data = rf.rm.lstat, zlab = "cmedv", drape = TRUE, 
          colorkey = TRUE, screen = list(z = 110, x = -60))
\end{example}

\subsection{The Boston Housing Data}

In \pkg{plotmo} the other variables are held fixed at their median values. This results in a less accurate, but much faster display. To illustrate we fit a multivariate adaptive regression spline (MARS) model to the Boston housing data. MARS models can be fit using the \code{mars} function from the \CRANpkg{mda} package, or using the \code{earth} function from the \CRANpkg{earth} package. Here we use the \pkg{earth} implementation which conveniently loads the \pkg{plotmo} package as well.
\begin{example}
# Fit a MARS model
fit.earth <- earth(cmedv ~ ., data = boston, degree = 3)

# Partial dependence function
earth.rm <- partial_1d(fit.earth, x.name = "rm")

# Compare partial and plotmo
pdf("plotmo_vs_partial.pdf", width = 7, height = 5)
plotmo(fit.earth, degree1 = "rm", degree2 = FALSE, do.par = FALSE, 
       ylim = c(16.79417, 41.87501), main = "", xlab = "rm", ylab = "cmedv")
lines(earth.rm, lty = 2)
legend("topleft", legend = c("plotmo", "partial"),
       lty = 1:2, inset = 0.01)
dev.off()
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{plotmo_vs_partial}
  \caption{The logo of R.}
  \label{figure:plotmo_vs_partial}
\end{figure}

\begin{example}
partial(fit.earth, pred.var = c("rm", "ptratio"), plot = TRUE)
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{earth_rm_ptratio}
  \caption{Partial dependence of \code{cmedv} on \code{rm} and \code{ptratio}.}
  \label{figure:earth_rm_ptratio}
\end{figure}

\code{partial} returns a\code{"data.frame"} object with an additioanl class called \code{"partial"}; the additional class is recognized by \code{plotPartial} for convenient default plotting. To create more elaborate plots, just work with the data frame directly. Also, the convenient \code{plot} option in \code{partial} returns a basic plot, but is not as customizable as uning plotPartial. The following snipet of code demonstrates:
\begin{example}
# Basic partial dependence plot
partial(fit.rf, "rm", plot = TRUE)

# Save the partial dependence data for customized plotting
pd.rm <- partial(fit.rf, "rm")
partialPlot(pd.rm, xlab = "Average number of rooms", ylab = "Median home value")
plot(pd.rm, type = "l")  # base R graphics

# Try using ggplot2
library(ggplot2)
ggplot(pd.rm, aes(rm, y)) +
  geom_point() +
  geom_line() +
  theme_light()
\end{example}

\subsection{Avoiding extrapolation}

There are few ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls. Rug displays are one-dimensional plots added to the axes. The \code{plotPartial} has a \code{rug} option that will display the deciles of the distribution  (as well as the minimum and maximum values) for the predictors on the horizontal or vertical axes. The following snippet of code produces the plot on the left side of Figure~\ref{fig:partial_extrap}. Notice that in both examples we had to provide the original training data to \code{plotPartial} via the \code{training.data} option.
\begin{example}
plotPartial(pf.lstat, rug = TRUE, training.data = boston)
\end{example}

In higher dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{convex.hull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the variables used for the horizontal and vertical axes) is added to the plot. Over interpreting the partial dependence plot outside of this region can be dangerous. The right side of Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
plotPartial(pd.lstat.rm, convex.hull = TRUE, training.data = boston)
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_extrap}
  \caption{Examples of partial dependence plots with the addition of a rug display (left) and a convex hull (right).}
  \label{fig:partial_extrap}
\end{figure}

\subsection{Addressing computational concerns}

Additional options are available to ease the computational intensity for large problems. For example, there was no need to compute partial dependence of median home value using each uniqe value of \code{rm} in the training data (which would require 446 passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points is controlled by the option \code{n.pts}. Alternatively, a specific set of values can be supplied (e.g., quantiles of interest) through the \code{pred.data} argument. For example, the following snippet of code re-computes the partial dependence of median home value on \code{rm} using both options; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(fit.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(fit.rf, "rm", pred.grid = 3:9, plot = TRUE)
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_manual}
  \caption{Examples of partial dependence plots using a reduced set of predictor values.}
  \label{fig:partial_manual}
\end{figure}

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than for loops. This makes it poosible to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \CRANpkg{foreach} package. To use this functionality, we must load and register a supported parallel backend (e.g., \CRANpkg{doParallel} or \CRANpkg{doMC}). The following snippet of code obtains the partial dependence of median home value on \code{rm}, \code{lstat}, and \code{ptratio} in parallel. The result is displayed in Figure~\ref{fig:partial_3}.
\begin{example}
library(doParallel)
cl <- makeCluster(2)  # using snow-like functionality
registerDoParallel(cl)  # use cores = 2 for multicore functionality
pd3 <- partial(fit.rf, pred.var = c("lstat", "rm", "ptratio"),
               grid.resolution = 10, .parallel = TRUE)
stopCluster(cl)  # required when using snow-like functionality
plotPartial(pd3, number = 4, overlap = 0.1)
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_3}
  \caption{Examples of partial dependence plots using a reduced set of predictor values.}
  \label{fig:partial_3}
\end{figure}

\subsection{Classification problems}

For classification problems, partial dependence functions are on a scale simliar to the logit; in particular, an average logit. Suppose the response is categorical with $K$ levels, then for each class we have
\begin{equation}
\label{eqn:avg-logit}
f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{i = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
\end{equation}
where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th cass depends on different subsets of the predictor variables.

To illustrate, consider the Edgar Anderson's iris data from the \CRANpkg{datasets} package. We fit a support vector machine with a radial basis function kernel to the data (the parameters were determined using 5-fold cross-validation).
\begin{example}
library(kernlab)
iris.svm <- ksvm(Species ~ ., data = iris, kernel = "rbfdot", C = 0.25,
                 kpar = list(sigma = 0.75), prob.model = TRUE)
\end{example}

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"), 
                 which.class = i, training.data = iris)
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}
levelplot(y ~ Petal.Width * Petal.Length | Species, data = pd)
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris}
  \caption{Partial dependence of species on petal width and petal length for the iris data.}
  \label{fig:partial_iris}
\end{figure}

\section{Summary}

In the future, this package will be expanded to appropriately handle other types of responses as well (e.g., countd data).

\bibliography{Greenwell}

\address{Brandon Greenwell\\
  Infoscitex, a dcs company\\
  4027 Colonel Glenn Highway\\
  Suite 210\\
  Dayton, OH 45431-1672\\
  United States of America\\}
\email{bgreenwell@infoscitex.com}

### 2016-07-11 manuscript/greenwell.tex
added manuscript

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Front matter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{partial: An R Package for Creating Partial Dependence Plots}
\author{by Brandon M. Greenwell}

\maketitle

\abstract{
An abstract of less than 150 words.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Introduction
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Predictor importance is an important task in any supervised learning problem. However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them and the response. This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details. PDPs help visualize the relationship between a subset (typically 1-3) of the predictors and the response while accounting for the average effect of the other independent variables in the model. They are particularly effective with black box models like random forest or support vector machines.
%PDPs offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.

Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_n\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$. If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_{c} = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
\begin{equation}
\label{eqn:pdf}
\bar{f}_s\left(\boldsymbol{z}_s\right) = \frac{1}{n}\sum_{i = 1}^n\widehat{f}\left(\boldsymbol{z}_s,\boldsymbol{z}_{i, c}\right),
\end{equation}
where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample. This can be quite computationally intensive since Equation~\eqref{eqn:pdf} involves a pass over the training records for each set of combinations of $\boldsymbol{z}_s$. Fortunately, this problem is embarrassingly parallel. Let $x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$. The partial dependence of the response on $x_1$ can be constructed as follows:
\begin{enumerate}
  \item For $i \in \left\{1, 2, \dots, k\right\}$:
  \begin{enumerate}
    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
    \item Compute the vector of predicted values from the modified copy of the training data.
    \item Compute the average prediction to obtain $\bar{f}_1\left(x_{1i}\right)$.
  \end{enumerate}
  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
\end{enumerate}
This algorithm can be easily extended to larger subsets of 2 or more features as well.

Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others. While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space; however, these implementations only apply to models fit using the respective package. Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., neural networks). However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively). The \CRANpkg{partial} \citep{partial-pkg} package tries to close this gap by offering a general framework for PDPs that can be applied to several types of fitted models.

The \CRANpkg{plotmo} package \citep{plotmo-pkg} has come a long way in this regard. According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors constant (continuous features are fixed at their median value, while factors are held at their first level). These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct! For additive models (i.e., no interactions), these plots are identical in shape to partial dependence plots.

PDPs are very useful but can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}. To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots -- available in the \CRANpkg{ICEbox} package. \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though, color can be used effectively to display information about an additional predictor. The ability to construct derivative plots is also available.

Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model. For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing partial-residual and marginal-model plots. The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest. However, these methods are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{partial} are mainly for nonparametric and black box models (though, they can be used for simple parametric models as well).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PDPs in R
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Constructing PDPs in R}

As described in the introduction, the \pkg{partial} package is useful for constructing PDPs for many types of fitted models in R. These are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest. The latest stable release is available from CRAN:
\begin{example}
install.packages("partial")
\end{example}
%low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$

Currently, only two functions are exported by \pkg{partial}:
\begin{itemize}
  \item \code{partial}
  \item \code{plotPartial}
\end{itemize}
The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values. If \code{plot = FALSE} (the default), \code{partial} returns a data frame with an additional class: \code{"partial"}. The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values. If \code{plot = TRUE}, a \code{"lattice"} object representing the PDP is returned. For more advanced plotting, the \code{plotPartial} functions will take a \code{"partial"} object and display a more customizable lattice plot. Currently supported models are described in Table~\ref{tab:models}.
\begin{table}[htbp]
  \begin{tabular}{p{4cm}ll}  
    \toprule
      Model & Package & Class \\
      \midrule
      Bagged decision trees     & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
      Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                & \pkg{partykit} & \code{"cforest"} \\
      Decision trees            & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
                                & \pkg{party}    & \code{"BinaryTree"} \\
                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
      Linear model              & \pkg{stats}    & \code{"lm"} \\
      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
      Support vector machines & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                              & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
      \bottomrule
  \end{tabular}
  \caption{Models specifically supported by the \pkg{package}.}
  \label{tab:models}
\end{table}
The \code{partial.default} method should be able to handle most other types of models not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a traditional random forest to the entire data set using the following code snippet: 
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes. The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{boston_rf_vimp}
  \caption{Dotchart of variable importance scores for the Boston housing data based on a random forest with 500 trees.}
  \label{fig:plotmo_vs_partial}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Univariate PDPs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Single predictor PDPs}

As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for displaying partial dependence of the response on a single predictor. For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
\begin{example}
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
\end{example}
The same plot (implemented in \pkg{lattice}) can be achieved using the \code{partial} function and setting \code{plot = TRUE}:
\begin{example}
library(partial)
partial(boston.rf, pred.var = "lstat", plot = TRUE)
\end{example}
For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function. \strong{Note:} the \dfn{pipe} operator provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is imported for writing more convenient code, as illustrated in snippet of code below.
\begin{example}
partial(boston.rf, pred.var = "lstat") %>%
  plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))
\end{example}
The results are displayed in Figure~\ref{fig:pd_lstat} below.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat}. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
  \label{fig:pd_lstat}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Higher order PDPs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Multi-predictor PDPs}

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} based on a conditional random forest using various options in \code{plotPartial}. The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(party)
boston.crf <- cforest(cmedv ~ ., data = boston)
pd.lstat.rm <- partial(boston.rf, pred.var = c("lstat", "rm"))
plotPartial(pd.lstat.rm)
plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("white", "blue")))
plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE, 
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a different color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}

% \sub\subsection{Multiple linear regression}
% Here we use the multivariate adaptive regression splines (See ????) algorithm to perform multiple linear regression with automatc variable selection and interaction detection.
% \begin{example}
% library(earth)
% boston.mars <- earth(cmedv ~ ., data = boston, degree = 3, linpreds = TRUE,
%                      pmethod = "exhaustive")
% partial(boston.mars, pred.var = c("nox", "rm"), grid.resolution = 10)
% \end{example}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Avoiding extrapolation
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Avoiding extrapolation}

There are few ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls. Rug displays are one-dimensional plots added to the axes. \code{plotPartial} has a \code{rug} option that will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal or vertical axes. The following snippet of code produces the plot on the left side of Figure~\ref{fig:partial_extrap}. Notice that in both examples we had to provide the original training data to \code{plotPartial} via the \code{training.data} option.
\begin{example}
plotPartial(pf.lstat, rug = TRUE, training.data = boston)
\end{example}

In higher dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on. When \code{convex.hull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot. Over interpreting the partial dependence plot outside of this region can be dangerous. The right side of Figure~\ref{fig:partial_extrap} was produced using:
\begin{example}
plotPartial(pd.lstat.rm, convex.hull = TRUE, training.data = boston)
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_extrap}
  \caption{Examples of partial dependence plots with the addition of a rug display (left) and a convex hull (right).}
  \label{fig:partial_extrap}
\end{figure}

\subsection{Addressing computational concerns}

Constructing PDPs can be quite computationally expensive. Additional options are available to ease the computational intensity for large problems. For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require 446 passes over the data!). We could get very reasonable results using a reduced number of points. Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{n.pts}. Alternatively, a specific set of values (e.g., quantiles of interest) can be supplied through the \code{pred.grid} argument. To illustrate, the following snippet of code computes the partial dependence of median home value on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
\begin{example}
partial(boston.rf, plot = TRUE)
partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
\end{example}

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_manual}
  \caption{Partial dependence of \code{cmedv} on \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a reduced grid size. \textit{Right}: Using a user-specified grid.}
  \label{fig:partial_manual}
\end{widefigure}

The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than for loops. This makes it possible to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel. In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package. To use this functionality, we must load and register a supported parallel backend (e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}). The following snippet of code obtains the partial dependence of median home value on \code{rm}, \code{lstat}, and \code{ptratio} in parallel. The result is displayed in Figure~\ref{fig:partial_par}.

Setting up a parallel backend is rather straightforward. To illustrate, the following snippet of sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel}.
\begin{example}
library(doParallel)  # load parallel backend
registerDoParallel(cores = 4)  # use 4 cores
\end{example}
Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} the parallel backend we loaded:
\begin{example}
pd <- partial(boston.rf, pred.var = c("lstat", "rm", "ptratio"),
              grid.resolution = 20, .parallel = TRUE)
plotPartial(pd, number = 4, overlap = 0.1)
\end{example}

It is important to note that when using more than two variables, \code{plotPartial} produces a trellis display. The first two variables given to \code{pred.var} are used for the horizontal and vertical axes; additional variables define the panels. If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle. The intervals may be allowed to overlap.} are produced first using the equal count algorithm. Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.8\linewidth]{partial_par}
  \caption{Partial dependence of \code{cmedv} on \code{lstat}, \code{rm}, and \code{ptratio}. Since \code{ptratio} is contiuous, it is first converted to a shingle; in this case, four groups with with 10\% overlap.}
  \label{fig:partial_par}
\end{figure}

\subsection{Classification problems}

For classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp. 369--370]{hastie-elements-2009}. Suppose the response is categorical with $K$ levels, then for each class we compute
\begin{equation}
\label{eqn:avg-logit}
f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{i = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
\end{equation}
where $p_k(x)$ is the predicted probability for the $k$-th class. Plotting $f_k(x)$ helps us understand how the log-odds for the $k$-th class depends on different subsets of the predictor variables.

To illustrate, consider Edgar Anderson's iris data from the \pkg{datasets} package. We fit a support vector machine with a radial basis function kernel to the data usinv the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
\begin{example}
library(e1071)
iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75, 
                cost = 0.25, probability = TRUE)
\end{example}
The \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to invoke the \code{prob.model = TRUE} option in the call to \code{svm}.

Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
\begin{example}
pd <- NULL
for (i in 1:3) {
  tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"), 
                 which.class = i, training.data = iris)
  pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
}
lattice::levelplot(y ~ Petal.Width * Petal.Length | Species, data = pd)
\end{example}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{partial_iris_svm}
  \caption{Partial dependence of species on petal width and petal length for the iris data.}
  \label{fig:partial_iris}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Summary
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Summary}

In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{partial} package. We discussed similar approaches in other packages. Ways to avoid extrpolation and high excecution times were demonstrated via examples. There is a bright future for \pkg{partial} in terms of growth. For example, we would like to include the ability to construct PDPs for various types of survival models (e.g. conditional random forests with censored response).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Summary
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Acknowledgments}

TBD.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Back matter
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliography{Greenwell}

\address{Brandon M. Greenwell\\
  Infoscitex Corporation\\
  4027 Colonel Glenn Highway\\
  Suite 210\\
  Dayton, OH 45431-1672\\
  United States of America\\}
\email{bgreenwell@infoscitex.com}

### 2016-07-11 manuscript/greenwell (Kristi-HP's conflicted copy 2016-07-06).R

@@ -1,0 +1,218 @@
+################################################################################
+# R code for "partial: partial: An R Package for Creating Partial Dependence
+# Plots"
+################################################################################
+
+
+################################################################################
+# Setup
+################################################################################
+
+# Set working directory
+setwd("/home/w108bmg/Desktop/Dropbox/Projects/greenwell_partial_2016")
+
+
+################################################################################
+# Code for paper
+################################################################################
+
+# Install the latest stable version from CRAN
+# install.packages("partial")
+
+# Install development version from GitHub
+# devtools::install_github("bgreenwell/partial")
+
+# Load the data
+data(BostonHousing2, package = "mlbench")  # load the data
+boston <- BostonHousing2[, -c(1, 2, 5)]
+
+# Load required packages
+library(partial)
+library(randomForest)
+
+# Fit a random forest
+set.seed(101)  # for reproducibility
+fit.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
+print(fit.rf)  # check model results
+pdf("boston_rf_vimp.pdf", width = 7, height = 5)
+varImpPlot(fit.rf, main = "")  # dotchart of variable importance scores
+dev.off()
+
+# Call:
+#   randomForest(formula = cmedv ~ ., data = boston, importance = TRUE)
+# Type of random forest: regression
+# Number of trees: 500
+# No. of variables tried at each split: 5
+#
+# Mean of squared residuals: 9.114248
+# % Var explained: 89.17
+
+pdf("pd_lstat.pdf", width = 7, height = 5)
+partial(fit.rf, pred.var = "lstat", plot = TRUE)
+dev.off()
+
+# Bivariate partial dependence
+pd.lstat.rm <- partial(fit.rf, pred.var = c("lstat", "rm"))
+pdf("pd_lstat_rm.pdf", width = 12, height = 6)
+p1 <- plotPartial(pd.lstat.rm)
+p2 <- plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
+                  colorkey = TRUE, screen = list(z = -20, x = -60))
+print(p1, position = c(0, 0, 0.5, 1), more = TRUE)
+print(p2, position = c(0.5, 0, 1, 1))
+dev.off()
+
[…]

### 2016-07-11 manuscript/greenwell.R

@@ -1,0 +1,141 @@
+# R code for "partial: partial: An R Package for Creating Partial Dependence
+# Plots"
+#
+#        Author: Brandon M. Greenwell
+# Date Modified: 06Jul2016 
+
+
+################################################################################
+# Setup
+################################################################################
+
+# Set working directory
+# setwd("C:\\Users\\Brandon.Greenwell\\Desktop\\greenwell_partial_2016")
+setwd("/home/w108bmg/Desktop/Dropbox/Projects/greenwell_partial_2016")
+
+# Install required packages
+# install.packages(c("doParallel", "e1071", "mlbench", "partial", "party", "randomForest"))
+
+# Load required packages
+# library(caret)
+library(doParallel)
+library(e1071)
+# library(kernlab)
+library(partial)
+library(party)
+library(randomForest)
+
+
+################################################################################
+# Boston housing example
+################################################################################
+
+# Load the (corrected) Boston housing data
+data(BostonHousing2, package = "mlbench")
+boston <- BostonHousing2[, -c(1, 2, 5)]  # remove unused columns
+
+# Fit a random forest and a conditional random forest using default settings
+set.seed(101)  # for reproducibility
+boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
+print(boston.rf)  # check model results
+# Call:
+#   randomForest(formula = cmedv ~ ., data = boston, importance = TRUE)
+# Type of random forest: regression
+# Number of trees: 500
+# No. of variables tried at each split: 5
+#
+# Mean of squared residuals: 9.114248
+# % Var explained: 89.17
+
+# Figure 1
+pdf("boston_rf_vimp.pdf", width = 7, height = 5)
+varImpPlot(boston.rf)  # dotchart of variable importance scores
+dev.off()
+
+# Figure 2
+pd.lstat <- partial(boston.rf, pred.var = "lstat")
+pdf("pd_lstat.pdf", width = 8, height = 4)
+pdp1 <- plotPartial(pd.lstat)
+pdp2 <- plotPartial(pd.lstat, lwd = 2, smooth = TRUE, span = 0.5, 
+                    ylab = expression(f(lstat)))
[…]

### 2016-07-12 manuscript/greenwell.tex
first draft -- rough

@@
 \abstract{
-An abstract of less than 150 words.
+Complex predictive models (e.g., neural networks, random forests, and support vector machines) offer much promise in the world of data analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g.
+multiple linear regression).
+Unfortunately though, it can be challenging for the uninitiated to understand the results of such models and explain them to management.
+Partial dependence plots offer a simple solution.
+Partial dependence plots offer low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood.
+These plots are especially useful in explaining the output from complex black box models like support vector machines or neural networks or constructing simpler parsimonious models on newly obtained data.
 }
@@
 \end{example}
-%low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$
 

### 2016-07-12 manuscript/greenwell (Kristi-HP's conflicted copy 2016-07-06).R
clean up a bit

@@ -1,218 +1,0 @@
-################################################################################
-# R code for "partial: partial: An R Package for Creating Partial Dependence
-# Plots"
-################################################################################
-
-
-################################################################################
-# Setup
-################################################################################
-
-# Set working directory
-setwd("/home/w108bmg/Desktop/Dropbox/Projects/greenwell_partial_2016")
-
-
-################################################################################
-# Code for paper
-################################################################################
-
-# Install the latest stable version from CRAN
-# install.packages("partial")
-
-# Install development version from GitHub
-# devtools::install_github("bgreenwell/partial")
-
-# Load the data
-data(BostonHousing2, package = "mlbench")  # load the data
-boston <- BostonHousing2[, -c(1, 2, 5)]
-
-# Load required packages
-library(partial)
-library(randomForest)
-
-# Fit a random forest
-set.seed(101)  # for reproducibility
-fit.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
-print(fit.rf)  # check model results
-pdf("boston_rf_vimp.pdf", width = 7, height = 5)
-varImpPlot(fit.rf, main = "")  # dotchart of variable importance scores
-dev.off()
-
-# Call:
-#   randomForest(formula = cmedv ~ ., data = boston, importance = TRUE)
-# Type of random forest: regression
-# Number of trees: 500
-# No. of variables tried at each split: 5
-#
-# Mean of squared residuals: 9.114248
-# % Var explained: 89.17
-
-pdf("pd_lstat.pdf", width = 7, height = 5)
-partial(fit.rf, pred.var = "lstat", plot = TRUE)
-dev.off()
-
-# Bivariate partial dependence
-pd.lstat.rm <- partial(fit.rf, pred.var = c("lstat", "rm"))
-pdf("pd_lstat_rm.pdf", width = 12, height = 6)
-p1 <- plotPartial(pd.lstat.rm)
-p2 <- plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
-                  colorkey = TRUE, screen = list(z = -20, x = -60))
-print(p1, position = c(0, 0, 0.5, 1), more = TRUE)
-print(p2, position = c(0.5, 0, 1, 1))
-dev.off()
-
[…]

### 2016-07-13 manuscript/greenwell.tex
clean up a bit
added example dataset descriptions

@@
 For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}.
+These data contain the median value of owner-occupied homes in 506 U.S.
+census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town.
 We begin by loading the data and omitting unimportant columns.
@@
 To illustrate, consider Edgar Anderson's iris data from the \pkg{datasets} package.
+The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica.
 We fit a support vector machine with a radial basis function kernel to the data usinv the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
@@
 
-\bibliography{Greenwell}
+\bibliography{greenwell}
 

### 2016-07-14 manuscript/greenwell.tex
minor updates

@@
 multiple linear regression).
-Unfortunately though, it can be challenging for the uninitiated to understand the results of such models and explain them to management.
-Partial dependence plots offer a simple solution.
-Partial dependence plots offer low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood.
+Unfortunately, it can be challenging for the uninitiated to understand the results of such models and explain them to management.
+Partial dependence plots (PDPs) offer a simple solution.
+Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood.
 These plots are especially useful in explaining the output from complex black box models like support vector machines or neural networks or constructing simpler parsimonious models on newly obtained data.
@@
 In higher dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on.
-When \code{convex.hull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot.
+When \code{convex.hull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot; alternatively, you can set \code{convex.hull = TRUE} in the call to \code{partial}, in which case only the region within the convex hull of the first two variables is plotted.
 Over interpreting the partial dependence plot outside of this region can be dangerous.
-The right side of Figure~\ref{fig:partial_extrap} was produced using:
+The right two displays in Figure~\ref{fig:partial_extrap} were produced, respectively, using:
 \begin{example}
 plotPartial(pd.lstat.rm, convex.hull = TRUE, training.data = boston)
+partial(boston.rf, pred.var = c("lstat", "rm"), convex.hull = TRUE, plot = TRUE)
 \end{example}

### 2016-07-14 R/plotPartial.R

@@ -9,7 +9,8 @@
 #'   (\code{FALSE}). Default is \code{TRUE}.
 #' @param rug Logical indicating whether or not to include a rug representation
 #'   to the plot. If \code{TRUE} the user must supply the original data.
-#' @param convex.hull Logical
+#' @param convex.hull Logical indicating whether or not to draw the convex hull
+#'   around the first two variables. Default is \code{FALSE}.
 #' @param number Integer
 #' @param overlap Proportion
 #' @param training.data Data frame containing the original training data. Only
@@ -17,7 +18,7 @@
 #' @param ... Additional optional arguments to be passed onto \code{levelplot},
 #'   \code{wireframe}, or \code{xyplot}.
 #'
-#' @importFrom lattice equal.count levelplot panel.levelplot panel.lines
+#' @importFrom lattice dotplot equal.count levelplot panel.levelplot panel.lines
 #' @importFrom lattice panel.loess panel.xyplot panel.rug wireframe xyplot
 #' @importFrom grDevices chull
 #' @importFrom stats as.formula quantile
@@ -63,7 +64,7 @@ plotPartial.partial <- function(x, smooth = FALSE, contour = TRUE, rug = FALSE,
                                      probs = 0:10/10, na.rm = TRUE))
                }
              }
-      })  
+      })
     }
   } else if (nx == 2) {
     form <- as.formula(paste("y ~", paste(names(x)[1L:2L], collapse = "*")))

### 2016-07-31 manuscript/greenwell.tex
minor tweaks

@@
 The following snippet of code produces the plot on the left side of Figure~\ref{fig:partial_extrap}.
-Notice that in both examples we had to provide the original training data to \code{plotPartial} via the \code{training.data} option.
+Notice that in both examples we had to provide the original training data to \code{plotPartial} via the \code{train} option.
 \begin{example}
-plotPartial(pf.lstat, rug = TRUE, training.data = boston)
+plotPartial(pf.lstat, rug = TRUE, train = boston)
 \end{example}
@@
 In higher dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on.
-When \code{convex.hull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot; alternatively, you can set \code{convex.hull = TRUE} in the call to \code{partial}, in which case only the region within the convex hull of the first two variables is plotted.
+When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot; alternatively, you can set \code{chull = TRUE} in the call to \code{partial}, in which case only the region within the convex hull of the first two variables is plotted.
 Over interpreting the partial dependence plot outside of this region can be dangerous.
@@
 \begin{example}
-plotPartial(pd.lstat.rm, convex.hull = TRUE, training.data = boston)
-partial(boston.rf, pred.var = c("lstat", "rm"), convex.hull = TRUE, plot = TRUE)
+plotPartial(pd.lstat.rm, chull = TRUE, train = boston)
+partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE, plot = TRUE)
 \end{example}
@@
   tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, training.data = iris)
+                 which.class = i, train = iris)
   pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))

### 2016-07-31 R/plotPartial.R

@@ -3,18 +3,21 @@
 #' Plots partial dependence functions using \code{lattice} graphics.
 #'
 #' @param x An object of class{"partial_1d"} or \code{"partial_2d"}.
-#' @param smooth Logical
+#' @param smooth Logical indicating whether or nor to overlay a loess smoother.
+#'   Default is \code{FALSE}.
 #' @param contour Logical indicating whether or not to use
 #'   \code{lattice::levelplot} (\code{TRUE}) or \code{lattice::wireframe}
 #'   (\code{FALSE}). Default is \code{TRUE}.
 #' @param rug Logical indicating whether or not to include a rug representation
 #'   to the plot. If \code{TRUE} the user must supply the original data.
-#' @param convex.hull Logical indicating whether or not to draw the convex hull
+#' @param chull Logical indicating whether or not to draw the convex hull
 #'   around the first two variables. Default is \code{FALSE}.
-#' @param number Integer
+#' @param number Integer specifying the number of conditional intervals for the
+#'   panel variables.
 #' @param overlap Proportion
-#' @param training.data Data frame containing the original training data. Only
-#'   required if \code{rug = TRUE} or \code{convex.hull = TRUE}.
+#' @param train Data frame containing the original training data. Only
+#'   required if \code{rug = TRUE} or \code{chull = TRUE}.
+#' @param col.regions Color vector to be used if \code{contour} is \code{TRUE}.
 #' @param ... Additional optional arguments to be passed onto \code{levelplot},
 #'   \code{wireframe}, or \code{xyplot}.
 #'
@@ -33,8 +36,9 @@ plotPartial <- function(x, ...) {
 #' @rdname plotPartial
 #' @export
 plotPartial.partial <- function(x, smooth = FALSE, contour = TRUE, rug = FALSE,
-                                convex.hull = FALSE, number = 4, overlap = 0.1,
-                                training.data = NULL, ...) {
+                                chull = FALSE, number = 4, overlap = 0.1,
+                                train = NULL, col.regions = viridis::viridis,
+                                ...) {
 
   # Determine number of variables to plot
   nx <- ncol(x) - 1  # don't count response
@@ -57,10 +61,10 @@ plotPartial.partial <- function(x, smooth = FALSE, contour = TRUE, rug = FALSE,
              }
              # Add a rug display
              if (rug) {
-               if (is.null(training.data)) {
+               if (is.null(train)) {
                  stop("The training data must be supplied for rug display.")
[…]

### 2016-07-31 manuscript/greenwell.R

@@ -20,6 +20,7 @@ setwd("/home/w108bmg/Desktop/Dropbox/devel/partial/manuscript")
 # library(caret)
 library(doParallel)
 library(e1071)
+library(earth)
 # library(kernlab)
 library(partial)
 library(party)
@@ -69,7 +70,7 @@ boston.crf <- cforest(cmedv ~ ., data = boston)
 boston.mars <- earth(cmedv ~ ., data = boston, degree = 3,
                      pmethod = "exhaustive")
 pd <- partial(boston.mars, c("rm", "lstat"), .progress = "text")
-plotPartial(pd, convex.hull = TRUE, training.data = boston)
+plotPartial(pd, chull = TRUE, train = boston)
 
 # Figure 3
 pd.lstat.rm <- partial(boston.rf, pred.var = c("lstat", "rm"))
@@ -87,10 +88,10 @@ dev.off()
 # Figure 4
 pd.lstat <- partial(boston.rf, pred.var = "lstat")
 pdf("partial_extrap.pdf", width = 12, height = 4)
-pdp1 <- plotPartial(pd.lstat, rug = TRUE, training.data = boston)
-pdp2 <- plotPartial(pd.lstat.rm, convex.hull = TRUE, training.data = boston)
+pdp1 <- plotPartial(pd.lstat, rug = TRUE, train = boston)
+pdp2 <- plotPartial(pd.lstat.rm, chull = TRUE, train = boston)
 pdp3 <- plotPartial(partial(boston.rf, pred.var = c("lstat", "rm"),
-                            convex.hull = TRUE))
+                            chull = TRUE))
 gridExtra::grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
 dev.off()
 
@@ -115,20 +116,20 @@ dev.off()
 # Edgar Anderson's iris data
 ################################################################################
 
-# Train an SVM to Edgar Anderson's iris data using 5-fold classification
-set.seed(101)
-ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)
-iris.svm.tune <- train(x = subset(iris, select = -Species),
-                       y = iris$Species,
-                       method = "svmRadial",
-                       trControl = ctrl,
-                       tuneLength = 100)
+# # Train an SVM to Edgar Anderson's iris data using 5-fold classification
+# set.seed(101)
+# ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)
[…]

### 2016-08-01 manuscript/greenwell.tex
major tweak to imports

@@
 \abstract{
-Complex predictive models (e.g., neural networks, random forests, and support vector machines) offer much promise in the world of data analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g.
+Complex nonparametric models (e.g., neural networks, random forests, and support vector machines) are often used in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g.
 multiple linear regression).
@@
 Predictor importance is an important task in any supervised learning problem.
-However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them and the response.
+However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response.
 This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details.
 PDPs help visualize the relationship between a subset (typically 1-3) of the predictors and the response while accounting for the average effect of the other independent variables in the model.
-They are particularly effective with black box models like random forest or support vector machines.
+They are particularly effective with black box models like random forest and support vector machines.
 %PDPs offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.
@@
 \end{equation}
-where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample.
-This can be quite computationally intensive since Equation~\eqref{eqn:pdf} involves a pass over the training records for each set of combinations of $\boldsymbol{z}_s$.
-Fortunately, this problem is embarrassingly parallel.
+where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, weaverage out the effects of all the other predictors in the model.
 Let $x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$.
@@
 \end{enumerate}
-This algorithm can be easily extended to larger subsets of 2 or more features as well.
+This can be quite computationally intensive since the algorithm involves a pass over the training records for each set of combinations of $\boldsymbol{z}_s$.
+Fortunately, this parallelized quite easily.
+This algorithm can also be easily extended to larger subsets of 2 or more features as well.
 
-Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others.
+Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using these packages.
 While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space; however, these implementations only apply to models fit using the respective package.
@@
 The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values.
-If \code{plot = TRUE}, a \code{"lattice"} object representing the PDP is returned.
+If \code{plot = TRUE}, then \code{partial} calls \code{plotPartial} and returns a \code{"lattice"} object representing the PDP.
 For more advanced plotting, the \code{plotPartial} functions will take a \code{"partial"} object and display a more customizable lattice plot.
@@
 The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}.
-Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes.
+Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}).
 The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.
@@
 
-As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for displaying partial dependence of the response on a single predictor.
+As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor.
 For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
@@
 \end{example}
-The same plot (implemented in \pkg{lattice}) can be achieved using the \code{partial} function and setting \code{plot = TRUE}:
+The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}) as in the folling example:
 \begin{example}
@@
 \end{example}
+The only difference is that \pkg{partia} uses \CRANpkg{lattice} \citep{lattice-pkg} to produce all of its displays.
+
 For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function.
@@
 \end{example}
-The results are displayed in Figure~\ref{fig:pd_lstat} below.
+The result is displayed in the right side of Figure~\ref{fig:pd_lstat} below.
 
@@
 \end{example}
+Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.
 

### 2016-08-01 R/pdRegression.R

@@ -1,29 +1,29 @@
 #' @keywords internal
-pdRegression <- function(object, pred.var, pred.grid, training.data, ...) {
+pdRegression <- function(object, pred.var, pred.grid, train, ...) {
   UseMethod("pdRegression")
 }
 
 
 #' @keywords internal
-pdRegression.default <- function(object, pred.var, pred.grid, training.data,
+pdRegression.default <- function(object, pred.var, pred.grid, train,
                                  ...) {
-  adply(pred.grid, .margins = 1, .fun = function(x) {
-    temp <- training.data
+  plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
+    temp <- train
     temp[pred.var] <- x
-    mean(predict(object, newdata = temp), na.rm = TRUE)
+    mean(stats::predict(object, newdata = temp), na.rm = TRUE)
   }, ...)
 }
 
 
 #' @keywords internal
 #' @importFrom utils capture.output
-pdRegression.gbm <- function(object, pred.var, pred.grid, training.data,
+pdRegression.gbm <- function(object, pred.var, pred.grid, train,
                                  ...) {
   # Necessary to avoid silly printing from predict.gbm
-  adply(pred.grid, .margins = 1, .fun = function(x) {
-    temp <- training.data
+  plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
+    temp <- train
     temp[pred.var] <- x
-    log <- capture.output(z <- mean(predict(object, newdata = temp),
+    log <- capture.output(z <- mean(stats::predict(object, newdata = temp),
                                     na.rm = TRUE))
     z
   }, ...)

### 2016-08-01 R/plotPartial.R

@@ -23,8 +23,6 @@
 #'
 #' @importFrom lattice dotplot equal.count levelplot panel.levelplot panel.lines
 #' @importFrom lattice panel.loess panel.xyplot panel.rug wireframe xyplot
-#' @importFrom grDevices chull
-#' @importFrom stats as.formula quantile
 #'
 #' @rdname plotPartial
 #' @export
@@ -49,10 +47,10 @@ plotPartial.partial <- function(x, smooth = FALSE, contour = TRUE, rug = FALSE,
   # Plot the partial dependence function
   if (nx == 1) {
     if (is.factor(x[[1L]])) {
-      dotplot(as.formula(paste("y ~", names(x)[1L])), data = x, ...)
+      dotplot(stats::as.formula(paste("y ~", names(x)[1L])), data = x, ...)
     } else {
-      xyplot(as.formula(paste("y ~", names(x)[1L])), data = x, type = "l", ...,
-           panel = function(xx, yy, ...) {
+      xyplot(stats::as.formula(paste("y ~", names(x)[1L])), data = x,
+             type = "l", ..., panel = function(xx, yy, ...) {
              # Add basic PDP
              panel.xyplot(xx, yy, col = "black", ...)
              # Add a loess smoother
@@ -64,14 +62,15 @@ plotPartial.partial <- function(x, smooth = FALSE, contour = TRUE, rug = FALSE,
                if (is.null(train)) {
                  stop("The training data must be supplied for rug display.")
                } else {
-                  panel.rug(quantile(train[[names(x)[1L]]],
-                                     probs = 0:10/10, na.rm = TRUE))
+                  panel.rug(stats::quantile(train[[names(x)[1L]]],
+                                            probs = 0:10/10, na.rm = TRUE))
                }
              }
       })
     }
   } else if (nx == 2) {
-    form <- as.formula(paste("y ~", paste(names(x)[1L:2L], collapse = "*")))
+    form <- stats::as.formula(paste("y ~",
+                                    paste(names(x)[1L:2L], collapse = "*")))
     if (contour) {
       levelplot(form, data = x, col.regions = col.regions, ...,
                 panel = function(x1, y1, ...) {
@@ -83,10 +82,10 @@ plotPartial.partial <- function(x, smooth = FALSE, contour = TRUE, rug = FALSE,
                   }
                   # Add a rug display
                   if (rug) {
-                    panel.rug(quantile(train[[names(x)[1L]]],
[…]

### 2016-08-08 manuscript/greenwell.tex
minor edits

@@
 Unfortunately, it can be challenging for the uninitiated to understand the results of such models and explain them to management.
-Partial dependence plots (PDPs) offer a simple solution.
+Partial dependence plots offer a simple solution.
 Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood.
@@
 \end{enumerate}
-This can be quite computationally intensive since the algorithm involves a pass over the training records for each set of combinations of $\boldsymbol{z}_s$.
-Fortunately, this parallelized quite easily.
+This can be quite computationally intensive since the algorithm involves $k$ passes over the training records.
+Fortunately, this can be parallelized quite easily.
 This algorithm can also be easily extended to larger subsets of 2 or more features as well.
@@
 Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using these packages.
+For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"}.
 While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space; however, these implementations only apply to models fit using the respective package.

### 2016-08-08 manuscript/greenwell.tex
minor edits

@@
 
-The \CRANpkg{plotmo} package \citep{plotmo-pkg} has come a long way in this regard.
-According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors constant (continuous features are fixed at their median value, while factors are held at their first level).
-These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct!
+The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{partial}.
+According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
+These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct.
 For additive models (i.e., no interactions), these plots are identical in shape to partial dependence plots.

### 2016-08-08 manuscript/greenwell.tex
minor edits
minor updates

@@
 
-PDPs are very useful but can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}.
+PDPs can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}.
 To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots -- available in the \CRANpkg{ICEbox} package.
-\pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though, color can be used effectively to display information about an additional predictor.
+\pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor.
 The ability to construct derivative plots is also available.

### 2016-08-08 manuscript/greenwell.tex
minor edits

@@
 The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest.
-However, these methods are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{partial} are mainly for nonparametric and black box models (though, they can be used for simple parametric models as well).
+However, these methods are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{partial} are mainly for nonparametric and black box models (though they can be used for simple parametric models as well).
 

### 2016-08-08 manuscript/greenwell.tex
minor edits

@@
 These are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest.
-The latest stable release is available from CRAN:
+The latest stable release is available from \href{https://github.com/bgreenwell/partial}{CRAN}:
 \begin{example}
@@
 \end{example}
+The development version is located on GitHub: \url{https://github.com/bgreenwell/partial}.
+Bug reports and usggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/partial/issues}.
 

### 2016-08-08 manuscript/greenwell.tex
fixed link (still nonexistent though)
minor edits

@@
 These are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest.
-The latest stable release is available from \href{https://github.com/bgreenwell/partial}{CRAN}:
+The latest stable release is available from \href{https://cran.r-project.org/package=partial}{CRAN}:
 \begin{example}
@@
 If \code{plot = FALSE} (the default), \code{partial} returns a data frame with an additional class: \code{"partial"}.
-The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values.
-If \code{plot = TRUE}, then \code{partial} calls \code{plotPartial} and returns a \code{"lattice"} object representing the PDP.
-For more advanced plotting, the \code{plotPartial} functions will take a \code{"partial"} object and display a more customizable lattice plot.
+The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
+If \code{plot = TRUE}, then \code{partial} calls \code{plotPartial} and returns a \code{"trellis"} object representing the PDP.
+For more advanced plotting, the \code{plotPartial} function will take a \code{"partial"} object and display a more customizable lattice plot.
 Currently supported models are described in Table~\ref{tab:models}.

### 2016-08-08 manuscript/greenwell.tex
minor edits

@@
 The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
-If \code{plot = TRUE}, then \code{partial} calls \code{plotPartial} and returns a \code{"trellis"} object representing the PDP.
+If \code{plot = TRUE}, then \code{partial} returns a \CRANpkg{lattice} plot \citep{lattice-pkg}.
 For more advanced plotting, the \code{plotPartial} function will take a \code{"partial"} object and display a more customizable lattice plot.
@@
 \end{example}
-The only difference is that \pkg{partia} uses \CRANpkg{lattice} \citep{lattice-pkg} to produce all of its displays.
+The only difference is that \pkg{partial} uses the \pkg{lattice} graphics package to produce all of its displays.
 

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 For more advanced plotting, the \code{plotPartial} function will take a \code{"partial"} object and display a more customizable lattice plot.
+
 Currently supported models are described in Table~\ref{tab:models}.
+The \code{partial.default} method, however, should be able to handle most other types of models; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 \begin{table}[htbp]
@@
 \end{table}
-The \code{partial.default} method should be able to handle most other types of models not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 

### 2016-08-09 manuscript/greenwell.tex
edits to Table 1

@@
 Currently supported models are described in Table~\ref{tab:models}.
-The \code{partial.default} method, however, should be able to handle most other types of models; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+While more models will be added, the \code{partial.default} method should be able to handle many others not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 \begin{table}[htbp]
@@
     \toprule
-      Model & Package & Class \\
+      Type of model & R package & Object class \\
       \midrule
+      Decision trees            & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
+                                & \pkg{party}    & \code{"BinaryTree"} \\
+                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
       Bagged decision trees     & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
@@
                                 & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
+      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
       Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
                                 & \pkg{partykit} & \code{"cforest"} \\
-      Decision trees            & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
-                                & \pkg{party}    & \code{"BinaryTree"} \\
-                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
       Linear model              & \pkg{stats}    & \code{"lm"} \\
+      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
       Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
-      Random forest & \pkg{randomForest} & \code{"randomForest"} \\
       Support vector machines   & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 \end{example}
-Next, we fit a traditional random forest to the entire data set using the following code snippet:
+Next, we fit a random forest to the entire data set with default tuning parameters:
 \begin{example}
@@
 \end{example}
-The model fit is reasonable, with an \dfn{out-of-bag} $R^2$ of 0.89.
+The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89.
 The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}.

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 \end{example}
-The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}) as in the folling example:
+The same plot can be achieved using the \code{partial} function and setting \code{plot = TRUE} (see the left side of Figure~\ref{fig:pd_lstat}):
 \begin{example}

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function.
-\strong{Note:} the \dfn{pipe} operator provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is imported for writing more convenient code, as illustrated in snippet of code below.
+\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is imported for writing more convenient code, as illustrated in snippet of code below.
 \begin{example}

### 2016-08-09 manuscript/greenwell.tex
minor edits

--- 2016-08-09 manuscript/greenwell.tex
@@
 For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function.
-\strong{Note:} the \dfn{pipe} operator provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is imported for writing more convenient code, as illustrated in snippet of code below.
+\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code, as illustrated in snippet of code below which adds a LOESS smoother to the previously obtained PDP.
 \begin{example}
-partial(boston.rf, pred.var = "lstat") %>%
+boston.rf %>%
+  partial(pred.var = "lstat") %>%
   plotPartial(smooth = TRUE, lwd = 2, ylab = expression(f(lstat)))

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function.
-\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code, as illustrated in snippet of code below which adds a LOESS smoother to the previously obtained PDP.
+\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code, as illustrated in snippet of code below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label.
 \begin{example}

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 \end{example}
-Next, we fit a random forest to the entire data set with default tuning parameters:
+Next, we fit a random forest to the entire data set with default tuning parameters and 500 trees:
 \begin{example}
@@
   \includegraphics[width=1.0\linewidth]{pd_lstat}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat}. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
+  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
   \label{fig:pd_lstat}
@@
 boston.crf <- cforest(cmedv ~ ., data = boston)
-pd.lstat.rm <- partial(boston.rf, pred.var = c("lstat", "rm"))
+pd.lstat.rm <- partial(boston.crf, pred.var = c("lstat", "rm"))
 plotPartial(pd.lstat.rm)
@@
   \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a different color palette. \textit{Right}: Using a 3-D surface.}
+  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a conditional random forest.
+\textit{Left}: Default plot.
+\textit{Middle}: Using a different color palette.
+\textit{Right}: Using a 3-D surface.}
   \label{fig:pd_lstat_rm}

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 
-There are few ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
+There are a few ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
 Rug displays are one-dimensional plots added to the axes.
-\code{plotPartial} has a \code{rug} option that will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal or vertical axes.
+\code{plotPartial} has a \code{rug} option that will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and/or vertical axes.
 The following snippet of code produces the plot on the left side of Figure~\ref{fig:partial_extrap}.
@@
 \begin{example}
-plotPartial(pf.lstat, rug = TRUE, train = boston)
+partial(boston.rf, pred.var = "lstat", rug = TRUE, train = boston)
 \end{example}
@@
 \begin{example}
-plotPartial(pd.lstat.rm, chull = TRUE, train = boston)
 partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE, plot = TRUE)

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 \code{plotPartial} has a \code{rug} option that will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and/or vertical axes.
-The following snippet of code produces the plot on the left side of Figure~\ref{fig:partial_extrap}.
-Notice that in both examples we had to provide the original training data to \code{plotPartial} via the \code{train} option.
+The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
+%Notice that in both examples we had to provide the original training data to \code{plotPartial} via the \code{train} option.
 \begin{example}
-partial(boston.rf, pred.var = "lstat", rug = TRUE, train = boston)
+partial(boston.rf, pred.var = "lstat", rug = TRUE, plot = TRUE)
 \end{example}
@@
 Over interpreting the partial dependence plot outside of this region can be dangerous.
-The right two displays in Figure~\ref{fig:partial_extrap} were produced, respectively, using:
+The right right display in Figure~\ref{fig:partial_extrap} was produced using:
 \begin{example}

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 Constructing PDPs can be quite computationally expensive.
-Additional options are available to ease the computational intensity for large problems.
-For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require 446 passes over the data!).
+Several strategies are available to ease the computational burden for large problems.
+For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!).
 We could get very reasonable results using a reduced number of points.
-Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{n.pts}.
+Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{grid.resolution}.
 Alternatively, a specific set of values (e.g., quantiles of interest) can be supplied through the \code{pred.grid} argument.

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package.
-To use this functionality, we must load and register a supported parallel backend (e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}).
-The following snippet of code obtains the partial dependence of median home value on \code{rm}, \code{lstat}, and \code{ptratio} in parallel.
-The result is displayed in Figure~\ref{fig:partial_par}.
+To use this functionality, we must load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].
 
@@
 \end{example}
-Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} the parallel backend we loaded:
+Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} the parallel backend we loaded.
+This is illustrated in following snippet of code which obtains the partial dependence of median home value on $\boldsymbol{z}_s = \left\{\code{rm}, \code{ptratio}, \code{chas}\right\} in parallel.
+The result is displayed in Figure~\ref{fig:partial_par}.
 \begin{example}
-pd <- partial(boston.rf, pred.var = c("lstat", "rm", "ptratio"),
+pd <- partial(boston.rf, pred.var = c("rm", "ptratio", "chas"),
               grid.resolution = 20, .parallel = TRUE)
@@
 \end{example}
+
+%Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} the parallel backend we loaded.
+This is illustrated in following snippet of code which obtains the partial dependence of median home value on $\boldsymbol{z}_s = \left\{\code{rm}, \code{lstat}, \code{ptratio}\right\} in parallel.
+The result is displayed in Figure~\ref{fig:partial_par}.
+%\begin{example}
+%pd <- partial(boston.rf, pred.var = c("lstat", "rm", "ptratio"),
+%              grid.resolution = 20, .parallel = TRUE)
+%plotPartial(pd, number = 4, overlap = 0.1)
+%\end{example}
 
@@
 The intervals may be allowed to overlap.} are produced first using the equal count algorithm.
-Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible.
+Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible, as was done in the previous example.
 

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 \begin{example}
-pd <- partial(boston.rf, pred.var = c("rm", "ptratio", "chas"),
-              grid.resolution = 20, .parallel = TRUE)
-plotPartial(pd, number = 4, overlap = 0.1)
+partial(boston.rf, pred.var = c("rm", "ptratio", "chas"), grid.resolution = 20, 
+        chull = TRUE, .parallel = TRUE, plot = TRUE)
 \end{example}

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 
-There are a few ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
+There are a couple ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
 Rug displays are one-dimensional plots added to the axes.
-\code{plotPartial} has a \code{rug} option that will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and/or vertical axes.
+Both \code{partial} and \code{plotPartial} have a \code{rug} option that, when set to \code{TRUE}, will display the deciles of the distribution (as well as the minimum and maximum values) for the predictors on the horizontal and vertical axes.
 The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
@@
 
-In higher dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on.
+In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on.
 When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot; alternatively, you can set \code{chull = TRUE} in the call to \code{partial}, in which case only the region within the convex hull of the first two variables is plotted.
 Over interpreting the partial dependence plot outside of this region can be dangerous.
-The right right display in Figure~\ref{fig:partial_extrap} was produced using:
+The right display in Figure~\ref{fig:partial_extrap} was produced using:
 \begin{example}

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica.
-We fit a support vector machine with a radial basis function kernel to the data usinv the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
+We fit a support vector machine with a radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
 \begin{example}

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 \end{example}
-The \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to invoke the \code{prob.model = TRUE} option in the call to \code{svm}.
+The \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set the \code{probability = TRUE} option in the call to \code{svm}.
 

### 2016-08-09 manuscript/greenwell.tex
minor edits

@@
 
-Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
+Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes.
+The result is displayed in Figure~\ref{fig:partial_iris}.
+Note that we had to supply the original training data to \code{partial} through the \code{train} option.
+This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
 \begin{example}
@@
 In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{partial} package.
-We discussed similar approaches in other packages.
+We briefly discussed related approaches in other packages as well.
 Ways to avoid extrpolation and high excecution times were demonstrated via examples.

### 2016-08-10 manuscript/greenwell.tex
minor edits

@@
 The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
-For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} based on a conditional random forest using various options in \code{plotPartial}.
-The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
+For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} using a multivariate adaptive regression splines (MARS) model (fit using the \pkg{earth} package) allowing for up to three-way interactions.
+The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
 \begin{example}
-library(party)
-boston.crf <- cforest(cmedv ~ ., data = boston)
-pd.lstat.rm <- partial(boston.crf, pred.var = c("lstat", "rm"))
+library(earth)
+boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
+pd.lstat.rm <- partial(boston.mars, pred.var = c("lstat", "rm"))
 plotPartial(pd.lstat.rm)
@@
   \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a conditional random forest.
+  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a third degree MARS model.
 \textit{Left}: Default plot.

### 2016-08-10 manuscript/greenwell.tex
minor edits

@@
 The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
-For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} using a multivariate adaptive regression splines (MARS) model (fit using the \pkg{earth} package) allowing for up to three-way interactions.
+For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-mars-1988} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
 The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
 \begin{example}
-library(earth)
+library(earth)  # for fitting MARS models
 boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)

### 2016-08-10 manuscript/greenwell.tex
fixed citation

@@
 The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
-For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-mars-1988} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
+For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
 The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.

### 2016-08-10 manuscript/greenwell.tex
minor edits

@@
 plotPartial(pd.lstat.rm)
-plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("white", "blue")))
+plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("red", "white", "blue")))
 plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,

### 2016-08-10 manuscript/greenwell.tex
minor updates

@@
 Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} the parallel backend we loaded.
-This is illustrated in following snippet of code which obtains the partial dependence of median home value on $\boldsymbol{z}_s = \left\{\code{rm}, \code{ptratio}, \code{chas}\right\} in parallel.
+This is illustrated in following snippet of code which obtains the partial dependence of median home value on $\boldsymbol{z}_s = \{$\code{rm}, \code{ptratio}, \code{chas}$\}$ in parallel.
 The result is displayed in Figure~\ref{fig:partial_par}.

### 2016-08-11 manuscript/greenwell.tex
minor edits

@@
 
+To illustrate, we will use the Los Angeles ozone pollution data described in ????.
+The data are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.data}.
+The following code chunk loads the data into R:
+\begin{example}
+ozone <- read.csv(paste0("http://statweb.stanford.edu/~tibs/ElemStatLearn/",
+                         "datasets/LAozone.data"), header = TRUE)
+\end{example}
+
+Next, we use MARS to to study the relationship between atmospheric ozone (\code{ozone}) concentration and meteorology in the Los Angeles basin in 1976, allowing for up to three-way interactions.
+\begin{example}
+ozone.earth <- earth(ozone ~ ., data = ozone, degree = 3)
+summary(ozone.earth)
+\end{example}
+A single three-way interaction term between \code{wind}, \code{temp}, and \code{dpg} was selected by the MARS algorithm.
+To understand this interaction, we can use partial.
+However, since the partial dependece between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.
+
 Setting up a parallel backend is rather straightforward.
-To illustrate, the following snippet of sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel}.
+To illustrate, the following snippet of sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel} package.
 \begin{example}
@@
 \end{example}
-Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} the parallel backend we loaded.
-This is illustrated in following snippet of code which obtains the partial dependence of median home value on $\boldsymbol{z}_s = \{$\code{rm}, \code{ptratio}, \code{chas}$\}$ in parallel.
+
+Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded.
+This is illustrated in following snippet of code which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
 The result is displayed in Figure~\ref{fig:partial_par}.
 \begin{example}
-partial(boston.rf, pred.var = c("rm", "ptratio", "chas"), grid.resolution = 20,
-        chull = TRUE, .parallel = TRUE, plot = TRUE)
+partial(ozone.earth, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
+        chull = TRUE, .parallel = TRUE)
 \end{example}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=0.8\linewidth]{partial_par}
+  \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}.
+Since \code{dpg} is contiuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
+  \label{fig:partial_par}
+\end{figure}
+
+% \begin{itemize}
+%   \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
+%   \item \code{temp}: temperature (degrees F) at Sandburg Air Force Base
+%   \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
+% \end{itemize}
 
@@
 Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible, as was done in the previous example.
-
-\begin{figure}[htbp]
-  \centering
-  \includegraphics[width=0.8\linewidth]{partial_par}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat}, \code{rm}, and \code{ptratio}.
-Since \code{ptratio} is contiuous, it is first converted to a shingle; in this case, four groups with with 10\% overlap.}
-  \label{fig:partial_par}
-\end{figure}
 

### 2016-08-11 manuscript/greenwell.R

@@ -97,13 +97,29 @@ pdp3 <- plotPartial(partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9)))
 gridExtra::grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
 dev.off()
 
+
+# Load the LA ozone data
+ozone <- read.csv(paste0("http://statweb.stanford.edu/~tibs/ElemStatLearn/",
+                         "datasets/LAozone.data"), header = TRUE)
+
+# Fit a MARS model
+ozone.earth <- earth(ozone ~ ., data = ozone, degree = 3)
+
 # Figure 6
 registerDoParallel(cores = 4)  # use 4 cores
-pd <- partial(boston.rf, pred.var = c("rm", "ptratio", "chas"),
-              grid.resolution = 20, chull = TRUE, .parallel = TRUE)
+pd <- partial(ozone.earth, pred.var = c("wind", "temp", "dpg"), chull = TRUE,
+              .parallel = TRUE)
 pdf("partial_par.pdf", width = 7, height = 5)
 plotPartial(pd)
 dev.off()
+
+# # Figure 6
+# registerDoParallel(cores = 4)  # use 4 cores
+# pd <- partial(boston.rf, pred.var = c("rm", "ptratio", "chas"),
+#               grid.resolution = 20, chull = TRUE, .parallel = TRUE)
+# pdf("partial_par.pdf", width = 7, height = 5)
+# plotPartial(pd)
+# dev.off()
 #registerDoParallel(cores = 4)  # use 4 cores
 #pd <- partial(boston.rf, pred.var = c("lstat", "rm", "ptratio"),
 #              grid.resolution = 20, .parallel = TRUE)

### 2016-08-11 manuscript/greenwell.tex
minor edits

@@
 
-To illustrate, we will use the Los Angeles ozone pollution data described in ????.
+To illustrate, we will use the Los Angeles ozone pollution data described in \citet{breiman-1985-estimating}.
+The data contain daily measurements of ozone concentration (\code{ozone}) along with eight meteorological quantities for 330 days in the Los Angeles basin in 1976.
 The data are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.data}.
+Details, including variable information, are available from \url{http://statweb.stanford.edu/~tibs/ElemStatLearn/datasets/LAozone.info}.
 The following code chunk loads the data into R:
@@
 \end{example}
-A single three-way interaction term between \code{wind}, \code{temp}, and \code{dpg} was selected by the MARS algorithm.
-To understand this interaction, we can use partial.
+The model produced a generalized $R^2$ of $0.788$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
+\begin{itemize}
+  \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
+  \item \code{temp}: temperature ($^oF$) at Sandburg Air Force Base
+  \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
+\end{itemize}
+To understand this interaction, we can use a partial plot.
 However, since the partial dependece between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.

### 2016-08-11 manuscript/greenwell.tex
minor edit
minor edits

@@
 
-Next, we use MARS to to study the relationship between atmospheric ozone (\code{ozone}) concentration and meteorology in the Los Angeles basin in 1976, allowing for up to three-way interactions.
+Next, we use MARS to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
 \begin{example}
@@
 \end{example}
-The model produced a generalized $R^2$ of $0.788$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
+The model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
 \begin{itemize}

### 2016-08-11 manuscript/greenwell.tex
minor edit
fixed typo

@@
 \end{equation}
-where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, weaverage out the effects of all the other predictors in the model.
+where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, we average out the effects of all the other predictors in the model.
 Let $x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$.

### 2016-08-11 manuscript/greenwell.tex
minor edits

@@
 For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"}.
-While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space; however, these implementations only apply to models fit using the respective package.
-Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., neural networks).
+While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space.
+Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., multiple linear regression and neural networks).
 However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively).

### 2016-08-28 manuscript/greenwell.tex
first pass at a review

@@
 \abstract{
-Complex nonparametric models (e.g., neural networks, random forests, and support vector machines) are often used in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g.
+Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g.
 multiple linear regression).
@@
 This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details.
-PDPs help visualize the relationship between a subset (typically 1-3) of the predictors and the response while accounting for the average effect of the other independent variables in the model.
+PDPs help visualize the relationship between a subset of the predictors (typically 1-3) and the response while accounting for the average effect of the other independent variables in the model.
 They are particularly effective with black box models like random forest and support vector machines.
@@
 This can be quite computationally intensive since the algorithm involves $k$ passes over the training records.
-Fortunately, this can be parallelized quite easily.
+Fortunately, this can be parallelized quite easily (more on this in a later section).
 This algorithm can also be easily extended to larger subsets of 2 or more features as well.
 
-Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using these packages.
+Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using the respective package.
 For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% PDPs in R
+% Constructing PDPs in R
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{itemize}
-The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values.
-If \code{plot = FALSE} (the default), \code{partial} returns a data frame with an additional class: \code{"partial"}.
-The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the partial dependence function values $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
-If \code{plot = TRUE}, then \code{partial} returns a \CRANpkg{lattice} plot \citep{lattice-pkg}.
+The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values specified by the \code{pred.var} argument.
+If \code{plot = FALSE} (the default), \code{partial} returns an object of class inheriting from \code{"partial"} which inherits from the class \code{"data.frame"}.
+The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
+If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns a \CRANpkg{lattice} plot \citep{lattice-pkg}.
 For more advanced plotting, the \code{plotPartial} function will take a \code{"partial"} object and display a more customizable lattice plot.
@@
 Currently supported models are described in Table~\ref{tab:models}.
-While more models will be added, the \code{partial.default} method should be able to handle many others not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} or projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+While more models will be added as the package grows, the \code{partial.default} method should be able to handle many others not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 \begin{table}[htbp]
@@
       \midrule
-      Decision trees            & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
+      Decision tree             & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
                                 & \pkg{party}    & \code{"BinaryTree"} \\
@@
       Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
-      Support vector machines   & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
+      Support vector machine    & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\
                                 & \CRANpkg{kernlab} \citep{kernlab-pkg} & \code{"ksvm"} \\
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Univariate PDPs
+% Single predictor PDPs
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Higher order PDPs
+% Multi-predictor PDPs
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-% \sub\subsection{Multiple linear regression}
-% Here we use the multivariate adaptive regression splines (See ????) algorithm to perform multiple linear regression with automatc variable selection and interaction detection.
-% \begin{example}
-% library(earth)
-% boston.mars <- earth(cmedv ~ ., data = boston, degree = 3, linpreds = TRUE,
-%                      pmethod = "exhaustive")
-% partial(boston.mars, pred.var = c("nox", "rm"), grid.resolution = 10)
-% \end{example}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-There are a couple ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
+It is not wise to draw conclusions from PDPs in regions outside the range of the training data.
+Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
 Rug displays are one-dimensional plots added to the axes.
@@
 The following snippet of code produces the left display in Figure~\ref{fig:partial_extrap}.
-%Notice that in both examples we had to provide the original training data to \code{plotPartial} via the \code{train} option.
 \begin{example}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Addressing computational concerns
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Addressing computational concerns}
@@
 Constructing PDPs can be quite computationally expensive.
-Several strategies are available to ease the computational burden for large problems.
+Several strategies are available to ease the computational burden in larger problems.
 For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!).
@@
 In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package.
-To use this functionality, we must load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].
+To use this functionality, we must first load and register a supported parallel backend [e.g., \CRANpkg{doMC} \citep{doMC-pkg} or \CRANpkg{doParallel} \citep{doParallel-pkg}].
 
@@
 
-% \begin{itemize}
-%   \item \code{wind}: wind speed (mph) at Los Angeles International Airport (LAX)
-%   \item \code{temp}: temperature (degrees F) at Sandburg Air Force Base
-%   \item \code{dpg}: the pressure gradient (mm Hg) from LAX to Dagget, CA
-% \end{itemize}
-
-%Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} the parallel backend we loaded.
-This is illustrated in following snippet of code which obtains the partial dependence of median home value on $\boldsymbol{z}_s = \left\{\code{rm}, \code{lstat}, \code{ptratio}\right\} in parallel.
-The result is displayed in Figure~\ref{fig:partial_par}.
-%\begin{example}
-%pd <- partial(boston.rf, pred.var = c("lstat", "rm", "ptratio"),
-%              grid.resolution = 20, .parallel = TRUE)
-%plotPartial(pd, number = 4, overlap = 0.1)
-%\end{example}
-
-It is important to note that when using more than two variables, \code{plotPartial} produces a trellis display.
-The first two variables given to \code{pred.var} are used for the horizontal and vertical axes; additional variables define the panels.
+It is important to note that when using more than two predictor variables, \code{plotPartial} produces a trellis display.
+The first two variables given to \code{pred.var} are used for the horizontal and vertical axes, and additional variables define the panels.
 If the panel variables are continuous, then shingles\footnote{A shingle is a special Trellis data structure that consists of a numeric vector along with intervals that define the "levels" of the shingle.
-The intervals may be allowed to overlap.} are produced first using the equal count algorithm.
+The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}).
 Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible, as was done in the previous example.
@@
 The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica.
-We fit a support vector machine with a radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
+We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
 \begin{example}
@@
 \end{example}
-The \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set the \code{probability = TRUE} option in the call to \code{svm}.
+\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set the \code{probability = TRUE} option in the call to \code{svm}.
 
@@
 In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{partial} package.
-We briefly discussed related approaches in other packages as well.
-Ways to avoid extrpolation and high excecution times were demonstrated via examples.
+We also briefly discussed related approaches availble in other R packages as well.
+Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples.
 There is a bright future for \pkg{partial} in terms of growth.
-For example, we would like to include the ability to construct PDPs for various types of survival models (e.g.
-conditional random forests with censored response).
+For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response---and \CRANpkg{xgboost} \citep{xgboost-pkg} models.
 

### 2016-09-01 manuscript/greenwell.tex
note to add something
updated abstract
added xgboost to supported models tabel

@@
 These plots are especially useful in explaining the output from complex black box models like support vector machines or neural networks or constructing simpler parsimonious models on newly obtained data.
+In this paper, we introduce partial, a general R package for creating partial dependence plots.
 }
@@
 %PDPs offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.
+
+\strong{Mention the integral approximation!!}
 
@@
       Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
+                                & \CRANpkg{xgboost} \citep{xgboost-pkg} & \code{"xgb.Booster"}
                                 & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\

### 2016-09-01 manuscript/greenwell.tex
commented out example summary

@@
 There is a bright future for \pkg{partial} in terms of growth.
-For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response---and \CRANpkg{xgboost} \citep{xgboost-pkg} models.
+For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
+% We introduced the investr package for computing point estimates along with inversion and Waldbased confidence intervals for linear and nonlinear calibration problems with constant variance.
+We also showed how the boot package can be used for constructing approximate 100(1 􀀀 a)% calibration intervals, which for convenience, may be incorporated into a future release of investr.
+The authors are currently working on extending the package to handle the case of heteroscedastic errors, random coefficient models (e.g., objects of class "lme" from the recommended nlme package; Pinheiro et al., 2014)), and even multivariate calibration problems (e.g., objects of class "mlm").
 

### 2016-09-01 manuscript/greenwell.tex
minor edits

@@
 
+% FIXME %%%%%%%%%%%%%%%%%%%%%
+
 Currently supported models are described in Table~\ref{tab:models}.
-While more models will be added as the package grows, the \code{partial.default} method should be able to handle many others not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+Using the \code{partial.default} method it is still possible to handle many other models not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+However, for these cases, the user will need to specify additional options in the call to \code{partial}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{table}[htbp]

### 2016-09-01 manuscript/greenwell.tex
minor edits

@@
 Using the \code{partial.default} method it is still possible to handle many other models not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
-However, for these cases, the user will need to specify additional options in the call to \code{partial}.
+However, for these cases, the user will need to specify the additional options \code{super.type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with \code{"xgb.Booster"} models, which requires the user to specify \code{train}.
 
@@
 \end{figure}
+
+Finally, for further illustration, we will compare the random forest results with those obtained from gradient boosting; for this, we shal use the popular \pkg{xgboost} package.
+The example below, uses \pkg{xgboost} to model the boston housing data.
+The tuning parameters were predetermined using 5-fold cross-validation.
+\begin{example}
+set.seed(102)
+boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
+                      label = boston$cmedv,
+                      objective = "reg:linear",
+                      nrounds = 2000,
+                      max_depth = 3,
+                      eta = 0.01)
+\end{example}
+
+To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}:
+\begin{example}
+partial(boston.xgb, pred.var = "lstat", train = subset(boston, select = -cmedv), 
+        plot = TRUE)
+\end{example}
 

### 2016-09-01 manuscript/greenwell.tex
minor edits

@@
       Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
-                                & \CRANpkg{xgboost} \citep{xgboost-pkg} & \code{"xgb.Booster"}
+                                & \CRANpkg{xgboost}\footnote{Partially supported. See section ????} \citep{xgboost-pkg} & \code{"xgb.Booster"}
                                 & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
@@
 \end{figure}
-
-Finally, for further illustration, we will compare the random forest results with those obtained from gradient boosting; for this, we shal use the popular \pkg{xgboost} package.
-The example below, uses \pkg{xgboost} to model the boston housing data.
-The tuning parameters were predetermined using 5-fold cross-validation.
-\begin{example}
-set.seed(102)
-boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
-                      label = boston$cmedv,
-                      objective = "reg:linear",
-                      nrounds = 2000,
-                      max_depth = 3,
-                      eta = 0.01)
-\end{example}
-
-To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}:
-\begin{example}
-partial(boston.xgb, pred.var = "lstat", train = subset(boston, select = -cmedv), 
-        plot = TRUE)
-\end{example}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Using partial with xgboost
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Using partial with xgboost}
+
+XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable.
+The associated R package \pkg{xgboost} has been used in many winning Kaggle competitions.
+It has been shown to be many times faster than the well known \pkg{gbm} package.
+However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying partial dependence plots.
+Fortunately, the \pkg{partial} package can be used to fill this gap.
+
+Finally illustration, we return to the Boston housing data. The example below, fits a model using \pkg{xgboost}; the tuning parameters were predetermined using 5-fold cross-validation.
+\begin{example}
+set.seed(102)
+boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
+                      label = boston$cmedv,
+                      objective = "reg:linear",
+                      nrounds = 2000,
+                      max_depth = 3,
+                      eta = 0.01)
+\end{example}
+
+To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}.
+The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
+The results are displayed in Figure~\fig{fig:boston_xgb}.
+\begin{example}
+partial(boston.xgb, pred.var = "lstat", train = subset(boston, select = -cmedv), 
+        plot = TRUE)
+\end{example}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Summary

### 2016-09-01 manuscript/greenwell.tex
minor updates

@@
       Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
-                                & \CRANpkg{xgboost}\footnote{Partially supported. See section ????} \citep{xgboost-pkg} & \code{"xgb.Booster"}
+                                & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
                                 & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
@@
   \end{tabular}
-  \caption{Models specifically supported by the \pkg{package}.}
+  \caption{Models specifically supported by the \pkg{package}. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} fr details.}
   \label{tab:models}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Using partial with xgboost}
+\subsection{Using partial with the XGBoost library}
+\label{sec:xgboost}
 
 XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable.
-The associated R package \pkg{xgboost} has been used in many winning Kaggle competitions.
+The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}.
 It has been shown to be many times faster than the well known \pkg{gbm} package.
@@
 
-Finally illustration, we return to the Boston housing data. The example below, fits a model using \pkg{xgboost}; the tuning parameters were predetermined using 5-fold cross-validation.
+For illustration, we return to the Boston housing data.
+The example below, fits an \pkg{xgboost} model; the tuning parameters were predetermined using 5-fold cross-validation.
+(Make sure you are using the latest version of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)
 \begin{example}
@@
 boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
-                      label = boston$cmedv,
-                      objective = "reg:linear",
-                      nrounds = 2000,
-                      max_depth = 3,
-                      eta = 0.01)
+                      label = boston$cmedv, objective = "reg:linear",
+                      nrounds = 2000, max_depth = 3, eta = 0.01)
 \end{example}
@@
 The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
-The results are displayed in Figure~\fig{fig:boston_xgb}.
+The results are displayed in Figure~\ref{fig:boston_xgb}.
 \begin{example}
@@
 \end{example}
+
+\begin{figure}[htbp]
+  \centering
+  \includegraphics[width=1.0\linewidth]{boston_xgb}
+  \caption{Partial dependence plots for the top two most important variables in the Boston housing data using XGBoost.
+Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
+  \label{fig:boston_xgb}
+\end{figure}
 

### 2016-09-01 manuscript/greenwell.R

@@ -1,5 +1,5 @@
 # R code for the manuscript
-# 
+#
 #   partial: partial: An R Package for Creating Partial Dependence Plots
 #
 #        Author: Brandon M. Greenwell
@@ -11,14 +11,14 @@
 ################################################################################
 
 # Set working directory
-setwd("C:\\Users\\Brandon.Greenwell\\Desktop\\greenwell_partial_2016")
+setwd("/home/w108bmg/Desktop/Dropbox/devel/partial/manuscript")
 
 # List of packages required to run all the examples in this script
-pkgs <- c("doParallel", 
-          "e1071", 
-          "mlbench", 
-          "partial", 
-          "party", 
+pkgs <- c("doParallel",
+          "e1071",
+          "mlbench",
+          "partial",
+          "party",
           "randomForest",
           "xgboost")
 # install.packages(pkgs)
@@ -80,20 +80,20 @@ boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                       max_depth = 3,
                       eta = 0.01)
 
-pdf("boston_xgb.pdf", width = 7, height = 5)
+pdf("boston_xgb.pdf", width = 12, height = 4)
 X <- subset(boston, select = -cmedv)
 pdp1 <- plotPartial(partial(boston.xgb, pred.var = "lstat", train = X),
                     rug = TRUE, smooth = TRUE, train = X)
 pdp2 <- plotPartial(partial(boston.xgb, pred.var = "rm", train = X),
                     rug = TRUE, smooth = TRUE, train = X)
-pdp3 <- plotPartial(partial(boston.xgb, pred.var = c("lstat", "rm"), 
+pdp3 <- plotPartial(partial(boston.xgb, pred.var = c("lstat", "rm"),
                             chull = TRUE, train = X), train = X)
 gridExtra::grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
 dev.off()
 
 # Where can this figure go?
-partial(boston.xgb, pred.var = c("lstat", "rm"), chull = TRUE, 
-        train = subset(boston, select = -cmedv), plot = TRUE, 
+partial(boston.xgb, pred.var = c("lstat", "rm"), chull = TRUE,
+        train = subset(boston, select = -cmedv), plot = TRUE,
         .progress = "text")
 
 # Figure 3

### 2016-09-02 manuscript/greenwell.tex
minor fixes

@@
 
-\title{partial: An R Package for Creating Partial Dependence Plots}
+\title{pdp: An R Package for Creating Partial Dependence Plots}
 \author{by Brandon M. Greenwell}
@@
 However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively).
-The \CRANpkg{partial} \citep{partial-pkg} package tries to close this gap by offering a general framework for PDPs that can be applied to several types of fitted models.
+The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for PDPs that can be applied to several types of fitted models.
 
-The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{partial}.
+The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}.
 According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
@@
 The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest.
-However, these methods are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{partial} are mainly for nonparametric and black box models (though they can be used for simple parametric models as well).
+However, these methods are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are mainly for nonparametric and black box models (though they can be used for simple parametric models as well).
 
@@
 
-As described in the introduction, the \pkg{partial} package is useful for constructing PDPs for many types of fitted models in R.
+As described in the introduction, the \pkg{pdp} package is useful for constructing PDPs for many types of fitted models in R.
 These are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest.
-The latest stable release is available from \href{https://cran.r-project.org/package=partial}{CRAN}:
+The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
 \begin{example}
-install.packages("partial")
+install.packages("pdp")
 \end{example}
-The development version is located on GitHub: \url{https://github.com/bgreenwell/partial}.
-Bug reports and usggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/partial/issues}.
+The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}.
+Bug reports and usggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/pdp/issues}.
 
-Currently, only two functions are exported by \pkg{partial}:
+Currently, only two functions are exported by \pkg{pdp}:
 \begin{itemize}
@@
 \end{example}
-The only difference is that \pkg{partial} uses the \pkg{lattice} graphics package to produce all of its displays.
+The only difference is that \pkg{pdp} uses the \pkg{lattice} graphics package to produce all of its displays.
 
@@
 However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying partial dependence plots.
-Fortunately, the \pkg{partial} package can be used to fill this gap.
+Fortunately, the \pkg{pdp} package can be used to fill this gap.
 
@@
 
-In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{partial} package.
+In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package.
 We also briefly discussed related approaches availble in other R packages as well.
 Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples.
-There is a bright future for \pkg{partial} in terms of growth.
+There is a bright future for \pkg{pdp} in terms of growth.
 For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.

### 2016-09-02 manuscript/greenwell.tex
finished? XGBoost example

@@
 The results are displayed in Figure~\ref{fig:boston_xgb}.
+Note that while \code{xgboost} requires the training data to be a matrix (or an object of class \code{"dgCMatrix"} or \code{"xgb.DMatrix"}), \code{partial} requires the training data as a data frame with factors in tact!
 \begin{example}
-partial(boston.xgb, pred.var = "lstat", train = subset(boston, select = -cmedv),
+pred.list <- list("lstat", "rm", c("lstat", "rm"))
+for (x in pred.list) {
+  partial(boston.xgb, pred.var = x, train = subset(boston, select = -cmedv),
           plot = TRUE)
+}
 \end{example}

### 2016-09-02 manuscript/greenwell.tex
minor updates

@@
 
-% FIXME %%%%%%%%%%%%%%%%%%%%%
+Currently supported models are described in Table~\ref{tab:models}.
+However, the default method \code{partial.default} is flexible enough to handle many of the model types not listed in in this table; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+Although, for these cases, the user will need to specify the additional options \code{super.type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with \code{"xgb.Booster"} models, which requires the user to specify \code{train}.
 
-Currently supported models are described in Table~\ref{tab:models}.
-Using the \code{partial.default} method it is still possible to handle many other models not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
-However, for these cases, the user will need to specify the additional options \code{super.type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with \code{"xgb.Booster"} models, which requires the user to specify \code{train}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{table}[htbp]
@@
       Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
+      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
       Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\

### 2016-09-02 R/partial.R

@@ -14,7 +14,7 @@
 #'   the continuous independent variables listed in \code{pred.var}.
 #' @param super.type Character string specifying the type of supervised learning.
 #'   Current options are \code{"regression"} or \code{"classification"}. For some
-#'   objects (e.g., tree-based models like \code{"rpart"}), \code{partial} can usually 
+#'   objects (e.g., tree-based models like \code{"rpart"}), \code{partial} can usually
 #'   extract the necessary information from \code{object}.
 #' @param which.class Integer specifying which column of the matrix of predicted
 #'   probabilities to use as the "focus" class. Default is to use the first class.
@@ -40,7 +40,7 @@ partial <- function(object, ...) {
 #' @rdname partial
 #' @export
 partial.default <- function(object, pred.var, pred.grid, grid.resolution = NULL,
-                            super.type, which.class = 1L, plot = TRUE, 
+                            super.type, which.class = 1L, plot = FALSE,
                             rug = FALSE, chull = FALSE, train, ...) {
 
   # Data frame

### 2016-09-02 R/utils.R

@@ -40,10 +40,10 @@ superType.boosting <- function(object) {
 
 #' @keywords internal
 superType.earth <- function(object) {
-  if (!is.null(object$glm.list) && 
+  if (!is.null(object$glm.list) &&
       object$glm.list[[1L]]$family$family == "binomial") {
     "classification"
-  } else if (is.null(object$glm.list) || 
+  } else if (is.null(object$glm.list) ||
              object$glm.list[[1L]]$family$family == "gaussian") {
     "regression"
   } else {
@@ -55,6 +55,12 @@ superType.earth <- function(object) {
 #' @keywords internal
 superType.lm <- function(object) {
   # FIXME: What about multivariate response models?
+  "regression"
+}
+
+
+#' @keywords internal
+superType.nls <- function(object) {
   "regression"
 }
 
@@ -141,14 +147,14 @@ superType.gbm <- function(object) {
 superType.xgb.Booster <- function(object) {
   if (object$params$objective == "reg:linear") {
     "regression"
-  } else if (object$params$objective %in% 
+  } else if (object$params$objective %in%
              c("binary:logistic", "multi:softprob")) {
-    # FIXME: Throw a warning if objective function is classification, but does 
+    # FIXME: Throw a warning if objective function is classification, but does
     # not return the predicted probabilities (e.g., "binary:logitraw").
     "classification"
-  } else if (object$params$objective %in% 
+  } else if (object$params$objective %in%
              c("reg:logistic", "binary:logitraw", "multi:softmax")) {
-    stop(paste0("For classification, switch to an objective function", 
+    stop(paste0("For classification, switch to an objective function",
                 "that returns the predicted probabilities."))
   } else {
     "other"
@@ -162,7 +168,7 @@ superType.svm <- function(object) {
     "regression"
   } else {
     "classification"
-  } 
+  }
 }
 
 

### 2016-09-04 manuscript/greenwell.tex
minor updates

@@
 
-\strong{Mention the integral approximation!!}
-
 Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_n\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$.
-If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_{c} = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
+If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_c = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
 \begin{equation}
@@
 where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, we average out the effects of all the other predictors in the model.
-Let $x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$.
+Equation~\eqref{eqn:pdf} approximates the average function
+\begin{equation*}
+  E_{\boldsymbol{z}_c}\left[\widehat{f}\left(\boldsymbol{z}_s\right)\right] = \int \widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)p_{c}\left(\boldsymbol{z}_c\right)d\boldsymbol{z}_c,
+\end{equation*}
+where $p_{c}\left(\boldsymbol{z}_c\right)$ is the marginal probability density of $\boldsymbol{z}_c$: $p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s$.
+%\begin{equation*}
+%  p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s.
+%\end{equation*}
+
+To construct a PDP in practice, we can proceed as follows.
+To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$.
 The partial dependence of the response on $x_1$ can be constructed as follows:
@@
 The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}.
-Bug reports and usggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/pdp/issues}.
+Bug reports and suggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/pdp/issues}.
 
@@
       \midrule
-      Decision tree             & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
-                                & \pkg{party}    & \code{"BinaryTree"} \\
+      Decision tree             & \pkg{party}    & \code{"BinaryTree"} \\
                                 & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
-      Bagged decision trees     & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
-                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
-      Boosted decision trees    & \pkg{gbm}      & \code{"gbm"} \\
+                                & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
+      Bagged decision trees     & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"bagging"} \\
+                                & \CRANpkg{ipred} \citep{ipred-pkg} & \code{"classbagg"}, \code{"regbagg"} \\
+      Boosted decision trees    & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
+                                & \pkg{gbm}      & \code{"gbm"} \\
                                 & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
-                                & \CRANpkg{adabag} \citep{adabag-pkg} & \code{"boosting"} \\
       Random forest & \pkg{randomForest} & \code{"randomForest"} \\
@@
   \end{tabular}
-  \caption{Models specifically supported by the \pkg{package}. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} fr details.}
+  \caption{Models specifically supported by the \pkg{pdp} package. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} for details.}
   \label{tab:models}
@@
   tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, train = iris)
+                 which.class = i, .progress = "text",
+                 grid.resolution = 101)
   pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))

### 2016-09-04 manuscript/greenwell.R

@@ -171,7 +171,8 @@ iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
 pd <- NULL
 for (i in 1:3) {
   tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, .progress = "text")
+                 which.class = i, .progress = "text",
+                 grid.resolution = 101)
   pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
 }
 pdf("partial_iris_svm.pdf", width = 12, height = 4)

### 2016-09-05 manuscript/greenwell.tex
spell check fixes
final draft?

@@
 \abstract{
-Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g.
-multiple linear regression).
+Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression).
 Unfortunately, it can be challenging for the uninitiated to understand the results of such models and explain them to management.
@@
 Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood.
-These plots are especially useful in explaining the output from complex black box models like support vector machines or neural networks or constructing simpler parsimonious models on newly obtained data.
-In this paper, we introduce partial, a general R package for creating partial dependence plots.
+These plots are especially useful in explaining the output from black box models or constructing simpler parsimonious models on newly obtained data.
+In this paper, we introduce \pkg{partial}, a general R package for constructing partial dependence plots
 }
@@
 
-Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_n\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$.
-If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_c = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the partial dependence of the response on $\boldsymbol{z}_s$ is defined as
+Let $\boldsymbol{x} = \left\{x_1, x_2, \dots, x_p\right\}$ represent the predictors in a model whose prediction function is $\widehat{f}\left(\boldsymbol{x}\right)$.
+If we partition $\boldsymbol{x}$ into an interest set, $\boldsymbol{z}_s$, and its compliment, $\boldsymbol{z}_c = \boldsymbol{x} \setminus \boldsymbol{z}_s$, then the "partial dependence" of the response on $\boldsymbol{z}_s$ is defined as
+\begin{equation}
+\label{eqn:avg_fun}
+  f_s\left(\boldsymbol{z}_s\right) = E_{\boldsymbol{z}_c}\left[\widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)\right] = \int \widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)p_{c}\left(\boldsymbol{z}_c\right)d\boldsymbol{z}_c,
+\end{equation}
+where $p_{c}\left(\boldsymbol{z}_c\right)$ is the marginal probability density of $\boldsymbol{z}_c$: $p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s$.
+Equation~\eqref{eqn:avg_fun} can be estimated from a set of training data by
 \begin{equation}
@@
 where $\boldsymbol{z}_{i, c}$ $\left(i = 1, 2, \dots, n\right)$ are the values of $\boldsymbol{z}_c$ that occur in the training sample; that is, we average out the effects of all the other predictors in the model.
-Equation~\eqref{eqn:pdf} approximates the average function
-\begin{equation*}
-  E_{\boldsymbol{z}_c}\left[\widehat{f}\left(\boldsymbol{z}_s\right)\right] = \int \widehat{f}\left(\boldsymbol{z}_s, \boldsymbol{z}_c\right)p_{c}\left(\boldsymbol{z}_c\right)d\boldsymbol{z}_c,
-\end{equation*}
-where $p_{c}\left(\boldsymbol{z}_c\right)$ is the marginal probability density of $\boldsymbol{z}_c$: $p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s$.
-%\begin{equation*}
-%  p_{c}\left(\boldsymbol{z}_c\right) = \int p\left(\boldsymbol{x}\right)d\boldsymbol{z}_s.
-%\end{equation*}
 
-To construct a PDP in practice, we can proceed as follows.
+To construct a PDP \eqref{eqn:pdf} in practice, we can proceed as follows.
 To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$.
@@
 The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}.
-Bug reports and suggestions are appreciated and should be sumbitted to \url{https://github.com/bgreenwell/pdp/issues}.
+Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}.
 

### 2016-09-05 manuscript/greenwell.R

@@ -71,31 +71,6 @@ dev.off()
 # Fit a MARS model
 boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
 
-# Boosted regression trees
-set.seed(102)
-boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
-                      label = boston$cmedv,
-                      objective = "reg:linear",
-                      nrounds = 2000,
-                      max_depth = 3,
-                      eta = 0.01)
-
-pdf("boston_xgb.pdf", width = 12, height = 4)
-X <- subset(boston, select = -cmedv)
-pdp1 <- plotPartial(partial(boston.xgb, pred.var = "lstat", train = X),
-                    rug = TRUE, smooth = TRUE, train = X)
-pdp2 <- plotPartial(partial(boston.xgb, pred.var = "rm", train = X),
-                    rug = TRUE, smooth = TRUE, train = X)
-pdp3 <- plotPartial(partial(boston.xgb, pred.var = c("lstat", "rm"),
-                            chull = TRUE, train = X), train = X)
-gridExtra::grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
-dev.off()
-
-# Where can this figure go?
-partial(boston.xgb, pred.var = c("lstat", "rm"), chull = TRUE,
-        train = subset(boston, select = -cmedv), plot = TRUE,
-        .progress = "text")
-
 # Figure 3
 pd.lstat.rm <- partial(boston.mars, pred.var = c("lstat", "rm"))
 pdf("pd_lstat_rm.pdf", width = 12, height = 4)
@@ -140,6 +115,8 @@ ozone <- read.csv(paste0("http://statweb.stanford.edu/~tibs/ElemStatLearn/",
 # Fit a MARS model
 ozone.earth <- earth(ozone ~ ., data = ozone, degree = 3)
 
+# Note: the following example will not work on Windows!
+
 # Figure 6
 registerDoParallel(cores = 4)  # use 4 cores
 pd <- partial(ozone.earth, pred.var = c("wind", "temp", "dpg"), chull = TRUE,
@@ -167,7 +144,7 @@ dev.off()
 iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
                 cost = 0.25, probability = TRUE)
 
-# Plot partial dependence for each class
+# Figure 7
 pd <- NULL
[…]

### 2016-09-05 manuscript/greenwell.tex
better final draft?

@@
 These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct.
-For additive models (i.e., no interactions), these plots are identical in shape to partial dependence plots.
+For additive models (i.e., no interactions), these plots are identical in shape to PDPs.
 
@@
 \end{itemize}
-The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values specified by the \code{pred.var} argument.
-If \code{plot = FALSE} (the default), \code{partial} returns an object of class inheriting from \code{"partial"} which inherits from the class \code{"data.frame"}.
+The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument.
+If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}.
 The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
-If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns a \CRANpkg{lattice} plot \citep{lattice-pkg}.
-For more advanced plotting, the \code{plotPartial} function will take a \code{"partial"} object and display a more customizable lattice plot.
+If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
+For more advanced plotting, \code{plotPartial} function will take a \code{"partial"} object and display a more customizable lattice plot.
 
@@
 When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot; alternatively, you can set \code{chull = TRUE} in the call to \code{partial}, in which case only the region within the convex hull of the first two variables is plotted.
-Over interpreting the partial dependence plot outside of this region can be dangerous.
+Over interpreting the PDP outside of this region can be dangerous.
 The right display in Figure~\ref{fig:partial_extrap} was produced using:
@@
   \includegraphics[width=1.0\linewidth]{partial_extrap}
-  \caption{Examples of partial dependence plots with the addition of a rug display (left) and a convex hull (right).}
+  \caption{Examples of PDPs with the addition of a rug display (left) and a convex hull (right).}
   \label{fig:partial_extrap}
@@
 It has been shown to be many times faster than the well known \pkg{gbm} package.
-However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying partial dependence plots.
+However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying PDPs.
 Fortunately, the \pkg{pdp} package can be used to fill this gap.
@@
   \includegraphics[width=1.0\linewidth]{boston_xgb}
-  \caption{Partial dependence plots for the top two most important variables in the Boston housing data using XGBoost.
+  \caption{PDPs for the top two most important variables in the Boston housing data using XGBoost.
 Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}

### 2016-09-06 manuscript/greenwell.tex
even better final draft?

@@
 The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument.
-If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}.
+If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; in other words, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
 The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.

### 2016-09-06 manuscript/greenwell.tex
minor tweaks

@@
 \section{Constructing PDPs in R}
+\label{sec:boston}
 
@@
 For illustration, we return to the Boston housing data.
-The example below, fits an \pkg{xgboost} model; the tuning parameters were predetermined using 5-fold cross-validation.
+The code chunk below fits an \pkg{xgboost} model to the \code{boston} data frame from Section~\ref{sec:boston}; the tuning parameters were predetermined using 5-fold cross-validation.
 (Make sure you are using the latest version of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)

### 2016-09-06 manuscript/greenwell.tex
minor tweaks

@@
 We could get very reasonable results using a reduced number of points.
-Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{grid.resolution}.
+Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{grid.resolution} option in the call to \code{partial}.
 Alternatively, a specific set of values (e.g., quantiles of interest) can be supplied through the \code{pred.grid} argument.

### 2016-09-06 manuscript/greenwell.tex
minor edits

@@
 Fortunately, this can be parallelized quite easily (more on this in a later section).
-This algorithm can also be easily extended to larger subsets of 2 or more features as well.
+This algorithm can also be easily extended to larger subsets of two or more features as well.
 
@@
 If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
-For more advanced plotting, \code{plotPartial} function will take a \code{"partial"} object and display a more customizable lattice plot.
+The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options.
+Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results.
 

### 2016-09-07 manuscript/greenwell.tex
tweak abstract

@@
 Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood.
-These plots are especially useful in explaining the output from black box models or constructing simpler parsimonious models on newly obtained data.
-In this paper, we introduce \pkg{partial}, a general R package for constructing partial dependence plots
+These plots are especially useful in explaining the output from black box models.
+In this paper, we introduce \pkg{pdp}, a general R package for constructing partial dependence plots.
 }

### 2016-09-07 manuscript/greenwell.tex
add ggplot2 reference

@@
 The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options.
-Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results.
+Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results--\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
 

### 2016-09-07 manuscript/greenwell.tex
minor edits

@@
 For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function.
-\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code, as illustrated in snippet of code below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label.
+\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code, as illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label.
 \begin{example}

### 2016-09-07 manuscript/greenwell.tex
added C50 to table

@@
       \midrule
-      Decision tree             & \pkg{party}    & \code{"BinaryTree"} \\
+      Decision tree             & \CRANpkg{C50} \citep{C50-pkg} & \code{"C5.0"} \\
+                                & \pkg{party}    & \code{"BinaryTree"} \\
                                 & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\

### 2016-09-07 manuscript/greenwell.tex
added Cubist to table

--- 2016-09-07 manuscript/greenwell.tex
@@
                                 & \CRANpkg{xgboost} & \code{"xgb.Booster"} \\
+      Cubist                    & \CRANpkg{Cubist} \citep{Cubist-pkg} & \code{"cubist"} \\
       Random forest & \pkg{randomForest} & \code{"randomForest"} \\

### 2016-09-07 manuscript/greenwell.tex
switched to ggplot2 in SVM example

@@
 boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
-pd.lstat.rm <- partial(boston.mars, pred.var = c("lstat", "rm"))
-plotPartial(pd.lstat.rm)
-plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("red", "white", "blue")))
-plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
+pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
+plotPartial(pd)
+plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
+plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
             colorkey = TRUE, screen = list(z = -20, x = -60))
@@
 \begin{example}
-partial(boston.rf, plot = TRUE)
+partial(boston.rf, "rm", plot = TRUE)
 partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
@@
 }
-lattice::levelplot(y ~ Petal.Width * Petal.Length | Species, data = pd)
+
+library(ggplot2)
+ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
+  geom_tile() + 
+  # coord_equal() +
+  geom_contour(color = "white", alpha = 0.5) + 
+  scale_fill_distiller(palette = "Spectral", na.value = "white") + 
+  theme_bw() +
+  facet_grid(~ Species)
 \end{example}

### 2016-09-07 manuscript/greenwell.tex
fixed typo

@@
   geom_tile() + 
-  # coord_equal() +
   geom_contour(color = "white", alpha = 0.5) + 

### 2016-09-07 manuscript/greenwell.tex
fixed typo

@@
   geom_contour(color = "white", alpha = 0.5) + 
-  scale_fill_distiller(palette = "Spectral", na.value = "white") + 
+  scale_fill_distiller(palette = "Spectral") + 
   theme_bw() +

### 2016-09-08 manuscript/greenwell.tex
updated manuscript

@@
   tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, .progress = "text",
-                 grid.resolution = 101)
+                 which.class = i, grid.resolution = 101, .progress = "text")
   pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
@@
 for (x in pred.list) {
-  partial(boston.xgb, pred.var = x, train = subset(boston, select = -cmedv),
-          plot = TRUE)
+  partial(boston.xgb, pred.var = x, plot = TRUE,
+          train = subset(boston, select = -cmedv))
 }
@@
   United States of America\\}
-\email{bgreenwell@infoscitex.com}
+\email{greenwell.brandon@gmail.com}

### 2016-09-11 manuscript/greenwell.tex
minor edits

@@
 PDPs help visualize the relationship between a subset of the predictors (typically 1-3) and the response while accounting for the average effect of the other independent variables in the model.
-They are particularly effective with black box models like random forest and support vector machines.
-%PDPs offer a low-dimensional graphical rendering of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ for any type of fitted models and are especially useful in visualizing the relationships discovered by complex machine learning algorithms such as a \dfn{random forest}.
+They are particularly effective with black box models like random forests and support vector machines.
 
@@
 
-To construct a PDP \eqref{eqn:pdf} in practice, we can proceed as follows.
+Constructing a PDP \eqref{eqn:pdf} in practice is rather straightforward.
 To simplify, let $\boldsymbol{z}_s = x_1$ be the predictor variable of interest with unique values $\left\{x_{11}, x_{12}, \dots, x_{1k}\right\}$.
@@
 However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively).
-The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for PDPs that can be applied to several types of fitted models.
+The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for constructing PDPs that can be applied to several types of fitted models.
 
@@
 These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct.
-For additive models (i.e., no interactions), these plots are identical in shape to PDPs.
+For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs.
 
 PDPs can be misleading in the presence of substantial interactions \citep{goldstein-peeking-2015}.
-To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots -- available in the \CRANpkg{ICEbox} package.
+To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots---available in the \CRANpkg{ICEbox} package.
 \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor.
@@
 The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument.
-If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; in other words, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
-The columns are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
+If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
+The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
 If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
 The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options.
-Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results--\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
+Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
 
@@
 However, the default method \code{partial.default} is flexible enough to handle many of the model types not listed in in this table; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
-Although, for these cases, the user will need to specify the additional options \code{super.type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with \code{"xgb.Booster"} models, which requires the user to specify \code{train}.
+Although, for these cases, the user will need to specify the additional options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with the \code{"xgb.Booster"} class, which requires the user to specify \code{train}.
 
@@
 
-For a more customizable plot, call \code{partial} with \code{plot = FALSE} and use the \code{plotPartial} function.
-\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code, as illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label.
+For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame.
+This is illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label.
+The result is displayed in the right side of Figure~\ref{fig:pd_lstat}.
+\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code.
 \begin{example}
@@
 \end{example}
-The result is displayed in the right side of Figure~\ref{fig:pd_lstat} below.
 
@@
 
-The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
+The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
 For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
@@
 \end{example}
-Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.
+\strong{Note:} the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.
 
@@
 For classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp.
-369--370]{hastie-elements-2009}.
+369---370]{hastie-elements-2009}.
 Suppose the response is categorical with $K$ levels, then for each class we compute
@@
 
+To round out our discussion, we provide one last example using a recently popular (and successful!) machine learning tool.
 XGBoost, short for eXtreme Gradient Boosting, is a popular library providing optimized distributed gradient boosting that is specifically designed to be highly efficient, flexible and portable.
@@
 \begin{example}
-set.seed(102)
+set.seed(102)  # for reproducibility
 boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
@@
 for (x in pred.list) {
-  partial(boston.xgb, pred.var = x, plot = TRUE,
-          train = subset(boston, select = -cmedv))
+  partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
+          chull = TRUE, train = subset(boston, select = -cmedv))
 }
@@
 For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
-% We introduced the investr package for computing point estimates along with inversion and Waldbased confidence intervals for linear and nonlinear calibration problems with constant variance.
-We also showed how the boot package can be used for constructing approximate 100(1 􀀀 a)% calibration intervals, which for convenience, may be incorporated into a future release of investr.
-The authors are currently working on extending the package to handle the case of heteroscedastic errors, random coefficient models (e.g., objects of class "lme" from the recommended nlme package; Pinheiro et al., 2014)), and even multivariate calibration problems (e.g., objects of class "mlm").
 

### 2016-09-11 R/partial.R

@@ -21,6 +21,8 @@
 #' @param plot Logical indicating whether to return a data frame containing the
 #'   partial dependence values (\code{FALSE}) or plot the partial dependence
 #'   function directly (\code{TRUE}). Default is \code{FALSE}.
+#' @param smooth Logical indicating whether or not to overlay a LOESS smoother.
+#'   Default is \code{FALSE}.
 #' @param rug Logical indicating whether or not to include a rug representation
 #'   to the plot. Only used when \code{plot = TRUE}. Default is \code{FALSE}.
 #' @param chull Logical indicating wether or not to restrict the first
@@ -40,8 +42,9 @@ partial <- function(object, ...) {
 #' @rdname partial
 #' @export
 partial.default <- function(object, pred.var, pred.grid, grid.resolution = NULL,
-                            type, which.class = 1L, plot = FALSE, rug = FALSE, 
-                            chull = FALSE, train, ...) {
+                            type, which.class = 1L, plot = FALSE,
+                            smooth = FALSE, rug = FALSE, chull = FALSE, train,
+                            ...) {
 
   # Data frame
   if (missing(train)) {
@@ -49,9 +52,9 @@ partial.default <- function(object, pred.var, pred.grid, grid.resolution = NULL,
       train <- object@data@get("input")
     } else {
       if (is.null(object$call$data)) {
-        stop(paste0("The training data could not be extracted from ", 
+        stop(paste0("The training data could not be extracted from ",
                     deparse(substitute(object)), ". Please supply the raw ",
-                    "training data using the `train` argument in the call ", 
+                    "training data using the `train` argument in the call ",
                     "to `partial`."))
       } else {
         train <- eval(object$call$data)
@@ -64,8 +67,6 @@ partial.default <- function(object, pred.var, pred.grid, grid.resolution = NULL,
     pred.val <- lapply(pred.var, function(x) {
       if (is.factor(train[[x]])) {
         levels(train[[x]])
-      #} #else if (missing(grid.resolution)) {
-        #sort(unique(newdata[[x]]))
       } else {
         if (is.null(grid.resolution)) {
           grid.resolution <- min(length(unique(train[[x]])), 51)
@@ -92,13 +93,26 @@ partial.default <- function(object, pred.var, pred.grid, grid.resolution = NULL,
     }
   }
 
[…]

### 2016-09-11 R/plotPartial.R

@@ -3,7 +3,7 @@
 #' Plots partial dependence functions using \code{lattice} graphics.
 #'
 #' @param x An object of class{"partial_1d"} or \code{"partial_2d"}.
-#' @param smooth Logical indicating whether or nor to overlay a loess smoother.
+#' @param smooth Logical indicating whether or not to overlay a LOESS smoother.
 #'   Default is \code{FALSE}.
 #' @param contour Logical indicating whether or not to use
 #'   \code{lattice::levelplot} (\code{TRUE}) or \code{lattice::wireframe}
@@ -12,17 +12,17 @@
 #'   to the plot. If \code{TRUE} the user must supply the original training data
 #'   via the \code{train} option. Default is \code{FALSE}.
 #' @param chull Logical indicating whether or not to draw the convex hull
-#'   around the first two variables. If \code{TRUE} the user must supply the 
+#'   around the first two variables. If \code{TRUE} the user must supply the
 #'   original training data via the \code{train} option.Default is \code{FALSE}.
 #' @param number Integer specifying the number of conditional intervals for the
 #'   panel variables.
-#' @param overlap The fraction of overlap of the conditioning variables. See 
-#'   \code{?graphics::co.intervals} and \code{?lattice::equal.count} for further 
+#' @param overlap The fraction of overlap of the conditioning variables. See
+#'   \code{?graphics::co.intervals} and \code{?lattice::equal.count} for further
 #'   details.
 #' @param train Data frame containing the original training data. Only
 #'   required if \code{rug = TRUE} or \code{chull = TRUE}.
 #' @param col.regions Color vector to be used if \code{contour} is \code{TRUE}.
-#'   Defaults to the wonderful Matplotlib 'viridis' color map provided by the 
+#'   Defaults to the wonderful Matplotlib 'viridis' color map provided by the
 #'   \code{viridis} package. See \code{?viridis::viridis} for details.
 #' @param ... Additional optional arguments to be passed onto \code{levelplot},
 #'   \code{wireframe}, or \code{xyplot}.

### 2016-09-11 manuscript/greenwell.R

@@ -3,7 +3,7 @@
 #   partial: partial: An R Package for Creating Partial Dependence Plots
 #
 #        Author: Brandon M. Greenwell
-# Date Modified: 01Sep2016
+# Date Modified: 11Sep2016
 
 
 ################################################################################
@@ -18,8 +18,8 @@ pkgs <- c("doParallel",
           "e1071",
           "ggplot2",
           "mlbench",
-          "partial",
           "party",
+          "pdp",
           "randomForest",
           "xgboost")
 # install.packages(pkgs)
@@ -29,8 +29,8 @@ library(doParallel)
 library(e1071)
 library(earth)
 library(ggplot2)
-library(partial)
 library(party)
+library(pdp)
 library(randomForest)
 library(xgboost)
 
@@ -156,9 +156,9 @@ for (i in 1:3) {
 }
 pdf("partial_iris_svm.pdf", width = 12, height = 4)
 ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
-  geom_tile() + 
-  geom_contour(color = "white", alpha = 0.5) + 
-  scale_fill_distiller(palette = "Spectral") + 
+  geom_tile() +
+  geom_contour(color = "white", alpha = 0.5) +
+  scale_fill_distiller(palette = "Spectral") +
   theme_bw() +
   facet_grid(~ Species)
 dev.off()
@@ -196,6 +196,9 @@ boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                       max_depth = 3,
                       eta = 0.01)
 
+# partial(boston.xgb, pred.var = c("lon", "lat"), plot = TRUE, chull = TRUE,
+#         train = X, .progress = "text")
+
 # Figure 8
 pdf("boston_xgb.pdf", width = 12, height = 4)
 X <- subset(boston, select = -cmedv)
@@ -204,6 +207,6 @@ pdp1 <- plotPartial(partial(boston.xgb, pred.var = "lstat", train = X),
 pdp2 <- plotPartial(partial(boston.xgb, pred.var = "rm", train = X),
                     rug = TRUE, smooth = TRUE, train = X)
 pdp3 <- plotPartial(partial(boston.xgb, pred.var = c("lstat", "rm"),
-                            chull = TRUE, train = X), train = X)
+                            chull = TRUE, train = X), rug = TRUE, train = X)
 gridExtra::grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
 dev.off()

### 2016-09-13 manuscript/greenwell.tex
minor edits

@@
 The result is displayed in the right side of Figure~\ref{fig:pd_lstat}.
-\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported for writing more convenient code.
+\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded for writing more readable code.
 \begin{example}
-boston.rf %>%
+boston.rf %>%  # the %>% operator is read as "and then"
   partial(pred.var = "lstat") %>%
@@
 \end{example}
-\strong{Note:} the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.
+\strong{Note:} the default color map for contour plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.
 
@@
 \textit{Left}: Default plot.
-\textit{Middle}: Using a different color palette.
+\textit{Middle}: Using a black and white color palette.
 \textit{Right}: Using a 3-D surface.}
@@
 The results are displayed in Figure~\ref{fig:boston_xgb}.
-Note that while \code{xgboost} requires the training data to be a matrix (or an object of class \code{"dgCMatrix"} or \code{"xgb.DMatrix"}), \code{partial} requires the training data as a data frame with factors in tact!
+\strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
 \begin{example}

### 2016-09-13 manuscript/greenwell.tex
minor edits

@@
 
-It is not wise to draw conclusions from PDPs in regions outside the range of the training data.
+It is not wise to draw conclusions from PDPs in regions outside the area of the training data.
 Here we describe two ways to mitigate the risk of extrapolation in PDPs: rug displays and convex hulls.
@@
 \begin{example}
-partial(boston.rf, pred.var = "lstat", rug = TRUE, plot = TRUE)
+partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
 \end{example}
@@
 In two or more dimensions, plotting the convex hull is more informative; it outlines the region of the predictor space that the model was trained on.
-When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is added to the plot; alternatively, you can set \code{chull = TRUE} in the call to \code{partial}, in which case only the region within the convex hull of the first two variables is plotted.
+When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted.
 Over interpreting the PDP outside of this region can be dangerous.
@@
 \begin{example}
-partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE, plot = TRUE)
+partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
 \end{example}
@@
 We could get very reasonable results using a reduced number of points.
-Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points ban be controlled using the \code{grid.resolution} option in the call to \code{partial}.
-Alternatively, a specific set of values (e.g., quantiles of interest) can be supplied through the \code{pred.grid} argument.
-To illustrate, the following snippet of code computes the partial dependence of median home value on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
+Current options are to use a grid of equally spaced values in the range of the variable of interest; the number of points can be controlled using the \code{grid.resolution} option in the call to \code{partial}.
+Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument.
+To illustrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
 \begin{example}
@@
 \begin{example}
-ozone.earth <- earth(ozone ~ ., data = ozone, degree = 3)
-summary(ozone.earth)
+ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
+summary(ozone.mars)
 \end{example}
@@
 \end{itemize}
-To understand this interaction, we can use a partial plot.
-However, since the partial dependece between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.
+To understand this interaction, we can use a PDP. However, since the partial dependece between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.
 
@@
 Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded.
-This is illustrated in following snippet of code which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
+This is illustrated in code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
 The result is displayed in Figure~\ref{fig:partial_par}.
 \begin{example}
-partial(ozone.earth, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
+partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
         chull = TRUE, .parallel = TRUE)
@@
 The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}).
-Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible, as was done in the previous example.
+Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible.
 
@@
 
-To illustrate, consider Edgar Anderson's iris data from the \pkg{datasets} package.
+To illustrate, we consider Edgar Anderson's iris data from the \pkg{datasets} package.
 The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica.
-We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the model parameters were determined using 5-fold cross-validation).
+We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the tuning parameters were determined using 5-fold cross-validation).
 \begin{example}
@@
 \end{example}
-\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set the \code{probability = TRUE} option in the call to \code{svm}.
+\strong{Note:} the \code{partial} function has to be able to extract the predicted probabilities for each class, so it is necessary to set \code{probability = TRUE} in the call to \code{svm}.
 
@@
 The result is displayed in Figure~\ref{fig:partial_iris}.
-Note that we had to supply the original training data to \code{partial} through the \code{train} option.
+%Note that we had to supply the original training data to \code{partial} through the \code{train} option.
 This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
@@
 It has been shown to be many times faster than the well known \pkg{gbm} package.
-However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for displaying PDPs.
+However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs.
 Fortunately, the \pkg{pdp} package can be used to fill this gap.
@@
 Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples.
-There is a bright future for \pkg{pdp} in terms of growth.
+In terms of future development, \pkg{pdp} can be expanded in a number of ways.
 For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.

### 2016-09-13 manuscript/greenwell.tex
edited summary
more tweaks to summary

@@
 
+PDPs can be used to graphically examine the dependence of the response on low cardinality subsets of the features, accounting for the average effect of the other predictors.
 In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package.
@@
 For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
+It would also be nice to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.
 

### 2016-09-14 manuscript/greenwell.tex
minor edits

@@
 Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression).
-Unfortunately, it can be challenging for the uninitiated to understand the results of such models and explain them to management.
+Unfortunately, it can be challenging to understand the results of such models and explain them to management.
 Partial dependence plots offer a simple solution.
@@
 This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details.
-PDPs help visualize the relationship between a subset of the predictors (typically 1-3) and the response while accounting for the average effect of the other independent variables in the model.
+PDPs help visualize the relationship between a subset of the features (typically 1-3) and the response while accounting for the average effect of the other predictors in the model.
 They are particularly effective with black box models like random forests and support vector machines.
@@
 
-\begin{table}[htbp]
+\begin{table}[!htbp]
   \begin{tabular}{p{4cm}ll}
@@
 
-\begin{figure}[htbp]
+\begin{figure}[!htbp]
   \centering

### 2016-09-14 R/partial.R

@@ -28,6 +28,9 @@
 #' @param chull Logical indicating wether or not to restrict the first
 #'   two variables in \code{pred.var} to lie within the convex hull of their
 #'   data points; this effects \code{pred.grid}. Default is \code{FALSE}.
+#' @param contour Logical indicating whether or not to use
+#'   \code{lattice::levelplot} (\code{TRUE}) or \code{lattice::wireframe}
+#'   (\code{FALSE}). Default is \code{TRUE}.
 #' @param train An optional data frame containing the original training
 #'   data. This may be required depending on the class of \code{object}.
 #' @param ... Additional optional arguments to be passed onto \code{aaply}.

### 2016-09-16 manuscript/greenwell.R
minor edits

@@ -3,7 +3,7 @@
 #   partial: partial: An R Package for Creating Partial Dependence Plots
 #
 #        Author: Brandon M. Greenwell
-# Date Modified: 11Sep2016
+# Date Modified: 16Sep2016
 
 
 ################################################################################
@@ -11,18 +11,24 @@
 ################################################################################
 
 # Set working directory
-setwd("manuscript")
+# setwd("manuscript")
 
 # List of packages required to run all the examples in this script
-pkgs <- c("doParallel",
-          "e1071",
-          "ggplot2",
-          "mlbench",
-          "party",
-          "pdp",
-          "randomForest",
-          "xgboost")
+# pkgs <- c("doParallel",
+#           "e1071",
+#           "earth",
+#           "ggplot2",
+#           "mlbench",
+#           "party",
+#           "pdp",
+#           "randomForest")
 # install.packages(pkgs)
+
+# Install latest version of xgboost
+# install.packages("drat", repos = "https://cran.rstudio.com")
+# drat:::addRepo("dmlc")
+# install.packages("data.table")  # required when installing xgboost from source!
+# install.packages("xgboost", repos = "http://dmlc.ml/drat/", type = "source")
 
 # Load required packages
 library(doParallel)
@@ -77,9 +83,9 @@ boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
 pd.lstat.rm <- partial(boston.mars, pred.var = c("lstat", "rm"))
 pdf("pd_lstat_rm.pdf", width = 12, height = 4)
 pdp1 <- plotPartial(pd.lstat.rm)
-# pdp2 <- plotPartial(pd.lstat.rm,
-#                     col.regions = colorRampPalette(c("red", "white", "blue")))
-pdp2 <- plotPartial(pd.lstat.rm, col.regions = grey.colors)
+pdp2 <- plotPartial(pd.lstat.rm,
+                    col.regions = colorRampPalette(c("red", "white", "blue")))
+# pdp2 <- plotPartial(pd.lstat.rm, col.regions = grey.colors)
 pdp3 <- plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
                     colorkey = FALSE, screen = list(z = -20, x = -60))
 # print(p1, position = c(0, 0, 0.5, 1), more = TRUE)
[…]

### 2016-09-19 manuscript/greenwell.tex
minor edits

@@
 
-Predictor importance is an important task in any supervised learning problem.
+Determining predictor importance is a crucial task in any supervised learning problem.
 However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response.
@@
 
-Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are limited in the sense that they only apply to the models fit using the respective package.
-For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"}.
+Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are all limited in the sense that they only apply to the models fit using the respective package.
+For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"} and the \code{plot} function in \pkg{gbm} only applies to \code{"gbm"} objects.
 While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space.
@@
 \textit{Left}: Default plot.
-\textit{Middle}: Using a black and white color palette.
+\textit{Middle}: Using a user-specified color palette.
 \textit{Right}: Using a 3-D surface.}

### 2016-09-19 manuscript/greenwell.tex
minor edits

@@
 According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
-These plots allow for up to two variables at a time and are less accurate than PDPs, but are faster to construct.
+These plots allow for up to two variables at a time.
+They are also less accurate than PDPs, but are faster to construct.
 For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs.
@@
 For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing partial-residual and marginal-model plots.
-The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest.
-However, these methods are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are mainly for nonparametric and black box models (though they can be used for simple parametric models as well).
+The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest; it provides tabular and graphical displays for the terms in parametric models.
+However, these methods were designed for and are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).
 
@@
 If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
-The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options.
+The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options (e.g., the option to add a LOESS smooth).
 Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.

### 2016-09-19 manuscript/greenwell.tex
minor edits

@@
 \begin{example}
-library(partial)
+library(pdp)
 partial(boston.rf, pred.var = "lstat", plot = TRUE)
@@
 When \code{chull = TRUE}, the convex hull of the first two dimensions of $\boldsymbol{z}_s$ (i.e., the first two variables supplied to \code{pred.var}) is computed; for example, if you set \code{chull = TRUE} in the call to \code{partial} only the region within the convex hull of the first two variables is plotted.
-Over interpreting the PDP outside of this region can be dangerous.
+Over interpreting the PDP outside of this region is considered extrapolation and is ill-advised.
 The right display in Figure~\ref{fig:partial_extrap} was produced using:
@@
 Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument.
-To illustrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
+To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
 \begin{example}
@@
 
-The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than for loops.
-This makes it possible to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel.
+The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops.
+This makes it easy to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel.
 In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package.
@@
 The intervals may be allowed to overlap.} are produced first using the equal count algorithm (see, for example, \code{?lattice::equal.count}).
-Hence, it is probably better to use categorical variables to define the panels in higher dimensional displays when possible.
+Hence, it will be more effective to use categorical variables to define the panels in higher dimensional displays when possible.
 
@@
 
-To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response!) in the call to \code{partial}.
+To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}.
 The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
@@
 In this paper, we showed how to construct PDPs for various types of black box models in R using the \pkg{pdp} package.
-We also briefly discussed related approaches availble in other R packages as well.
-Suggestions to avoid extrpolation and high excecution times were discussed and demonstrated via examples.
+We also briefly discussed related approaches available in other R packages.
+Suggestions to avoid extrapolation and high exececution times were discussed and demonstrated via examples.
+
 In terms of future development, \pkg{pdp} can be expanded in a number of ways.
 For example, it would be useful to have the ability to construct PDPs for black box survival models---like conditional random forests with censored response.
-It would also be nice to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.
+It would also be worthwhile to implement the partial dependence-based $H$-statistic \citep{friedman-2008-predictive} for assessing the strength of interaction between predictors.
 

### 2016-09-19 manuscript/greenwell.tex
minor edits

@@
 Currently supported models are described in Table~\ref{tab:models}.
-However, the default method \code{partial.default} is flexible enough to handle many of the model types not listed in in this table; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
-Although, for these cases, the user will need to specify the additional options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data) in the call to \code{partial}---the exception in Table~\ref{tab:models} is using partial with the \code{"xgb.Booster"} class, which requires the user to specify \code{train}.
+In a lot of these cases, \code{partial} should be able to automatically determine \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data).
+In other cases, the user may need to specify these additional arguments.
+Using these additional arguments, it is possible to handle many of the model types not listed in in this table; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 
@@
 \end{table}
+
+When additional arguments are necessary, the user will be prompted with an informative error message. For instance, the user may occasionally see the following:
+\begin{example}
+Error: The training data could not be extracted from object. Please supply the raw training data using the `train` argument in the call to `partial`.
+\end{example}
+This will occure, for example, when using partial with the \code{"xgb.Booster"} class.
+In fact, using \code{partial} with any object that does not store the original training data will result in the above error message.
 

### 2016-09-19 manuscript/greenwell.tex
minor edits

@@
 In a lot of these cases, \code{partial} should be able to automatically determine \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data).
-In other cases, the user may need to specify these additional arguments.
-Using these additional arguments, it is possible to handle many of the model types not listed in in this table; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+In other situations, the user may need to specify these additional arguments.
+This allows \code{partial} to be flexible enough to handle many of the model types not listed in in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 

### 2016-09-19 manuscript/greenwell.tex
minor edits

@@
 Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}.
-
 Currently, only two functions are exported by \pkg{pdp}:
@@
 \end{itemize}
-The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the predictors are specified using the \code{pred.var} argument.
+
+The \code{partial} function evaluates the partial dependence \eqref{eqn:pdf} from a fitted model over a grid of predictor values; the fitted model and predictors are specified using the \code{object} and \code{pred.var} arguments, respectively---these are the only required arguments.
 If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
@@
 If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
+
 The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options (e.g., the option to add a LOESS smooth).

### 2016-09-20 manuscript/greenwell.tex
minor edits

@@
 
-The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options (e.g., the option to add a LOESS smooth).
-Of course, since the default output produced by \code{partial} is still a data frame, the user can use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
+The \code{plotPartial} function can be used for displaying more advanced PDPs; it operates on objects of class \code{"partial"} and has many useful plotting options.
+For example, \code{plotPartial} makes it straight forward to add a LOESS smooth, or produce a 3-D surface instead of a false color level plot (the default).
+Of course, since the default output produced by \code{partial} is still a data frame, the user can easily use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
 
 Currently supported models are described in Table~\ref{tab:models}.
-In a lot of these cases, \code{partial} should be able to automatically determine \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data).
-In other situations, the user may need to specify these additional arguments.
+In a lot of these cases, \code{partial} should be able to automatically determine appropriate values for the options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data).
+In other situations, the user may need to specify one or both of these additional arguments.
 This allows \code{partial} to be flexible enough to handle many of the model types not listed in in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
@@
 boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
+rwb <- colorRampPalette(c("red", "white", "blue"))
 pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
 plotPartial(pd)
-plotPartial(pd, col.regions = colorRampPalette(c("red", "white", "blue")))
-plotPartial(pd, contour = FALSE, zlab = "cmedv", drape = TRUE,
+plotPartial(pd, contour = TRUE, col.regions = rwb)
+plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
             colorkey = TRUE, screen = list(z = -20, x = -60))
 \end{example}
-\strong{Note:} the default color map for contour plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.
+\strong{Note:} the default color map for level plots is the color blind-friendly matplotlib \citep{hunter-matplotlib-2007} 'viridis' color map provided by the \CRANpkg{viridis} package \citep{viridis-pkg}.
 
@@
 \textit{Left}: Default plot.
-\textit{Middle}: Using a user-specified color palette.
+\textit{Middle}: Adds contour lines and uses a different color palette.
 \textit{Right}: Using a 3-D surface.}

### 2016-09-20 manuscript/greenwell.R

@@ -11,7 +11,7 @@
 ################################################################################
 
 # Set working directory
-# setwd("manuscript")
+setwd("manuscript")
 
 # List of packages required to run all the examples in this script
 # pkgs <- c("doParallel",
@@ -80,14 +80,15 @@ dev.off()
 boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
 
 # Figure 3
+rwb <- colorRampPalette(c("red", "white", "blue"))
 pd.lstat.rm <- partial(boston.mars, pred.var = c("lstat", "rm"))
 pdf("pd_lstat_rm.pdf", width = 12, height = 4)
 pdp1 <- plotPartial(pd.lstat.rm)
-pdp2 <- plotPartial(pd.lstat.rm,
-                    col.regions = colorRampPalette(c("red", "white", "blue")))
+pdp2 <- plotPartial(pd.lstat.rm, contour = TRUE, col.regions = rwb)
 # pdp2 <- plotPartial(pd.lstat.rm, col.regions = grey.colors)
-pdp3 <- plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
-                    colorkey = FALSE, screen = list(z = -20, x = -60))
+pdp3 <- plotPartial(pd.lstat.rm, levelplot = FALSE, zlab = "cmedv",
+                    drape = TRUE, colorkey = FALSE,
+                    screen = list(z = -20, x = -60))
 # print(p1, position = c(0, 0, 0.5, 1), more = TRUE)
 # print(p2, position = c(0.5, 0, 1, 1))
 gridExtra::grid.arrange(pdp1, pdp2, pdp3, ncol = 3)

### 2016-09-21 manuscript/greenwell.tex
minor edits

@@
   \begin{enumerate}
-    \item Create a copy of the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
+    \item Copy the training data and replace the original values of $x_1$ with the constant $x_{1i}$.
     \item Compute the vector of predicted values from the modified copy of the training data.
@@
 Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., multiple linear regression and neural networks).
-However, to our knowledge, there is no general package for constructing PDPs  in R (e.g., PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively).
-The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for constructing PDPs that can be applied to several types of fitted models.
+However, to our knowledge, there is no general package for constructing PDPs  in R.
+For example, PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively.
+The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for constructing PDPs that can be applied to several classes of fitted models.
 
 The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}.
-According to the author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
+According to its author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
 These plots allow for up to two variables at a time.
@@
 \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor.
-The ability to construct derivative plots is also available.
+The ability to construct derivative plots is also available in \pkg{ICEbox}.
 
@@
 For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing partial-residual and marginal-model plots.
-The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest; it provides tabular and graphical displays for the terms in parametric models.
-However, these methods were designed for and are orientated towards simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).
+The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest.
+It provides tabular and graphical displays for the terms in parametric models.
+However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).
 
@@
 
-As described in the introduction, the \pkg{pdp} package is useful for constructing PDPs for many types of fitted models in R.
-These are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest.
+The \pkg{pdp} package is useful for constructing PDPs for many classes of fitted models in R.
+PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest.
 The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
@@
 In other situations, the user may need to specify one or both of these additional arguments.
-This allows \code{partial} to be flexible enough to handle many of the model types not listed in in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+These two arguments allow \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 
@@
   \end{tabular}
-  \caption{Models specifically supported by the \pkg{pdp} package. \strong{Note:} package \pkg{xgboost} is only partially supported; see Section~\ref{sec:xgboost} for details.}
+  \caption{Models specifically supported by the \pkg{pdp} package. \strong{Note:} for some of these cases, the user may still need to supply additional arguments in the call to \code{partial}.}
   \label{tab:models}
@@
 \end{example}
-This will occure, for example, when using partial with the \code{"xgb.Booster"} class.
+This will occur, for example, when using partial with the \code{"xgb.Booster"} class.
 In fact, using \code{partial} with any object that does not store the original training data will result in the above error message.
@@
 
-As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor.
+As previously mentioned, the \code{randomForest} package has its own \code{partialPlot} function for visualizing the partial dependence of the response on a single predictor---the keywords here are "single predictor".
 For example, the following snippet of code plots the partial dependence of \code{cmedv} on \code{lstat}:
@@
 For a more customizable plot, we can set \code{plot = FALSE} in the call to \code{partial} and then use the \code{plotPartial} function on the resulting data frame.
-This is illustrated in the example below which increases the line width, adds a LOESS smoother, and customizes the $y$-axis label.
+This is illustrated in the example below which increases the line width, adds a LOESS smooth, and customizes the $y$-axis label.
 The result is displayed in the right side of Figure~\ref{fig:pd_lstat}.
-\strong{Note:} the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded for writing more readable code.
+\strong{Note:} to encourage writing more readable code, the \dfn{pipe} operator \code{\%>\%} provided by the \CRANpkg{magrittr} package \citep{magrittr-pkg} is exported whenever \pkg{pdp} is loaded.
 \begin{example}

### 2016-09-21 manuscript/greenwell.tex
final draft complete

@@
 Several strategies are available to ease the computational burden in larger problems.
-For example, there is no need to compute partial dependence of median home value using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!).
+For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!).
 We could get very reasonable results using a reduced number of points.
@@
 \end{itemize}
-To understand this interaction, we can use a PDP. However, since the partial dependece between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.
+To understand this interaction, we can use a PDP. However, since the partial dependence between three continuous variables can be computationally expensive, we will run \code{partial} in parallel.
 
 Setting up a parallel backend is rather straightforward.
-To illustrate, the following snippet of sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel} package.
+To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel} package.
 \begin{example}
@@
 \end{example}
-
 Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded.
-This is illustrated in code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
+This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
 The result is displayed in Figure~\ref{fig:partial_par}.
@@
 The \code{iris} data frame contains the sepal length, sepal width, petal length, and petal width (in centimeters) for 50 flowers from each of three species of iris:  setosa, versicolor, and virginica.
-We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} (the tuning parameters were determined using 5-fold cross-validation).
+We fit a support vector machine with a Gaussian radial basis function kernel to the data using the \code{svm} function in the \pkg{e1071} package (the tuning parameters were determined using 5-fold cross-validation).
 \begin{example}

### 2016-09-21 manuscript/greenwell.tex
switched to RF example

@@
 The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
-For example, the following code chunk uses a multivariate adaptive regression splines (MARS) model \citep{friedman-1991-mars} to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
+For example, the following code chunk assesses the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
 The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
 \begin{example}
-library(earth)  # for fitting MARS models
-boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
 rwb <- colorRampPalette(c("red", "white", "blue"))
-pd <- partial(boston.mars, pred.var = c("lstat", "rm"))
+pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
 plotPartial(pd)
@@
   \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a third degree MARS model.
+  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm}.
 \textit{Left}: Default plot.
-\textit{Middle}: Adds contour lines and uses a different color palette.
+\textit{Middle}: With contour lines and a different color palette.
 \textit{Right}: Using a 3-D surface.}

### 2016-09-21 manuscript/greenwell.tex
reworded sentense
minor edit

@@
 The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
-For example, the following code chunk assesses the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
+For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
 The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
@@
 
-Next, we use MARS to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
+Next, we use the multivariate adaptive regression splines (MARS) algorithm introduced in \citet{friedman-1991-mars} to model ozone concentration as a nonlinear function of the eight meteorological variables plus day of the year; we allow for up to three-way interactions.
 \begin{example}
@@
 \end{example}
-The model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
+The MARS model produced a generalized $R^2$ of $0.79$, similar to what was reported in \citet{breiman-1985-estimating}. A single three-way interaction was found involving the predictors
 \begin{itemize}

### 2016-09-22 manuscript/greenwell.tex
xgboost version warning
final draft

@@
   \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
-  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm}.
+  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a random forest.
 \textit{Left}: Default plot.
@@
 The code chunk below fits an \pkg{xgboost} model to the \code{boston} data frame from Section~\ref{sec:boston}; the tuning parameters were predetermined using 5-fold cross-validation.
-(Make sure you are using the latest version of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)
+(Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)
 \begin{example}

### 2016-09-22 manuscript/greenwell.tex
final draft

@@
 
-\title{pdp: An R Package for Creating Partial Dependence Plots}
+\title{pdp: An R Package for Constructing Partial Dependence Plots}
 \author{by Brandon M. Greenwell}
@@
 We also briefly discussed related approaches available in other R packages.
-Suggestions to avoid extrapolation and high exececution times were discussed and demonstrated via examples.
+Suggestions to avoid extrapolation and high execution times were discussed and demonstrated via examples.
 

### 2016-11-27 README.Rmd
better README

[![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/pdp)](https://cran.r-project.org/package=pdp)
[![Build Status](https://travis-ci.org/bgreenwell/pdp.svg?branch=master)](https://travis-ci.org/bgreenwell/pdp)
[![Coverage Status](https://img.shields.io/codecov/c/github/bgreenwell/pdp.svg)](https://codecov.io/github/bgreenwell/pdp?branch=master)
[![Downloads](http://cranlogs.r-pkg.org/badges/pdp)](http://cranlogs.r-pkg.org/badges/pdp)
[![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/pdp)](http://cranlogs.r-pkg.org/badges/grand-total/pdp)

pdp: An R Package for Constructing Partial Dependence Plots
================

The primary purpose of this package is to provide a general framework for constructing _partial dependence plots_ (PDPs) in R.

### Installation

The R package `pdp` is available from [CRAN](https://cran.r-project.org/package=pdp); the development version is hosted on [GitHub](https://github.com/bgreenwell/pdp). There are two ways to install:
```{r, eval=FALSE}
# Install latest release from CRAN (recommended)
install.packages("pdp")

# Install development version from GitHub repo
devtools::install_github("bgreenwell/pdp")
```

### Example usage

The examples below demonstrate various usages of the `pdp` package: regression, classification, and interfacing with the well-known `caret` package. To start, we need to install a few additional packages that will be required to run the examples.
```{r, eval=FALSE}
install.packages(c("caret", "ggmap", "kernlab", "randomForest"))
```

#### Regression example

In this example, we fit a random forest to the Boston housing data. Note, for any of the following examples, you can see a progress bar by simply specifying `.progress = "text"` in the call to `partial`.
```{r, message = FALSE, warning=FALSE}
# Load required packages
library(ggmap)  # for plotting data on maps
library(pdp)  # for constructing PDPs
library(randomForest)  # for random forest algorithm

# Load the Boston housing data
data (boston)  # included with the pdp package

# Fit a random forest to the boston housing data
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston)

# Using randomForest's partialPlot function
partialPlot(boston.rf, pred.data = boston, x.var = "lstat")

# Package pdp's partial function returns a data frame by default
head(partial(boston.rf, pred.var = "lstat"))  # returns a data frame

# Set `plot = TRUE` to return a "trellis" object
partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)

# The partial function allows for multiple predictors
partial(boston.rf, pred.var = c("lstat", "rm"), grid.resolution = 10,
        plot = TRUE, chull = TRUE)

# The plotPartial function offers more flexible plotting
pd <- partial(boston.rf, pred.var = c("lstat", "rm"), grid.resolution = 10)

# Default plot
plotPartial(pd)

# Customized plot
plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = FALSE, screen = list(z = -20, x = -60))

# Look at partial dependence of median home value on location
pd.loc <- partial(boston.rf, pred.var = c("lon", "lat"), chull = TRUE)

# Overlay predictions on a map of Boston
# ll <- c(range(boston$lon), range(boston$lat))
# map <- get_map(location = ll[c(1, 3, 2, 4)], zoom = 11, maptype = "toner-lite")
# ggmap(map) + 
#   geom_point(aes(x = lon, y = lat), data = boston, alpha = 0.2) +
#   geom_tile(aes(x = lon, y = lat, z = y, fill = y), 
#             data = pd.loc, alpha = 0.3) +
#   geom_contour(color = "white", alpha = 0.5) +
#   scale_fill_distiller(palette = "Spectral", name = "Median\nvalue") +
#   coord_fixed(ratio = 1) +
#   labs(x = "Longitude", y = "Latitude")
```

#### Classification example

In this example, we fit another random forest to the Pima Indians diabetes data.
```{r, message=FALSE, warning=FALSE}
# Fit a random forest to the Pima Indians diabetes data
data (pima)  # load the boston housing data
set.seed(102)  # for reproducibility
pima.rf <- randomForest(diabetes ~ ., data = pima, na.action = na.omit)

# Partial dependence of glucose on diabetes test result (neg/pos)
partial(pima.rf, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE)
```

#### Interface with `caret`

Finally, we demonstrate the seemless interface between the `pdp` package and the `caret` package. For illustration we use `caret`'s `train` function to tune a support vector machine (SVM) with a radial basis function kernel to the Pima Indians diabtes data. We then cinstruct PDPs from the final model for the most "important" variables. When training a model using `caret`'s `train` function, you can view tuning progress by setting `verboseIter = TRUE` in the call to `trainControl`.
```{r, message=FALSE, warning=FALSE}
# Load required packages
library(caret)  # for model training/tuning

# Set up for 5-fold cross-validation
ctrl <- trainControl(method = "cv", number = 5)

# Tune a support vector machine (SVM) using a radial basis function kerel to
# the Pima Indians diabetes data
set.seed(103)  # for reproducibility
pima.svm <- train(diabetes ~ ., data = pima, method = "svmRadial",
                  prob.model = TRUE, na.action = na.omit, trControl = ctrl,
                  tuneLength = 10)

# Partial dependence of glucose on diabetes test result (neg/pos)
partial(pima.svm, pred.var = "glucose", plot = TRUE, rug = TRUE)
```

### 2016-11-27 README.Rmd
oops, forgot .md file

@@
+---
+output: 
+  html_document: 
+    keep_md: yes
+---
 [![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/pdp)](https://cran.r-project.org/package=pdp)

### 2016-11-28 README.Rmd
better tests and README

@@
 Note, for any of the following examples, you can see a progress bar by simply specifying `.progress = "text"` in the call to `partial`.
+You may also reduce the computation time via the `grid.resolution` option in the call to `partial`.
 ```{r, message = FALSE, warning=FALSE}
@@
 # The partial function allows for multiple predictors
-partial(boston.rf, pred.var = c("lstat", "rm"), grid.resolution = 10,
-        plot = TRUE, chull = TRUE)
+partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
 
 # The plotPartial function offers more flexible plotting
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"), grid.resolution = 10)
+pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
 
@@
 
-Finally, we demonstrate the seemless interface between the `pdp` package and the `caret` package.
-For illustration we use `caret`'s `train` function to tune a support vector machine (SVM) with a radial basis function kernel to the Pima Indians diabtes data.
-We then cinstruct PDPs from the final model for the most "important" variables.
+Finally, we demonstrate the seamless interface between the `pdp` package and the `caret` package; `caret` is an extremetly useful package for classification and regression training that, essentially, has one function (`train`) for fitting all kinds of predictive models in R (e.g., `glmnet`, `svm`, `xgboost`, etc.).
+For illustration we use `caret`'s `train` function to tune a support vector machine (SVM) with a radial basis function kernel to the Pima Indians diabetes data.
+We then construct PDPs from the final model for the most "important" variables.
 When training a model using `caret`'s `train` function, you can view tuning progress by setting `verboseIter = TRUE` in the call to `trainControl`.

### 2016-11-28 tests/testthat/test-partial.R

@@ -1,27 +1,48 @@
 context("Partial dependence functions")
 
 # Data frames used in the following tests
-#   regression: mtcars
-#   classification: iris
+pima2 <- na.omit(pima)
+set.seed(101)
+df.reg <- boston[sample(nrow(boston), size = 50, replace = FALSE), ]
+df.class <- pima2[sample(nrow(pima2), size = 50, replace = FALSE), ]
+
+# Switch to generic response names
+df.reg$y <- df.reg$cmedv
+df.class$y <- df.class$diabetes
+df.reg$cmedv <- NULL
+df.class$diabetes <- NULL
 
 test_that("partial works correctly", {
 
   # Regression
-  mtcars.lm <- lm(mpg ~ ., data = mtcars)
-  pd1 <- partial(mtcars.lm, pred.var = "wt", grid.resolution = 1,
-                 type = "regression", train = mtcars)
-  pd2 <- partial(mtcars.lm, pred.var = "wt", pred.grid = 1, check.class = FALSE)
-  expect_is(pd1, "data.frame")
-  expect_is(pd2, "data.frame")
-  expect_is(partial(mtcars.lm, pred.var = "wt", plot = TRUE), "trellis")
+  df.reg.lm <- lm(y ~ .,
+                  data = df.reg)
 
   # Classification
-  if (require(randomForest, quietly = TRUE)) {
-    iris.rf <- randomForest(Species ~ ., data = iris)
-    iris.un <- randomForest( ~ ., data = iris)
-    pd <- partial(iris.rf, pred.var = "Petal.Width", grid.resolution = 1)
-    expect_is(pd, "data.frame")
-    expect_error(partial(iris.un, pred.var = "Petal.Width", grid.resolution = 1))
-  }
+  df.class.glm <- glm(y ~ .,
+                      data = df.class,
+                      family = binomial)
+
+  # PDPs
+  pd1 <- partial(df.reg.lm,
+                 pred.var = "rm",
+                 grid.resolution = 1,
+                 train = df.reg)
+  pd2 <- partial(df.reg.lm,
+                 pred.var = "rm",
+                 pred.grid = 1,
+                 check.class = FALSE,
+                 train = df.reg)
+  pd3 <- partial(df.class.glm,
+                 pred.var = "glucose",
+                 grid.resolution = 1,
+                 train = df.class)
+
+  # Expectations
+  expect_is(pd1, "data.frame")
[…]

### 2016-11-28 tests/testthat/test-utils.R

@@ -1,11 +1,20 @@
 context("Utility functions")
 
 # Data frames used in the following tests
-#   regression: mtcars
-#   classification: iris
+pima2 <- na.omit(pima)
+set.seed(101)
+df.reg <- boston[sample(nrow(boston), size = 50, replace = FALSE), ]
+df.class <- pima2[sample(nrow(pima2), size = 50, replace = FALSE), ]
+
+# Switch to generic response names
+df.reg$y <- df.reg$cmedv
+df.class$y <- df.class$diabetes
+df.reg$cmedv <- NULL
+df.class$diabetes <- NULL
 
 test_that("copyClasses works correctly", {
-  # Incorrect classes
+
+  # Data frame with incorrect classes
   set.seed(101)
   d1 <- data.frame(x1 = 1:3 * 1.0,
                    x2 = rnorm(3),
@@ -15,7 +24,7 @@ test_that("copyClasses works correctly", {
                    x6 = c(1, 0, 1),
                    stringsAsFactors = TRUE)
 
-  # Correct classes
+  # Data frame with correct classes
   set.seed(101)
   d2 <- data.frame(x1 = 1:3,
                    x2 = rnorm(3),
@@ -28,13 +37,20 @@ test_that("copyClasses works correctly", {
   # Copy classes from d2 to d1
   d3 <- copyClasses(d1, d2)
   d4 <- copyClasses(d1[1:2, ], d2)
+
+  # Expectations
   expect_identical(d2, d3)
   expect_identical(sapply(d2, levels), sapply(d4, levels))
+
 })
 
 test_that("avgLogit works correctly", {
+
+  # Probabilitymatrix/data frame
   pm <- matrix(c(0.1, 0.3, 0.6), nrow = 1, ncol = 3, byrow = TRUE)
   pm.df <- as.data.frame(pm)
+
+  # Expectations
   expect_identical(avgLogit(pm), avgLogit(pm.df))
   expect_identical(avgLogit(pm, which.class = 1L),
                    log(0.1) - (log(0.1) + log(0.3) + log(0.6)) / 3)
@@ -42,60 +58,178 @@ test_that("avgLogit works correctly", {
                    log(0.3) - (log(0.1) + log(0.3) + log(0.6)) / 3)
   expect_identical(avgLogit(pm, which.class = 3L),
                    log(0.6) - (log(0.1) + log(0.3) + log(0.6)) / 3)
+
+})
+
[…]

### 2016-11-28 README.Rmd
minor tweaks

@@
 In this example, we fit a random forest to the Boston housing data.
-Note, for any of the following examples, you can see a progress bar by simply specifying `.progress = "text"` in the call to `partial`.
+(See `?boston` for a brief explanation of the data.) Note, for any of the following examples, you can see a progress bar by simply specifying `.progress = "text"` in the call to `partial`.
 You may also reduce the computation time via the `grid.resolution` option in the call to `partial`.
@@
 
-In this example, we fit another random forest to the Pima Indians diabetes data.
+In this example, we fit a random forest to the Pima Indians diabetes data. (See `?pima` for a brief explanation of the data.)
 ```{r, message=FALSE, warning=FALSE}
@@
 
-# Partial dependence of glucose on diabetes test result (neg/pos)
+# Partial dependence of glucose on diabetes test result (neg/pos). This may take
+# a couple minutes!
 partial(pima.rf, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE)
@@
 
-Finally, we demonstrate the seamless interface between the `pdp` package and the `caret` package; `caret` is an extremetly useful package for classification and regression training that, essentially, has one function (`train`) for fitting all kinds of predictive models in R (e.g., `glmnet`, `svm`, `xgboost`, etc.).
-For illustration we use `caret`'s `train` function to tune a support vector machine (SVM) with a radial basis function kernel to the Pima Indians diabetes data.
-We then construct PDPs from the final model for the most "important" variables.
-When training a model using `caret`'s `train` function, you can view tuning progress by setting `verboseIter = TRUE` in the call to `trainControl`.
+Finally, we demonstrate the construction of PDPs from models fit using the `caret` package; `caret` is an extremetly useful package for classification and regression training that, essentially, has one function (`train`) for fitting all kinds of predictive models in R (e.g., `glmnet`, `svm`, `xgboost`, etc.).
+
+For illustration we use `caret`'s `train` function to tune a support vector machine (SVM) with a radial basis function kernel to the Pima Indians diabetes data using 5-fold cross-validation.
+We then use the final model to construct a PDPs for `glucose` and `age`.
+Note, when training a model using `caret`'s `train` function, you can view tuning progress by setting `verboseIter = TRUE` in the call to `trainControl`.
 ```{r, message=FALSE, warning=FALSE}
@@
 partial(pima.svm, pred.var = "glucose", plot = TRUE, rug = TRUE)
+
+# Partial dependence of age on diabetes test result (neg/pos)
+partial(pima.svm, pred.var = "age", plot = TRUE, rug = TRUE)
+
+# Partial dependence of glucose and age on diabetes test result (neg/pos). This
+# may take a couple minutes!
+partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE,
+        train = pima)
 ```

### 2016-11-30 README.Rmd
removed unnecessary call to train option
new badge

@@
 [![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/pdp)](http://cranlogs.r-pkg.org/badges/grand-total/pdp)
+[![Rdoc](http://www.rdocumentation.org/badges/version/pdp)](http://www.rdocumentation.org/packages/pdp)
 
@@
 # may take a couple minutes!
-partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE,
-        train = pima)
+partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE)
 ```

### 2016-12-04 README.Rmd
minor tweaks

@@
 You may also reduce the computation time via the `grid.resolution` option in the call to `partial`.
-```{r, message = FALSE, warning=FALSE}
+```{r, message = FALSE, warning=FALSE, fig.show="hold"}
 # Load required packages
-library(ggmap)  # for plotting data on maps
 library(pdp)  # for constructing PDPs
@@
 
-# Using randomForest's partialPlot function
-partialPlot(boston.rf, pred.data = boston, x.var = "lstat")
-
-# Package pdp's partial function returns a data frame by default
-head(partial(boston.rf, pred.var = "lstat"))  # returns a data frame
-
-# Set `plot = TRUE` to return a "trellis" object
+# Partial dependence of lstat and rm on cmedv
 partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
-
-# The partial function allows for multiple predictors
+partial(boston.rf, pred.var = "age", plot = TRUE, rug = TRUE)
 partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
-
-# The plotPartial function offers more flexible plotting
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
-
-# Default plot
-plotPartial(pd)
-
-# Customized plot
-plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
-            colorkey = FALSE, screen = list(z = -20, x = -60))
-
-# Look at partial dependence of median home value on location
-pd.loc <- partial(boston.rf, pred.var = c("lon", "lat"), chull = TRUE)
-
-# Overlay predictions on a map of Boston
-# ll <- c(range(boston$lon), range(boston$lat))
-# map <- get_map(location = ll[c(1, 3, 2, 4)], zoom = 11, maptype = "toner-lite")
-# ggmap(map) + 
-#   geom_point(aes(x = lon, y = lat), data = boston, alpha = 0.2) +
-#   geom_tile(aes(x = lon, y = lat, z = y, fill = y), 
-#             data = pd.loc, alpha = 0.3) +
-#   geom_contour(color = "white", alpha = 0.5) +
-#   scale_fill_distiller(palette = "Spectral", name = "Median\nvalue") +
-#   coord_fixed(ratio = 1) +
-#   labs(x = "Longitude", y = "Latitude")
 ```
@@
 
-In this example, we fit a random forest to the Pima Indians diabetes data. (See `?pima` for a brief explanation of the data.)
-```{r, message=FALSE, warning=FALSE}
-# Fit a random forest to the Pima Indians diabetes data
+In this example, we fit a support vector machine with a radial basis function kernel to the Pima Indians diabetes data. (See `?pima` for a brief explanation of the data.)
+```{r, message=FALSE, warning=FALSE, fig.show="hold"}
+# Load required packages
+library(kernlab)  # for fitting SVMs
+
+# Fit an SVM to the Pima Indians diabetes data
 data (pima)  # load the boston housing data
-set.seed(102)  # for reproducibility
-pima.rf <- randomForest(diabetes ~ ., data = pima, na.action = na.omit)
+pima.svm <- ksvm(diabetes ~ ., data = pima, type = "C-svc", kernel = "rbfdot",
+                 C = 0.5, prob.model = TRUE)
 
-# Partial dependence of glucose on diabetes test result (neg/pos). This may take
-# a couple minutes!
-partial(pima.rf, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE)
+# Partial dependence of glucose and age on diabetes test result (neg/pos). 
+partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE)
 ```
@@
 
-For illustration we use `caret`'s `train` function to tune a support vector machine (SVM) with a radial basis function kernel to the Pima Indians diabetes data using 5-fold cross-validation.
-We then use the final model to construct a PDPs for `glucose` and `age`.
+For illustration we use `caret`'s `train` function to tune an [XGBoost](https://github.com/dmlc/xgboost) model to the Pima Indians diabetes data using 5-fold cross-validation.
+We then use the final model to construct PDPs for `glucose` and `age`.
 Note, when training a model using `caret`'s `train` function, you can view tuning progress by setting `verboseIter = TRUE` in the call to `trainControl`.
@@
 
+# Grid of tuning parameter values
+xgb.grid <- expand.grid(
+  nrounds = 500,
+  max_depth = 1:6,
+  eta = c(0.001, 0.01, 0.1, 0.2, 0.3, 0.5),
+  gamma = 0, 
+  colsample_bytree = 1,
+  min_child_weight = 1,
+  subsample = 1
+)
+
 # Tune a support vector machine (SVM) using a radial basis function kerel to
-# the Pima Indians diabetes data
+# the Pima Indians diabetes data. This may take a few minutes!
 set.seed(103)  # for reproducibility
-pima.svm <- train(diabetes ~ ., data = pima, method = "svmRadial",
+pima.xgb <- train(diabetes ~ ., data = pima, method = "xgbTree",
                   prob.model = TRUE, na.action = na.omit, trControl = ctrl,
-                  tuneLength = 10)
+                  tuneGrid = xgb.grid)
 
-# Partial dependence of glucose on diabetes test result (neg/pos)
-partial(pima.svm, pred.var = "glucose", plot = TRUE, rug = TRUE)
-
-# Partial dependence of age on diabetes test result (neg/pos)
-partial(pima.svm, pred.var = "age", plot = TRUE, rug = TRUE)
-
-# Partial dependence of glucose and age on diabetes test result (neg/pos). This
-# may take a couple minutes!
-partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE)
+# Partial dependence of glucose and age on diabetes test result (neg/pos)
+partial(pima.xgb, pred.var = "glucose", plot = TRUE, rug = TRUE)
+partial(pima.xgb, pred.var = "age", plot = TRUE, rug = TRUE)
+partial(pima.xgb, pred.var = "mass", plot = TRUE, chull = TRUE)
 ```

### 2016-12-06 README.Rmd
... now refers to stats::predict arguments

@@
 In this example, we fit a random forest to the Boston housing data.
-(See `?boston` for a brief explanation of the data.) Note, for any of the following examples, you can see a progress bar by simply specifying `.progress = "text"` in the call to `partial`.
+(See `?boston` for a brief explanation of the data.) Note, for any of the following examples, you can see a progress bar by simply specifying `progress = "text"` in the call to `partial`.
 You may also reduce the computation time via the `grid.resolution` option in the call to `partial`.
@@
 partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
-partial(boston.rf, pred.var = "age", plot = TRUE, rug = TRUE)
+partial(boston.rf, pred.var = "rm", plot = TRUE, rug = TRUE)
 partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)

### 2016-12-06 R/pdClassification.R

@@ -1,94 +1,95 @@
 #' @keywords internal
 pdClassification <- function(object, pred.var, pred.grid, which.class,
-                             train, ...) {
+                             train, progress, parallel, paropts, ...) {
   UseMethod("pdClassification")
 }
 
 
 #' @keywords internal
 pdClassification.default <- function(object, pred.var, pred.grid, which.class,
-                                     train, ...) {
+                                     train, progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     # FIXME: Does this copy need to be made?
     temp <- train
     temp[pred.var] <- x
-    pr <- stats::predict(object, newdata = temp, type = "prob")
+    pr <- stats::predict(object, newdata = temp, type = "prob", ...)
     avgLogit(pr, which.class = which.class)
-  }, ...)
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
 
 #' @keywords internal
 pdClassification.glm <- function(object, pred.var, pred.grid, which.class,
-                                     train, ...) {
+                                     train, progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
-    pr <- stats::predict(object, newdata = temp, type = "response")
+    pr <- stats::predict(object, newdata = temp, type = "response", ...)
     # Binary regression returns a vector of predicted probabilities!
     avgLogit(cbind(pr, 1 - pr), which.class = which.class)
-  }, ...)
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
 
 #' @keywords internal
 pdClassification.bagging <- function(object, pred.var, pred.grid, which.class,
-                                      train, ...) {
+                                      train, progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
-    pr <- stats::predict(object, newdata = temp)$prob
+    pr <- stats::predict(object, newdata = temp, ...)$prob
     avgLogit(pr, which.class = which.class)
-  }, ...)
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
 
 #' @keywords internal
 pdClassification.boosting <- function(object, pred.var, pred.grid, which.class,
-                                      train, ...) {
[…]

### 2016-12-06 R/pdRegression.R

@@ -1,45 +1,48 @@
 #' @keywords internal
-pdRegression <- function(object, pred.var, pred.grid, train, ...) {
+pdRegression <- function(object, pred.var, pred.grid, train, progress, parallel,
+                         paropts, ...) {
   UseMethod("pdRegression")
 }
 
 
 #' @keywords internal
-pdRegression.default <- function(object, pred.var, pred.grid, train,
-                                 ...) {
+pdRegression.default <- function(object, pred.var, pred.grid, train, progress,
+                                 parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
-    pred <- stats::predict(object, newdata = temp)
+    pred <- stats::predict(object, newdata = temp, ...)
     # Some fitting function return a matrix (e.g., mda::mars)
     if (is.matrix(pred) || is.data.frame(pred)) {
       pred <- pred[, 1L, drop = TRUE]
     }
     mean(pred, na.rm = TRUE)
-  }, ...)
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
 
 #' @keywords internal
-pdRegression.gbm <- function(object, pred.var, pred.grid, train,
-                                 ...) {
+pdRegression.gbm <- function(object, pred.var, pred.grid, train, progress,
+                             parallel, paropts, ...) {
   # Necessary to avoid silly printing from predict.gbm
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
     log <- utils::capture.output(z <- mean(stats::predict(object,
-                                                          newdata = temp),
+                                                          newdata = temp, ...),
                                            na.rm = TRUE))
     z
-  }, ...)
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
 
 #' @keywords internal
-pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, train, ...) {
+pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, train,
+                                     progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
-    mean(stats::predict(object, newdata = data.matrix(temp)), na.rm = TRUE)
-  }, ...)
[…]

### 2016-12-07 manuscript/greenwell.tex
started working on reviewers comments

@@
 \abstract{
-Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression).
+Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes $E\left[\boldsymbol{\mathcal{Y}}\right] = \boldsymbol{X}\boldsymbol{\beta}$).
 Unfortunately, it can be challenging to understand the results of such models and explain them to management.
 Partial dependence plots offer a simple solution.
-Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be easily understood.
+Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be
+more easily understood.
 These plots are especially useful in explaining the output from black box models.
@@
   \end{enumerate}
-  \item Plot the pairs $\left[x_{1i}, \bar{f}_1\left(x_{1i}\right)\right]$ for $i = 1, 2, \dotsc, k$.
+  \item Plot the pairs $\left\{x_{1i}, \bar{f}_1\left(x_{1i}\right)\right\}$ for $i = 1, 2, \dotsc, k$.
 \end{enumerate}
@@
 \begin{example}
-Error: The training data could not be extracted from object. Please supply the raw training data using the `train` argument in the call to `partial`.
+Error: The training data could not be extracted from object. Please supply
+the raw training data using the `train` argument in the call to `partial`.
 \end{example}
@@
 The \code{partial} function relies on the \CRANpkg{plyr} package \citep{plyr-pkg}, rather than R's built-in \code{for} loops.
-This makes it easy to request progress bars (e.g., \code{.progress = "text"}) or run \code{partial} in parallel.
+This makes it easy to request progress bars (e.g., \code{progress = "text"}) or run \code{partial} in parallel.
 In fact, \code{partial} can use any of the parallel backends supported by the \pkg{foreach} package.
@@
 \end{example}
-Now, to run \code{partial} in parallel, all we have to do is invoke the \code{.parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded.
+Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded.
 This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
@@
 partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
-        chull = TRUE, .parallel = TRUE)
+        chull = TRUE, parallel = TRUE)
 \end{example}
@@
   tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, grid.resolution = 101, .progress = "text")
+                 which.class = i, grid.resolution = 101, progress = "text")
   pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
@@
   geom_contour(color = "white", alpha = 0.5) +
-  scale_fill_distiller(palette = "Spectral") +
+  scale_fill_distiller(name = "Average\nlogit", palette = "Spectral") +
   theme_bw() +
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Using partial with caret
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Summary

### 2016-12-07 R/pdClassification.R

@@ -13,7 +13,7 @@ pdClassification.default <- function(object, pred.var, pred.grid, which.class,
     temp <- train
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, type = "prob", ...)
-    avgLogit(pr, which.class = which.class)
+    setNames(avgLogit(pr, which.class = which.class), "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -26,7 +26,7 @@ pdClassification.glm <- function(object, pred.var, pred.grid, which.class,
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, type = "response", ...)
     # Binary regression returns a vector of predicted probabilities!
-    avgLogit(cbind(pr, 1 - pr), which.class = which.class)
+    setNames(avgLogit(cbind(pr, 1 - pr), which.class = which.class), "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -38,7 +38,7 @@ pdClassification.bagging <- function(object, pred.var, pred.grid, which.class,
     temp <- train
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, ...)$prob
-    avgLogit(pr, which.class = which.class)
+    setNames(avgLogit(pr, which.class = which.class), "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -50,7 +50,7 @@ pdClassification.boosting <- function(object, pred.var, pred.grid, which.class,
     temp <- train
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, ...)$prob
-    avgLogit(pr, which.class = which.class)
+    setNames(avgLogit(pr, which.class = which.class), "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -64,7 +64,7 @@ pdClassification.gbm <- function(object, pred.var, pred.grid, which.class,
     temp[pred.var] <- x
     log <- utils::capture.output(pr <- stats::predict(object, newdata = temp,
                                                       type = "response", ...))
-    avgLogit(cbind(pr, 1 - pr), which.class = which.class)
+    setNames(avgLogit(cbind(pr, 1 - pr), which.class = which.class), "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
[…]

### 2016-12-07 R/pdp.R

@@ -1,0 +1,19 @@
+#' pdp: A general framework for constructing partial dependence (i.e., marginal
+#' effect) plots from various types machine learning models in R.
+#'
+#' Partial dependence plots (PDPs) help visualize the relationship between a
+#' subset of the features (typically 1-3) and the response while accounting for
+#' the average effect of the other predictors in the model. They are
+#' particularly effective with black box models like random forests and support
+#' vector machines.
+#'
+#' The development version can be found on GitHub: https://github.com/bgreenwell/pdp.
+#' As of right now, \code{pdp} only exports two functions:
+#' \itemize{
+#'   \item{\code{partial}} --- Compute partial dependence functions (i.e., marginal effects) for various model fitting objects.
+#'   \item{\code{plotPartial}} --- Plot partial dependence functions (i.e., marginal effects) using lattice graphics.
+#' }
+#'
+#' @docType package
+#' @name pdp
+NULL

### 2016-12-07 manuscript/greenwell.R

@@ -144,6 +144,16 @@ pdf("partial_par.pdf", width = 7, height = 5)
 plotPartial(pd)
 dev.off()
 
+# Figure 6
+cl <- makeCluster(4)  # use 4 cores
+registerDoParallel(cl)
+pd <- partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), chull = TRUE,
+              parallel = TRUE, paropts = list(.packages = "earth"))
+stopCluster(cl)
+pdf("partial_par.pdf", width = 7, height = 5)
+plotPartial(pd)
+dev.off()
+
 
 ################################################################################
 # Edgar Anderson's iris data

### 2016-12-11 manuscript/reviews/response.Rmd
tweaks based on review comments

---
title: "Response to Comments"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
  word_document: default
---

## Reviewer 1

#### Comment
Bug: A graph is produced as expected if the code below is cut-and-pasted at the command line .  However, if the code is put into a file (say `foo.R`) and `source("foo.R")` is executed at the command line, no graph is produced.  (But note that if `plot=TRUE` is added to the call to `partial()`, then a plot is produced as expected.)
```{r, eval=FALSE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pd <- partial(ozone.rf, pred.var="Temp")
plotPartial(pd)
```

#### Response
This is simply a side effect of using `lattice` to produce the plots. If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))`).

#### Comment
I couldn't get `partial` to work with `gbm` (version 2.1.1) models (although I didn't try very hard).  Perhaps this is because the `gbm` maintainer changed the `gbm` interface in a non-backcompatible way.  The following (where `gbm.mod` is a model built using the gbm package)
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age")
```
gives
```{r, eval=FALSE}
Error in paste("Using", n.trees, "trees...\n") :
argument "n.trees" is missing, with no default
```
I also tried
```{r, eval=FALSE}
partial(gbm.mod, pred.var="age", n.trees=100)
```
but that gives
```{r, eval=FALSE}
Error in .fun(piece, ...) : unused argument (n.trees = 100)
```

#### Response
This been fixed in `pdp` (version 0.3.0), but requires the user to supply `n.trees` in the call to `partial`. So, `partial(gbm.mod, pred.var="age", n.trees=100)` should work now. The test suite is being expanded, as suggested in a later comment, to help catch more errors like this.

#### Comment
Standard partial matching isn't implemented (e.g., `partial(..., type="r")`).

#### Response
Partial matching is now implemented, but only for the `type` argument.

#### Comment
The following code incorrectly gives an error message
```
"Error in names(pd_df) <- c(pred.var, "y") :
  'names' attribute [2] must be the same length as the vector [1]"
```
It seems to work ok if you change the variable names.
```{r, eval=FALSE}
library(pdp)
data <- data.frame(V1=1:10, V2=1:10)
mod <- lm(V2~V1, data=data)
partial(mod, pred.var="V1", plot=TRUE) # error
```

#### Response
This has been fixed in `pdp` (version 0.3.0). This was simply a side effect of `partial`'s internal use of `plyr::adply` which uses default variable names (e.g., `V1`, `V2`, etc.). This caused the column `V1` in the output data frame to be overwritten, producing the error. 
```{r, eval=TRUE}
library(pdp)
data <- data.frame(V1=1:10, V2=1:10)
mod <- lm(V2~V1, data=data)
partial(mod, pred.var="V1") # no more error
```

#### Comment
I think you should mention in the documentation something about the assumptions you make about accessing the model data. The following (somewhat contrived) example fails with
`"Error in train[[x]] : subscript out of bounds":`
```{r, eval=FALSE}
data(trees)
foo <- function(data)
{
    lm(Volume~., data=data)
}
mod <- foo(trees)
library(pdp)
partial(mod, pred.var="Girth", plot=TRUE) # fails
```
and the following silently gives misleading results:

```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var="Girth", plot=TRUE) # ok
trees <- trees[1:2,]
partial(mod, pred.var="Girth", plot=TRUE) # wrong xlim
```
I'm not necessarily saying that you need to resolve the above two
issues, but at least mention in the documentation that this kind of
thing can happen, and how to avoid it (basically call partial in the
same environment used to build the model, and don't change any of the
data used to build the model).

#### Comment
The following code gives an obscure error message:
```
"Error in seq.default(from = min(train[[x]], na.rm = TRUE),
   to = max(train[[x]],  :
  'from' cannot be NA, NaN or infinite"
```
Some better handholding for the user may be helpful here.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var="nosuchvariable", plot=TRUE) # obscure err msg
```

#### Response
This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
```{r, eval=FALSE}
library(pdp)
data(trees)
mod <- lm(Volume~., data=trees)
partial(mod, pred.var=c("Girth", "foo", "bar"), plot=TRUE)
```
```
Error in partial.default(mod, pred.var = "nosuchvariable", plot = TRUE) : 
  nosuchvariable not found in the training data.
```

#### Comment
On page 2 of the paper it says "The columns of the data frame are labeled in the same order as the features supplied to `pred.var`, and the last column is always labeled `y`". What happens if `y` is one of the predictors?

#### Response
The last column is labeled `y` for consistency with `gbm` and `randomForest` behavior. Only `gbm` does not fail in this case.

#### Comment
Hastie et al. in the book cited in the paper say "Although such a collection [of partial dependence plots] can seldom provide a comprehensive depiction of the approximation, it can often produce helpful clues".  Therefore in in the abstract for the paper, to avoid overselling consider changing "relationship between the outcome and predictors of interest can be easily understood" to "relationship between the outcome and predictors of interest can be more easily understood".

#### Response
Fixed.

#### Comment
On page 3 in the PDF of the paper, add a newline to the error message in the paper:
```
Error: The training data could not be extracted from object. Please
supply the raw training data using the ...
```

#### Response
Fixed.

#### Comment
Is there any reason the package is called `pdp` but the main function is called `partial`?  Consider giving the function the same name as the package, so e.g the following works (to help the user get started)
```{r, eval=FALSE}
library(pdp)
?pdp # fails
```

#### Response
The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letetrs). At the time I decided to change the name of the package to `pdp`. I've personally never liked the idea of the main function name to be the same as the package name. However, I did add a help file for the page describing the package, main functions, etc. So, `?pdp` no longer fails.

#### Comment
`partial` should return the same value regardless of whether or not the `plot` argument is used?

#### Response
I dont think so and this seems to be the same behavior for the PDP functions in `randomForest` and `gbm`. However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that the user can extract the data easily from that if necessary (e.g., if they do not like the plot and don't want to re-run partial). For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
pd.Girth <- partial(mod, pred.var = "Girth")  # data only
pdp.Girth <- partial(mod, pred.var = "Girth", plot = TRUE)  # plot
identical(pd.Girth, attr(pdp.Girth, "partial.data"))  # should be identical
```
Also the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded. This will be useful when the user does not store the resulting plot in the call to `partial` but needs to retrieve the outout data to avoid re-running `partial` when computation time is a concern. For example,
```{r}
library(pdp)
data(trees)
mod <- lm(Volume ~ ., data = trees)
partial(mod, pred.var = "Girth", plot = TRUE)
pdp.Girth <- lattice::trellis.last.object()
pd.Girth <- attr(pdp.Girth, "partial.data")
head(pd.Girth)
```

#### Comment
A call to say `par(mfrow = c(2, 2))` before calling `plotPartial` is ignored by `plotPartial`.  This is a pity because it would allow multiple plots to be put on one page like this:
```{r, eval=FALSE}
par(mfrow=c(2,2))
partial(ozone.rf, pred.var = "Temp", plot = TRUE)
partial(ozone.rf, pred.var = "Wind", plot = TRUE)
```

#### Response
The command `par(mfrow = c(2, 2))` does not work with `"trellice"` objects, such as those produced by `lattice` which `partial` relies on for producing its plots. Other methods are available. For example, using the `gridExtra` package:
```{r, eval=TRUE}
library(randomForest)
data(airquality)
ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
library(pdp)
pdp1 <- partial(ozone.rf, pred.var = "Temp", plot = TRUE)
pdp2 <- partial(ozone.rf, pred.var = "Wind", plot = TRUE)
gridExtra::grid.arrange(pdp1, pdp2, ncol = 2)
```
However, for convenience, I have included a `multiplot` function that is just a simple wrapper for `gridExtra::grid.arrange` and should satisfy in most cases. For example,
```{r, eval=FALSE}
multiplot(pdp1, pdp2, ncol = 2)
?multiplot
```

#### Comment
It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.

#### Response 
I agree, this would be a nice feature, but the problem I see is that not all models emit a "variable importance". And those that do often have many different ways. For example, random forest offeres two types, which can then be scaled, etc. At this point, I think it is simpler to have the user select which variables they would like to see plotted. But perhaps in the near future. The best way I can see implementing this feature is to have the option when the first argument to `partial` is a `"train"` object (i.e., a model fitted using the `caret` package). `caret` has a generic `varImp` function that applies to a lot of model classes. Perhaps in the near future, a new argument to `partial`, say `top` can be added. So, something like
```{r, eval=FALSE}
library(caret)
library(pdp)
model <- train(x, y, method = "xgbTree", ...)
partial(model, pred.var = 5, plot = TRUE)  # integer indicating using top 5
```
I added this as a feature request on the GitHub page: https://github.com/bgreenwell/pdp/issues/19.

#### Comment
It would be nice if the functions automatically got the data from the model used to build the model in more cases without forcing the user to explicitly specify the data.  Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only -- one can't access the data stored in `xgb.DMatrix`).

#### Response
This is also true for the fitting functions in other package (e.g., `kernlab::ksvm`). As time goes on, I will try to increase coverage to make it easier on the user. For example, `partial` now works with `caret` which always stores the taining data in a component called `trainingData`. I suspect that package maintainers, especially those for `xgboost`, would be hesitent in storing a copy of the training data as the package is often used for modeling large data sets which they do not

#### Comment
Consider adding a `slowtests` directory (below `pdp/inst/` or elsewhere) that has tests for all the models supported by the package (Table 1 in the paper) to easily check back compatibility when packages change.  (For justification for such tests see for example my
comments on `gbm` back compatibility above.)  These tests will probably be quite slow, but that won't be an issue because they won't be invoked by CRAN check; the slow tests would be just for the maintainer of the `pdp` package to run manually.

#### Response
This is a good suggestion and is currently being put together.

## Reviewer 2

#### Comment 
The Introduction immediately introduces the PDP. You must first motivate why it is important. In a linear model, they are not important. In a linear model with simple interactions, likewise. They first become important if you have non-parametric macine- optimizing techniques. Further, they are not the only idea extant when trying to explain a predictor's impact on the response. This paper would benefit if you can give a bit more of a lit review here.

### Important Clarifications Required:

#### Comment
Top of p2. The `ICEbox` package allows for general construction and visualization of PDP's. However, it is limited as you only get one dimensional representations. The selling point of your package should focus on the flexibility of graphics and the illustration of non- extrapoloation zones.

#### Comment
All `pdp` figures throughout the paper: Can you make explicit in the caption what code produces the left and right subfigures? I would use the subfigure package in latex and then put a comment in the R code e.g. # Fig 3b.

### Edits:

#### Comment
p1 abstract: strict assumptions imposed by traditional statistical models

#### Comment
p1 pairs should be using curly braces of angle braces, not square braces as this is easily confused with closed intervals.

#### Response
Fixed. Switched to curly braces.

#### Comment
p2 unless the editor disagrees, there is no reason to add "install.packages("pdp")" in the text of the paper as it is assumed a user knows how to install a package.

#### Comment
p3 "Error: The training..." should be wrapped.

#### Response
Fixed.

#### Comment
Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?

#### Comment
Fig 7: as beautiful as this is, you should label the sidebar as "logit" or "centered logit" instead of "y".
#### Response
Fixed. Color key labeled "Average logit".

### Suggestions for the software

#### Comment
I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP. I realize that the default is `FALS`E. However, I still think it's a disservice to the user as constructing PDP's is computationally intense and if they get the plot wrong or it's not to their liking, they will have to regenerate the whole thing again. So, it is my opinion that you should take it out and force two steps: generation and illustration.

#### Comment
The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use ranfom forests to illustrate classification probabilities.

#### Comment
p7 I believe there are packages allowing for parallelization in Windows. Can you switch the example so this works on all platforms?

### 2016-12-11 manuscript/greenwell.tex
tweaks based on review comments

@@
 Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}.
-Currently, only two functions are exported by \pkg{pdp}:
+Currently, only three functions are exported by \pkg{pdp}:
 \begin{itemize}
@@
   \item \code{plotPartial}
+  \item \code{multiplot}
 \end{itemize}
@@
 Of course, since the default output produced by \code{partial} is still a data frame, the user can easily use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
+
+Finally, \code{multiplot} is a simple helper function that makes it easier to display multiple PDPs on a single plot.
 
@@
 
+The \code{partial} function also supports object of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}.
+This means that \code{partial} can be used with any classification or regression model that \code{caret} supports; see \url{http://topepo.github.io/caret/available-models.html} for a current list of supported models.
+An example is given in Section~?.
+
 For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}.
@@
 library(pdp)
-partial(boston.rf, pred.var = "lstat", plot = TRUE)
+partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
 \end{example}
@@
 \begin{example}
+# Figure 2 (right)
 boston.rf %>%  # the %>% operator is read as "and then"
@@
 pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
-plotPartial(pd)
-plotPartial(pd, contour = TRUE, col.regions = rwb)
+plotPartial(pd)  # Figure 3 (left)
+plotPartial(pd, contour = TRUE, col.regions = rwb)  # Figure 3 (middle)
 plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
-            colorkey = TRUE, screen = list(z = -20, x = -60))
+            colorkey = TRUE, screen = list(z = -20, x = -60))  # Figure 3 (right)
 \end{example}
@@
 \begin{example}
+# Figure 4 (left)
 partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
@@
 \begin{example}
+# Figure 4 (right)
 partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
@@
 \begin{example}
-partial(boston.rf, "rm", plot = TRUE)
-partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
-partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
+multiplot(
+  # Figure 5 (left)
+  partial(boston.rf, "rm", plot = TRUE),
+
+  # Figure 5 (middle)
+  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
+
+  # Figure 5 (right)
+  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
+
+  # One column for each PDP
+  ncol = 3
+)
 \end{example}
@@
 \label{eqn:avg-logit}
-f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{i = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
+f_k(x) = \log\left[p_k(x)\right] - \frac{1}{K}\sum_{k = 1}^K\log\left[p_k(x)\right], \quad k = 1, 2, \dots, K,
 \end{equation}
@@
 (Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)
+
+The code chunk below uses \pkg{caret} to train an XGBoost model of a grid of tuning values using 5-fold cross-validation.
 \begin{example}
-set.seed(102)  # for reproducibility
-boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
-                      label = boston$cmedv, objective = "reg:linear",
-                      nrounds = 2000, max_depth = 3, eta = 0.01)
+# Optimize tuning parameters using 5-fold cross-validation
+library(caret)
+
+# Setup for 5-fold cross-validation
+ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)
+
+# Grid of tuning values for XGBoost
+xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
+                        max_depth = 1:6,
+                        eta = c(0.001, 0.01, 0.1, 0.5, 1),
+                        gamma = 0,
+                        colsample_bytree = 1,
+                        min_child_weight = 1)
+
+# Train an XGBoost model
+set.seed(202)
+boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
+                         y = boston$cmedv,
+                         method = "xgbTree",
+                         metric = "RMSE",
+                         trControl = ctrl,
+                         tuneGrid = xgb.grid)
 \end{example}
-
-To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}.
+The final model achieves a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively.
 The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
 The results are displayed in Figure~\ref{fig:boston_xgb}.
-\strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
 \begin{example}
-pred.list <- list("lstat", "rm", c("lstat", "rm"))
-for (x in pred.list) {
-  partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
-          chull = TRUE, train = subset(boston, select = -cmedv))
-}
+multiplot(
+  partial(boston.xgb.tune, pred.var = "rm", plot = T, rug = T),
+  partial(boston.xgb.tune, pred.var = "lstat", plot = T, rug = T),
+  partial(boston.xgb.tune, pred.var = c("lstat", "rm"), plot = T, chull = T),
+  ncol = 3)
 \end{example}
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Using partial with caret
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+Since \code{train} always stores a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}.
+However, this is not the case when using the \pkg{xgboost} package directly.
+For example, the following code fits the same model using the \code{xgboost} function.
+\begin{example}
+set.seed(102)  # for reproducibility
+boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
+                      label = boston$cmedv, objective = "reg:linear",
+                      nrounds = 2000, max_depth = 3, eta = 0.01)
+\end{example}
+To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}.
+The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown).
+\strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
+\begin{example}
+partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE,
+        train = subset(boston, select = -cmedv))
+\end{example}
+
+% \begin{example}
+% pred.list <- list("lstat", "rm", c("lstat", "rm"))
+% for (x in pred.list) {
+%   partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
+%           chull = TRUE, train = subset(boston, select = -cmedv))
+% }
+% \end{example}
 

### 2016-12-11 R/pdRegression.R

@@ -16,7 +16,7 @@ pdRegression.default <- function(object, pred.var, pred.grid, train, progress,
     if (is.matrix(pred) || is.data.frame(pred)) {
       pred <- pred[, 1L, drop = TRUE]
     }
-    setNames(mean(pred, na.rm = TRUE), "y")
+    stats::setNames(mean(pred, na.rm = TRUE), "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -31,7 +31,7 @@ pdRegression.gbm <- function(object, pred.var, pred.grid, train, progress,
     log <- utils::capture.output(z <- mean(stats::predict(object,
                                                           newdata = temp, ...),
                                            na.rm = TRUE))
-    setNames(z, "y")
+    stats::setNames(z, "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -42,7 +42,7 @@ pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, train,
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
-    setNames(mean(stats::predict(object, newdata = data.matrix(temp), ...),
+    stats::setNames(mean(stats::predict(object, newdata = data.matrix(temp), ...),
                   na.rm = TRUE), "y")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }

### 2016-12-11 manuscript/greenwell.R

@@ -17,8 +17,8 @@ setwd("manuscript")
 # release will be on CRAN soon. I am waiting out the grace period from my last
 # submission---the only change in this version is the name of an argument in
 # the two exported functions.
-install.packages("devtools")
-devtools::install_github("bgreenwell/pdp")
+# install.packages("devtools")
+# devtools::install_github("bgreenwell/pdp")
 
 
 # List of packages required to run all the examples in this script
@@ -64,7 +64,7 @@ set.seed(101)
 mushroom.rf <- train(x = subset(mushroom, select = -Edibility),
                      y = mushroom$Edibility,
                      method = "rf",
-                     trControl = trainControl(method = "cv", number = 10, 
+                     trControl = trainControl(method = "cv", number = 10,
                                               verboseIter = TRUE),
                      tuneLength = 22)
 
@@ -233,22 +233,23 @@ dev.off()
 ################################################################################
 
 # Optimize tuning parameters using 5-fold cross-validation
-# library(caret)
-# ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)
-# xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
-#                         max_depth = 1:6,
-#                         eta = c(0.001, 0.01, 0.1, 0.5, 1),
-#                         gamma = 0,
-#                         colsample_bytree = 1,
-#                         min_child_weight = 1)
-# set.seed(202)
-# boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
-#                          y = boston$cmedv,
-#                          method = "xgbTree",
-#                          metric = "RMSE",
-#                          trControl = ctrl,
-#                          tuneGrid = xgb.grid)
-# plot(boston.xgb.tune)
+library(caret)
+ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)
+xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
+                        max_depth = 1:6,
+                        eta = c(0.001, 0.01, 0.1, 0.5, 1),
+                        gamma = 0,
+                        colsample_bytree = 1,
+                        min_child_weight = 1,
[…]

### 2016-12-11 slowtests/slowtests.R

@@ -1,0 +1,89 @@
+#-------------------------------------------------------------------------------
+# Slow tests for the pdp package
+#-------------------------------------------------------------------------------
+
+
+################################################################################
+# Setup
+################################################################################
+
+# Load required packages
+library(caret)
+library(pdp)
+
+# Load data
+data(boston)  # regression
+data(pima)  # classification
+
+
+################################################################################
+# randomForest package
+################################################################################
+
+# Load required package
+library(randomForest)
+
+# Regression
+set.seed(101)
+boston.randomForest <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
+
+# Partial dependence of rm and lstat on cmedv
+pdp1 <- partial(boston.randomForest, pred.var = "rm", plot = TRUE, 
+                rug = TRUE, progress = "text")
+pdp2 <- partial(boston.randomForest, pred.var = "lstat", plot = TRUE, 
+                rug = TRUE, progress = "text")
+pdp3 <- partial(boston.randomForest, pred.var = "lon", plot = TRUE, 
+                rug = TRUE, progress = "text")
+pdp4 <- partial(boston.randomForest, pred.var = c("rm", "lstat"), plot = TRUE, 
+                rug = TRUE, chull = TRUE, progress = "text")
+multiplot(pdp1, pdp2, pdp3, pdp4, ncol = 2)
+
+
+
+ctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 10, verboseIter = TRUE)
+set.seed(101)
+boston.mod <- train(x = data.matrix(subset(boston, select = -cmedv)),
+                    y = boston$cmedv,
+                    method = "svmRadial",
+                    metric = "Rsquared",
+                    trControl = ctrl,
+                    tuneLength = 10)
+plot(boston.mod)
+plot(varImp(boston.mod))
+partial(boston.mod, pred.var = "rm", plot = TRUE)
+
+topPredictors <- function(object, n = 1L, ...) {
+  UseMethod("topPredictors")
+}
+topPredictors.train <- function(object, n = 1L, ...) {
+  imp <- varImp(object)$importance
+  imp <- imp[order(imp$Overall, decreasing = TRUE), , drop = FALSE]
+  rownames(imp)[seq_len(n)]
+}
+topPredictors(boston.mod, 2)
+pdp.top <- boston.mod %>%
[…]

### 2016-12-11 manuscript/greenwell.tex
tweaks based on review comments

@@
 Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}.
-Currently, only three functions are exported by \pkg{pdp}:
+Currently, only two functions are exported by \pkg{pdp}:
 \begin{itemize}
@@
   \item \code{plotPartial}
-  \item \code{multiplot}
 \end{itemize}
@@
 
-Finally, \code{multiplot} is a simple helper function that makes it easier to display multiple PDPs on a single plot.
+\strong{Note:} as mentioned above, \pkg{pdp} relies on \pkg{lattice} for its graphics.
+\pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}.
+\pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making:
+\begin{enumerate}
+  \item \pkg{lattice} functions create a \code{"trellis"} object, but does not display it; the \code{print} method produces the actual display.
+The result is automatically printed when using these functions in the command line, but in \code{source} or inside a function, an explicit \code{print} statement is required to display the resulting graph.
+  \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots.
+Instead, \pkg{lattice} provides its own \code{trellis.par.set} function for modifying graphical parameters.
+\end{enumerate}
+A consequence of second point is that the \code{par} function can not be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots.
+Simple solutions are available in packages \CRANpkg{latticeExtra} and \CRANpkg{gridExtra}.
+For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.
 
@@
 \begin{example}
-multiplot(
+grid.arrange(
   # Figure 5 (left)
@@
 \begin{example}
-multiplot(
+grid.arrange(
+  # Figure 8 (left)
   partial(boston.xgb.tune, pred.var = "rm", plot = T, rug = T),
+
+  # Figure 8 (middle)
   partial(boston.xgb.tune, pred.var = "lstat", plot = T, rug = T),
+
+  # Figure 8 (right)
   partial(boston.xgb.tune, pred.var = c("lstat", "rm"), plot = T, chull = T),
-  ncol = 3)
+
+  # One column for each PDP
+  ncol = 3
+)
 \end{example}

### 2016-12-11 R/multiplot.R

@@ -1,27 +1,0 @@
-#' Arrange Multiple grobs on a Page
-#'
-#' Set up a gtable layout to place multiple grobs on a page.
-#'
-#' @param ... grobs, gtables, ggplot or trellis objects.
-#' @param nrow Integer specifying the number of rows of plots.
-#' @param ncol Integer specifying the number of columns of plots.
-#' @return An object of class \code{c("gtable", "gTree", "grob", "gDesc")}.
-#' @details \code{multiplot} is just a simple wrapper around
-#'   \code{\link[gridExtra]{grid.arrange}}. For details see the documentation
-#'   for that function.
-#' @export
-#' @examples
-#' \dontrun{
-#' # Fit a random forest to the boston housing data
-#' library(randomForest)
-#' data (boston)  # load the boston housing data
-#' set.seed(101)  # for reproducibility
-#' boston.rf <- randomForest(cmedv ~ ., data = boston)
-#' pdp.rm <- partial(boston.rf, pred.var = "rm", plot = TRUE)
-#' pdp.lstat <- partial(boston.rf, pred.var = "lstat", plot = TRUE)
-#' pdp.rm.lstat <- partial(boston.rf, pred.var = c("rm", "lstat"), plot = TRUE)
-#' multiplot(pdp.rm, pdp.lstat, pdp.rm.lstat, ncol = 3)
-#' }
-multiplot <- function(..., nrow = NULL, ncol = NULL) {
-  gridExtra::grid.arrange(..., nrow = nrow, ncol = ncol)
-}

### 2016-12-11 manuscript/reviews/response.Rmd
tweaks based on review comments

@@
 output:
+  html_document: default
+  html_notebook: default
   pdf_document: default
-  html_notebook: default
-  html_document: default
   word_document: default
 ---
+
+A few additional changes have been made to the package since the original submission.
+The Boston housing data no longer has to be imported from the `mlbench` package; it is loaded automatically with `pdp`.
+`pdp` now interfaces with the `caret` package.
+This is breifely discussed on page X and has been incorporated into the last example.
+
+A few of the comments were related to the plotting behavior of `plotPartial`. I added a paragraph on page X that should help clarify many of these issues.
 
@@
 
+#### Response
+Fixed. Code chunk was removed.
+
 #### Comment
@@
 #### Response
-Fixed.
+Fixed. The text was wrapped appropriately.
 
@@
 Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?
+
+#### Response
 

### 2016-12-11 manuscript/greenwell.tex
tweaks based on review comments

@@
 PDPs are especially useful for visualizing the relationships discovered by complex machine learning algorithms such as a random forest.
-The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}:
-\begin{example}
-install.packages("pdp")
-\end{example}
+The latest stable release is available from \href{https://cran.r-project.org/package=pdp}{CRAN}.
 The development version is located on GitHub: \url{https://github.com/bgreenwell/pdp}.
@@
 \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}.
-\pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making:
+\pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making (see \code{?lattice} for more details):
 \begin{enumerate}
-  \item \pkg{lattice} functions create a \code{"trellis"} object, but does not display it; the \code{print} method produces the actual display.
-The result is automatically printed when using these functions in the command line, but in \code{source} or inside a function, an explicit \code{print} statement is required to display the resulting graph.
+  \item \pkg{lattice} functions return a \code{"trellis"} object, but do not display it; the \code{print} method produces the actual display.
+However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line.
+If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph.
   \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots.
@@
 \end{enumerate}
-A consequence of second point is that the \code{par} function can not be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots.
+A consequence of second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots.
 Simple solutions are available in packages \CRANpkg{latticeExtra} and \CRANpkg{gridExtra}.
@@
 Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument.
-To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; the results are displayed in Figure~\ref{fig:partial_manual}.
+To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
 \begin{example}
+p1 <- partial(boston.rf, "rm", plot = TRUE)
+p2 <- partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
+p3 <- partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
 grid.arrange(
-  # Figure 5 (left)
-  partial(boston.rf, "rm", plot = TRUE),
-
-  # Figure 5 (middle)
-  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
-
-  # Figure 5 (right)
-  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
-
-  # One column for each PDP
+  p1,  # Figure 5 (left)
+  p2,  # Figure 5 (middle)
+  p3,  # Figure 5 (right)
   ncol = 3
@@
 Setting up a parallel backend is rather straightforward.
-To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on Unix-like systems\footnote{This example will not run on Windows.} using the \pkg{doParallel} package.
+To demonstrate, the following snippet of code sets up the \code{partial} function to run in parallel on both Windows and Unix-like systems using the \pkg{doParallel} package.
 \begin{example}
 library(doParallel)  # load parallel backend
-registerDoParallel(cores = 4)  # use 4 cores
+cl <- makeCluster(4)  # use 4 workers
+registerDoParallel(cl)
 \end{example}
-Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} option and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded.
+Now, to run \code{partial} in parallel, all we have to do is invoke the \code{parallel = TRUE} and \code{paropts} options and the rest is taken care of by the internal call to \pkg{plyr} and the parallel backend we loaded\footnote{Notice we have to pass the names of external packages that the tasks depend on via the \code{paropts} argument; in this case, \code{"earth"}.
+See \code{?plyr::adply} for details.}.
 This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
 The result is displayed in Figure~\ref{fig:partial_par}.
+\strong{Note:} it is considered good practice to shut down the workers by calling \code{stopCluster} when finished.
 \begin{example}
 partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
-        chull = TRUE, parallel = TRUE)
+        chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
+stopCluster(cl)  # good practice
 \end{example}
@@
 
+# Figure 7
 library(ggplot2)
@@
   geom_contour(color = "white", alpha = 0.5) +
-  scale_fill_distiller(name = "Average\nlogit", palette = "Spectral") +
+  scale_fill_distiller(name = "Logit", palette = "Spectral") +
   theme_bw() +
@@
 
-The code chunk below uses \pkg{caret} to train an XGBoost model of a grid of tuning values using 5-fold cross-validation.
+The code chunk below uses \pkg{caret} to train an XGBoost model over a grid of tuning values using 5-fold cross-validation.
 \begin{example}
@@
 set.seed(202)
-boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
+boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
                     y = boston$cmedv,
@@
 \end{example}
-The final model achieves a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively.
-The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
+The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 2000}, \code{max\_depth = 3}, \code{eta = 0.01}, and achieved a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively.
+The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display al three PDPs on the same page.
 The results are displayed in Figure~\ref{fig:boston_xgb}.
 \begin{example}
+p1 <- partial(boston.xgb, pred.var = "rm", plot = T, rug = T)
+p2 <- partial(boston.xgb, pred.var = "lstat", plot = T, rug = T)
+p3 <- partial(boston.xgb, pred.var = c("lstat", "rm"), plot = T, chull = T),
 grid.arrange(
-  # Figure 8 (left)
-  partial(boston.xgb.tune, pred.var = "rm", plot = T, rug = T),
-
-  # Figure 8 (middle)
-  partial(boston.xgb.tune, pred.var = "lstat", plot = T, rug = T),
-
-  # Figure 8 (right)
-  partial(boston.xgb.tune, pred.var = c("lstat", "rm"), plot = T, chull = T),
-
-  # One column for each PDP
+  p1,  # Figure 8 (left)
+  p2,  # Figure 8 (middle)
+  p3,  # Figure 8 (right)
   ncol = 3
@@
 \begin{example}
-set.seed(102)  # for reproducibility
+set.seed(203)  # for reproducibility
 boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
@@
 \end{example}
-
-% \begin{example}
-% pred.list <- list("lstat", "rm", c("lstat", "rm"))
-% for (x in pred.list) {
-%   partial(boston.xgb, pred.var = x, plot = TRUE, smooth = TRUE, rug = TRUE,
-%           chull = TRUE, train = subset(boston, select = -cmedv))
-% }
-% \end{example}
 

### 2016-12-11 manuscript/greenwell.R

@@ -88,7 +88,7 @@ p2 <- ggplot(pd.odor, aes(x = odor, y = y)) +
   theme_bw()
 
 # Display both plots side by side
-multiplot(p1, p2, ncol = 2)
+grid.arrange(p1, p2, ncol = 2)
 
 
 ################################################################################
@@ -122,7 +122,7 @@ pdf("pd_lstat.pdf", width = 8, height = 4)
 pdp1 <- plotPartial(pd.lstat)
 pdp2 <- plotPartial(pd.lstat, lwd = 2, smooth = TRUE,
                     ylab = expression(f(lstat)))
-multiplot(pdp1, pdp2, ncol = 2)
+grid.arrange(pdp1, pdp2, ncol = 2)
 dev.off()
 
 # Fit a MARS model
@@ -132,31 +132,31 @@ boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
 rwb <- colorRampPalette(c("red", "white", "blue"))
 pd.lstat.rm <- partial(boston.rf, pred.var = c("lstat", "rm"))
 pdf("pd_lstat_rm.pdf", width = 12, height = 4)
-pdp1 <- plotPartial(pd.lstat.rm)
-pdp2 <- plotPartial(pd.lstat.rm, contour = TRUE, col.regions = rwb)
-pdp3 <- plotPartial(pd.lstat.rm, levelplot = FALSE, zlab = "cmedv",
-                    drape = TRUE, colorkey = FALSE,
-                    screen = list(z = -20, x = -60))
-multiplot(pdp1, pdp2, pdp3, ncol = 3)
+p1 <- plotPartial(pd.lstat.rm)
+p2 <- plotPartial(pd.lstat.rm, contour = TRUE, col.regions = rwb)
+p3 <- plotPartial(pd.lstat.rm, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
+                  colorkey = FALSE, screen = list(z = -20, x = -60))
+grid.arrange(p1, p2, p3, ncol = 3)
 dev.off()
 
 # Figure 4
-pd.lstat <- partial(boston.rf, pred.var = "lstat")
-# pdf("partial_extrap.pdf", width = 12, height = 4)
 pdf("partial_extrap.pdf", width = 8, height = 4)
-pdp1 <- plotPartial(pd.lstat, rug = TRUE, train = boston)
-pdp2 <- plotPartial(pd.lstat.rm, chull = TRUE, train = boston)
[…]

### 2016-12-13 manuscript/reviews/response.Rmd
tweaks based on review comments

--- 2016-12-11 manuscript/reviews/response.Rmd
@@
 output:
+  html_document: default
+  html_notebook: default
   pdf_document: default
-  html_notebook: default
-  html_document: default
   word_document: default
 ---
+
+Thanks to the reviewers and editor for the well thought out comments.
+I believe that they have greatly imporved, not only the manuscript, but the package it describes as well.
+I hope that my responses are at least satisfactory.
+Please find below a copy of the reviewers comments along with my responses.
 
@@
 #### Response
-The last column is labeled `y` for consistency with `gbm` and `randomForest` behavior. Only `gbm` does not fail in this case.
+The last column is labeled `y` for consistency with `gbm` and `randomForest` behavior.
+Unfortunately, both `pdp::partial` and `randomForest::partialPlot` fail in this case; `gbm`, however, does not fail.
+I think regardless of what the default column name is, there is always the possibility of this issue.
+I filed an issue on the GitHub issue page: https://github.com/bgreenwell/pdp/issues/20.
+In the mean time, `partial` will throw an error describing the problem so that the user can take action.
 
@@
 Is there any reason the package is called `pdp` but the main function is called `partial`?
-Consider giving the function the same name as the package, so e.g the following works (to help the user get started)
+Consider giving the function the same name as the package, so e.g., the following works (to help the user get started)
 ```{r, eval=FALSE}
@@
 The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letetrs).
-At the time I decided to change the name of the package to `pdp`.
-I've personally never liked the idea of the main function name to be the same as the package name.
-However, I did add a help file for the page describing the package, main functions, etc.
+So, I decided to change the name of the package to `pdp`; besides, I think `partial` and `plotPartial` work well together as function names.
+However, I did add a help page describing the package, main functions, etc.
 So, `?pdp` no longer fails.
+I am, however, not opposed to creating an alias to `partial` called `pdp` in the future.
 
@@
 #### Response
-I dont think so and this seems to be the same behavior for the PDP functions in `randomForest` and `gbm`.
+I dont think so and this is the same behavior for the PDP functions in `randomForest` and `gbm` (not that `pdp` needs to immitate the behavior of these packages PDP functions).
 However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that the user can extract the data easily from that if necessary (e.g., if they do not like the plot and don't want to re-run partial).
@@
 ```
-However, for convenience, I have included a `multiplot` function that is just a simple wrapper for `gridExtra::grid.arrange` and should satisfy in most cases. For example,
-```{r, eval=FALSE}
-multiplot(pdp1, pdp2, ncol = 2)
-?multiplot
-```
+For convenience, `pdp` now imports the `grid.extra` function and this has been incorporated into the manuscript where appropriate.
 
@@
 I agree, this would be a nice feature, but the problem I see is that not all models emit a "variable importance".
-And those that do often have many different ways.
-For example, random forest offeres two types, which can then be scaled, etc.
-At this point, I think it is simpler to have the user select which variables they would like to see plotted.
-But perhaps in the near future.
-The best way I can see implementing this feature is to have the option when the first argument to `partial` is a `"train"` object (i.e., a model fitted using the `caret` package).
-`caret` has a generic `varImp` function that applies to a lot of model classes.
-Perhaps in the near future, a new argument to `partial`, say `top` can be added.
-So, something like
-```{r, eval=FALSE}
-library(caret)
-library(pdp)
-model <- train(x, y, method = "xgbTree", ...)
-partial(model, pred.var = 5, plot = TRUE)  # integer indicating using top 5
-```
-I added this as a feature request on the GitHub page: https://github.com/bgreenwell/pdp/issues/19.
+The `caret` package has a come a loong way in this regard.
+At this point, I think it is simpler to have the user select which variables they would like to see plotted (it forces them to explore the results).
+I am not opposed to including this feature in the future.
+I added this as a feature request: https://github.com/bgreenwell/pdp/issues/19.
 
@@
 This is also true for the fitting functions in other package (e.g., `kernlab::ksvm`).
-As time goes on, I will try to increase coverage to make it easier on the user.
+As time goes on, the coverage will certainly be inncreased to make it easier on the user.
 For example, `partial` now works with `caret` which always stores the taining data in a component called `trainingData`.
-I suspect that package maintainers, especially those for `xgboost`, would be hesitent in storing a copy of the training data as the package is often used for modeling large data sets which they do not
+I suspect that the package maintainers for `xgboost`, would be hesitent in storing a copy of the training data, especially since the package is often used for modeling large data sets; they probably do not want to store a copy of the data with the fitted model object for memory reasons.
 
@@
 Consider adding a `slowtests` directory (below `pdp/inst/` or elsewhere) that has tests for all the models supported by the package (Table 1 in the paper) to easily check back compatibility when packages change.
-(For justification for such tests see for example my
-comments on `gbm` back compatibility above.)  These tests will probably be quite slow, but that won't be an issue because they won't be invoked by CRAN check; the slow tests would be just for the maintainer of the `pdp` package to run manually.
+(For justification for such tests see for example my comments on `gbm` back compatibility above.)  These tests will probably be quite slow, but that won't be an issue because they won't be invoked by CRAN check; the slow tests would be just for the maintainer of the `pdp` package to run manually.
 
 #### Response
-This is a good suggestion and is currently being put together.
+This is a good suggestion and is currently under development. In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.
 
@@
 In a linear model with simple interactions, likewise.
-They first become important if you have non-parametric macine- optimizing techniques.
+They first become important if you have non-parametric machine-learning techniques.
 Further, they are not the only idea extant when trying to explain a predictor's impact on the response.
 This paper would benefit if you can give a bit more of a lit review here.
+
+#### Response
+I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.).
+In fact, linear models are still common in machine learning/computer science/data science (they just typically contain many terms).
+I also agree that PDPs are not the only, or sometimes best, approach to understanding the impact of a subset of the predictor's on the response.
+Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few others available in the `car` package.
+Unless the editor disagrees, I don't think it is necessary to go into more detail than what is already provided at the end of the introduction.
 
@@
 However, it is limited as you only get one dimensional representations.
-The selling point of your package should focus on the flexibility of graphics and the illustration of non- extrapoloation zones.
+The selling point of your package should focus on the flexibility of graphics and the illustration of non-extrapoloation zones.
+
+#### Response
+ICEbox does not actually construct PDPs.
+Only graphics trying to do the same thing: visualize the relationship between a predictor and the response.
+I think flexible graphics and non-extrapoloation zones are only half of the selling point here.
+I think that being able to run `partial` in parallel is huge, and also that this packge works with many types of machine learning models---not just tree-based models.
+For example, as described in the manuscript, `pdp` can produce single or multi-predictor PDPs for conditional random forests and XGBoost models (among many others), and in parallel.
+I expanded the last example to describe the interface with `caret`, which I think opens the door for wonderful exploratory data analysis pipe lines (e.g., choose model --> plot importance --> plot relationships, etc.).
+Most, if any, of these other software packages do not describe or support these important components (e.g., parallel execution), not even ICEbox.
 
@@
 # Fig 3b.
+
+#### Response
+Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produce which figure---in addition to in the text.
+Also, in response to another reviewer's comments, the code which produces figures containing multiple plots is better described and demonstrated.
+See, for example, the section called Multi-predictor PDPs.
 
@@
 p1 pairs should be using curly braces of angle braces, not square braces as this is easily confused with closed intervals.
-
 #### Response
@@
 p2 unless the editor disagrees, there is no reason to add "install.packages("pdp")" in the text of the paper as it is assumed a user knows how to install a package.
+#### Response
+Fixed. Code chunk was removed.
 
@@
 p3 "Error: The training..." should be wrapped.
-
 #### Response
-Fixed.
+Fixed. The text is now wrapped appropriately.
 
@@
 Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?
+#### Response
+The sidebar (i.e., colorkey) has been labeled, but for this example only.
+As sad as it sounds, labeling the colorkey is not as trivial as we'd expect for lattice plots.
+The code for doing so is incorporated in the example in question.
+Also, the text describing the first code chunck that produces a display with a colorkey now contains a footnote pointing to the example where the code to produce a label can be found.
+This is also motivation for why one of the examples showed how to use `ggplot2` for plotting the results from `partial`---`ggplot2` makes it a trivial matter to label the colorkey.
 
@@
 So, it is my opinion that you should take it out and force two steps: generation and illustration.
+#### Response
+While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them.
+However, I agree with your concern which is why `partial` has `plot = FALSE` as the default.
+Furthermore, I have made it possible to retrieve the raw PDP data even when \code{plot = TRUE}.
+This is because the "trellis" object returned by `partial` \ `plotPartial` now contains an additional `partial.data` attribute.
+This is described in the package documentation.
+For example,
+```{r, eval=TRUE}
+library(pdp)
+data(trees)
+mod <- lm(Volume ~ ., data = trees)
+partial(mod, pred.var = "Girth", plot = TRUE)
+pdp.Girth <- lattice::trellis.last.object()
+pd.Girth <- attr(pdp.Girth, "partial.data")
+head(pd.Girth)
+```
 
 #### Comment
-The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use ranfom forests to illustrate classification probabilities.
+The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.
+#### Response
+I agree, and I have changed the behavior of `...` to refer to additional parameters to `stats::predict` (e.g., the `n.trees` argument in `predict.gbm`.)
 
@@
 p7 I believe there are packages allowing for parallelization in Windows. Can you switch the example so this works on all platforms?
+#### Response
+Good suggestion. I tweaked the example to run in both Windows and Unix-like systems. This is also reflected in the manuscript text.

### 2016-12-13 manuscript/greenwell.tex
tweaks based on review comments

@@
 \abstract{
-Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes $E\left[\boldsymbol{\mathcal{Y}}\right] = \boldsymbol{X}\boldsymbol{\beta}$).
+Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity and normality).
 Unfortunately, it can be challenging to understand the results of such models and explain them to management.
@@
 \pkg{lattice} itself is built on top of \CRANpkg{grid} \citep{grid-pkg}.
-\pkg{grid} graphics behave a little differently than traditional R graphics and two points are worth making (see \code{?lattice} for more details):
+\pkg{grid} graphics behave a little differently than traditional R graphics, and two points are worth making (see \code{?lattice} for more details):
 \begin{enumerate}
@@
 \end{enumerate}
-A consequence of second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots.
-Simple solutions are available in packages \CRANpkg{latticeExtra} and \CRANpkg{gridExtra}.
+A consequence of the second point is that the \code{par} function cannot be used to control the layout of multiple \pkg{lattice} (and hence \pkg{pdp}) plots.
+Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}.
 For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.
@@
 \end{example}
-This will occur, for example, when using partial with the \code{"xgb.Booster"} class.
+This will occur, for example, when using partial with the \code{"xgb.Booster"} class (see Section~\ref{sec:xgboost}).
 In fact, using \code{partial} with any object that does not store the original training data will result in the above error message.
@@
 The \code{partial} function also supports object of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}.
-This means that \code{partial} can be used with any classification or regression model that \code{caret} supports; see \url{http://topepo.github.io/caret/available-models.html} for a current list of supported models.
-An example is given in Section~?.
+This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}.
+An example is given in Section~\ref{sec:xgboost}.
 
@@
 boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
-varImpPlot(boston.rf)  # variable importance plot
+varImpPlot(boston.rf)  # Figure 1
 \end{example}
@@
 For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
-The results, which make use of various \code{plotPartial} options, are displayed in Figure~\ref{fig:pd_lstat_rm}.
+The \code{grid.arrange} function is then used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph.
+The result is displayed in Figure~\ref{fig:pd_lstat_rm}.
 \begin{example}
@@
 pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
-plotPartial(pd)  # Figure 3 (left)
-plotPartial(pd, contour = TRUE, col.regions = rwb)  # Figure 3 (middle)
+grid.arrange(
+  # Figure 3 (left)
+  plotPartial(pd),
+  # Figure 3 (middle)
+  plotPartial(pd, contour = TRUE, col.regions = rwb),
+  # Figure 3 (right)
   plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
-            colorkey = TRUE, screen = list(z = -20, x = -60))  # Figure 3 (right)
+              colorkey = TRUE, screen = list(z = -20, x = -60)),
+  ncol = 3
+)
 \end{example}
@@
 \subsection{Addressing computational concerns}
+\label{sec:computational}
 
@@
 \begin{example}
-p1 <- partial(boston.rf, "rm", plot = TRUE)
-p2 <- partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE)
-p3 <- partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE)
 grid.arrange(
-  p1,  # Figure 5 (left)
-  p2,  # Figure 5 (middle)
-  p3,  # Figure 5 (right)
+  # Figure 5 (left)
+  partial(boston.rf, "rm", plot = TRUE),
+  # Figure 5 (middle)
+  partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
+  # Figure 5 (right)
+  partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
   ncol = 3
@@
 \begin{example}
-library(doParallel)  # load parallel backend
+library(doParallel)  # load the parallel backend
 cl <- makeCluster(4)  # use 4 workers
-registerDoParallel(cl)
+registerDoParallel(cl)  # register the parallel backend
 \end{example}
@@
 This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
+The last three lines of code simply add a label to the colorkey.
 The result is displayed in Figure~\ref{fig:partial_par}.
@@
 stopCluster(cl)  # good practice
+
+# Add a label to the colorkey
+lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
+grid::grid.text("ozone", x = 0.2, y = 1.05, hjust = 0.5, vjust = 1)
+lattice::trellis.unfocus()
 \end{example}
@@
 The associated R package \pkg{xgboost} has been used to win a number of \href{https://www.kaggle.com/}{Kaggle competitions}.
-It has been shown to be many times faster than the well known \pkg{gbm} package.
+It has been shown to be many times faster than the well-known \pkg{gbm} package.
 However, unlike \pkg{gbm}, \pkg{xgboost} does not have built-in functions for constructing PDPs.
@@
 For illustration, we return to the Boston housing data.
-The code chunk below fits an \pkg{xgboost} model to the \code{boston} data frame from Section~\ref{sec:boston}; the tuning parameters were predetermined using 5-fold cross-validation.
-(Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.)
-
-The code chunk below uses \pkg{caret} to train an XGBoost model over a grid of tuning values using 5-fold cross-validation.
+The code chunk below uses \pkg{caret} to train an XGBoost model to the \code{boston} data frame from Section~\ref{sec:boston} over a grid of tuning values (see \code{?xgboost::xgboost} for an explanation of the various tuning parameters).
+The optimal combination of tuning parameter values is selected via 5-fold cross-validation.
 \begin{example}
-# Optimize tuning parameters using 5-fold cross-validation
 library(caret)
@@
 
-# Grid of tuning values for XGBoost
+# Grid of tuning values for XGBoost.
 xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
@@
 The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 2000}, \code{max\_depth = 3}, \code{eta = 0.01}, and achieved a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively.
-The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display al three PDPs on the same page.
+The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page.
 The results are displayed in Figure~\ref{fig:boston_xgb}.
 \begin{example}
-p1 <- partial(boston.xgb, pred.var = "rm", plot = T, rug = T)
-p2 <- partial(boston.xgb, pred.var = "lstat", plot = T, rug = T)
-p3 <- partial(boston.xgb, pred.var = c("lstat", "rm"), plot = T, chull = T),
 grid.arrange(
-  p1,  # Figure 8 (left)
-  p2,  # Figure 8 (middle)
-  p3,  # Figure 8 (right)
+  # Figure 8 (left)
+  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
+  # Figure 8 (middle)
+  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
+  # Figure 8 (right)
+  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
   ncol = 3
@@
 The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown).
-\strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
+(Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
 \begin{example}

### 2016-12-13 manuscript/reviews/response.Rmd
tweaks based on review comments

@@
 #### Response
-I agree, and I have changed the behavior of `...` to refer to additional parameters to `stats::predict` (e.g., the `n.trees` argument in `predict.gbm`.)
+I agree, and I have changed the behavior of `...` to refer to additional parameters to `stats::predict` (e.g., the `n.trees` argument in `predict.gbm`.).
+However, I'm not sure yet about the scale on which the predictions are plotted.
+For now, I am more comfortable leaving as the usual logit scale defined for PDPs (mostly so that they are defined the same whatever the number of class outcomes).
+But I am not opposed to making this (simple) change in the future.
+I have added this as a feature request: https://github.com/bgreenwell/pdp/issues/21.
 

### 2016-12-15 README.Rmd
tweaks based on review comments

@@
 You may also reduce the computation time via the `grid.resolution` option in the call to `partial`.
-```{r, message = FALSE, warning=FALSE, fig.show="hold"}
+```{r, message = FALSE, warning=FALSE, fig.width=12, fig.height=4}
 # Load required packages
@@
 # Partial dependence of lstat and rm on cmedv
-partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE)
-partial(boston.rf, pred.var = "rm", plot = TRUE, rug = TRUE)
-partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE)
+grid.arrange(
+  partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE),
+  partial(boston.rf, pred.var = "rm", plot = TRUE, rug = TRUE),
+  partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
+  ncol = 3
+)
+
 ```
@@
 # Partial dependence of glucose and age on diabetes test result (neg/pos). 
-partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE)
+partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE,
+        train = pima)
 ```
@@
 Note, when training a model using `caret`'s `train` function, you can view tuning progress by setting `verboseIter = TRUE` in the call to `trainControl`.
-```{r, message=FALSE, warning=FALSE}
+```{r, message=FALSE, warning=FALSE, fig.width=15, fig.height=5}
 # Load required packages
@@
 # Partial dependence of glucose and age on diabetes test result (neg/pos)
-partial(pima.xgb, pred.var = "glucose", plot = TRUE, rug = TRUE)
-partial(pima.xgb, pred.var = "age", plot = TRUE, rug = TRUE)
-partial(pima.xgb, pred.var = "mass", plot = TRUE, chull = TRUE)
+grid.arrange(
+  partial(pima.xgb, pred.var = "glucose", plot = TRUE, rug = TRUE),
+  partial(pima.xgb, pred.var = "age", plot = TRUE, rug = TRUE),
+  partial(pima.xgb, pred.var = "mass", plot = TRUE, rug = TRUE),
+  ncol = 3 
+)
 ```

### 2016-12-15 R/pdClassification.R

@@ -13,7 +13,7 @@ pdClassification.default <- function(object, pred.var, pred.grid, which.class,
     temp <- train
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, type = "prob", ...)
-    stats::setNames(avgLogit(pr, which.class = which.class), "y")
+    stats::setNames(avgLogit(pr, which.class = which.class), "yhat")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -26,7 +26,8 @@ pdClassification.glm <- function(object, pred.var, pred.grid, which.class,
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, type = "response", ...)
     # Binary regression returns a vector of predicted probabilities!
-    stats::setNames(avgLogit(cbind(pr, 1 - pr), which.class = which.class), "y")
+    stats::setNames(avgLogit(cbind(pr, 1 - pr),
+                             which.class = which.class), "yhat")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -38,7 +39,7 @@ pdClassification.bagging <- function(object, pred.var, pred.grid, which.class,
     temp <- train
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, ...)$prob
-    stats::setNames(avgLogit(pr, which.class = which.class), "y")
+    stats::setNames(avgLogit(pr, which.class = which.class), "yhat")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -50,7 +51,7 @@ pdClassification.boosting <- function(object, pred.var, pred.grid, which.class,
     temp <- train
     temp[pred.var] <- x
     pr <- stats::predict(object, newdata = temp, ...)$prob
-    stats::setNames(avgLogit(pr, which.class = which.class), "y")
+    stats::setNames(avgLogit(pr, which.class = which.class), "yhat")
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
@@ -64,7 +65,8 @@ pdClassification.gbm <- function(object, pred.var, pred.grid, which.class,
     temp[pred.var] <- x
     log <- utils::capture.output(pr <- stats::predict(object, newdata = temp,
                                                       type = "response", ...))
-    stats::setNames(avgLogit(cbind(pr, 1 - pr), which.class = which.class), "y")
[…]

### 2016-12-15 R/plotPartial.R

@@ -106,7 +106,7 @@ plotPartial.partial <- function(x, smooth = FALSE, rug = FALSE, chull = FALSE,
     }
 
   }
-  
+
   # Print and return (invisibly) the "trellis" object
   p
 
@@ -115,7 +115,7 @@ plotPartial.partial <- function(x, smooth = FALSE, rug = FALSE, chull = FALSE,
 
 #' @keywords internal
 pdpNumeric <- function(x, smooth, rug, train = NULL, ...) {
-  xyplot(stats::as.formula(paste("y ~", names(x)[1L])), data = x,
+  xyplot(stats::as.formula(paste("yhat ~", names(x)[1L])), data = x,
          type = "l", ..., panel = function(xx, yy, ...) {
            panel.xyplot(xx, yy, col = "black", ...)
            if (smooth) {
@@ -135,13 +135,13 @@ pdpNumeric <- function(x, smooth, rug, train = NULL, ...) {
 
 #' @keywords internal
 pdpFactor <- function(x, ...) {
-  dotplot(stats::as.formula(paste("y ~", names(x)[1L])), data = x, ...)
+  dotplot(stats::as.formula(paste("yhat ~", names(x)[1L])), data = x, ...)
 }
 
 
 #' @keywords internal
 pdpFactorFactor <- function(x, ...) {
-  dotplot(stats::as.formula(paste("y ~",
+  dotplot(stats::as.formula(paste("yhat ~",
                                   paste(names(x)[1L:2L], collapse = "|"))),
           data = x, ...)
 }
@@ -152,9 +152,9 @@ pdpNumericFactor <- function(x, smooth, rug, train, ...) {
 
   # Lattice plot formula
   form <- if (is.factor(x[[1L]])) {
-    stats::as.formula(paste("y ~", paste(names(x)[2L:1L], collapse = "|")))
+    stats::as.formula(paste("yhat ~", paste(names(x)[2L:1L], collapse = "|")))
   } else {
-    stats::as.formula(paste("y ~", paste(names(x)[1L:2L], collapse = "|")))
+    stats::as.formula(paste("yhat ~", paste(names(x)[1L:2L], collapse = "|")))
   }
 
   # Produce a paneled lineplot
@@ -182,7 +182,7 @@ pdpNumericNumeric <- function(x, levelplot, rug, chull, train, contour,
                               col.regions, ...) {
 
   # Lattice plot formula
-  form <- stats::as.formula(paste("y ~",
+  form <- stats::as.formula(paste("yhat ~",
[…]

### 2016-12-15 manuscript/greenwell.tex
tweaks based on review comments

@@
 \abstract{
-Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity and normality).
+Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality).
 Unfortunately, it can be challenging to understand the results of such models and explain them to management.
@@
 \section{Introduction}
+
+\citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data.
+One of their goals was to find a housing value equation using data on median housing values from 506 census tracts in the suburbs of Boston from the 1970 census.
+The prediction equation from one of there regression analyses is reproduced in Equation~\eqref{eqn:boston}.
+\citet{harrison-1978-hedonic} took great care and time to find this model which obtained an $R^2$ of roughly 0.81.
+Part of the payoff for there efforts was a reaonable fitting model with an interpretable equation.
+For example, the equation tells us that an increase in median home value is associated with an increase in the average number of rooms per dwelling.
+\begin{equation}
+\label{eqn:boston}
+\begin{split}
+\widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
+  &\quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
+  &\quad - 0.012 CRIM + 8.03\time10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS - 0.0064 NOX^2
+\end{split}
+\end{equation}
+
+Many modern nonparametric models, however, can fit the data automatically in seconds---typically with higher accuracy.
+The downfall, however, is some loss of interpretation; for example, there is usually no prediction formula like in Equation~\eqref{eqn:boston}.
+Fortunately, methods like boosting and random forest have a natural way to assign variable importance scroes to the predictors which helps gives some insight into the data.
 

### 2016-12-15 manuscript/greenwell.tex
tweaks based on review comments

@@
 \citet{harrison-1978-hedonic} took great care and time to find this model which obtained an $R^2$ of roughly 0.81.
-Part of the payoff for there efforts was a reaonable fitting model with an interpretable equation.
+Part of the payoff for there efforts was a reasonable fitting model with an interpretable equation.
 For example, the equation tells us that an increase in median home value is associated with an increase in the average number of rooms per dwelling.
@@
 The downfall, however, is some loss of interpretation; for example, there is usually no prediction formula like in Equation~\eqref{eqn:boston}.
-Fortunately, methods like boosting and random forest have a natural way to assign variable importance scroes to the predictors which helps gives some insight into the data.
+Fortunately, these methods can still provide insight into the data, but it is not in the form of simple equations.
+For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors.
 

### 2016-12-15 manuscript/greenwell.tex
tweaks based on review comments

@@
 If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
-The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{y} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
+The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{yhat} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
 If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
@@
 library(ggplot2)
-ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
+ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = yhat, fill = yhat)) +
   geom_tile() +

### 2016-12-15 manuscript/greenwell.R

@@ -56,42 +56,6 @@ library(xgboost)
 
 
 ################################################################################
-# Mushroom example
-################################################################################
-
-# Train a random forest to the mushroom data using 10-fold cross-validation
-set.seed(101)
-mushroom.rf <- train(x = subset(mushroom, select = -Edibility),
-                     y = mushroom$Edibility,
-                     method = "rf",
-                     trControl = trainControl(method = "cv", number = 10,
-                                              verboseIter = TRUE),
-                     tuneLength = 22)
-
-# Extract partial dependence of odor on odds of edibility
-pd.odor <- partial(mushroom.rf, pred.var = "odor")
-
-# Variable importance plot
-p1 <- ggplot(varImp(mushroom.rf), color = "black") +
-  theme_light() +
-  xlab("") +
-  ylab("Predictor importance") +
-  theme_bw()
-
-# Partial dependence plot
-p2 <- ggplot(pd.odor, aes(x = odor, y = y)) +
-  geom_bar(stat = "identity") +
-  geom_hline(yintercept = 0) +
-  xlab("Odor") +
-  ylab("Log odds of being edible") +
-  coord_flip() +
-  theme_bw()
-
-# Display both plots side by side
-grid.arrange(p1, p2, ncol = 2)
-
-
-################################################################################
 # Boston housing example
 ################################################################################
 
@@ -224,12 +188,12 @@ iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
 pd <- NULL
 for (i in 1:3) {
   tmp <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-                 which.class = i, .progress = "text",
+                 which.class = i, progress = "text",
                  grid.resolution = 101)
   pd <- rbind(pd, cbind(tmp, Species = levels(iris$Species)[i]))
 }
 pdf("partial_iris_svm.pdf", width = 12, height = 4)
-ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = y, fill = y)) +
+ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = yhat, fill = yhat)) +
   geom_tile() +
   geom_contour(color = "white", alpha = 0.5) +
   scale_fill_distiller(name = "Logit", palette = "Spectral") +

### 2016-12-16 manuscript/reviews/response.Rmd
tweaks based on review comments

@@
 
-## Reviewer 1
+## ----------Reviewer 1----------
 
@@
 ```
-
 #### Response
-This is simply a side effect of using `lattice` to produce the plots. If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))`).
+This is simply a side effect of using `lattice` to produce the plots.
+If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))`).
+This has been explained in both the documentation and the manuscript in a **Note:** on 3.
 
@@
 ```
-
 #### Response
-This been fixed in `pdp` (version 0.3.0), but requires the user to supply `n.trees` in the call to `partial`.
-So, `partial(gbm.mod, pred.var="age", n.trees=100)` should work now.
+The `gbm` maintainer did contact me a while back and this has been fixed in `pdp` (version 0.3.0).
+The user will still have to supply `n.trees` in the call to `partial`.
 The test suite is being expanded, as suggested in a later comment, to help catch more errors like this.
@@
 Standard partial matching isn't implemented (e.g., `partial(..., type="r")`).
-
 #### Response
-Partial matching is now implemented, but only for the `type` argument.
+Partial matching is now implemented, but only for the `type` argument. Thanks for the suggestion.
 
@@
 ```
-
 #### Response
@@
 data used to build the model).
+#### Response
+Aggred!
+Unfortunately, some functions just don't store a copy of the training data.
+I have expanded the code to work harder, especially for `S4` methods, to inform the user to supply the training data when necessary.
+The documentation for `partial` has also been expanded to discuss this.
 
@@
 ```
-Error in partial.default(mod, pred.var = "nosuchvariable", plot = TRUE) : 
-  nosuchvariable not found in the training data.
+Error in partial.default(mod, pred.var = c("Girth", "foo", "bar"), plot = TRUE) : foo, bar not found in the training data.
 ```
@@
 What happens if `y` is one of the predictors?
-
 #### Response
-The last column is labeled `y` for consistency with `gbm` and `randomForest` behavior.
-Unfortunately, both `pdp::partial` and `randomForest::partialPlot` fail in this case; `gbm`, however, does not fail.
+The last column is now labeled `yhat`; which is less likely to be a predictor name.
+When one of the predictors is labeled `"y"` `randomForest::partialPlot` fails while `gbm::plot.gbm` does not.
 I think regardless of what the default column name is, there is always the possibility of this issue.
-I filed an issue on the GitHub issue page: https://github.com/bgreenwell/pdp/issues/20.
-In the mean time, `partial` will throw an error describing the problem so that the user can take action.
+This has been filed as an issue on the `pdp` GitHib page :
 
@@
 Therefore in in the abstract for the paper, to avoid overselling consider changing "relationship between the outcome and predictors of interest can be easily understood" to "relationship between the outcome and predictors of interest can be more easily understood".
-
 #### Response
-Fixed.
+Fixed. Thanks for the suggestion.
 
@@
 ```
-
 #### Response
-Fixed.
+Fixed. Thanks for pointing that out.
 
@@
 `partial` should return the same value regardless of whether or not the `plot` argument is used?
-
 #### Response
-I dont think so and this is the same behavior for the PDP functions in `randomForest` and `gbm` (not that `pdp` needs to immitate the behavior of these packages PDP functions).
+I dont think so, and this is the same behavior for the PDP functions in `randomForest` and `gbm` (not that `pdp` needs to immitate the behavior of these packages PDP functions).
 However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that the user can extract the data easily from that if necessary (e.g., if they do not like the plot and don't want to re-run partial).
@@
 ```
-Also the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded.
+Also, the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded.
 This will be useful when the user does not store the resulting plot in the call to `partial` but needs to retrieve the outout data to avoid re-running `partial` when computation time is a concern.
@@
 ```
+Thanks for the suggestion!
 
@@
 ```
-For convenience, `pdp` now imports the `grid.extra` function and this has been incorporated into the manuscript where appropriate.
+For convenience, `pdp` now imports the `grid.extra` function and this has been incorporated into the manuscripts examples where appropriate.
+
+However, since `partial` returns a data frame by default, using `par` would work in the following case
+```{r, eval=FALSE}
+par(mfrow=c(2,2))
+plot(partial(ozone.rf, pred.var = "Temp"), type = "l")
+plot(partial(ozone.rf, pred.var = "Wind"), type = "l")
+```
 
@@
 It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.
-
 #### Response 
 I agree, this would be a nice feature, but the problem I see is that not all models emit a "variable importance".
-The `caret` package has a come a loong way in this regard.
+The `caret` package has a come a long way in this regard.
 At this point, I think it is simpler to have the user select which variables they would like to see plotted (it forces them to explore the results).
 I am not opposed to including this feature in the future.
-I added this as a feature request: https://github.com/bgreenwell/pdp/issues/19.
+I added added a possible solution along with a feature request: https://github.com/bgreenwell/pdp/issues/19.
 
@@
 Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only -- one can't access the data stored in `xgb.DMatrix`).
-
 #### Response
@@
 As time goes on, the coverage will certainly be inncreased to make it easier on the user.
-For example, `partial` now works with `caret` which always stores the taining data in a component called `trainingData`.
-I suspect that the package maintainers for `xgboost`, would be hesitent in storing a copy of the training data, especially since the package is often used for modeling large data sets; they probably do not want to store a copy of the data with the fitted model object for memory reasons.
+For example, `partial` now works with `caret` which, by default, stores the taining data in a component called `trainingData`.
+I suspect that the package maintainers for `xgboost`  would be hesitent in storing a copy of the training data, especially since the package is often used for modeling large data sets; they probably do not want to store a copy of the data with the fitted model object for memory reasons.
 
@@
 
-## Reviewer 2
+## ----------Reviewer 2----------
 

### 2016-12-16 manuscript/greenwell.tex
tweaks based on review comments

@@
 Partial dependence plots offer a simple solution.
-Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be
-more easily understood.
+Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood.
 These plots are especially useful in explaining the output from black box models.
@@
 \citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data.
-One of their goals was to find a housing value equation using data on median housing values from 506 census tracts in the suburbs of Boston from the 1970 census.
-The prediction equation from one of there regression analyses is reproduced in Equation~\eqref{eqn:boston}.
-\citet{harrison-1978-hedonic} took great care and time to find this model which obtained an $R^2$ of roughly 0.81.
-Part of the payoff for there efforts was a reasonable fitting model with an interpretable equation.
-For example, the equation tells us that an increase in median home value is associated with an increase in the average number of rooms per dwelling.
+One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see Table IV of \citet{harrison-1978-hedonic} for a description of each variable.
+The data violate many classical assumptions like linearity, normality, and constant variance.
+Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$).
+Part of the payoff for there efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
 \begin{equation}
@@
 \widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
-  &\quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
-  &\quad - 0.012 CRIM + 8.03\time10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS - 0.0064 NOX^2
+  & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
+  & \quad - 0.012 CRIM + 8.03\time10^{-5} ZN + 2.41\times10^{-4}  INDUS + 0.088 CHAS - 0.0064 NOX^2
 \end{split}
@@
 
-Many modern nonparametric models, however, can fit the data automatically in seconds---typically with higher accuracy.
-The downfall, however, is some loss of interpretation; for example, there is usually no prediction formula like in Equation~\eqref{eqn:boston}.
+Nowadays, supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
+(We will revisit the Boston housing data in Section~\ref{sec:boston}) The downfall, however, is some loss of interpretation; for example, there is usually no simple prediction formula like in Equation~\eqref{eqn:boston}.
 Fortunately, these methods can still provide insight into the data, but it is not in the form of simple equations.
-For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors.
+For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors in the training data.
 
@@
 However, due to R's automatic printing rule, the result is automatically printed when using these functions in the command line.
-If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph.
+If \code{plotPartial} is called inside of \code{source} or inside a loop (e.g., \code{for} or \code{while}), an explicit \code{print} statement is required to display the resulting graph; hence, the same is true when using \code{partial} with \code{plot = TRUE}.
   \item Setting graphical parameters via \code{par} typically has no effect on \pkg{lattice} plots.
@@
 
-The \code{partial} function also supports object of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}.
+The \code{partial} function also supports objects of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}.
 This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}.
@@
 
-Since \code{train} always stores a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}.
+%Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}.
+However, this is not the case when using the \pkg{xgboost} package directly.
+For example, the following code fits the same model using the \code{xgboost} function.
+Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object there is no need to supply it in the call to \code{partial}.
 However, this is not the case when using the \pkg{xgboost} package directly.

### 2016-12-16 manuscript/reviews/response.Rmd
tweaks based on review comments

@@
 This paper would benefit if you can give a bit more of a lit review here.
-
 #### Response
@@
 Unless the editor disagrees, I don't think it is necessary to go into more detail than what is already provided at the end of the introduction.
+Nonetheless, I added a bit more of a motivational introduction that links to the examples using the Boston housing data.
 
@@
 The selling point of your package should focus on the flexibility of graphics and the illustration of non-extrapoloation zones.
-
 #### Response
@@
 # Fig 3b.
-
 #### Response
-Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produce which figure---in addition to in the text.
+Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produce which figure.
 Also, in response to another reviewer's comments, the code which produces figures containing multiple plots is better described and demonstrated.
@@
 p1 abstract: strict assumptions imposed by traditional statistical models
+#### Response
+Added some clarification here; in particular, normality, constant variance, and linearity.
 
@@
 I agree, and I have changed the behavior of `...` to refer to additional parameters to `stats::predict` (e.g., the `n.trees` argument in `predict.gbm`.).
-However, I'm not sure yet about the scale on which the predictions are plotted.
-For now, I am more comfortable leaving as the usual logit scale defined for PDPs (mostly so that they are defined the same whatever the number of class outcomes).
+However, I'm not sure yet about the scale on which the predictions are plotted for classification problems.
+For now, I am more comfortable leaving it as the usual logit scale defined for PDPs (mostly so that they are defined the same whatever the number of class outcomes).
 But I am not opposed to making this (simple) change in the future.

### 2016-12-16 manuscript/greenwell.tex
tweaks based on review comments

@@
   & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
-  & \quad - 0.012 CRIM + 8.03\time10^{-5} ZN + 2.41\times10^{-4}  INDUS + 0.088 CHAS - 0.0064 NOX^2
+  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS - 0.0064 NOX^2
 \end{split}
@@
 Nowadays, supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
-(We will revisit the Boston housing data in Section~\ref{sec:boston}) The downfall, however, is some loss of interpretation; for example, there is usually no simple prediction formula like in Equation~\eqref{eqn:boston}.
+(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, there is usually no simple prediction formula like in Equation~\eqref{eqn:boston}.
 Fortunately, these methods can still provide insight into the data, but it is not in the form of simple equations.
@@
 census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town.
+These are the same data analyzed in \citet{harrison-1978-hedonic}, but with a few corrections and additional spatial variables.
 We begin by loading the data and omitting unimportant columns.

### 2016-12-18 manuscript/reviews/response.Rmd
rm Date field in Description

@@
 
-## ----------Reviewer 1----------
+# -----------Reviewer 1-----------
 
-#### Comment
-Bug: A graph is produced as expected if the code below is cut-and-pasted at the command line .
+### Comment 1.1
+**Bug**: A graph is produced as expected if the code below is cut-and-pasted at the command line .
 However, if the code is put into a file (say `foo.R`) and `source("foo.R")` is executed at the command line, no graph is produced.
@@
 ```
-#### Response
+
+### Response 1.1
 This is simply a side effect of using `lattice` to produce the plots.
-If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))`).
-This has been explained in both the documentation and the manuscript in a **Note:** on 3.
+If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))` or `print(partial(fit, pred.var = "x", plot = TRUE))`).
+Additional text has been added to manuscript (see **Note:** on page 3) and the package documentation explaining this behavior.
+The fact that this worked when calling `partial` with `plot = TRUE` was a bug (which is now fixed).
 
-#### Comment
+### Comment 1.2
 I couldn't get `partial` to work with `gbm` (version 2.1.1) models (although I didn't try very hard).
@@
 ```
-#### Response
-The `gbm` maintainer did contact me a while back and this has been fixed in `pdp` (version 0.3.0).
+
+### Response 1.2
+This has been fixed in `pdp` (version 0.3.0).
 The user will still have to supply `n.trees` in the call to `partial`.
-The test suite is being expanded, as suggested in a later comment, to help catch more errors like this.
+Also, as suggested in another comment, a `/slowtests` directory is underway and will help prevent future issues like this.
+Thanks for cathing this.
 
-#### Comment
+### Comment 1.3
 Standard partial matching isn't implemented (e.g., `partial(..., type="r")`).
-#### Response
-Partial matching is now implemented, but only for the `type` argument. Thanks for the suggestion.
 
-#### Comment
+### Response 1.3
+Partial matching is now implemented (as of `pdp` version 0.3.0), but only for the `type` argument. Thanks for the suggestion.
+
+### Comment 1.4
 The following code incorrectly gives an error message
@@
 library(pdp)
-data <- data.frame(V1=1:10, V2=1:10)
-mod <- lm(V2~V1, data=data)
-partial(mod, pred.var="V1", plot=TRUE) # error
+data <- data.frame(V1 = 1:10, V2 = 1:10)
+mod <- lm(V2 ~ V1, data = data)
+partial(mod, pred.var = "V1")  # error
 ```
-#### Response
+
+### Response 1.4
 This has been fixed in `pdp` (version 0.3.0).
@@
 library(pdp)
-data <- data.frame(V1=1:10, V2=1:10)
-mod <- lm(V2~V1, data=data)
-partial(mod, pred.var="V1") # no more error
+data <- data.frame(V1 = 1:10, V2 = 1:10)
+mod <- lm(V2 ~ V1, data = data)
+partial(mod, pred.var = "V1")  # no more error
 ```
 
-#### Comment
+### Comment 1.5
 I think you should mention in the documentation something about the assumptions you make about accessing the model data. The following (somewhat contrived) example fails with
@@
 {
-    lm(Volume~., data=data)
+    lm(Volume ~ ., data = data)
 }
@@
 library(pdp)
-partial(mod, pred.var="Girth", plot=TRUE) # fails
+partial(mod, pred.var = "Girth", plot = TRUE)  # fails
 ```
@@
 data(trees)
-mod <- lm(Volume~., data=trees)
-partial(mod, pred.var="Girth", plot=TRUE) # ok
-trees <- trees[1:2,]
-partial(mod, pred.var="Girth", plot=TRUE) # wrong xlim
+mod <- lm(Volume ~ ., data = trees)
+partial(mod, pred.var = "Girth", plot = TRUE)  # ok
+trees <- trees[1:2, ]
+partial(mod, pred.var = "Girth", plot = TRUE)  # wrong xlim
 ```
@@
 data used to build the model).
-#### Response
-Aggred!
+
+### Response 1.5
+Agreed!
+The documentation for `partial` has also been expanded to discuss this.
 Unfortunately, some functions just don't store a copy of the training data.
-I have expanded the code to work harder, especially for `S4` methods, to inform the user to supply the training data when necessary.
-The documentation for `partial` has also been expanded to discuss this.
+However, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary.
 
-#### Comment
+### Comment 1.6
 The following code gives an obscure error message:
@@
 ```
-#### Response
+
+### Response 1.6
 This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
@@
 
-#### Comment
+### Comment 1.7
 On page 2 of the paper it says "The columns of the data frame are labeled in the same order as the features supplied to `pred.var`, and the last column is always labeled `y`".
 What happens if `y` is one of the predictors?
-#### Response
+
+### Response 1.7
 The last column is now labeled `yhat`; which is less likely to be a predictor name.
@@
 I think regardless of what the default column name is, there is always the possibility of this issue.
-This has been filed as an issue on the `pdp` GitHib page :
+This has been filed as an issue on the `pdp` GitHib page: https://github.com/bgreenwell/pdp/issues/22.
 
-#### Comment
+### Comment 1.8
 Hastie et al.
@@
 Therefore in in the abstract for the paper, to avoid overselling consider changing "relationship between the outcome and predictors of interest can be easily understood" to "relationship between the outcome and predictors of interest can be more easily understood".
-#### Response
+
+### Response 1.8
 Fixed. Thanks for the suggestion.
 
-#### Comment
+### Comment 1.9
 On page 3 in the PDF of the paper, add a newline to the error message in the paper:
@@
 ```
-#### Response
+
+### Response 1.9
 Fixed. Thanks for pointing that out.
 
-#### Comment
+### Comment 1.10
 Is there any reason the package is called `pdp` but the main function is called `partial`?
@@
 ```
-#### Response
+
+### Response 1.10
 The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letetrs).
@@
 So, `?pdp` no longer fails.
-I am, however, not opposed to creating an alias to `partial` called `pdp` in the future.
+However, I am not opposed to creating an alias to `partial` called `pdp` in the future.
 
-#### Comment
+### Comment 1.11
 `partial` should return the same value regardless of whether or not the `plot` argument is used?
-#### Response
-I dont think so, and this is the same behavior for the PDP functions in `randomForest` and `gbm` (not that `pdp` needs to immitate the behavior of these packages PDP functions).
-However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that the user can extract the data easily from that if necessary (e.g., if they do not like the plot and don't want to re-run partial).
+
+### Response 1.11
+I dont think so, because I want `partial` to behave like a `lattice` function when using `plot = TRUE` or `plotPartial`; that is, I want it to return a `"trellis"` object.
+However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that it can easily be extracted if necessary (e.g., if they do not like the plot and don't want to re-run partial).
 For example,
@@
 Also, the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded.
-This will be useful when the user does not store the resulting plot in the call to `partial` but needs to retrieve the outout data to avoid re-running `partial` when computation time is a concern.
+This will be useful when the user does not store the resulting plot in the call to `partial` but needs to retrieve the outout data to avoid re-running `partial` (e.g., when computation time is a concern).
 For example,
-```{r}
+```{r, fig.width=4, fig.height=4}
 library(pdp)
@@
 ```
-Thanks for the suggestion!
+This is described in the package documentation. Thanks for the idea!
 
-#### Comment
+### Comment 1.12
 A call to say `par(mfrow = c(2, 2))` before calling `plotPartial` is ignored by `plotPartial`.  This is a pity because it would allow multiple plots to be put on one page like this:
 ```{r, eval=FALSE}
-par(mfrow=c(2,2))
+par(mfrow = c(2, 2))
 partial(ozone.rf, pred.var = "Temp", plot = TRUE)
@@
 ```
-#### Response
-The command `par(mfrow = c(2, 2))` does not work with `"trellice"` objects, such as those produced by `lattice` which `partial` relies on for producing its plots.
+
+### Response 1.12
+The command `par(mfrow = c(2, 2))` does not work with `"trellis"` objects, such as those produced by `lattice`---which `partial` relies on for producing its plots.
 Other methods are available.
-For example, using the `gridExtra` package:
-```{r, eval=TRUE}
+For convenience, `pdp` now imports the `grid.arrange` function from the `gridxtra` package (this has been discussed in the manuscript and incorporated into the examples where appropriate).
+```{r, include=FALSE}
 library(randomForest)
 data(airquality)
-ozone.rf <- randomForest(Ozone ~ ., data=airquality, na.action=na.omit)
+ozone.rf <- randomForest(Ozone ~ ., data = airquality, na.action = na.omit)
 library(pdp)
-pdp1 <- partial(ozone.rf, pred.var = "Temp", plot = TRUE)
-pdp2 <- partial(ozone.rf, pred.var = "Wind", plot = TRUE)
-gridExtra::grid.arrange(pdp1, pdp2, ncol = 2)
 ```
-For convenience, `pdp` now imports the `grid.extra` function and this has been incorporated into the manuscripts examples where appropriate.
-
+```{r, eval=TRUE, fig.width=8, fig.height=4}
+grid.arrange(
+  partial(ozone.rf, pred.var = "Temp", plot = TRUE),
+  partial(ozone.rf, pred.var = "Wind", plot = TRUE),
+  ncol = 2
+)
+```
 However, since `partial` returns a data frame by default, using `par` would work in the following case
@@
 
-#### Comment
+### Comment 1.13
 It would be nice if the functions automatically plotted all the (important) variables in a grid of plots, without forcing the user to explicitly specify them.
-#### Response 
-I agree, this would be a nice feature, but the problem I see is that not all models emit a "variable importance".
+
+### Response 1.13
+I agree, this would be a nice feature, but the problem I see is that not all models naturally emit a "variable importance" score.
 The `caret` package has a come a long way in this regard.
@@
 I added added a possible solution along with a feature request: https://github.com/bgreenwell/pdp/issues/19.
+Thanks for the suggestion!
 
-#### Comment
+### Comment 1.14
 It would be nice if the functions automatically got the data from the model used to build the model in more cases without forcing the user to explicitly specify the data.
-Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only -- one can't access the data stored in `xgb.DMatrix`).
-#### Response
-This is also true for the fitting functions in other package (e.g., `kernlab::ksvm`).
-As time goes on, the coverage will certainly be inncreased to make it easier on the user.
+Since XGBoost seems now quite popular, perhaps the authors of that package could be persuaded to make the original model data available for functions like partial (currently XGBoost uses a custom object `xgb.DMatrix` which is currently write-only---one can't access the data stored in `xgb.DMatrix`).
+
+### Response 1.14
+As mentioned in `Response 1.?`, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary.
+As another example, `kernlab::ksvm` does not store a copy of the training data, at least not as far as I can tell.
+As time goes on, the coverage will certainly improve, which will make it easier on the user.
 For example, `partial` now works with `caret` which, by default, stores the taining data in a component called `trainingData`.
-I suspect that the package maintainers for `xgboost`  would be hesitent in storing a copy of the training data, especially since the package is often used for modeling large data sets; they probably do not want to store a copy of the data with the fitted model object for memory reasons.
 
-#### Comment
+I suspect that the package maintainers for `xgboost` (and similar packages for designed for working with "large" data sets) would be hesitent in storing a copy of the training data with the fitted model object for memory reasons.
+
+### Comment 1.15
 Consider adding a `slowtests` directory (below `pdp/inst/` or elsewhere) that has tests for all the models supported by the package (Table 1 in the paper) to easily check back compatibility when packages change.
@@
 
-#### Response
-This is a good suggestion and is currently under development. In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.
+### Response 1.15
+This is a good suggestion and is currently under development: https://github.com/bgreenwell/pdp/blob/master/slowtests/slowtests.R.
+In fact, the `gbm` maintainers have already contacted me about a breaking change caused by the next planned release of their package.
 
-## ----------Reviewer 2----------
+# -----------Reviewer 2-----------
 
-#### Comment 
+### Comment
 The Introduction immediately introduces the PDP.
@@
 This paper would benefit if you can give a bit more of a lit review here.
-#### Response
+
+### Response
 I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.).
@@
 
-### Important Clarifications Required:
+## Important Clarifications Required:
 
-#### Comment
+### Comment
 Top of p2.
@@
 The selling point of your package should focus on the flexibility of graphics and the illustration of non-extrapoloation zones.
-#### Response
-ICEbox does not actually construct PDPs.
+
+### Response
+`ICEbox` does not actually construct PDPs.
 Only graphics trying to do the same thing: visualize the relationship between a predictor and the response.
@@
 
-#### Comment
+### Comment
 All `pdp` figures throughout the paper: Can you make explicit in the caption what code produces the left and right subfigures?
@@
 # Fig 3b.
-#### Response
+### Response
 Rather than using subfigures, which I think would look messy with all the code, I made quite an attempt to comment which pieces of code produce which figure.
@@
 
-#### Comment
+### Comment
 p1 abstract: strict assumptions imposed by traditional statistical models
-#### Response
+### Response
 Added some clarification here; in particular, normality, constant variance, and linearity.
 
-#### Comment
+### Comment 2.5
 p1 pairs should be using curly braces of angle braces, not square braces as this is easily confused with closed intervals.
-#### Response
+
+### Response 2.5
 Fixed. Switched to curly braces.
 
-#### Comment
+### Comment 2.6
 p2 unless the editor disagrees, there is no reason to add "install.packages("pdp")" in the text of the paper as it is assumed a user knows how to install a package.
-#### Response
+
+### Response 2.6
 Fixed. Code chunk was removed.
 
-#### Comment
+### Comment 2.7
 p3 "Error: The training..." should be wrapped.
-#### Response
-Fixed. The text is now wrapped appropriately.
 
-#### Comment
+### Response 2.7
+Fixed. The text is now wrapped appropriately. Thanks for catching this!
+
+### Comment 2.8
 Fig 6: the sidebar should be labeled "ozone". Also, would it be terrible to label all axes even though it is redundant?
-#### Response
+
+### Response 2.8
 The sidebar (i.e., colorkey) has been labeled, but for this example only.
-As sad as it sounds, labeling the colorkey is not as trivial as we'd expect for lattice plots.
-The code for doing so is incorporated in the example in question.
+As sad as it sounds, labeling the colorkey is not as trivial as we'd expect for `lattice` plots.
+The code for doing so has been incorporated in the example in question (see pages 8-9).
 Also, the text describing the first code chunck that produces a display with a colorkey now contains a footnote pointing to the example where the code to produce a label can be found.
@@
 
-#### Comment
+### Comment 2.9
 Fig 7: as beautiful as this is, you should label the sidebar as "logit" or "centered logit" instead of "y".
-#### Response
-Fixed. Color key labeled "Average logit".
+
+### Response 2.9
+Fixed. Color key labeled "Logit".
 
@@
 
-#### Comment
+### Comment
 I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP.
@@
 So, it is my opinion that you should take it out and force two steps: generation and illustration.
-#### Response
+### Response
 While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them.
@@
 
-#### Comment
+### Comment
 The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.
-#### Response
+### Response
 I agree, and I have changed the behavior of `...` to refer to additional parameters to `stats::predict` (e.g., the `n.trees` argument in `predict.gbm`.).
@@
 
-#### Comment
+### Comment 2.12
 p7 I believe there are packages allowing for parallelization in Windows. Can you switch the example so this works on all platforms?
-#### Response
-Good suggestion. I tweaked the example to run in both Windows and Unix-like systems. This is also reflected in the manuscript text.
+
+### Response 2.12
+Good suggestion. I tweaked the example to run on both Windows and Unix-like systems. This is also reflected in the manuscript text.

### 2016-12-18 R/partial.R

@@ -70,6 +70,9 @@
 #' avoided when the training data are passed to the \code{train} argument in the
 #' call to \code{partial}.
 #'
+#' It is poosible to retrieve the most recent PDP constructed by \code{partial}
+#' even after it is deleted. Simply use \code{trellis.last.object()}.
+#'
 #' @references
 #' J. H. Friedman. Greedy function approximation: A gradient boosting machine.
 #' \emph{Annals of Statistics}, \bold{29}: 1189-1232, 2000.
@@ -172,6 +175,13 @@ partial.default <- function(object, pred.var, pred.grid, grid.resolution = NULL,
         stop(mssg)
       } else {
         train <- eval(object$call$data)
+        if (!(is.data.frame(train))) {
+          if (is.matrix(train)) {
+            train <- as.data.frame(train)
+          } else {
+            stop(mssg)
+          }
+        }
       }
     }
   }

### 2016-12-18 R/utils.R

@@ -24,6 +24,19 @@ NULL
 NULL
 
 
+#' Retrieve the last Trellis Object
+#'
+#' See \code{\link[lattice]{trellis.last.object}} for more details.
+#'
+#' @name trellis.last.object
+#' @rdname trellis.last.object
+#' @keywords internal
+#' @export
+#' @importFrom lattice  trellis.last.object
+#' @usage trellis.last.object(..., prefix)
+NULL
+
+
 #' @keywords internal
 copyClasses <- function(x, y) {
   x.names <- names(x)

### 2016-12-19 R/partial.R
much faster for  objects

@@ -197,46 +197,76 @@ partial.default <- function(object, pred.var, pred.grid, grid.resolution = NULL,
     stop("\"y\" cannot be a predictor name.")
   }
 
-  # NOTE: It is worth considering the approach taken by the plotmo package,
-  # which now has the option to compute PDPs. plotmo seems to make one large
-  # pred.grid which increases memory usage, but decreases the number of calls to
-  # stats::predict.
+  # Construct partial dependence data
+  if (inherits(object, "gbm") && missing(pred.grid)) {
 
-  # Predictor values of interest
-  if (missing(pred.grid)) {
-    pred.grid <- predGrid(object, pred.var = pred.var, train = train,
-                          grid.resolution = grid.resolution)
-  }
+    # Assign value to grid.resolution
+    if (is.null(grid.resolution)) {
+      grid.resolution <- 100
+    }
 
-  # Make sure each column has the correct class, factor levels, etc.
-  if (check.class) {
-    pred.grid <- copyClasses(pred.grid, train)
-  }
+    # FIXME: This behavior needs to be documented!
 
-  # Restrict grid to covext hull of first two columns
-  if (chull) {
-    pred.grid <- trainCHull(pred.var, pred.grid = pred.grid, train = train)
-  }
+    # Call gbm::plot.gbm directly; this is MUCH FASTER!
+    pd.df <- gbm::plot.gbm(object, i.var = pred.var,
+                           continuous.resolution = grid.resolution,
+                           return.grid = TRUE)
 
-  # Determine the type of supervised learning used
-  type <- match.arg(type)
-  if (type == "auto") {
-    type <- superType(object)
-  }
+    # Restrict to convex hull of first two predictors, if requested
+    if (chull) {
+      pd.df <- if (length(pred.var) >= 2 && is.numeric(pd.df[[1L]]) &&
+          is.numeric(pd.df[[2L]])) {
+        X <- stats::na.omit(data.matrix(train[pred.var[1L:2L]]))
+        Y <- stats::na.omit(data.matrix(pd.df[, -ncol(pd.df),
+                                              drop = FALSE][1L:2L]))
+        hpts <- grDevices::chull(X)
+        hpts <- c(hpts, hpts[1L])
+        keep <- mgcv::in.out(X[hpts, ], Y)
+        pd.df[keep, ]
+      } else {
+        pd.df
+      }
+    }
 
-  # Calculate partial dependence values
-  if (type == "regression") {
[…]

### 2016-12-19 slowtests/slowtests.R

@@ -7,13 +7,77 @@
 # Setup
 ################################################################################
 
+# Make sure all packges are up to date
+# update.packages()
+
 # Load required packages
-library(caret)
 library(pdp)
 
 # Load data
 data(boston)  # regression
 data(pima)  # classification
+
+
+################################################################################
+# gbm package
+################################################################################
+
+# Load required package
+library(gbm)
+
+
+# Regression -------------------------------------------------------------------
+
+# Fit a gbm to the Boston housing data
+set.seed(101)
+boston.gbm <- gbm(cmedv ~ .,
+                  data = boston,
+                  distribution = "gaussian",
+                  n.trees = 10000,
+                  cv.folds = 10,
+                  interaction.depth = 3,
+                  shrinkage = 0.01,
+                  verbose = TRUE)
+best.iter <- gbm.perf(boston.gbm, method = "cv")
+
+# Compare plots for `lstat`
+par(mfrow = c(1, 2))
+plot(partial(boston.gbm, pred.var = "lstat", n.trees = best.iter), type = "l",
+     main = "pdp::partial")
+plot(boston.gbm, i.var = "lstat", n.trees = best.iter,
+     continuous.resolution = 51, main = "gbm::plot.gbm")
+
+
+# Classification ---------------------------------------------------------------
+
+# Fit a gbm to the Pima indians diabetes data
+set.seed(101)
+pima.gbm <- gbm(unclass(diabetes) - 1 ~ .,
+                data = pima,
+                distribution = "bernoulli",
+                n.trees = 5000,
+                interaction.depth = 3,
+                shrinkage = 0.001,
+                verbose = TRUE)
+best.iter <- gbm.perf(pima.gbm, method = "OOB")
+summary(pima.gbm, n.trees = best.iter, las = 1)
+
+# Compare plots for `glucose`
+par(mfrow = c(1, 2))
+plot(partial(pima.gbm, pred.var = "glucose", n.trees = best.iter), type = "l",
+     main = "pdp::partial")
+plot(pima.gbm, i.var = "glucose", n.trees = best.iter,
+     continuous.resolution = 51, main = "gbm::plot.gbm")
+
+# Compare plots for `glucose` and `age`
+p <- plot(pima.gbm, i.var = c("glucose", "mass"), n.trees = best.iter,
+          continuous.resolution = 51)
[…]

### 2016-12-28 manuscript/greenwell.tex
added pred.fun argument to partial

@@
 However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response.
-This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2000-greedy} for details.
+This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2001-greedy} for details.
 PDPs help visualize the relationship between a subset of the features (typically 1-3) and the response while accounting for the average effect of the other predictors in the model.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% User-defined prediction functions
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{User-defined prediction functions}
+\label{sec:prediction}
+
+By default, \code{partial} constructs PDPs for regression on the response scale, and PDPs for classification on the logit\footnote{Add footnote here!} scale.
+However, in some cases it may be more informative to view PDPs on the probability scale.
+
+It is possible to provide your own prediction function to via the \code{pred.fun} argument in the call to \code{partial}.
+This is useful if, for example, the user requires PDPs for classification problems on the probability scale.
+
+By default, this function computes Equation~\eqref{eqn:pdf} and requires two arguments: \code{object} and \code{newdata}.
+This makes it possible to replace the sample mean in Equation~\eqref{eqn:pdf} with the median, or the logit in Equation~\eqref{eqn:avg-logit} with the predicted probaility of a particular class.
+
+The prediction function provided to \code{pred.fun} must have two specific arguments: \code{object} and \code{newdata}.
+Additionally, the function must return a single value---typically the mean of the predicted values.
+
+For regression problems, the default prediction function is given by
+\begin{example}
+pred.fun <- function(object, newdata) {
+  mean(predict(object, newdata), na.rm = TRUE)
+}
+\end{example}
+which corresponds to Equation~\eqref{eqn:pdf}.
+
+To obtain PDPs on the probability scale we just need to write a function that extracts the predicted class probabilities of interest and the averages the result.
+The snippet of code below does uses the fitted SVM from the previous example to extract the predicted probability of belonging to the setosa species.
+\begin{example}
+probFun <- function(object, newdata) {
+  pred <- predict(object, newdata, probability = TRUE)
+  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
+  mean(prob.setosa)
+}
+\end{example}
+Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. This result is displayed in Figure~\ref{fig:partial_iris_prob}.
+\begin{example}
+# Figure 8
+partial(iris.svm, pred.var = "Petal.Width", pred.fun = probFun, plot = TRUE,
+        rug = TRUE)
+\end{example}
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
+  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probabiltity of belonging to the setosa species.}
+  \label{fig:partial_iris_prob}
+\end{figure}
+
+This argument makes \code{partial} quite flexible.
+For instance, it is possible to obtain the indivudal conditional expectation (ICE) curves described in \citet{goldstein-peeking-2015} and made available in the \pkg{ICEbox} package.
+To obtain ICE curves, the user must supply \code{pred.fun} with a function that simply returns the desired type of prediction for each observation.
+This is demonstarted for the Boston housing example in the code snippet below.
+\begin{example}
+pred.ice <- function(object, newdata) {  # all predictions
+  predict(object, newdata)
+}
+lstat.ice <- partial(boston.randomForest, pred.var = "lstat",
+                     pred.fun = pred.ice)  # Figure 9 (left)
+\end{example}
+Whenever returns a data frame that also inherits from class \code{"partial.ice"}.
+The only difference is the addition of a new column called \code{"obs"} which indicates which observation a curve belongs to; in this example there is one for each unique observation.
+To further illustrate, the code below constructs ICE curves for specific quantiles of the predictions.
+The results, along with the typical PDP for \code{lstat}, are displayed in Figure~\ref{fig:partial_boston_ice}.
+\begin{example}
+pred.ice.quan <- function(object, newdata) {  # quantiles of predictions
+  quantile(predict(object, newdata), probs = 1:9/10)
+}
+lstat.ice.quan <- partial(boston.randomForest, pred.var = "lstat",
+                          pred.fun = pred.ice.quan)  # Figure 9 (middle)
+\end{example}
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=1.0\linewidth]{partial_boston_ice}
+  \caption{ICE plots and PDP for the Boston housing data. \textit{Left}: ICE curves. \textit{Middle}: Specific quantiles of ICE curves. \textit{Right}: Standard PDP.}
+  \label{fig:partial_boston_ice}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Using partial with xgboost

### 2016-12-28 R/partial.R

@@ -10,6 +10,10 @@
 #'   should include no more than three variables.
 #' @param pred.grid Data frame containing the joint values of the variables
 #'   listed in \code{pred.var}.
+#' @param pred.fun Optional prediction function that requires two arguments:
+#'   \code{object} and \code{newdata}. If specified, then the function must
+#'   return a single prediction or a vector of predictions (i.e., not a matrix
+#'   or data frame). Default is \code{NULL}.
 #' @param grid.resolution Integer giving the number of equally spaced points to
 #'   use (only used for the continuous variables listed in \code{pred.var} when
 #'   \code{pred.grid} is not supplied). If left \code{NULL}, it will default to
@@ -63,24 +67,38 @@
 #' training data from \code{object}. In these cases an error message is
 #' displayed requesting the user to supply the training data via the
 #' \code{train} argument in the call to \code{partial}. In most cases where
-#' \code{partial} can extract the needed training data from \code{object},
-#' it is taken from the same environment in which \code{partial} was called.
-#' Therefore, it is important to not change the data used to construct
+#' \code{partial} can extract the required training data from \code{object},
+#' it is taken from the same environment in which \code{partial} is called.
+#' Therefore, it is important to not change the training data used to construct
 #' \code{object} before calling \code{partial}. This problem is completely
 #' avoided when the training data are passed to the \code{train} argument in the
 #' call to \code{partial}.
 #'
-#' It is poosible to retrieve the most recent PDP constructed by \code{partial}
-#' even after it is deleted. Simply use \code{trellis.last.object()}.
+#' It is possible to retrieve the most recent PDP constructed by \code{partial}
+#' after it is deleted (as long as no other \code{"trellis"} objects have
+#' been created since); simply use \code{trellis.last.object()}.
 #'
-#' If \code{object} inherits from class \code{"gbm"} and \code{pred.grid} is not
-#' specified, then \code{partial} makes an internal call to \code{gbm::plot.gbm}
-#' in order to exploit \code{gbm}'s implementation of the weighted tree
-#' traversal method described in Friedman (2001).
+#' It is possible for \code{partial} to run much faster if \code{object}
+#' inherits from class \code{"gbm"}. In particular, if \code{object} inherits
[…]

### 2016-12-28 R/pdRegression.R

@@ -1,49 +1,53 @@
 #' @keywords internal
-pdRegression <- function(object, pred.var, pred.grid, train, progress, parallel,
-                         paropts, ...) {
+pdRegression <- function(object, pred.var, pred.grid, pred.fun, train,
+                         progress, parallel, paropts, ...) {
   UseMethod("pdRegression")
 }
 
 
 #' @keywords internal
-pdRegression.default <- function(object, pred.var, pred.grid, train, progress,
-                                 parallel, paropts, ...) {
+pdRegression.default <- function(object, pred.var, pred.grid, pred.fun,
+                                 train, progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
-    pred <- stats::predict(object, newdata = temp, ...)
-    # Some fitting functions return a matrix (e.g., mda::mars)
-    if (is.matrix(pred) || is.data.frame(pred)) {
-      pred <- pred[, 1L, drop = TRUE]
+    out <- if (is.null(pred.fun)) {
+      pred <- stats::predict(object, newdata = temp, ...)
+      if (is.matrix(pred) || is.data.frame(pred)) {
+        pred <- pred[, 1L, drop = TRUE]
+      }
+      mean(pred, na.rm = TRUE)
+    } else {
+      pred.fun(object, newdata = temp)
     }
-    stats::setNames(mean(pred, na.rm = TRUE), "yhat")
+    if (length(out) == 1) {
+      stats::setNames(out, "yhat")
+    } else {
+      stats::setNames(out, paste0("yhat.", 1L:length(out)))
+    }
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
 
 
 #' @keywords internal
-pdRegression.gbm <- function(object, pred.var, pred.grid, train, progress,
-                             parallel, paropts, ...) {
-  # Necessary to avoid silly printing from predict.gbm
+pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, pred.fun,
+                                     train, progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
     temp <- train
     temp[pred.var] <- x
-    log <- utils::capture.output(z <- mean(stats::predict(object,
-                                                          newdata = temp, ...),
-                                           na.rm = TRUE))
[…]

### 2016-12-28 R/plotPartial.R

@@ -113,6 +113,41 @@ plotPartial.partial <- function(x, smooth = FALSE, rug = FALSE, chull = FALSE,
 }
 
 
+#' @rdname plotPartial
+#' @export
+plotPartial.partial.ice <- function(x, rug = FALSE, train = NULL, ...) {
+  if (is.factor(x[[1L]])) {
+    dotplot(stats::as.formula(paste("yhat ~", names(x)[1L])), data = x,
+            groups = x$obs, type = "l", ...,
+            panel = function(xx, yy, ...) {
+              panel.xyplot(xx, yy, col = "black", ...)
+              if (rug) {
+                if (is.null(train)) {
+                  stop("The training data must be supplied for rug display.")
+                } else {
+                  panel.rug(stats::quantile(train[[names(x)[1L]]],
+                                            probs = 0:10/10, na.rm = TRUE))
+                }
+              }
+            })
+  } else {
+    xyplot(stats::as.formula(paste("yhat ~", names(x)[1L])), data = x,
+           groups = x$obs, type = "l", ...,
+           panel = function(xx, yy, ...) {
+             panel.xyplot(xx, yy, col = "black", ...)
+             if (rug) {
+               if (is.null(train)) {
+                 stop("The training data must be supplied for rug display.")
+               } else {
+                 panel.rug(stats::quantile(train[[names(x)[1L]]],
+                                           probs = 0:10/10, na.rm = TRUE))
+               }
+             }
+           })
+  }
+}
+
+
 #' @keywords internal
 pdpNumeric <- function(x, smooth, rug, train = NULL, ...) {
   xyplot(stats::as.formula(paste("yhat ~", names(x)[1L])), data = x,

### 2016-12-28 manuscript/greenwell.R

@@ -201,6 +201,44 @@ ggplot(pd, aes(x = Petal.Width, y = Petal.Length, z = yhat, fill = yhat)) +
   facet_grid(~ Species)
 dev.off()
 
+# Figure 8
+probFun <- function(object, newdata) {
+  pred <- predict(object, newdata, probability = TRUE)
+  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
+  mean(prob.setosa)
+}
+pdf("partial_iris_svm_prob.pdf", width = 12, height = 4)
+grid.arrange(
+  partial(iris.svm, pred.var = "Petal.Width", pred.fun = probFun, plot = TRUE),
+  partial(iris.svm, pred.var = "Petal.Length", pred.fun = probFun, plot = TRUE),
+  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
+          pred.fun = probFun, plot = TRUE),
+  ncol = 3
+)
+dev.off()
+
+# Figure 9
+pred.ice <- function(object, newdata) {
+  predict(object, newdata)
+}
+pred.ice.quan <- function(object, newdata) {
+  quantile(predict(object, newdata), probs = 1:9/10)
+}
+lstat.ice <- partial(boston.randomForest, pred.var = "lstat",
+                     pred.fun = pred.ice)
+lstat.ice.quan <- partial(boston.randomForest, pred.var = "lstat",
+                          pred.fun = pred.ice.quan)
+lstat.pdp <- partial(boston.randomForest, pred.var = "lstat")
+ylim <- c(3.196822, 52.149681)
+pdf("partial_boston_ice.pdf", width = 12, height = 4)
+grid.arrange(
+  plotPartial(lstat.ice, alpha = 0.1, ylim = ylim),
+  plotPartial(lstat.ice.quan, ylim = ylim),
+  plotPartial(lstat.pdp, ylim = ylim),
+  ncol = 3
+)
+dev.off()
+
 
 ################################################################################
 # Using partial with the XGBoost library

### 2016-12-28 slowtests/slowtests-classification.R

@@ -1,0 +1,191 @@
+#-------------------------------------------------------------------------------
+# Slow tests for the pdp package
+#-------------------------------------------------------------------------------
+
+
+################################################################################
+# Setup
+################################################################################
+
+# Make sure all packges are up to date
+# update.packages()
+
+# Load required packages
+library(pdp)
+
+# Load data
+data(pima)
+
+
+################################################################################
+# "bagging" objects: adabag package
+################################################################################
+
+# Load required package
+library(adabag)
+
+# Fit a glm to the Pima indians diabetes data
+set.seed(101)  # for reproducibility
+pima.bagging <- bagging(diabetes ~ ., data = pima, fmfinal = 500)
+
+# Construct a PDP for glucose
+partial(pima.bagging, pred.var = "glucose", plot = TRUE, progress = "text",
+        rug = TRUE)
+
+# Construct a PDP for glucose (probability scale)
+partial(pima.bagging, pred.var = "glucose", plot = TRUE, progress = "text",
+        pred.fun = function(object, newdata) {
+          mean(predict(object, newdata)$prob[, 1L], na.rm = TRUE)
+        })
+
+
+################################################################################
+# "bagging" objects: adabag package
+################################################################################
+
+# Load required package
+library(adabag)
+
+# Fit a glm to the Pima indians diabetes data
+set.seed(101)  # for reproducibility
+pima.boosting <- boosting(diabetes ~ ., data = pima, fmfinal = 500)
+
+# Construct a PDP for glucose
+partial(pima.boosting, pred.var = "glucose", plot = TRUE, progress = "text",
+        rug = TRUE)
+
+# Construct a PDP for glucose (probability scale)
+partial(pima.boosting, pred.var = "glucose", plot = TRUE, progress = "text",
+        pred.fun = function(object, newdata) {
+          mean(predict(object, newdata)$prob[, 1L], na.rm = TRUE)
+        })
+
+
+################################################################################
+# "glm" objects: stats package
+################################################################################
+
+# Fit a glm to the Pima indians diabetes data
+pima.glm <- glm(diabetes ~ ., data = pima, family = binomial)
+
+# Construct a PDP for glucose
[…]

### 2016-12-28 slowtests/slowtests-regression.R

@@ -1,0 +1,135 @@
+#-------------------------------------------------------------------------------
+# Slow tests for the pdp package
+#-------------------------------------------------------------------------------
+
+
+################################################################################
+# Setup
+################################################################################
+
+# Make sure all packges are up to date
+# update.packages()
+
+# Load required packages
+library(pdp)
+
+# Load data
+data(boston)
+
+
+################################################################################
+# "lm" objects: stats package
+################################################################################
+
+# Fit a glm to the boston indians cmedv data
+boston.lm <- lm(cmedv ~ ., data = boston)
+
+# Construct a PDP for lstat
+partial(boston.lm, pred.var = "lstat", plot = TRUE)
+
+
+################################################################################
+# "gbm" object: gbm package
+################################################################################
+
+# Load required package
+library(gbm)
+
+# Fit a gbm to the boston indians cmedv data
+set.seed(101)
+boston.gbm <- gbm(cmedv ~ .,
+                data = boston,
+                distribution = "gaussian",
+                n.trees = 5000,
+                interaction.depth = 3,
+                shrinkage = 0.001,
+                verbose = TRUE)
+best.iter <- gbm.perf(boston.gbm, method = "OOB")
+summary(boston.gbm, n.trees = best.iter, las = 1)
+
+# Compare plots for `lstat`
+par(mfrow = c(1, 2))
+plot(partial(boston.gbm, pred.var = "lstat", n.trees = best.iter), type = "l",
+     main = "pdp::partial")
+plot(boston.gbm, i.var = "lstat", n.trees = best.iter,
+     continuous.resolution = 51, main = "gbm::plot.gbm")
+
+# Compare plots for `lstat` and `age`
+p <- plot(boston.gbm, i.var = c("lstat", "mass"), n.trees = best.iter,
+          continuous.resolution = 51)
+grid.arrange(
+  partial(boston.gbm, pred.var = c("lstat", "mass"), plot = TRUE,
+          chull = TRUE, progress = "text", n.trees = best.iter),
+  p,
+  ncol = 2
+)
+
+
+# Try user-specified response scale
+par(mfrow = c(1, 2))
+plot(partial(boston.gbm, pred.var = "lstat",
+             pred.fun = function(object, newdata) {
+               median(predict(object, newdata, type = "response",
+                            n.trees = best.iter))
[…]

### 2016-12-28 slowtests/slowtests.R

@@ -42,8 +42,8 @@ best.iter <- gbm.perf(boston.gbm, method = "cv")
 
 # Compare plots for `lstat`
 par(mfrow = c(1, 2))
-plot(partial(boston.gbm, pred.var = "lstat", n.trees = best.iter), type = "l",
-     main = "pdp::partial")
+plot(partial(boston.gbm, pred.var = "lstat", n.trees = best.iter,
+             grid.resolution = 51), type = "l", main = "pdp::partial")
 plot(boston.gbm, i.var = "lstat", n.trees = best.iter,
      continuous.resolution = 51, main = "gbm::plot.gbm")
 
@@ -78,6 +78,19 @@ grid.arrange(
   p,
   ncol = 2
 )
+
+
+# Try user-specified response scale
+par(mfrow = c(1, 2))
+plot(partial(pima.gbm, pred.var = "glucose",
+             pred.fun = function(object, newdata) {
+               mean(predict(object, newdata, type = "response", n.trees = best.iter))
+             },
+             grid.resolution = 51, progress = "text"),
+     type = "l", main = "pdp::partial")
+plot(pima.gbm, i.var = "glucose", type = "response", n.trees = best.iter,
+     continuous.resolution = 51)
+title("gbm::plot.gbm")
 
 
 ################################################################################

### 2016-12-28 manuscript/greenwell.tex
fixed bug in getTrainingData.train

--- 2016-12-16 manuscript/greenwell.tex
@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 The partial dependence of the response on $x_1$ can be constructed as follows:
+
+\begin{algorithm}
 \begin{enumerate}
@@
 \end{enumerate}
-This can be quite computationally intensive since the algorithm involves $k$ passes over the training records.
+\caption{A simple algorithm for consructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
+\end{algorithm}
+Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records.
 Fortunately, this can be parallelized quite easily (more on this in a later section).
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Classification problems
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Classification problems}
+\label{sec:class}
 
@@
   \label{fig:partial_iris}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% User-defined prediction functions
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{User-defined prediction functions}
+\label{sec:prediction}
+
+PDPs are essentially just averaged predictions; see, for example, step 1.
+(c) in Algorithm~\ref{alg:pdp}.
+As pointed at by \citet{goldstein-peeking-2015}, since traditional PDPs are based on the sample mean, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest.
+This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE curves.
+
+With \code{partial} it is possible to replace the mean in step 1.
+(c) with, say, the median, or obtain PDPs for classification problems on the raw probability scale.
+It is even possible to obtain ICE curves.
+This flexibility is due to the new \code{pred.fun} argument in \code{partial} (starting with \pkg{pdp} version 0.4.0).
+This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}.
+The supplied prediction function must return either (i) a single prediction---typically an averaged prediction---or  (ii) a vector of predictions.
+The two examples below illustrate.
+
+Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result.
+The snippet of code below uses the fitted SVM from Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
+\begin{example}
+pred.prob <- function(object, newdata) {  # see ?predict.svm
+  pred <- predict(object, newdata, probability = TRUE)
+  prob.setosa <- attr(pred, which = "probabilities")[, "setosa"]
+  mean(prob.setosa)
+}
+\end{example}
+Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The result is displayed in Figure~\ref{fig:partial_iris_prob}.
+\begin{example}
+# Figure 8
+partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE,
+        rug = TRUE)
+\end{example}
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
+  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probabiltity of belonging to the \code{Setosa} species.}
+  \label{fig:partial_iris_prob}
+\end{figure}
+
+\newpage
+
+For regression problems, the default prediction function is essentially
+\begin{example}
+pred.fun <- function(object, newdata) {
+  mean(predict(object, newdata), na.rm = TRUE)
+}
+\end{example}
+This corresponds to step step 1.
+(c) in Algorithm~\ref{alg:pdp}.
+Suppose we would like ICE curves instead.
+To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}.
+This is demonstarted for the Boston housing example in the code snippet below.
+\begin{example}
+pred.ice <- function(object, newdata) {  # all predictions
+  predict(object, newdata)
+}
+lstat.ice <- partial(boston.randomForest, pred.var = "lstat",
+                     pred.fun = pred.ice)  # Figure 9 (left)
+\end{example}
+
+Recall that \code{partial} constructs a data frame that also inherits from class \code{"partial"}.
+When the function supplied to \code{pred.fun} returns multiple predictions, \code{partial} instead constructs a data frame that also inherits from class \code{"partial.ice"}.
+The only difference is the addition of a new column called \code{"obs"} which indicates which observation a curve belongs to; in this example, there is one curve for each unique observation in \code{boston}.
+To further illustrate, the code below constructs ICE curves for specific quantiles of the predictions.
+The results, along with the traditional PDP for \code{lstat}, are displayed in Figure~\ref{fig:partial_boston_ice}.
+\begin{example}
+pred.ice.quan <- function(object, newdata) {  # quantiles of predictions
+  quantile(predict(object, newdata), probs = 1:9/10)
+}
+lstat.ice.quan <- partial(boston.randomForest, pred.var = "lstat",
+                          pred.fun = pred.ice.quan)  # Figure 9 (middle)
+\end{example}
+
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=1.0\linewidth]{partial_boston_ice}
+  \caption{ICE plots and PDP for the Boston housing data. \textit{Left}: ICE curves. \textit{Middle}: Specific quantiles of ICE curves. \textit{Right}: Standard PDP.}
+  \label{fig:partial_boston_ice}
 \end{figure}

### 2016-12-28 manuscript/greenwell.R

@@ -247,13 +247,13 @@ dev.off()
 # Optimize tuning parameters using 5-fold cross-validation
 library(caret)
 ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)
-xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
-                        max_depth = 1:6,
-                        eta = c(0.001, 0.01, 0.1, 0.5, 1),
+xgb.grid <- expand.grid(nrounds = c(500, 1000, 2000, 5000),
+                        max_depth = 1:4,
+                        eta = c(0.001, 0.01, 0.1),
                         gamma = 0,
                         colsample_bytree = 1,
                         min_child_weight = 1,
-                        subsample = 1)
+                        subsample = c(0.8, 1))
 set.seed(202)
 boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
                          y = boston$cmedv,
@@ -262,6 +262,7 @@ boston.xgb.tune <- train(x = data.matrix(subset(boston, select = -cmedv)),
                          trControl = ctrl,
                          tuneGrid = xgb.grid)
 plot(boston.xgb.tune)
+print(boston.xgb.tune$bestTune)
 # boston.xgb.tune$bestTune
 #    nrounds max_depth  eta gamma colsample_bytree min_child_weight
 # 44    2000         3 0.01     0                1                1

### 2016-12-28 manuscript/greenwell.tex
fixed bug in pdRegression.xgb.Booster

@@
   \includegraphics[width=1.0\linewidth]{partial_boston_ice}
-  \caption{ICE plots and PDP for the Boston housing data. \textit{Left}: ICE curves. \textit{Middle}: Specific quantiles of ICE curves. \textit{Right}: Standard PDP.}
+  \caption{ICE curves and PDP depicting the relationship between \code{cmedv} and \code{lstat} for the Boston housing example.
+\textit{Left}: ICE curves.
+\textit{Middle}: Specific quantiles of ICE curves.
+\textit{Right}: Standard PDP.}
   \label{fig:partial_boston_ice}
@@
 # Grid of tuning values for XGBoost.
-xgb.grid <- expand.grid(nrounds = c(100, 500, 1000, 2000, 5000),
-                        max_depth = 1:6,
-                        eta = c(0.001, 0.01, 0.1, 0.5, 1),
+xgb.grid <- expand.grid(nrounds = c(1000, 2000, 3000),
+                        max_depth = 1:4,
+                        eta = c(0.001, 0.01, 0.1),
                         gamma = 0,
                         colsample_bytree = 1,
-                        min_child_weight = 1)
+                        min_child_weight = 1,
+                        subsample = c(0.5, 0.75, 1))
 
@@
 \end{example}
-The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 2000}, \code{max\_depth = 3}, \code{eta = 0.01}, and achieved a cross-validated RMSE and $R^2$ of 2.819 and 0.905, respectively.
+The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 3000}, \code{max\_depth = 3}, \code{eta = 0.01}, and \code{subsample = 0.75}; the final model achieved a cross-validated RMSE and $R^2$ of 2.789 and 0.906, respectively.
 The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page.

### 2016-12-28 R/pdRegression.R

@@ -36,7 +36,7 @@ pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, pred.fun,
     temp <- train
     temp[pred.var] <- x
     out <- if (is.null(pred.fun)) {
-      pred <- stats::predict(object, newdata = temp, ...)
+      pred <- stats::predict(object, newdata = data.matrix(temp), ...)
       if (is.matrix(pred) || is.data.frame(pred)) {
         pred <- pred[, 1L, drop = TRUE]
       }

### 2016-12-28 manuscript/greenwell.tex
updated manuscript

@@
 The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown).
-(Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"}.
+(Make sure you are using version 0.6-0 or later of \pkg{xgboost}: \url{https://github.com/dmlc/xgboost/tree/master/R-package}.) \strong{Note:} while \code{xgboost} requires the training data to be an object of class \code{"matrix"}, \code{"dgCMatrix"}, or \code{"xgb.DMatrix"}, \code{partial} requires a \code{"data.frame"} that does not contain the response column.
 \begin{example}

### 2016-12-28 manuscript/greenwell.R

@@ -287,5 +287,11 @@ dev.off()
 # # Figure 8
 # pdf("boston_xgb.pdf", width = 12, height = 4)
 X <- subset(boston, select = -cmedv)
-partial(boston.xgb, pred.var = "lstat", train = X, rug = TRUE)
+grid.arrange(
+  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE, train = X),
+  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE, train = X),
+  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE,
+          train = X),
+  ncol = 3
+)
 # dev.off()

### 2016-12-28 README.Rmd
updated plots and code

@@
 In this example, we fit a support vector machine with a radial basis function kernel to the Pima Indians diabetes data. (See `?pima` for a brief explanation of the data.)
-```{r, message=FALSE, warning=FALSE, fig.show="hold"}
+```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
 # Load required packages
@@
 # Partial dependence of glucose and age on diabetes test result (neg/pos).
-partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, chull = TRUE,
-        train = pima)
+grid.arrange(
+  partial(pima.svm, pred.var = "glucose", plot = TRUE, train = pima),
+  partial(pima.svm, pred.var = "age", plot = TRUE, train = pima),
+  partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, train = pima),
+  ncol = 3
+)
 ```
@@
 xgb.grid <- expand.grid(
-  nrounds = 500,
-  max_depth = 1:6,
-  eta = c(0.001, 0.01, 0.1, 0.2, 0.3, 0.5),
+  nrounds = c(1000, 2000),
+  max_depth = 2:4,
+  eta = c(0.001, 0.01, 0.1),
   gamma = 0, 
@@
   min_child_weight = 1,
-  subsample = 1
+  subsample = c(0.5, 0.75, 1)
 )
 
-# Tune a support vector machine (SVM) using a radial basis function kerel to
-# the Pima Indians diabetes data. This may take a few minutes!
+# Tune am XGBoost model to the Pima Indians diabetes data. This may take a few 
+# minutes!
 set.seed(103)  # for reproducibility

### 2016-12-28 manuscript/greenwell.tex
updated examples

@@
 
-Nowadays, supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
+Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
 (We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, there is usually no simple prediction formula like in Equation~\eqref{eqn:boston}.
-Fortunately, these methods can still provide insight into the data, but it is not in the form of simple equations.
+Fortunately, these models can still provide insight into the data, but it is not in the form of simple equations.
 For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors in the training data.
 
-Determining predictor importance is a crucial task in any supervised learning problem.
-However, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response.
+While determining predictor importance is a crucial task in any supervised learning problem, ranking variables is only part of the story and once a subset of "important" features is identified it is often necessary to assess the relationship between them (or subset thereof) and the response.
 This can be done in many ways, but in machine learning it is often accomplished by constructing \dfn{partial dependence plots} (PDPs); see \citet{friedman-2001-greedy} for details.
@@
 To overcome this issue \citeauthor*{goldstein-peeking-2015} developed the concept of \dfn{individual conditional expectation} (ICE) plots---available in the \CRANpkg{ICEbox} package.
+ICE plots display the estimated relationship between the response and a predictor of interest for each observation.
+Consequently, the PDP for a predictor of interest can be obtained by averaging the corresponding ICE curves across all observations.
+In Section~\ref{sec:prediction}, it is shown how to obtain ICE curves using the \pkg{pdp} package (which is slightly faster than \pkg{ICEbox}).
 \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor.
-The ability to construct derivative plots is also available in \pkg{ICEbox}.
+The ability to construct centered ICE (c-ICE) plots and derivative ICE (d-ICE) plots is also available in \pkg{ICEbox}; c-ICE plots help visualize heterogeneity in the modeled relationship between observations, and d-ICE plots help to explore interaction effects.
 
@@
 Bug reports and suggestions are appreciated and should be submitted to \url{https://github.com/bgreenwell/pdp/issues}.
-Currently, only two functions are exported by \pkg{pdp}:
+The two most important functions exported by \pkg{pdp} are:
 \begin{itemize}
@@
 If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
-The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{yhat} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
+The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{yhat}\footnote{There is one exception to this.
+When a function supplied via the \code{pred.fun} argument (this argument is discussed in Section~\ref{sec:prediction}) returns multiple predictions, the last two columns will be labeled \code{yhat} and \code{yhat.id}.} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
 If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
@@
   geom_contour(color = "white", alpha = 0.5) +
-  scale_fill_distiller(name = "Logit", palette = "Spectral") +
+  scale_fill_distiller(name = "Centered\nlogit", palette = "Spectral") +
   theme_bw() +
@@
 (c) in Algorithm~\ref{alg:pdp}.
-As pointed at by \citet{goldstein-peeking-2015}, since traditional PDPs are based on the sample mean, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest.
+As pointed out by \citet{goldstein-peeking-2015}, since traditional PDPs are based on the sample mean, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest.
 This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE curves.
@@
 With \code{partial} it is possible to replace the mean in step 1.
-(c) with, say, the median, or obtain PDPs for classification problems on the raw probability scale.
+(c) with, say, the median or trimmed mean, or obtain PDPs for classification problems on the raw probability scale.
 It is even possible to obtain ICE curves.
@@
 This argument accepts an optional prediction function that requires two arguments: \code{object} and \code{newdata}.
-The supplied prediction function must return either (i) a single prediction---typically an averaged prediction---or  (ii) a vector of predictions.
+The supplied prediction function must return either a single prediction or a vector of predictions.
+If the mean of all the predictions is returned, this is equivalent to a traditional PDP.
+If a vector of predictions is returned (i.e., one for each observation), then this is equivalent to the ICE plot procedure.
 The two examples below illustrate.
@@
 Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result.
-The snippet of code below uses the fitted SVM from Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
+The snippet of code below uses the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
 \begin{example}
@@
 \end{example}
-Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}. The result is displayed in Figure~\ref{fig:partial_iris_prob}.
+Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}.
+The following chunk of code uses \code{partial} and \code{grid.arrange} to construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale.
+The results are displayed in Figure~\ref{fig:partial_iris_prob}.
 \begin{example}
-# Figure 8
-partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE,
-        rug = TRUE)
+grid.arrange(
+  # Figure 8 (left)
+  partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE),
+  # Figure 8 (middle)
+  partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob, plot = TRUE),
+  # Figure 8 (right)
+  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
+          pred.fun = pred.prob, plot = TRUE),
+  ncol = 3
+)
 \end{example}
@@
 
-Recall that \code{partial} constructs a data frame that also inherits from class \code{"partial"}.
-When the function supplied to \code{pred.fun} returns multiple predictions, \code{partial} instead constructs a data frame that also inherits from class \code{"partial.ice"}.
-The only difference is the addition of a new column called \code{"obs"} which indicates which observation a curve belongs to; in this example, there is one curve for each unique observation in \code{boston}.
+When the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which observation a curve belongs to; in this example, there is one curve for each unique observation in \code{boston}.
 To further illustrate, the code below constructs ICE curves for specific quantiles of the predictions.
@@
   \includegraphics[width=1.0\linewidth]{partial_boston_ice}
-  \caption{ICE curves and PDP depicting the relationship between \code{cmedv} and \code{lstat} for the Boston housing example.
+  \caption{ICE curves and PDP depicting the relationship between \code{cmedv} and \code{lstat} for the Boston housing example (the same $y$-axis limits are used for each plot).
 \textit{Left}: ICE curves.
@@
   \label{fig:partial_boston_ice}
+\end{figure}
+
+Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process ICE curves to obtain c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}.
+The code snippet below illustrates this for the Boston housing example using the predictor \code{age}.
+\begin{example}
+# Use partial to obtain ICE curves
+pred.ice <- function(object, newdata) predict(object, newdata)
+age.ice <- partial(boston.randomForest, pred.var = "age", pred.fun = pred.ice)
+
+# Post-process age.ice to obtain c-ICE curves
+library(dplyr)
+age.ice <- age.ice %>%
+  group_by(yhat.id) %>%
+  mutate(yhat.centered = yhat - first(yhat))
+\end{example}
+Furthermore, using \pkg{ggplot2}, we can easily display ICE curves along with the PDP.
+This can be accomplished by using the \code{stat\_summary} function to average the ICE curves together.
+The code snippet below uses \pkg{ggplot2} and \code{grid.arrange} to plot ICE curves and c-ICE curves for \code{age} in the Boston housing example.
+The results are displayed in Figure~\ref{fig:partial_boston_ice_pdp}; compare these to Figures 3-4 in \citet{goldstein-peeking-2015}.
+\begin{example}
+# ICE curves
+p1 <- ggplot(age.ice, aes(age, yhat)) +
+  geom_line(aes(group = yhat.id), alpha = 0.3) +
+  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)
+
+# c-ICE curves
+p2 <- ggplot(age.ice, aes(age, yhat.centered)) +
+  geom_line(aes(group = yhat.id), alpha = 0.3) +
+  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)
+
+# Figure 10
+grid.arrange(p1, p2, ncol = 2)
+\end{example}
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=1.0\linewidth]{partial_boston_ice_pdp}
+  \caption{ICE curves (black curves) and PDP (red curve) depicting the relationship between \code{cmedv} and \code{age} for the Boston housing example.
+\textit{Left}: Uncentered.
+\textit{Right}: Centered.}
+  \label{fig:partial_boston_ice_pdp}
 \end{figure}

### 2016-12-28 R/pdRegression.R

@@ -23,7 +23,11 @@ pdRegression.default <- function(object, pred.var, pred.grid, pred.fun,
     if (length(out) == 1) {
       stats::setNames(out, "yhat")
     } else {
-      stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      if (is.null(names(out))) {
+        stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      } else {
+        stats::setNames(out, paste0("yhat.", names(out)))
+      }
     }
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }
@@ -47,7 +51,11 @@ pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, pred.fun,
     if (length(out) == 1) {
       stats::setNames(out, "yhat")
     } else {
-      stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      if (is.null(names(out))) {
+        stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      } else {
+        stats::setNames(out, paste0("yhat.", names(out)))
+      }
     }
   }, .progress = progress, .parallel = parallel, .paropts = paropts)
 }

### 2016-12-28 R/plotPartial.R

@@ -52,26 +52,43 @@ plotPartial.partial <- function(x, smooth = FALSE, rug = FALSE, chull = FALSE,
                                 col.regions = viridis::viridis,
                                 ...) {
 
+  # Determine of x contains multiple PDPs
+  multi <- if ("yhat.id" %in% names(x)) {
+    TRUE
+  } else {
+    FALSE
+  }
+
   # Determine number of variables to plot
-  nx <- ncol(x) - 1  # don't count response
-  if (!(nx %in% 1:3)) {
+  nx <- if (multi) {
+    ncol(x) - 2  # don't count yhat or yhat.id
+  } else {
+    ncol(x) - 1  # don't count yhat
+  }
+
+  # Throw error if too difficult to plot
+  if ((!multi && !(nx %in% 1:3)) || (multi && (nx > 1))) {
     stop("Too many variables to plot. Try using lattice or ggplot2 directly.")
   }
 
-  # Single predictor
-  if (nx == 1) {
+  # Determine which type of plot to produce
+  if (multi) {
 
-    # PDPs for a single predictor
+    # Multiple PDPs for a single predictor
+    p <- pdpMulti(x, rug = rug, train = train, ...)
+
+  } else if (nx == 1) {
+
+    # PDP for a single predictor
     p <- if (is.factor(x[[1L]])) {
       pdpFactor(x, ...)
     } else {
       pdpNumeric(x, rug = rug, smooth = smooth, train = train, ...)
     }
 
-    # Two predictors
   } else if (nx == 2) {
 
-    # PDPs for two predictors
+    # PDP for two predictors
     p <- if (is.factor(x[[1L]]) && is.factor(x[[2L]])) {
       pdpFactorFactor(x, ...)
     } else if (is.factor(x[[1L]]) || is.factor(x[[2L]])) {
@@ -82,7 +99,6 @@ plotPartial.partial <- function(x, smooth = FALSE, rug = FALSE, chull = FALSE,
                         ...)
     }
 
-    # More than two predictors
   } else {
 
     # Convert additional predictors to factors using the equal count algorithm
@@ -92,7 +108,7 @@ plotPartial.partial <- function(x, smooth = FALSE, rug = FALSE, chull = FALSE,
       }
     }
 
-    # PDPs for more than two predictors
+    # PDP for more than two predictors
     p <- if (is.factor(x[[1L]]) && is.factor(x[[2L]])) {
       pdpFactorFactorShingle(x, nx = nx, ...)
[…]

### 2016-12-28 slowtests/slowtests-regression.R

@@ -79,40 +79,31 @@ plot(boston.gbm, i.var = "lstat", type = "response", n.trees = best.iter,
 title("gbm::plot.gbm")
 
 
-# Plot ICE curves
-pred.ice <- function(object, newdata) {
-  predict(object, newdata, n.trees = best.iter)
-}
-pred.ice.quan <- function(object, newdata) {
-  quantile(predict(object, newdata, n.trees = best.iter), probs = 1:9/10)
-}
-lstat.ice <- partial(boston.gbm, pred.var = "lstat",
-                     pred.fun = pred.ice)
-lstat.ice.quan <- partial(boston.gbm, pred.var = "lstat",
-                          pred.fun = pred.ice.quan)
-lstat.pdp <- partial(boston.gbm, pred.var = "lstat")
-ylim <- c(5.475361, 52.642910)
+# PDPs for lstat based on the mean (left) and median (right)
 grid.arrange(
-  plotPartial(lstat.ice, alpha = 0.1, main = "ICE curves", ylim = ylim),
-  plotPartial(lstat.ice.quan, main = "ICE curve quantiles", ylim = ylim),
-  plotPartial(lstat.pdp, main = "PDP", ylim = ylim),
-  ncol = 3
+  partial(boston.gbm, pred.var = "lstat", plot = TRUE),
+  partial(boston.gbm, pred.var = "lstat", plot = TRUE,
+          pred.fun = function(object, newdata) {
+            median(predict(object, newdata, n.trees = best.iter), na.rm = TRUE)
+          }),
+  ncol = 2
 )
 
-# Compare to results from ICEbox package
-plot(
-  ICEbox::ice(boston.randomForest,
-              X = subset(boston, select = -cmedv),
-              y = boston$cmedv,
-              predictor = "lstat")
-)
+# ICE curves
+partial(boston.gbm, pred.var = "lstat", plot = TRUE,
+        pred.fun = function(object, newdata) {
+          predict(object, newdata, n.trees = best.iter)
+        })
 
-pred.ice.trim <- function(object, newdata) {
-  mean(predict(object, newdata, n.trees = best.iter), trim = 0.2)
-}
-lstat.ice.trim <- partial(boston.gbm, pred.var = "lstat",
-                          pred.fun = pred.ice.trim)
-plotPartial(lstat.ice.trim, main = "PDP (trimmed mean)", ylim = ylim)
+# PDPs based on the mean and median
[…]

### 2016-12-29 manuscript/greenwell.tex
minor edits

@@
 Nonetheless, \citeauthor{harrison-1978-hedonic}---using a combination of transformations, significance testing, and grid searches---were able to find a reasonable fitting model ($R^2 = 0.81$).
-Part of the payoff for there efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
+Part of the payoff for there time and efforts was an interpretable prediction equation which is reproduced in Equation~\eqref{eqn:boston}.
 \begin{equation}
 \label{eqn:boston}
-\begin{split}
+\begin{aligned}
 \widehat{\log\left(MV\right)} &= 9.76 + 0.0063 RM^2 + 8.98\times10^{-5} AGE - 0.19\log\left(DIS\right) + 0.096\log\left(RAD\right) \\
   & \quad - 4.20\times10^{-4} TAX - 0.031 PTRATIO + 0.36\left(B - 0.63\right)^2 - 0.37\log\left(LSTAT\right) \\
-  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS - 0.0064 NOX^2
-\end{split}
+  & \quad - 0.012 CRIM + 8.03\times10^{-5} ZN + 2.41\times10^{-4} INDUS + 0.088 CHAS \\
+  & \quad - 0.0064 NOX^2
+\end{aligned}
 \end{equation}
@@
 Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records.
-Fortunately, this can be parallelized quite easily (more on this in a later section).
+Fortunately, this can be parallelized quite easily (more on this in Section~\ref{sec:computational}).
 This algorithm can also be easily extended to larger subsets of two or more features as well.
@@
 For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
-The \code{grid.arrange} function is then used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph.
-The result is displayed in Figure~\ref{fig:pd_lstat_rm}.
+The \code{grid.arrange} function is used to display three PDPs, which make use of various \code{plotPartial} options\footnote{See Section~\ref{sec:computational} for an example of how to add a label to the colorkey in these types of graphs.}, on the same graph.
+The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
 \begin{example}
+# Compute partial dependence data for lstat and rm
+pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
+
+# Default PDP
+pdp1 <- plotPartial(pd)
+
+# Add contour lines and use a different color palette
 rwb <- colorRampPalette(c("red", "white", "blue"))
-pd <- partial(boston.rf, pred.var = c("lstat", "rm"))
-grid.arrange(
-  # Figure 3 (left)
-  plotPartial(pd),
-  # Figure 3 (middle)
-  plotPartial(pd, contour = TRUE, col.regions = rwb),
-  # Figure 3 (right)
-  plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
-              colorkey = TRUE, screen = list(z = -20, x = -60)),
-  ncol = 3
-)
+pdp2 <- plotPartial(pd, contour = TRUE, col.regions = rwb)
+
+# 3-D surface
+pdp3 <- plotPartial(pd, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
+                    colorkey = TRUE, screen = list(z = -20, x = -60))
+
+# Display plots side by side (Figure 3)
+grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
 \end{example}
@@
 
-Constructing PDPs can be quite computationally expensive.
-Several strategies are available to ease the computational burden in larger problems.
+Constructing PDPs can be quite computationally expensive\footnote{The exception is regression trees based on single-variable splits which can make use of the efficient weighted tree traversal method described in \citet{friedman-2001-greedy}, however, only the \pkg{gbm} package seems to make use of this approach; consequently, \pkg{pdp} can also exploit this strategy when used with \pkg{gbm} models (see \code{?partial} for details).} Several strategies are available to ease the computational burden in larger problems.
 For example, there is no need to compute partial dependence of \code{cmedv} using each unique value of \code{rm} in the training data (which would require $k = 446$ passes over the data!).
@@
 \begin{example}
+# Display plots side by side (Figure 5)
 grid.arrange(
-  # Figure 5 (left)
   partial(boston.rf, "rm", plot = TRUE),
-  # Figure 5 (middle)
   partial(boston.rf, "rm", grid.resolution = 30, plot = TRUE),
-  # Figure 5 (right)
   partial(boston.rf, "rm", pred.grid = data.frame(rm = 3:9), plot = TRUE),
@@
 \begin{example}
+# Display plots side by side (Figure 8)
 grid.arrange(
-  # Figure 8 (left)
   partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE),
-  # Figure 8 (middle)
   partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob, plot = TRUE),
-  # Figure 8 (right)
   partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
@@
 
-\newpage
-
 For regression problems, the default prediction function is essentially
@@
 \begin{example}
-pred.ice <- function(object, newdata) {  # all predictions
+lstat.ice <- partial(boston.rf, pred.var = "lstat",
+                     pred.fun = function(object, newdata) {
                        predict(object, newdata)
-}
-lstat.ice <- partial(boston.randomForest, pred.var = "lstat",
-                     pred.fun = pred.ice)  # Figure 9 (left)
+                     })  # see Figure 9 (left)
 \end{example}
@@
 \begin{example}
-pred.ice.quan <- function(object, newdata) {  # quantiles of predictions
+lstat.quan <- partial(boston.rf, pred.var = "lstat",
+                      pred.fun = function(object, newdata) {
                         quantile(predict(object, newdata), probs = 1:9/10)
-}
-lstat.ice.quan <- partial(boston.randomForest, pred.var = "lstat",
-                          pred.fun = pred.ice.quan)  # Figure 9 (middle)
+                      })  # see Figure 9 (middle)
 \end{example}
-
 \begin{figure}[!htbp]
@@
 pred.ice <- function(object, newdata) predict(object, newdata)
-age.ice <- partial(boston.randomForest, pred.var = "age", pred.fun = pred.ice)
+age.ice <- partial(boston.rf, pred.var = "age", pred.fun = pred.ice)
 
@@
 p1 <- ggplot(age.ice, aes(age, yhat)) +
-  geom_line(aes(group = yhat.id), alpha = 0.3) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)
+  geom_line(aes(group = yhat.id), alpha = 0.2) +
+  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 
@@
 p2 <- ggplot(age.ice, aes(age, yhat.centered)) +
-  geom_line(aes(group = yhat.id), alpha = 0.3) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)
+  geom_line(aes(group = yhat.id), alpha = 0.2) +
+  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 
-# Figure 10
+# Display plots side by side (Figure 10)
 grid.arrange(p1, p2, ncol = 2)
@@
 \begin{example}
+# Display plots side by side (Figure 11)
 grid.arrange(
-  # Figure 8 (left)
   partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
-  # Figure 8 (middle)
   partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
-  # Figure 8 (right)
   partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),

### 2016-12-29 manuscript/greenwell.R

@@ -225,11 +225,11 @@ pred.ice <- function(object, newdata) {
 pred.ice.quan <- function(object, newdata) {
   quantile(predict(object, newdata), probs = 1:9/10)
 }
-lstat.ice <- partial(boston.randomForest, pred.var = "lstat",
+lstat.ice <- partial(boston.rf, pred.var = "lstat",
                      pred.fun = pred.ice)
-lstat.ice.quan <- partial(boston.randomForest, pred.var = "lstat",
+lstat.ice.quan <- partial(boston.rf, pred.var = "lstat",
                           pred.fun = pred.ice.quan)
-lstat.pdp <- partial(boston.randomForest, pred.var = "lstat")
+lstat.pdp <- partial(boston.rf, pred.var = "lstat")
 ylim <- c(3.196822, 52.149681)
 pdf("partial_boston_ice.pdf", width = 12, height = 4)
 grid.arrange(
@@ -242,7 +242,7 @@ dev.off()
 
 # Use partial to obtain ICE curves for age
 pred.ice <- function(object, newdata) predict(object, newdata)
-age.ice <- partial(boston.randomForest, pred.var = "age", pred.fun = pred.ice)
+age.ice <- partial(boston.rf, pred.var = "age", pred.fun = pred.ice)
 
 # Post-process age.ice to obtain c-ICE curves
 age.ice <- age.ice %>%
@@ -251,21 +251,18 @@ age.ice <- age.ice %>%
 
 # ICE curves
 p1 <- ggplot(age.ice, aes(age, yhat)) +
-  geom_line(aes(group = yhat.id), alpha = 0.3) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)
+  geom_line(aes(group = yhat.id), alpha = 0.2) +
+  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 
 # c-ICE curves
 p2 <- ggplot(age.ice, aes(age, yhat.centered)) +
-  geom_line(aes(group = yhat.id), alpha = 0.3) +
-  stat_summary(fun.y = mean, geom = "line", col = "red", size = 2)
+  geom_line(aes(group = yhat.id), alpha = 0.2) +
+  stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 
 # Figure 10
 pdf("partial_boston_ice_pdp.pdf", width = 8, height = 4)
 grid.arrange(p1, p2, ncol = 2)
 dev.off()
-
-
-
 
 
 ################################################################################

### 2016-12-30 manuscript/greenwell.tex
minor edits

@@
 Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
-(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, there is usually no simple prediction formula like in Equation~\eqref{eqn:boston}.
+(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, these algorithms typically do not produce simple prediction formulas like in Equation~\eqref{eqn:boston}.
 Fortunately, these models can still provide insight into the data, but it is not in the form of simple equations.
@@
 With \code{partial} it is possible to replace the mean in step 1.
-(c) with, say, the median or trimmed mean, or obtain PDPs for classification problems on the raw probability scale.
+(c) with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale.
 It is even possible to obtain ICE curves.
@@
 The supplied prediction function must return either a single prediction or a vector of predictions.
-If the mean of all the predictions is returned, this is equivalent to a traditional PDP.
-If a vector of predictions is returned (i.e., one for each observation), then this is equivalent to the ICE plot procedure.
-The two examples below illustrate.
+If the mean of all the predictions is returned, the result will be equivalent to the traditional PDP.
+If a vector of predictions is returned (i.e., one for each observation), the result will be equivalent to the ICE plot procedure.
+The examples below illustrate.
 
 Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result.
-The snippet of code below uses the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
+The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
 \begin{example}
@@
 To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}.
-This is demonstarted for the Boston housing example in the code snippet below.
-\begin{example}
-lstat.ice <- partial(boston.rf, pred.var = "lstat",
-                     pred.fun = function(object, newdata) {
-                       predict(object, newdata)
-                     })  # see Figure 9 (left)
-\end{example}
-
-When the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which observation a curve belongs to; in this example, there is one curve for each unique observation in \code{boston}.
-To further illustrate, the code below constructs ICE curves for specific quantiles of the predictions.
-The results, along with the traditional PDP for \code{lstat}, are displayed in Figure~\ref{fig:partial_boston_ice}.
-\begin{example}
-lstat.quan <- partial(boston.rf, pred.var = "lstat",
-                      pred.fun = function(object, newdata) {
-                        quantile(predict(object, newdata), probs = 1:9/10)
-                      })  # see Figure 9 (middle)
-\end{example}
-\begin{figure}[!htbp]
-  \centering
-  \includegraphics[width=1.0\linewidth]{partial_boston_ice}
-  \caption{ICE curves and PDP depicting the relationship between \code{cmedv} and \code{lstat} for the Boston housing example (the same $y$-axis limits are used for each plot).
-\textit{Left}: ICE curves.
-\textit{Middle}: Specific quantiles of ICE curves.
-\textit{Right}: Standard PDP.}
-  \label{fig:partial_boston_ice}
-\end{figure}
-
-Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process ICE curves to obtain c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}.
-The code snippet below illustrates this for the Boston housing example using the predictor \code{age}.
+The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}.
+The result is displayed in Figure~\ref{fig:ice_boston}.
+\strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in this example, there is one curve for each observation in \code{boston}.
 \begin{example}
@@
 
-# Post-process age.ice to obtain c-ICE curves
+# Figure 9
+plotPartial(age.ice, rug = TRUE, train = boston, alpha = 0.3)
+\end{example}
+\begin{figure}[!htbp]
+  \centering
+  \includegraphics[width=0.8\linewidth]{ice_boston}
+  \caption{ICE curves depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example. Each curve corresponds to a different observation.}
+  \label{fig:ice_boston}
+\end{figure}
+
+The curves in Figure~\ref{fig:ice_boston} indicate some heterogeneity in the fitted model (i.e., some of the curves depict the opposite relationship).
+Such heterogeneity can be easier to spot using c-ICE curves; see Equation (4) on page 49 of \citet{goldstein-peeking-2015}.
+Using \CRANpkg{dplyr} \citep{dplyr-pkg}, it is rather straightforward to post-process the output from \code{partial} to obtain c-ICE curves (similar to the construction of \dfn{raw change scores} \citep[pg.
+130]{fitzmaurice-2011-applied} for longitudinal data).
+This is shown below.
+\begin{example}
+# Post-process rm.ice to obtain c-ICE curves
 library(dplyr)
-age.ice <- age.ice %>%
+rm.ice <- rm.ice %>%
   group_by(yhat.id) %>%
@@
 \end{example}
-Furthermore, using \pkg{ggplot2}, we can easily display ICE curves along with the PDP.
-This can be accomplished by using the \code{stat\_summary} function to average the ICE curves together.
-The code snippet below uses \pkg{ggplot2} and \code{grid.arrange} to plot ICE curves and c-ICE curves for \code{age} in the Boston housing example.
-The results are displayed in Figure~\ref{fig:partial_boston_ice_pdp}; compare these to Figures 3-4 in \citet{goldstein-peeking-2015}.
+
+Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot.
+This can be accomplished by using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together.
+The code snippet below uses \pkg{ggplot2} and \code{grid.arrange} to plot ICE curves and c-ICE curves for the predictor \code{rm} in the Boston housing example.
+The results are displayed in Figure~\ref{fig:ice_cice_boston}.
 \begin{example}
-# ICE curves
-p1 <- ggplot(age.ice, aes(age, yhat)) +
+# ICE curves with their average
+p1 <- ggplot(rm.ice, aes(rm, yhat)) +
   geom_line(aes(group = yhat.id), alpha = 0.2) +
@@
 
-# c-ICE curves
-p2 <- ggplot(age.ice, aes(age, yhat.centered)) +
+# c-ICE curves with their average
+p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
   geom_line(aes(group = yhat.id), alpha = 0.2) +
@@
   \centering
-  \includegraphics[width=1.0\linewidth]{partial_boston_ice_pdp}
-  \caption{ICE curves (black curves) and PDP (red curve) depicting the relationship between \code{cmedv} and \code{age} for the Boston housing example.
-\textit{Left}: Uncentered.
+  \includegraphics[width=1.0\linewidth]{ice_cice_boston}
+  \caption{ICE curves (black curves) and their average (red curve) depicting the relationship between \code{cmedv} and \code{rm} for the Boston housing example.
+\textit{Left}: Uncentered (here the red curve is just the traditional PDP).
 \textit{Right}: Centered.}
-  \label{fig:partial_boston_ice_pdp}
+  \label{fig:ice_cice_boston}
 \end{figure}

### 2016-12-30 R/partial.R

@@ -184,7 +184,13 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
   # Construct partial dependence data
   if (inherits(object, "gbm") && missing(pred.grid) && is.null(pred.fun)) {
 
-    # Assign value to grid.resolution
+    # If not supplied, try to extract training data from object when plot = TRUE
+    if (plot && missing(train)) {
+      train <- getTrainingData(object)
+    }
+
+    # Make sure grid.resolution is not NULL. If it is, use the same default as
+    # continuous.resolution in gbm::plot.gbm.
     if (is.null(grid.resolution)) {
       grid.resolution <- 100
     }
@@ -192,7 +198,7 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
     # Call gbm::plot.gbm directly; this is MUCH FASTER!
     pd.df <- gbm::plot.gbm(object, i.var = pred.var,
                            continuous.resolution = grid.resolution,
-                           return.grid = TRUE)
+                           return.grid = TRUE, ...)
 
     # Restrict to convex hull of first two predictors, if requested
     if (chull) {

### 2016-12-30 manuscript/greenwell.R

@@ -240,27 +240,32 @@ grid.arrange(
 )
 dev.off()
 
-# Use partial to obtain ICE curves for age
+# Use partial to obtain ICE curves for rm
 pred.ice <- function(object, newdata) predict(object, newdata)
-age.ice <- partial(boston.rf, pred.var = "age", pred.fun = pred.ice)
+rm.ice <- partial(boston.rf, pred.var = "rm", pred.fun = pred.ice)
 
-# Post-process age.ice to obtain c-ICE curves
-age.ice <- age.ice %>%
+# Figure 9
+pdf("ice_boston.pdf", width = 7, height = 5)
+plotPartial(rm.ice, rug = TRUE, train = boston, alpha = 0.2)
+dev.off()
+
+# Post-process rm.ice to obtain c-ICE curves
+rm.ice <- rm.ice %>%
   group_by(yhat.id) %>%
   mutate(yhat.centered = yhat - first(yhat))
 
-# ICE curves
-p1 <- ggplot(age.ice, aes(age, yhat)) +
+# ICE curves with PDP
+p1 <- ggplot(rm.ice, aes(rm, yhat)) +
   geom_line(aes(group = yhat.id), alpha = 0.2) +
   stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 
-# c-ICE curves
-p2 <- ggplot(age.ice, aes(age, yhat.centered)) +
+# c-ICE curves with centered PDP
+p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
   geom_line(aes(group = yhat.id), alpha = 0.2) +
   stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 
 # Figure 10
-pdf("partial_boston_ice_pdp.pdf", width = 8, height = 4)
+pdf("ice_cice_boston.pdf", width = 8, height = 4)
 grid.arrange(p1, p2, ncol = 2)
 dev.off()
 

### 2016-12-30 slowtests/slowtests-regression.R

@@ -16,6 +16,44 @@ library(pdp)
 # Load data
 data(boston)
 
+# Prediction functions
+pred.ice <- function(object, newdata) predict(object, newdata)
+pred.med <- function(object, newdata) median(predict(object, newdata))
+
+
+################################################################################
+# "cubist" objects: Cubist package
+################################################################################
+
+# Load required packages
+library(Cubist)
+
+# Fit a glm to the boston indians cmedv data
+boston.cubist <- cubist(x = subset(boston, select = -cmedv),
+                        y = boston$cmedv,
+                        committees = 100)
+
+# PDPs for lstat and rm
+grid.arrange(
+  partial(boston.cubist, pred.var = "lstat", plot = TRUE, rug = TRUE,
+          train = boston),
+  partial(boston.cubist, pred.var = "rm", plot = TRUE, rug = TRUE,
+          train = boston),
+  partial(boston.cubist, pred.var = c("lstat", "rm"), plot = TRUE,
+          chull = TRUE, train = boston),
+  ncol = 3
+)
+
+# User-specified prediction functions
+grid.arrange(
+  partial(boston.cubist, pred.var = "lstat", pred.fun = pred.ice, plot = TRUE,
+          rug = TRUE, train = boston),
+  partial(boston.cubist, pred.var = "lstat", pred.fun = pred.med, plot = TRUE,
+          rug = TRUE, train = boston),
+  ncol = 2
+)
+
+
 
 ################################################################################
 # "lm" objects: stats package
@@ -24,8 +62,47 @@ data(boston)
 # Fit a glm to the boston indians cmedv data
 boston.lm <- lm(cmedv ~ ., data = boston)
 
-# Construct a PDP for lstat
-partial(boston.lm, pred.var = "lstat", plot = TRUE)
+# PDPs for lstat and rm
+grid.arrange(
+  partial(boston.lm, pred.var = "lstat", plot = TRUE, rug = TRUE),
+  partial(boston.lm, pred.var = "rm", plot = TRUE, rug = TRUE),
+  partial(boston.lm, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
+  ncol = 3
+)
+
+# User-specified prediction functions
+grid.arrange(
+  partial(boston.lm, pred.var = "lstat", pred.fun = pred.ice, plot = TRUE,
+          rug = TRUE),
[…]

### 2016-12-31 manuscript/greenwell.tex
minor edits

@@
 The result is displayed in Figure~\ref{fig:ice_boston}.
-\strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in this example, there is one curve for each observation in \code{boston}.
+\strong{Note:} when the function supplied to \code{pred.fun} returns multiple predictions, the data frame returned by \code{partial} includes an additional column, \code{yhat.id}, that indicates which curve a point belongs to; in the following code chunk, there will be one curve for each observation in \code{boston}.
 \begin{example}
@@
 
-For illustration, we return to the Boston housing data.
-The code chunk below uses \pkg{caret} to train an XGBoost model to the \code{boston} data frame from Section~\ref{sec:boston} over a grid of tuning values (see \code{?xgboost::xgboost} for an explanation of the various tuning parameters).
-The optimal combination of tuning parameter values is selected via 5-fold cross-validation.
+For illustration, we return to the Boston housing example.
+The code chunk below uses \pkg{caret} to tune an \pkg{xgboost} model using 10-fold cross-validation.
+(After loading \pkg{caret}, use \code{getModelInfo("xgbTree")} for information on tuning \pkg{xgboost} models.) \strong{Warning:} The following code chunk may take a few minutes to run.
 \begin{example}
-library(caret)
-
-# Setup for 5-fold cross-validation
-ctrl <- trainControl(method = "cv", number = 5, verboseIter = TRUE)
-
-# Grid of tuning values for XGBoost.
-xgb.grid <- expand.grid(nrounds = c(1000, 2000, 3000),
-                        max_depth = 1:4,
-                        eta = c(0.001, 0.01, 0.1),
-                        gamma = 0,
-                        colsample_bytree = 1,
-                        min_child_weight = 1,
-                        subsample = c(0.5, 0.75, 1))
-
-# Train an XGBoost model
-set.seed(202)
+# Tune an XGBoost model using 10-fold cross-validation
+library(caret)  # functions related to classification and regression training
+set.seed(202)  # for reproducibility
 boston.xgb <- train(x = data.matrix(subset(boston, select = -cmedv)),
-                    y = boston$cmedv,
-                    method = "xgbTree",
-                    metric = "RMSE",
-                    trControl = ctrl,
-                    tuneGrid = xgb.grid)
+                    y = boston$cmedv, method = "xgbTree", metric = "Rsquared",
+                    trControl = trainControl(method = "cv", number = 10),
+                    tuneLength = 10)
 \end{example}
-The model with the smallest root-mean-squared error (RMSE) had \code{nrounds = 3000}, \code{max\_depth = 3}, \code{eta = 0.01}, and \code{subsample = 0.75}; the final model achieved a cross-validated RMSE and $R^2$ of 2.789 and 0.906, respectively.
+The optimal model had a cross-validated $R^2$ of $0.902$ which was achieved using \code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}.
 The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page.
@@
 \end{example}
-
 \begin{figure}[!htbp]
@@
   \includegraphics[width=1.0\linewidth]{boston_xgb}
-  \caption{PDPs for the top two most important variables in the Boston housing data using XGBoost.
+  \caption{PDPs for the top two most important variables in the Boston housing data using \pkg{xgboost}.
 Compare this to the random forest results displayed in Figures~\ref{fig:partial_extrap}-\ref{fig:partial_manual}.}
@@
 
-%Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object in a component called \code{trainingData} (e.g., \code{boston.xgb.tune\$trainingData}) there is no need to supply the training data in the call to \code{partial}.
-However, this is not the case when using the \pkg{xgboost} package directly.
-For example, the following code fits the same model using the \code{xgboost} function.
 Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object there is no need to supply it in the call to \code{partial}.
@@
                       label = boston$cmedv, objective = "reg:linear",
-                      nrounds = 2000, max_depth = 3, eta = 0.01)
+                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
+                      colsample_bytree = 0.8, min_child_weight = 1,
+                      subsample = 0.9444444)
 \end{example}

### 2017-01-01 manuscript/reviews/response.Rmd
much faster for  objects
minor changes

@@
 This has been fixed in `pdp` (version 0.3.0).
-The user will still have to supply `n.trees` in the call to `partial`.
+The user may still have to supply `n.trees` in the call to `partial` (depending on which version of `gbm` they are using).
 Also, as suggested in another comment, a `/slowtests` directory is underway and will help prevent future issues like this.
@@
 I also agree that PDPs are not the only, or sometimes best, approach to understanding the impact of a subset of the predictor's on the response.
-Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few others available in the `car` package.
+Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few other techniques available in the `car` package.
 Unless the editor disagrees, I don't think it is necessary to go into more detail than what is already provided at the end of the introduction.
-Nonetheless, I added a bit more of a motivational introduction that links to the examples using the Boston housing data.
+Nonetheless, I have expanded the introduction (which links to the Boston housing example used throughout the paper) and included more background on the `ICEbox` package.
 
@@
 # Fig 3b.
+
 ### Response
@@
 p1 abstract: strict assumptions imposed by traditional statistical models
+
 ### Response
@@
 ### Response 2.9
-Fixed. Color key labeled "Logit".
+Fixed. Color key labeled "Centered logit".
 
@@
 I understand that you like the feature of having a one-shot function partial to both generate and plot the PDP.
-I realize that the default is `FALS`E.
+I realize that the default is `FALSE`.
 However, I still think it's a disservice to the user as constructing PDP's is computationally intense and if they get the plot wrong or it's not to their liking, they will have to regenerate the whole thing again.
 So, it is my opinion that you should take it out and force two steps: generation and illustration.
+
 ### Response
@@
 The package could be more powerful if you allowed a predict function as a parameter. For instance, you cannot currently use random forests to illustrate classification probabilities.
+
 ### Response
-I agree, and I have changed the behavior of `...` to refer to additional parameters to `stats::predict` (e.g., the `n.trees` argument in `predict.gbm`.).
-However, I'm not sure yet about the scale on which the predictions are plotted for classification problems.
-For now, I am more comfortable leaving it as the usual logit scale defined for PDPs (mostly so that they are defined the same whatever the number of class outcomes).
-But I am not opposed to making this (simple) change in the future.
-I have added this as a feature request: https://github.com/bgreenwell/pdp/issues/21.
+This is a good suggestion.
+First, I have changed the behavior of `...` to refer to additional arguments in `predict` (e.g., the `n.trees` argument in `predict.gbm`.).
+Second, `partial` gained a new argument, `pred.fun`, where the user can provide an optional prediction function.
+This makes `partial` far more flexible.
+For example, you can use this argument to construct PDPs on the probability scale, or to construct ICE curves.
+These are illusrated in the new section labeled: **User-defined prediction functions**
 

### 2017-01-01 manuscript/greenwell.tex
minor changes

@@
 (c) in Algorithm~\ref{alg:pdp}.
-As pointed out by \citet{goldstein-peeking-2015}, since traditional PDPs are based on the sample mean, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest.
-This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE curves.
+Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest.
+This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE plot procedure.
 
 With \code{partial} it is possible to replace the mean in step 1.
-(c) with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale.
+(c) of Algorithm~\ref{alg:pdp} with any other function (e.g., the median or trimmed mean), or obtain PDPs for classification problems on the probability scale.
 It is even possible to obtain ICE curves.
@@
 The supplied prediction function must return either a single prediction or a vector of predictions.
-If the mean of all the predictions is returned, the result will be equivalent to the traditional PDP.
-If a vector of predictions is returned (i.e., one for each observation), the result will be equivalent to the ICE plot procedure.
+Returing the mean of all the predictions will result in the traditional PDP.
+Returing a vector of predictions (i.e., one for each observation) will result in a set of ICE curves.
 The examples below illustrate.
 
-Using the \code{pred.fun} argument it is possible to obtain PDPs for classification problems on the probability scale; we just need to write a function that extracts the predicted class probabilities of interest and then averages the result.
+Using the \code{pred.fun} argument, it is possible to obtain PDPs for classification problems on the probability scale.
+We just need to write a function that computes the predicted class probability of interest averaged across all observations.
 The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
@@
 Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}.
-The following chunk of code uses \code{partial} and \code{grid.arrange} to construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale.
+The following chunk of code construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale.
 The results are displayed in Figure~\ref{fig:partial_iris_prob}.
 \begin{example}
-# Display plots side by side (Figure 8)
-grid.arrange(
-  partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob, plot = TRUE),
-  partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob, plot = TRUE),
-  partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
-          pred.fun = pred.prob, plot = TRUE),
-  ncol = 3
-)
+# PDPs for Petal.Width and Petal.Length on the probability scale
+pdp.pw <- partial(iris.svm, pred.var = "Petal.Width", pred.fun = pred.prob,
+                  plot = TRUE)
+pdp.pl <- partial(iris.svm, pred.var = "Petal.Length", pred.fun = pred.prob,
+                  plot = TRUE)
+pdp.pw.pl <- partial(iris.svm, pred.var = c("Petal.Width", "Petal.Length"),
+                     pred.fun = pred.prob, plot = TRUE)
+
+# Figure 8
+grid.arrange(pdp.pw, pdp.pl, pdp.pw.pl, ncol = 3)
 \end{example}
@@
 This can be accomplished by using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together.
-The code snippet below uses \pkg{ggplot2} and \code{grid.arrange} to plot ICE curves and c-ICE curves for the predictor \code{rm} in the Boston housing example.
+The code snippet below plots the ICE curves and c-ICE curves, along with their averages, for the predictor \code{rm} in the Boston housing example.
 The results are displayed in Figure~\ref{fig:ice_cice_boston}.
@@
 
-# Display plots side by side (Figure 10)
+# Figure 10
 grid.arrange(p1, p2, ncol = 2)
@@
 \end{example}
-The optimal model had a cross-validated $R^2$ of $0.902$ which was achieved using \code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}.
-The following snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together, and then uses \code{grid.arrange} to display all three PDPs on the same page.
+The optimal model had a cross-validated $R^2$ of $0.902$ (use \code{print(boston.xgb\$bestTune)} to view the optimum tuning parameters).
+The next snippet of code computes the partial dependence of \code{cmedv} on both \code{rm} and \code{lstat}, individually and together.
 The results are displayed in Figure~\ref{fig:boston_xgb}.
 \begin{example}
-# Display plots side by side (Figure 11)
-grid.arrange(
-  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
-  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
-  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
-  ncol = 3
-)
+# PDPs for lstat and rm
+pdp.lstat <- partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE)
+pdp.rm <- partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE)
+pdp.lstat.rm <- partial(boston.xgb, pred.var = c("lstat", "rm"),
+                        plot = TRUE, chull = TRUE)
+
+# Figure 11
+grid.arrange(pdp.lstat, pdp.rm, pdp.lstat.rm, ncol = 3)
 \end{example}
@@
 \end{figure}
+%\code{nrounds = 100}, \code{max\_depth = 5}, \code{eta = 0.3}, \code{gamma = 0}, \code{colsample\_bytree = 0.8}, \code{min\_child\_weight = 1}, and \code{subsample = 0.9444444}
 
-Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object there is no need to supply it in the call to \code{partial}.
+The \code{train} function creates objects of class \code{"train"}, whereas the \code{xgboost} function creates objects of class \code{"xgb.Booster"}.
+Since \code{train} defaults to storing a copy of the training data as part of the \code{"train"} object, there is no need to supply it in the call to \code{partial} in this example.
 However, this is not the case when using the \pkg{xgboost} package directly.
-For example, the following code fits the same model using the \code{xgboost} function.
+To illustrate, we fit the same model using the \code{xgboost} function with the tuning parameters found previously using \pkg{caret}.
 \begin{example}
@@
 \end{example}
-To use \code{partial} with the \code{"xgb.Booster"} class, we need to supply the original training data (minus the response) in the call to \code{partial}.
+To use \code{partial} with \code{"xgb.Booster"} objects, we need to supply the original training data (minus the response) in the call to \code{partial}.
 The following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} (plot not shown).

### 2017-01-01 R/pdRegression.R

@@ -34,6 +34,31 @@ pdRegression.default <- function(object, pred.var, pred.grid, pred.fun,
 
 
 #' @keywords internal
+pdRegression.ksvm <- function(object, pred.var, pred.grid, pred.fun,
+                                 train, progress, parallel, paropts, ...) {
+  plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
+    temp <- train
+    temp[pred.var] <- x
+    out <- if (is.null(pred.fun)) {
+      pred <- kernlab::predict(object, newdata = temp, ...)[, 1L, drop = TRUE]
+      mean(pred, na.rm = TRUE)
+    } else {
+      pred.fun(object, newdata = temp)
+    }
+    if (length(out) == 1) {
+      stats::setNames(out, "yhat")
+    } else {
+      if (is.null(names(out))) {
+        stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      } else {
+        stats::setNames(out, paste0("yhat.", names(out)))
+      }
+    }
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
+}
+
+
+#' @keywords internal
 pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, pred.fun,
                                      train, progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
@@ -41,9 +66,6 @@ pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, pred.fun,
     temp[pred.var] <- x
     out <- if (is.null(pred.fun)) {
       pred <- stats::predict(object, newdata = data.matrix(temp), ...)
-      if (is.matrix(pred) || is.data.frame(pred)) {
-        pred <- pred[, 1L, drop = TRUE]
-      }
       mean(pred, na.rm = TRUE)
     } else {
       pred.fun(object, newdata = data.matrix(temp))

### 2017-01-01 R/utils.R

@@ -388,7 +388,7 @@ superType.xgb.Booster <- function(object) {
 
 #' @keywords internal
 superType.svm <- function(object) {
-  if (is.null(object$levels)) {
+  if (object$type %in% c(3, 4)) {
     "regression"
   } else {
     "classification"

### 2017-01-01 manuscript/greenwell.R

@@ -288,33 +288,32 @@ print(boston.xgb)
 # Tuning parameter 'min_child_weight' was
 # held constant at a value of 1
 # Rsquared was used to select the optimal model using  the largest value.
-# The final values used for the model were
-#   nrounds = 100
-#   max_depth = 5
-#   eta = 0.3
-#   gamma = 0
+# The final values used for the model were:
+#            nrounds = 100
+#          max_depth = 5
+#                eta = 0.3
+#              gamma = 0
 #   colsample_bytree = 0.8
 #   min_child_weight = 1
-#   subsample = 0.9444444
-# filter(boston.xgb$results,
-#        nrounds == 100 &
-#          max_depth == 5 &
-#          eta == 0.3 &
-#          gamma == 0 &
-#          colsample_bytree == 0.8 &
-#          min_child_weight == 1 &
-#          subsample == 0.94444444444444444444444)
+#          subsample = 0.9444444
 
-# Figure 8
+# Print the cross-validated resuts
+# filter(boston.xgb$results, nrounds == 100 & max_depth == 5 & eta == 0.3 &
+#        gamma == 0 & colsample_bytree == 0.8 & min_child_weight == 1 &
+#        subsample == 0.94444444444444444444444)
+
+# PDPs for lstat and rm
+pdp.lstat <- partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE)
+pdp.rm <- partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE)
+pdp.lstat.rm <- partial(boston.xgb, pred.var = c("lstat", "rm"),
+                        plot = TRUE, chull = TRUE)
+
+# Figure 11
 pdf("boston_xgb.pdf", width = 12, height = 4)
-grid.arrange(
-  partial(boston.xgb, pred.var = "lstat", plot = TRUE, rug = TRUE),
-  partial(boston.xgb, pred.var = "rm", plot = TRUE, rug = TRUE),
-  partial(boston.xgb, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
-  ncol = 3)
+grid.arrange(pdp.lstat, pdp.rm, pdp.lstat.rm, ncol = 3)
 dev.off()
 
-# Use xgboost function directly
+# Use use xgboost function directly
 set.seed(203)
 boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
                       label = boston$cmedv, objective = "reg:linear",

### 2017-01-01 manuscript/reviews/response.Rmd
minor edits

@@
 The documentation for `partial` has also been expanded to discuss this.
-Unfortunately, some functions just don't store a copy of the training data.
+Unfortunately, some functions don't store a copy of the training data.
 However, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary.
@@
 ### Response 1.7
-The last column is now labeled `yhat`; which is less likely to be a predictor name.
+The last column is now labeled `yhat`, which is less likely to be a predictor name.
 When one of the predictors is labeled `"y"` `randomForest::partialPlot` fails while `gbm::plot.gbm` does not.

### 2017-01-01 manuscript/greenwell.tex
minor edits

@@
 \citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data.
-One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see Table IV of \citet{harrison-1978-hedonic} for a description of each variable.
+One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see \citet[Table IV]{harrison-1978-hedonic} for a description of each variable.
 The data violate many classical assumptions like linearity, normality, and constant variance.
@@
 Algorithm~\ref{alg:pdp} can be quite computationally intensive since it involves $k$ passes over the training records.
-Fortunately, this can be parallelized quite easily (more on this in Section~\ref{sec:computational}).
-This algorithm can also be easily extended to larger subsets of two or more features as well.
+Fortunately, the algorithm can be parallelized quite easily (more on this in Section~\ref{sec:computational}).
+It can also be easily extended to larger subsets of two or more features as well.
 
-Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm}, among others; these are all limited in the sense that they only apply to the models fit using the respective package.
+Limited implementations of Friedman's PDPs are available in packages \CRANpkg{randomForest} \citep{randomForest-pkg} and \CRANpkg{gbm} \citep{gbm-pkg}, among others; these are limited in the sense that they only apply to the models fit using the respective package.
 For example, the \code{partialPlot} function in \pkg{randomForest} only applies to objects of class \code{"randomForest"} and the \code{plot} function in \pkg{gbm} only applies to \code{"gbm"} objects.
 While the \pkg{randomForest} implementation will only allow for a single predictor, the \pkg{gbm} implementation can deal with any subset of the predictor space.
-Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., multiple linear regression and neural networks).
+Partial dependence functions are not restricted to tree-based models; they can be applied to any supervised learning algorithm (e.g., generalized additive models and neural networks).
 However, to our knowledge, there is no general package for constructing PDPs in R.
-For example, PDPs for a \dfn{conditional random forest} as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively.
+For example, PDPs for a conditional random forest as implemented by the \code{cforest} function in the \CRANpkg{party} and \CRANpkg{partykit} packages; see \citet{party-pkg} and \citet{partykit-pkg}, respectively.
 The \CRANpkg{pdp} \citep{pdp-pkg} package tries to close this gap by offering a general framework for constructing PDPs that can be applied to several classes of fitted models.
@@
 The \CRANpkg{plotmo} package \citep{plotmo-pkg} is one alternative to \pkg{pdp}.
-According to its author, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
+According to \citeauthor{plotmo-pkg}, \pkg{plotmo} constructs "a poor man's partial dependence plot." In particular, it plots a model's response when varying one or two predictors while holding the other predictors in the model constant (continuous features are fixed at their median value, while factors are held at their first level).
 These plots allow for up to two variables at a time.
@@
 For additive models (i.e., models with no interactions), these plots are identical in shape to PDPs.
+As of \pkg{plotmo} version 3.3.0, there is now support for constructing PDPs, but it is not the default.
+The main difference is that \pkg{plotmo}, rather than applying step 1.
+(a)-(c) in Algorithm~\ref{alg:pdp}, accumulates all the data at once thereby reducing the number of internal calls to \code{predict}.
+The trade-off is a slight increase in speed at the expense of using more memory.
+So, why use the \pkg{pdp} package?
+As will be discussed in the upcoming sections, \pkg{pdp}:
+\begin{itemize}
+  \item contains only a few functions with relatively few arguments;
+  \item does \underline{not} produce a plot by default;
+  \item can be used more efficiently with \code{"gbm"} objects (see Section~\ref{sec:computational});
+  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are far more flexible than base R graphcs;
+  \item defaults to using false color level plots for multivariate displays (see Section~\ref{sec:multi});
+  \item contains options to mitigate the risks associated with extrapolation (see Section~\ref{sec:computational});
+  \item has the option to construct PDPs in parallel (see Section~\ref{sec:computational});
+    \item is extremely flexible in the types of PDPs that can be produced (see Section~\ref{sec:prediction}),
+\end{itemize}
 
@@
 Many other techniques exist for visualizing relationships between the predictors and the response based on a fitted model.
-For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing partial-residual and marginal-model plots.
-The \CRANpkg{effects} package \citep{fox-effects-2003} is also of interest.
-It provides tabular and graphical displays for the terms in parametric models.
+For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots.
+\dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provides tabular and graphical displays for the terms in parametric models while holding all other predcitors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots.
 However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).
@@
 If \code{plot = FALSE} (the default), \code{partial} returns an object of class \code{"partial"} which inherits from the class \code{"data.frame"}; put another way, by default, \code{partial} returns a data frame with an additional class that is recognized by the \code{plotPartial} function.
-The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is always labeled \code{yhat}\footnote{There is one exception to this.
-When a function supplied via the \code{pred.fun} argument (this argument is discussed in Section~\ref{sec:prediction}) returns multiple predictions, the last two columns will be labeled \code{yhat} and \code{yhat.id}.} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
-If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} and returns the PDP in the form of a \CRANpkg{lattice} plot \citep{lattice-pkg}.
+The columns of the data frame are labeled in the same order as the features supplied to \code{pred.var}, and the last column is labeled \code{yhat}\footnote{There is one exception to this.
+When a function supplied via the \code{pred.fun} argument returns multiple predictions, the second to last and last columns will be labeled \code{yhat} and \code{yhat.id}, respectively (see Section~\ref{sec:prediction}).} and contains the values of the partial dependence function $\bar{f}_s\left(\boldsymbol{z}_s\right)$.
+If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} (with fewer plotting options) and returns the PDP in the form of a \pkg{lattice} plot (i.e., a \code{"trellis"} object).
+\strong{Note:}  it is reccomended to call \code{partial} with \code{plot = FALSE} and store the results; this allows for more flexible plotting, and the user will not have to waste time calling \code{partial} again if the default plot is not sufficient.
 
@@
 For example, \code{plotPartial} makes it straight forward to add a LOESS smooth, or produce a 3-D surface instead of a false color level plot (the default).
-Of course, since the default output produced by \code{partial} is still a data frame, the user can easily use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance.
+Of course, since the default output produced by \code{partial} is still a data frame, the user can easily use any plotting package he/she desires to visualize the results---\CRANpkg{ggplot2} \citep{ggplot2-pkg}, for instance (see Section~\ref{sec:classification} and Section~\ref{sec:prediction} for examples).
 
@@
 Simple solutions are available in packages \CRANpkg{latticeExtra} \citep{latticeExtra-pkg} and \CRANpkg{gridExtra} \citep{gridExtra-pkg}.
-For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{pdp} plots on a single page.
+For convenience, \pkg{pdp} imports the \code{grid.arrange} function from \pkg{gridExtra} which makes it easy to display multiple \pkg{grid}-based graphical objects on a single plot (these include graphics produced using \pkg{lattice} (hence, \pkg{pdp}) and \pkg{ggplot2}).
+This is demonstrated in multiple examples throughout this paper.
 
 Currently supported models are described in Table~\ref{tab:models}.
-In a lot of these cases, \code{partial} should be able to automatically determine appropriate values for the options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train} (the original training data).
+In a lot of these cases, \code{partial} should be able to automatically determine appropriate values for the options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train}\footnote{For reasons described in the package documentation, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}; see \code{?partial for details}} (the original training data).
 In other situations, the user may need to specify one or both of these additional arguments.
@@
 \subsection{Multi-predictor PDPs}
+\label{sec:multi}
 
@@
 \subsection{Avoiding extrapolation}
+\label{sec:extrapolation}
 
@@
 \subsection{Classification problems}
-\label{sec:class}
+\label{sec:classification}
 
@@
 We just need to write a function that computes the predicted class probability of interest averaged across all observations.
-The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:class} to extract the average predicted probability of belonging to the \code{Setosa} class.
+The function below can be used with the fitted SVM from the iris example of Section~\ref{sec:classification} to extract the average predicted probability of belonging to the \code{Setosa} class.
 \begin{example}

### 2017-01-02 manuscript/greenwell.tex
minor edits

@@
 Currently supported models are described in Table~\ref{tab:models}.
-In a lot of these cases, \code{partial} should be able to automatically determine appropriate values for the options \code{type} (i.e., \code{"regression"} or \code{"classification"}) and \code{train}\footnote{For reasons described in the package documentation, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}; see \code{?partial for details}} (the original training data).
-In other situations, the user may need to specify one or both of these additional arguments.
-These two arguments allow \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+In these cases, \code{partial} should be able to automatically determine an appropriate value for \code{type} (\code{"regression"} or \code{"classification"}).
+In other situations, the user may need to specify this argument.
+This allows \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+
+If \code{train = NULL}, \code{partial} tries to extract the original training data from the fitted model object.
+For objects that typically store a copy of the training data (e.g., objects of class \code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is straightforward.
+Otherwise, \code{partial} will attempt to extract the call stored in \code{object} (if available) and use that to evaluate the training data in the same environment from which \code{partial} was called.
+This can cause problems when, for example, the training data have been changed after fitting the model, but before calling \code{partial}.
+Hence, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}.
 

### 2017-01-02 manuscript/greenwell.tex
add ranger support

@@
 Currently supported models are described in Table~\ref{tab:models}.
-In these cases, \code{partial} should be able to automatically determine an appropriate value for \code{type} (\code{"regression"} or \code{"classification"}).
+In these cases, \code{partial} should be able to automatically determine an appropriate value for the argument \code{type} (i.e., \code{"regression"} or \code{"classification"}).
 In other situations, the user may need to specify this argument.
 This allows \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
-
-If \code{train = NULL}, \code{partial} tries to extract the original training data from the fitted model object.
-For objects that typically store a copy of the training data (e.g., objects of class \code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is straightforward.
-Otherwise, \code{partial} will attempt to extract the call stored in \code{object} (if available) and use that to evaluate the training data in the same environment from which \code{partial} was called.
-This can cause problems when, for example, the training data have been changed after fitting the model, but before calling \code{partial}.
-Hence, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}.
 
@@
       Random forest             & \pkg{randomForest} & \code{"randomForest"} \\
-      Conditional random forest & \pkg{party}    & \code{"RandomForest"} \\
+                                & \pkg{party}        & \code{"RandomForest"} \\
                                 & \pkg{partykit} & \code{"cforest"} \\
+                                & \CRANpkg{ranger} \citep{ranger-pkg} & \code{"ranger"} \\
+
       Linear model              & \pkg{stats}    & \code{"lm"} \\
@@
 
-When additional arguments are necessary, the user will be prompted with an informative error message. For instance, the user may occasionally see the following:
+Another important argument is \code{train}.
+If \code{train = NULL} (the default), \code{partial} tries to extract the original training data from the fitted model object.
+For objects that typically store a copy of the training data (e.g., objects of class \code{"BinaryTree"}, \code{"RandomForest"}, and \code{"train"}), this is straightforward.
+Otherwise, \code{partial} will attempt to extract the call stored in \code{object} (if available) and use that to evaluate the training data in the same environment from which \code{partial} was called.
+This can cause problems when, for example, the training data have been changed after fitting the model, but before calling \code{partial}.
+Hence, it is good practice to always supply the training data via the \code{train} argument in the call to \code{partial}\footnote{For brevity, we ignore this option in most of the examples in this paper.}.
+If \code{train = NULL} and the training data can not be extracted from the fitted model, the user will be prompted with an informative error message (this will occur, for example, when using \code{partial} with \code{"ksvm"} and \code{"xgb.Booster"} objects):
 \begin{example}
@@
 \end{example}
-This will occur, for example, when using partial with the \code{"xgb.Booster"} class (see Section~\ref{sec:xgboost}).
-In fact, using \code{partial} with any object that does not store the original training data will result in the above error message.
 

### 2017-01-02 R/pdClassification.R

@@ -231,6 +231,36 @@ pdClassification.nnet <- function(object, pred.var, pred.grid, pred.fun,
 
 
 #' @keywords internal
+pdClassification.ranger <- function(object, pred.var, pred.grid, pred.fun,
+                                    which.class, train,
+                                    progress, parallel, paropts, ...) {
+  if (is.null(getCall(object)$probability)) {
+    stop(paste("Cannot obtain predicted probabilities from",
+               deparse(substitute(object))))
+  }
+  plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
+    temp <- train
+    temp[pred.var] <- x
+    out <- if (is.null(pred.fun)) {
+      avgLogit(stats::predict(object, data = temp, ...)$predictions,
+               which.class = which.class)
+    } else {
+      pred.fun(object, newdata = temp)
+    }
+    if (length(out) == 1) {
+      stats::setNames(out, "yhat")
+    } else {
+      if (is.null(names(out))) {
+        stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      } else {
+        stats::setNames(out, paste0("yhat.", names(out)))
+      }
+    }
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
+}
+
+
+#' @keywords internal
 pdClassification.svm <- function(object, pred.var, pred.grid, pred.fun,
                                  which.class, train,
                                  progress, parallel, paropts, ...) {

### 2017-01-02 R/pdRegression.R

@@ -59,6 +59,30 @@ pdRegression.ksvm <- function(object, pred.var, pred.grid, pred.fun,
 
 
 #' @keywords internal
+pdRegression.ranger <- function(object, pred.var, pred.grid, pred.fun, train,
+                                progress, parallel, paropts, ...) {
+  plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
+    temp <- train
+    temp[pred.var] <- x
+    out <- if (is.null(pred.fun)) {
+      mean(stats::predict(object, data = temp, ...)$predictions, na.rm = TRUE)
+    } else {
+      pred.fun(object, newdata = temp)
+    }
+    if (length(out) == 1) {
+      stats::setNames(out, "yhat")
+    } else {
+      if (is.null(names(out))) {
+        stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      } else {
+        stats::setNames(out, paste0("yhat.", names(out)))
+      }
+    }
+  }, .progress = progress, .parallel = parallel, .paropts = paropts)
+}
+
+
+#' @keywords internal
 pdRegression.xgb.Booster <- function(object, pred.var, pred.grid, pred.fun,
                                      train, progress, parallel, paropts, ...) {
   plyr::adply(pred.grid, .margins = 1, .fun = function(x) {

### 2017-01-03 manuscript/reviews/response.Rmd
MAJOR code cleanup!

@@
 
-Thanks to the reviewers and editor for the well thought out comments.
-I believe that they have greatly imporved, not only the manuscript, but the package it describes as well.
-I hope that my responses are at least satisfactory.
+Thanks to the reviewers and editor for their time and and comments.
+I believe that they have greatly improved, not only the manuscript, but the package it describes as well.
 Please find below a copy of the reviewers comments along with my responses.
@@
 If sourcing a script, then the plot needs to be explicitly printed (e.g., `print(plotPartial(pd))` or `print(partial(fit, pred.var = "x", plot = TRUE))`).
-Additional text has been added to manuscript (see **Note:** on page 3) and the package documentation explaining this behavior.
+This has been clarified in the package documentation and the manuscript (see **Note:** at the bottom of page 3).
 The fact that this worked when calling `partial` with `plot = TRUE` was a bug (which is now fixed).

### 2017-01-03 manuscript/greenwell.tex
MAJOR code cleanup!

@@
 \end{enumerate}
-\caption{A simple algorithm for consructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
+\caption{A simple algorithm for constructing the partial dependence of the response on a single predictor $x_1$. \label{alg:pdp}}
 \end{algorithm}
@@
   \item can be used more efficiently with \code{"gbm"} objects (see Section~\ref{sec:computational});
-  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are far more flexible than base R graphcs;
+  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are far more flexible than base R graphics;
   \item defaults to using false color level plots for multivariate displays (see Section~\ref{sec:multi});
@@
 For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots.
-\dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provides tabular and graphical displays for the terms in parametric models while holding all other predcitors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots.
+\dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provides tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots.
 However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).
@@
 If \code{plot = TRUE}, then \code{partial} makes an internal call to \code{plotPartial} (with fewer plotting options) and returns the PDP in the form of a \pkg{lattice} plot (i.e., a \code{"trellis"} object).
-\strong{Note:}  it is reccomended to call \code{partial} with \code{plot = FALSE} and store the results; this allows for more flexible plotting, and the user will not have to waste time calling \code{partial} again if the default plot is not sufficient.
+\strong{Note:}  it is recommended to call \code{partial} with \code{plot = FALSE} and store the results; this allows for more flexible plotting, and the user will not have to waste time calling \code{partial} again if the default plot is not sufficient.
 
@@
 
-Another important argument is \code{train}.
+The \code{partial} function also supports objects of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}.
+This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}.
+An example is given in Section~\ref{sec:xgboost}.
+
+Another important argument to \code{partial} is \code{train}.
 If \code{train = NULL} (the default), \code{partial} tries to extract the original training data from the fitted model object.
@@
 \end{example}
-
-The \code{partial} function also supports objects of class \code{"train"} produced using the \code{train} function from the well-known \CRANpkg{caret} package \citep{caret-pkg}.
-This means that \code{partial} can be used with any classification or regression model that has been fit using \pkg{caret}'s \code{train} function; see \url{http://topepo.github.io/caret/available-models.html} for a current list of models supported by \pkg{caret}.
-An example is given in Section~\ref{sec:xgboost}.
 
@@
 \begin{example}
-library(randomForest)
+library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
 set.seed(101)  # for reproducibility
@@
 \begin{example}
-library(pdp)
+library(pdp)  # for partial, plotPartial, and grid.arrange functions
 partial(boston.rf, pred.var = "lstat", plot = TRUE)  # Figure 2 (left)
@@
 Alternatively, a user-specified grid of values (e.g., containing specific quantiles of interest) can be supplied through the \code{pred.grid} argument.
-To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option, then uses \code{grid.arrange} to display all three PDPs on the same page; the results are displayed in Figure~\ref{fig:partial_manual}.
+To demonstrate, the following snippet of code computes the partial dependence of \code{cmedv} on \code{rm} using each option; \code{grid.arrange} is used to display all three PDPs on the same graph, side by side.
+The results are displayed in Figure~\ref{fig:partial_manual}.
 \begin{example}
-# Display plots side by side (Figure 5)
+# Figure 5
 grid.arrange(
@@
 \begin{example}
+library(earth)  # for earth function (i.e., MARS algorithm)
 ozone.mars <- earth(ozone ~ ., data = ozone, degree = 3)
@@
   \caption{Partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg}.
-Since \code{dpg} is contiuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
+Since \code{dpg} is continuous, it is first converted to a shingle; in this case, four groups with 10\% overlap.}
   \label{fig:partial_par}
@@
 \begin{example}
-library(e1071)
+library(e1071)  # for svm function
 iris.svm <- svm(Species ~ ., data = iris, kernel = "radial", gamma = 0.75,
@@
 
-Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes.
-The result is displayed in Figure~\ref{fig:partial_iris}.
-%Note that we had to supply the original training data to \code{partial} through the \code{train} option.
-This is necessary for fitted model objects that don't store a copy of the training data (e.g., model fitting functions that don't rely on a formula method).
+Next, we plot the partial dependence of \code{Species} on both \code{Petal.Width} and \code{Petal.Length} for each of the three classes. The result is displayed in Figure~\ref{fig:partial_iris}.
 \begin{example}
@@
   \includegraphics[width=1.0\linewidth]{partial_iris_svm}
-  \caption{Partial dependence of species on petal width and petal length for the iris data.}
+  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} for the iris data.}
   \label{fig:partial_iris}
@@
 The supplied prediction function must return either a single prediction or a vector of predictions.
-Returing the mean of all the predictions will result in the traditional PDP.
-Returing a vector of predictions (i.e., one for each observation) will result in a set of ICE curves.
+Returning the mean of all the predictions will result in the traditional PDP.
+Returning a vector of predictions (i.e., one for each observation) will result in a set of ICE curves.
 The examples below illustrate.
@@
   \includegraphics[width=1.0\linewidth]{partial_iris_svm_prob}
-  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probabiltity of belonging to the \code{Setosa} species.}
+  \caption{Partial dependence of \code{Species} on \code{Petal.Width} and \code{Petal.Length} plotted on the probability scale; in this case, the probability of belonging to the setosa species.}
   \label{fig:partial_iris_prob}
@@
 # Post-process rm.ice to obtain c-ICE curves
-library(dplyr)
+library(dplyr)  # for group_by and mutate functions
 rm.ice <- rm.ice %>%
@@
 \begin{example}
+library(xgboost)  # for xgboost function
 set.seed(203)  # for reproducibility

### 2017-01-03 R/pdClassification.R

@@ -1,291 +1,130 @@
 #' @keywords internal
 pdClassification <- function(object, pred.var, pred.grid, pred.fun, which.class,
                              train, progress, parallel, paropts, ...) {
-  UseMethod("pdClassification")
-}
 
+  # Use plyr::adply, rather than a for loop
+  plyr::adply(pred.grid, .margins = 1, .progress = progress,
+              .parallel = parallel, .paropts = paropts, .fun = function(x) {
 
-#' @keywords internal
-pdClassification.default <- function(object, pred.var, pred.grid, pred.fun,
-                                     which.class, train,
-                                     progress, parallel, paropts, ...) {
-  plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
+    # Copy training data and replace pred.var with constant
     temp <- train
     temp[pred.var] <- x
-    out <- if (is.null(pred.fun)) {
-      pr <- stats::predict(object, newdata = temp, type = "prob", ...)
-      avgLogit(pr, which.class = which.class)
+
+    # Get prediction(s)
+    if (is.null(pred.fun)) {
+      stats::setNames(pdPredictClassification(object, newdata = temp,
+                                              which.class = which.class, ...),
+                      "yhat")
     } else {
-      pred.fun(object, newdata = temp)
-    }
-    if (length(out) == 1) {
-      stats::setNames(out, "yhat")
-    } else {
-      if (is.null(names(out))) {
-        stats::setNames(out, paste0("yhat.", 1L:length(out)))
+      out <- pred.fun(object, newdata = temp)
+      if (length(out) == 1) {
+        stats::setNames(out, "yhat")
       } else {
-        stats::setNames(out, paste0("yhat.", names(out)))
+        if (is.null(names(out))) {
+          stats::setNames(out, paste0("yhat.", 1L:length(out)))
+        } else {
+          stats::setNames(out, paste0("yhat.", names(out)))
+        }
       }
     }
-  }, .progress = progress, .parallel = parallel, .paropts = paropts)
+
+  })
+
 }
 
 
 #' @keywords internal
-pdClassification.glm <- function(object, pred.var, pred.grid, pred.fun,
-                                 which.class, train,
-                                 progress, parallel, paropts, ...) {
-  plyr::adply(pred.grid, .margins = 1, .fun = function(x) {
-    temp <- train
-    temp[pred.var] <- x
[…]

### 2017-01-03 R/plotPartial.R

@@ -31,8 +31,9 @@
 #' @param ... Additional optional arguments to be passed onto \code{dotplot},
 #'   \code{levelplot}, \code{xyplot}, or \code{wireframe}.
 #'
-#' @importFrom lattice dotplot equal.count levelplot panel.levelplot panel.lines
-#' @importFrom lattice panel.loess panel.xyplot panel.rug wireframe xyplot
+#' @importFrom lattice dotplot equal.count levelplot panel.dotplot
+#' @importFrom lattice panel.levelplot panel.lines panel.loess panel.xyplot
+#' @importFrom lattice panel.rug wireframe xyplot
 #'
 #' @rdname plotPartial
 #' @export

### 2017-01-03 manuscript/greenwell.R

@@ -46,7 +46,7 @@ setwd("manuscript")
 # Load required packages
 library(doParallel)
 library(caret)
-library(dplyr)
+# library(dplyr)  # use dplyr:: instead to avoid conflictions with plyr!
 library(e1071)
 library(earth)
 library(ggplot2)
@@ -57,7 +57,7 @@ library(xgboost)
 
 
 ################################################################################
-# Boston housing example
+# Constructing PDPs in R
 ################################################################################
 
 # Load the (corrected) Boston housing data
@@ -81,6 +81,11 @@ pdf("boston_rf_vimp.pdf", width = 7, height = 5)
 varImpPlot(boston.rf)  # dotchart of variable importance scores
 dev.off()
 
+
+################################################################################
+# Single predictor PDPs
+################################################################################
+
 # Figure 2
 pd.lstat <- partial(boston.rf, pred.var = "lstat")
 pdf("pd_lstat.pdf", width = 8, height = 4)
@@ -90,8 +95,10 @@ pdp2 <- plotPartial(pd.lstat, lwd = 2, smooth = TRUE,
 grid.arrange(pdp1, pdp2, ncol = 2)
 dev.off()
 
-# Fit a MARS model
-boston.mars <- earth(cmedv ~ ., data = boston, degree = 3)
+
+################################################################################
+# Multi-predictor PDPs
+################################################################################
 
 # Figure 3
 rwb <- colorRampPalette(c("red", "white", "blue"))
@@ -103,6 +110,11 @@ p3 <- plotPartial(pd.lstat.rm, levelplot = FALSE, zlab = "cmedv", drape = TRUE,
                   colorkey = FALSE, screen = list(z = -20, x = -60))
 grid.arrange(p1, p2, p3, ncol = 3)
 dev.off()
+
+
+################################################################################
+# Avoiding extrapolation
+################################################################################
 
 # Figure 4
 pdf("partial_extrap.pdf", width = 8, height = 4)
@@ -126,7 +138,7 @@ dev.off()
 
 
 ################################################################################
-# Los Angeles ozone example
+# Addressing computational concerns
 ################################################################################
 
 # Load the LA ozone data
[…]

### 2017-01-03 slowtests/slowtests-regression.R

@@ -20,7 +20,9 @@ library(nnet)
 library(party)
 library(pdp)
 library(randomForest)
+library(ranger)
 library(rpart)
+library(xgboost)
 
 # Load data
 data(boston)
@@ -48,12 +50,12 @@ pred.med <- function(object, newdata) {
 # Fit models
 ################################################################################
 
-# Cubist
+# Cubist::cubist
 boston.cubist <- cubist(x = subset(boston, select = -cmedv),
                         y = boston$cmedv,
                         committees = 100)
 
-# e1071
+# e1071::svm
 boston.svm <- svm(cmedv ~ ., data = boston, type = "eps-regression")
 
 # earth
@@ -70,17 +72,17 @@ boston.gbm <- gbm(cmedv ~ .,
                   verbose = TRUE)
 best.iter <- gbm.perf(boston.gbm, method = "OOB")
 
-# kernlab
+# kernlab::svm
 boston.ksvm <- ksvm(cmedv ~ ., data = boston, type = "eps-svr")
 
 # nnet
 boston.nnet <- nnet(cmedv ~ ., data = boston, size = 6, decay = 0.1,
                     linout = TRUE)
 
-# party (ctree)
+# party::ctree
 boston.ctree <- ctree(cmedv ~ ., data = boston)
 
-# party (cforest)
+# party::cforest
 set.seed(101)
 boston.crf <- cforest(cmedv ~ ., data = boston)
 
@@ -88,77 +90,181 @@ boston.crf <- cforest(cmedv ~ ., data = boston)
 set.seed(101)
 boston.rf <- randomForest(cmedv ~ ., data = boston)
 
+# ranger
+set.seed(101)
+boston.ranger <- ranger(cmedv ~ ., data = boston)
+
 # rpart
 boston.rpart <- rpart(cmedv ~ ., data = boston)
 
-# stats (lm)
+# stats::lm
 boston.lm <- lm(cmedv ~ ., data = boston)
 
-# stats (glm)
+# stats::glm
 boston.glm <- glm(cmedv ~ ., data = boston, family = gaussian)
 
+# xgboost
+set.seed(101)
+boston.xgb <- xgboost(data = data.matrix(subset(boston, select = -cmedv)),
+                      label = boston$cmedv, objective = "reg:linear",
+                      nrounds = 100, max_depth = 5, eta = 0.3, gamma = 0,
+                      colsample_bytree = 0.8, min_child_weight = 1,
[…]

### 2017-01-04 manuscript/reviews/response.Rmd
minor edits

@@
 
+```{r, include=FALSE}
+rm(data)
+```
+
 ### Comment 1.5
@@
 data(trees)
-foo <- function(data)
-{
+foo <- function(data) {
   lm(Volume ~ ., data = data)
@@
 Agreed!
-The documentation for `partial` has also been expanded to discuss this.
+Both the manuscript (see page 3) and the help page for `partial` have been expanded to discuss this.
 Unfortunately, some functions don't store a copy of the training data.
 However, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary.
+Re-running your first example should give an informative error message:
+```{r, eval=TRUE, error=TRUE}
+data(trees)
+foo <- function(data)
+{
+  lm(Volume ~ ., data = data)
+}
+mod <- foo(trees)
+library(pdp)
+partial(mod, pred.var = "Girth", plot = TRUE)  # fails with better error mssg
+```
 
@@
 This has been fixed in `pdp` (version 0.3.0) by adding an informative error message listing which variables are the cause of the issue.
-```{r, eval=FALSE}
+```{r, error=TRUE}
 library(pdp)
@@
 partial(mod, pred.var=c("Girth", "foo", "bar"), plot=TRUE)
-```
-```
-Error in partial.default(mod, pred.var = c("Girth", "foo", "bar"), plot = TRUE) : foo, bar not found in the training data.
 ```
@@
 When one of the predictors is labeled `"y"` `randomForest::partialPlot` fails while `gbm::plot.gbm` does not.
-I think regardless of what the default column name is, there is always the possibility of this issue.
+I think regardless of what the default column name is, there is always the possibility of this problem.
 This has been filed as an issue on the `pdp` GitHib page: https://github.com/bgreenwell/pdp/issues/22.
@@
 ### Response 1.10
-The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letetrs).
+The package was originally called `partial`, but at the time of its original submission this conflicted with a CRAN package of the same name (but with mixed lowercase/uppercase letters).
 So, I decided to change the name of the package to `pdp`; besides, I think `partial` and `plotPartial` work well together as function names.
@@
 ### Response 1.11
-I dont think so, because I want `partial` to behave like a `lattice` function when using `plot = TRUE` or `plotPartial`; that is, I want it to return a `"trellis"` object.
-However, I did add the partial dependence data as an attribute to the returned `"trellis"` object (i.e., plot) so that it can easily be extracted if necessary (e.g., if they do not like the plot and don't want to re-run partial).
+I dont think so, because I want `partial` to behave like a `lattice` function when using `plot = TRUE`; that is, I want it to return a `"trellis"` object.
+However, I did add the partial dependence data as an attribute to the returned `"trellis"` object so that it can easily be extracted if necessary (e.g., if they do not like the plot and don't want to re-run partial).
 For example,
@@
 Also, the `lattice` function `trellis.last.object` is now exported when `pdp` is loaded.
-This will be useful when the user does not store the resulting plot in the call to `partial` but needs to retrieve the outout data to avoid re-running `partial` (e.g., when computation time is a concern).
+This would be useful if the user did not store the PDP but needs to retrieve the outout data to avoid re-running `partial` (e.g., when computation time is a concern).
 For example,
@@
 ```
-This is described in the package documentation. Thanks for the idea!
+This has been noted in the package documentation (see `?partial`). Thanks for the idea!
 
@@
 ```{r, eval=FALSE}
-par(mfrow = c(2, 2))
+par(mfrow = c(1, 2))
 partial(ozone.rf, pred.var = "Temp", plot = TRUE)
@@
 ```
+
 However, since `partial` returns a data frame by default, using `par` would work in the following case
-```{r, eval=FALSE}
-par(mfrow=c(2,2))
+```{r, eval=TRUE, fig.width=8, fig.height=4}
+par(mfrow = c(1, 2))
 plot(partial(ozone.rf, pred.var = "Temp"), type = "l")
@@
 ### Response 1.13
-I agree, this would be a nice feature, but the problem I see is that not all models naturally emit a "variable importance" score.
-The `caret` package has a come a long way in this regard.
-At this point, I think it is simpler to have the user select which variables they would like to see plotted (it forces them to explore the results).
+I agree, this would be a nice feature, but the problems I see are (1) not all models naturally emit a "variable importance" score; the `caret` package has come a long way in this regard, and (2) too many plots will look cluttered.
+At this point, I'd rather encourage the user to select which variables they would like to see plotted.
 I am not opposed to including this feature in the future.
-I added added a possible solution along with a feature request: https://github.com/bgreenwell/pdp/issues/19.
+I added this as a feature request: https://github.com/bgreenwell/pdp/issues/19.
 Thanks for the suggestion!
@@
 ### Response 1.14
-As mentioned in `Response 1.?`, I have expanded the code to work harder---especially for `S4` methods---to inform the user to supply the training data when necessary.
-As another example, `kernlab::ksvm` does not store a copy of the training data, at least not as far as I can tell.
+`kernlab::ksvm` does not store a copy of the training data either, as far as I can tell.
 As time goes on, the coverage will certainly improve, which will make it easier on the user.
 For example, `partial` now works with `caret` which, by default, stores the taining data in a component called `trainingData`.
-
-I suspect that the package maintainers for `xgboost` (and similar packages for designed for working with "large" data sets) would be hesitent in storing a copy of the training data with the fitted model object for memory reasons.
+Also, I suspect that the package maintainers for `xgboost` (and similar packages designed for working with "large" data sets) would be hesitent in storing a copy of the training data with the fitted model for memory reasons.
 
@@
 ### Response
+I expanded the introduction a bit to try and motivate why PDPs are important (the new material links directly to the Boston housing example used throughout the manuscript).
+I also included more background on the `plotmo` and `ICEbox` packages (and there corresponding methodologies).
+`plotmo` very recently introduced a `partdep` option for displaying PDPs.
+Because of this, more detail was added to differentiate `pdp` and motivate the use of `partial`.
+
 I agree that PDPs are not as useful in "simple" linear models (e.g., few variables, only two-way interactions, etc.), but certainly there will be cases where PDPs will be extremely useful (e.g., many terms, complex interactions, etc.).
@@
 Brief descriptions and references are given for many alternatives: marginal model plots (`plotmo`), individual conditional expectation plots (`ICEbox`)---which are described as an alternatve to PDPs when strong interactions are suspected, effect plots (`effects`), and a few other techniques available in the `car` package.
-Unless the editor disagrees, I don't think it is necessary to go into more detail than what is already provided at the end of the introduction.
-Nonetheless, I have expanded the introduction (which links to the Boston housing example used throughout the paper) and included more background on the `ICEbox` package.
 
@@
 ### Response
-`ICEbox` does not actually construct PDPs.
-Only graphics trying to do the same thing: visualize the relationship between a predictor and the response.
+`ICEbox` does not actually construct PDPs, rather individual curves for each observation (which can be difficults to interpret when you have millions of observations).
 I think flexible graphics and non-extrapoloation zones are only half of the selling point here.
 I think that being able to run `partial` in parallel is huge, and also that this packge works with many types of machine learning models---not just tree-based models.
-For example, as described in the manuscript, `pdp` can produce single or multi-predictor PDPs for conditional random forests and XGBoost models (among many others), and in parallel.
+For example, as described in the manuscript, `pdp` can produce single or multi-predictor PDPs for conditional random forests and XGBoost models (among many others) in parallel.
 I expanded the last example to describe the interface with `caret`, which I think opens the door for wonderful exploratory data analysis pipe lines (e.g., choose model --> plot importance --> plot relationships, etc.).
-Most, if any, of these other software packages do not describe or support these important components (e.g., parallel execution), not even ICEbox.
+Also, a new section was added to the manuscript: **User-defined prediction functions**.
+This section describes the new `pred.fun` argument (as suggested by a later comment) that makes `partial` even more flexible.
 
@@
 Also, in response to another reviewer's comments, the code which produces figures containing multiple plots is better described and demonstrated.
-See, for example, the section called Multi-predictor PDPs.
+See, for example, the **Multi-predictor PDPs** section.
 
@@
 This is also motivation for why one of the examples showed how to use `ggplot2` for plotting the results from `partial`---`ggplot2` makes it a trivial matter to label the colorkey.
+I don't think it would be terrible to label all the axes.
+However, in my option, labelling all the axes would (1) looks cluttered, and (2) require more unnecessary code that would probably distract the reader.
+This is the standard axis labelling scheme used in trellis plots from S-plus to R (via `lattice`).
 
@@
 ### Response
-While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them.
+While I am not suggesting that the `partial` function needs to mimic the behavior of the PDP fucntions in `randomForest` and `gbm`, I do find it useful to be able to plot the results without forcing the user to store them first.
 However, I agree with your concern which is why `partial` has `plot = FALSE` as the default.
 Furthermore, I have made it possible to retrieve the raw PDP data even when \code{plot = TRUE}.
-This is because the "trellis" object returned by `partial` \ `plotPartial` now contains an additional `partial.data` attribute.
+This is because the `"trellis"` object returned by `partial` \ `plotPartial` now contains an additional `partial.data` attribute.
 This is described in the package documentation.
 For example,
-```{r, eval=TRUE}
+```{r, eval=TRUE, fig.width=5, fig.height=3}
 library(pdp)
@@
 ### Response
-This is a good suggestion.
+This is an excellent suggestion.
 First, I have changed the behavior of `...` to refer to additional arguments in `predict` (e.g., the `n.trees` argument in `predict.gbm`.).
@@
 For example, you can use this argument to construct PDPs on the probability scale, or to construct ICE curves.
-These are illusrated in the new section labeled: **User-defined prediction functions**
+This is illusrated in the new section labeled **User-defined prediction functions**
 

### 2017-01-04 manuscript/greenwell.tex
minor edits

@@
 \citet{harrison-1978-hedonic} were among the first to analyze the well-known Boston housing data.
-One of their goals was to find a housing value equation using data on median housing values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see \citet[Table IV]{harrison-1978-hedonic} for a description of each variable.
+One of their goals was to find a housing value equation using data on median home values from $n = 506$ census tracts in the suburbs of Boston from the 1970 census; see \citet[Table IV]{harrison-1978-hedonic} for a description of each variable.
 The data violate many classical assumptions like linearity, normality, and constant variance.
@@
 Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
-(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, these algorithms typically do not produce simple prediction formulas like in Equation~\eqref{eqn:boston}.
-Fortunately, these models can still provide insight into the data, but it is not in the form of simple equations.
-For example, many modern methods, like boosting and random forests, can assign importance scores to all of the predictors in the training data.
+(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, these algorithms typically do not produce simple prediction formulas like Equation~\eqref{eqn:boston}.
+These models can still provide insight into the data, but it is not in the form of simple equations.
+For example, quantifying predictor importance has become a crucial task in the analysis of "big data", and many supervised learning algorithms, like tree-based methods, can naturally assign variable importance scores to all of the predictors in the training data.
 
@@
 For example, the \CRANpkg{car} package \citep{fox-car-2011} contains many functions for constructing \dfn{partial-residual} and \dfn{marginal-model} plots.
-\dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provides tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots.
+\dfn{Effect displays}, available in the \CRANpkg{effects} package \citep{fox-effects-2003}, provide tabular and graphical displays for the terms in parametric models while holding all other predictors at some constant value---similar in spirit to \pkg{plotmo}'s marginal model plots.
 However, these methods were designed for simpler parametric models (e.g., linear and generalized linear models), whereas \pkg{plotmo}, \pkg{ICEbox}, and \pkg{pdp} are more useful for black box models (although, they can be used for simple parametric models as well).
@@
 
-# Display plots side by side (Figure 3)
+# Figure 3
 grid.arrange(pdp1, pdp2, pdp3, ncol = 3)
@@
 This is illustrated in the code chunk below which obtains the partial dependence of \code{ozone} on \code{wind}, \code{temp}, and \code{dpg} in parallel.
-The last three lines of code simply add a label to the colorkey.
+The last three lines of code add a label to the colorkey.
 The result is displayed in Figure~\ref{fig:partial_par}.
@@
 \address{Brandon M. Greenwell\\
-  Infoscitex Corporation\\
+  Infoscitex, a DCS Corporation\\
   4027 Colonel Glenn Highway\\

### 2017-01-04 R/partial.R

@@ -74,9 +74,8 @@
 #' avoided when the training data are passed to the \code{train} argument in the
 #' call to \code{partial}.
 #'
-#' It is possible to retrieve the most recent PDP constructed by \code{partial}
-#' after it is deleted (as long as no other \code{"trellis"} objects have
-#' been created since); simply use \code{trellis.last.object()}.
+#' It is possible to retrieve the last printed \code{"trellis"} object, such as
+#' those produced by \code{plotPartial}, using \code{trellis.last.object()}.
 #'
 #' It is possible for \code{partial} to run much faster if \code{object}
 #' inherits from class \code{"gbm"}. In particular, if \code{object} inherits

### 2017-01-04 manuscript/greenwell.tex
fixed bugs in pdClassification

@@
 In other situations, the user may need to specify this argument.
-This allows \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002} and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
+This allows \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002}.% and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
 
@@
       Cubist                    & \CRANpkg{Cubist} \citep{Cubist-pkg} & \code{"cubist"} \\
+      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
+      Linear model              & \pkg{stats}    & \code{"lm"} \\
+      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
+      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
+      Projection pursuit regression & \pkg{stats} & \code{"ppr"} \\
       Random forest             & \pkg{randomForest} & \code{"randomForest"} \\
@@
                                 & \CRANpkg{ranger} \citep{ranger-pkg} & \code{"ranger"} \\
-
-      Linear model              & \pkg{stats}    & \code{"lm"} \\
-      Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
-      Nonlinear least squares   & \pkg{stats}    & \code{"nls"} \\
-      Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
       Support vector machine    & \CRANpkg{e1071} \citep{e1071-pkg} & \code{"svm"} \\

### 2017-01-04 R/pdClassification.R

@@ -47,6 +47,14 @@ pdPredictClassification.default <- function(object, newdata, which.class, ...) {
 
 
 #' @keywords internal
+pdPredictClassification.BinaryTree <- function(object, newdata, which.class,
+                                               ...) {
+  pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
+  avgLogit(do.call(rbind, pr), which.class = which.class)
+}
+
+
+#' @keywords internal
 pdPredictClassification.bagging <- function(object, newdata, which.class, ...) {
   avgLogit(stats::predict(object, newdata = newdata, ...)$prob,
            which.class = which.class)
@@ -58,6 +66,13 @@ pdPredictClassification.boosting <- function(object, newdata, which.class,
                                              ...) {
   avgLogit(stats::predict(object, newdata = newdata, ...)$prob,
            which.class = which.class)
+}
+
+
+#' @keywords internal
+pdPredictClassification.earth <- function(object, newdata, which.class, ...) {
+  pr <- stats::predict(object, newdata = newdata, type = "response", ...)
+  avgLogit(cbind(pr, 1 - pr), which.class = which.class)
 }
 
 
@@ -94,12 +109,32 @@ pdPredictClassification.nnet <- function(object, newdata, which.class, ...) {
   } else {
     stats::predict(object, newdata = newdata, type = "raw", ...)
   }
-  avgLogit(pr, which.class = which.class)
+  # It seems that when the response has more than two levels, predict.nnet
+  # returns a matrix whose column names are the same as the factor levels. When
+  # the response is binary, a single-columned matrix with no column name is
+  # returned.
+  if (ncol(pr) == 1) {
+    avgLogit(cbind(pr, 1 - pr), which.class = which.class)
+  } else {
+    avgLogit(pr, which.class = which.class)
+  }
+}
+
+
+#' @keywords internal
+pdPredictClassification.RandomForest <- function(object, newdata, which.class,
+                                                 ...) {
+  pr <- stats::predict(object, newdata = newdata, type = "prob", ...)
+  avgLogit(do.call(rbind, pr), which.class = which.class)
 }
 
 
 #' @keywords internal
 pdPredictClassification.ranger <- function(object, newdata, which.class, ...) {
+  if (object$treetype != "Probability estimation") {
+    stop(paste("Cannot obtain predicted probabilities from",
+               deparse(substitute(object))))
+  }
   avgLogit(stats::predict(object, data = newdata, ...)$predictions,
            which.class = which.class)
 }

### 2017-01-04 R/utils.R

@@ -248,6 +248,18 @@ superType.boosting <- function(object) {
 
 
 #' @keywords internal
+superType.classbagg <- function(object) {
+  "classification"
+}
+
+
+#' @keywords internal
+superType.regbagg <- function(object) {
+  "regression"
+}
+
+
+#' @keywords internal
 superType.cubist <- function(object) {
   "regression"
 }
@@ -282,6 +294,15 @@ superType.lm <- function(object) {
 
 #' @keywords internal
 superType.nls <- function(object) {
+  "regression"
+}
+
+
+#' @keywords internal
+superType.ppr <- function(object) {
+  if (object$q > 1) {
+    stop("`partial` does not currently support multivariate response models.")
+  }
   "regression"
 }
 
@@ -346,7 +367,7 @@ superType.randomForest <- function(object) {
 superType.ranger <- function(object) {
   if (object$treetype == "Regression") {
     "regression"
-  } else if (object$treetype %in% c("Classification", "Probability estimation")) {
+  } else if (object$treetype == "Probability estimation") {
     "classification"
   } else {
     "other"
@@ -390,8 +411,8 @@ superType.xgb.Booster <- function(object) {
     "classification"
   } else if (object$params$objective %in%
              c("reg:logistic", "binary:logitraw", "multi:softmax")) {
-    stop(paste0("For classification, switch to an objective function",
-                "that returns the predicted probabilities."))
+    stop(paste("For classification, switch to an objective function",
+               "that returns the predicted probabilities."))
   } else {
     "other"
   }

### 2017-01-04 slowtests/slowtests-classification.R

@@ -11,82 +11,51 @@
 # update.packages()
 
 # Load required packages
+library(adabag)
+library(C50)
+library(earth)
+library(e1071)
+library(gbm)
+library(ipred)
+library(kernlab)
+library(nnet)
+library(party)
 library(pdp)
+library(randomForest)
+library(ranger)
+library(rpart)
+library(xgboost)
 
 # Load data
 data(pima)
+pima <- na.omit(pima)
 
 
 ################################################################################
-# "bagging" objects: adabag package
+# Fit models
 ################################################################################
 
-# Load required package
-library(adabag)
+# adabag::bagging
+set.seed(101)
+pima.bagging <- bagging(diabetes ~ ., data = pima)
 
-# Fit a glm to the Pima indians diabetes data
-set.seed(101)  # for reproducibility
-pima.bagging <- bagging(diabetes ~ ., data = pima, fmfinal = 500)
+# adabag::boosting
+set.seed(101)
+pima.boosting <- boosting(diabetes ~ ., data = pima)
 
-# Construct a PDP for glucose
-partial(pima.bagging, pred.var = "glucose", plot = TRUE, progress = "text",
-        rug = TRUE)
+# C50::C5.0
+set.seed(101)
+pima.C5.0 <- C5.0(diabetes ~ ., data = pima, trials = 100)
 
-# Construct a PDP for glucose (probability scale)
-partial(pima.bagging, pred.var = "glucose", plot = TRUE, progress = "text",
-        pred.fun = function(object, newdata) {
-          mean(predict(object, newdata)$prob[, 1L], na.rm = TRUE)
-        })
+# e1071::svm
+pima.svm <- svm(diabetes ~ ., data = pima, type = "C-classification",
+                probability = TRUE)
 
+# earth
+pima.earth <- earth(diabetes ~ ., data = pima, degree = 2,
+                    glm = list(family = binomial))
 
-################################################################################
-# "bagging" objects: adabag package
-################################################################################
-
-# Load required package
-library(adabag)
-
-# Fit a glm to the Pima indians diabetes data
-set.seed(101)  # for reproducibility
-pima.boosting <- boosting(diabetes ~ ., data = pima, fmfinal = 500)
-
-# Construct a PDP for glucose
[…]

### 2017-01-04 slowtests/slowtests-regression.R

@@ -15,6 +15,7 @@ library(Cubist)
 library(earth)
 library(e1071)
 library(gbm)
+library(ipred)
 library(kernlab)
 library(nnet)
 library(party)
@@ -71,6 +72,10 @@ boston.gbm <- gbm(cmedv ~ .,
                   shrinkage = 0.001,
                   verbose = TRUE)
 best.iter <- gbm.perf(boston.gbm, method = "OOB")
+
+# ipred::bagging
+set.seed(101)
+boston.ipred.bagging <- ipred::bagging(cmedv ~ ., data = boston, nbagg = 500)
 
 # kernlab::svm
 boston.ksvm <- ksvm(cmedv ~ ., data = boston, type = "eps-svr")
@@ -132,6 +137,9 @@ pdp.earth <- partial(boston.earth, pred.var = x)
 pdp.gbm <- partial(boston.gbm, pred.var = x)
 
 # kernlab::ksvm
+pdp.ipred.bagging <- partial(boston.ipred.bagging, pred.var = x)
+
+# kernlab::ksvm
 pdp.ksvm <- partial(boston.ksvm, pred.var = x, train = boston)
 
 # nnet
@@ -169,6 +177,7 @@ grid.arrange(
   plotPartial(pdp.crf, main = "party::cforest"),
   plotPartial(pdp.earth, main = "earth"),
   plotPartial(pdp.gbm, main = "gbm"),
+  plotPartial(pdp.ipred.bagging, main = "ipred::bagging"),
   plotPartial(pdp.rf, main = "randomForest"),
   plotPartial(pdp.ranger, main = "ranger"),
   plotPartial(pdp.rpart, main = "rpart"),

### 2017-01-04 manuscript/greenwell.tex
minor edits

@@
 Nowadays, many supervised learning algorithms can fit the data automatically in seconds---typically with higher accuracy.
-(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation; for example, these algorithms typically do not produce simple prediction formulas like Equation~\eqref{eqn:boston}.
+(We will revisit the Boston housing data in Section~\ref{sec:boston}.) The downfall, however, is some loss of interpretation since these algorithms typically do not produce simple prediction formulas like Equation~\eqref{eqn:boston}.
 These models can still provide insight into the data, but it is not in the form of simple equations.

### 2017-01-08 manuscript/greenwell.tex
minor edits

@@
 These models can still provide insight into the data, but it is not in the form of simple equations.
-For example, quantifying predictor importance has become a crucial task in the analysis of "big data", and many supervised learning algorithms, like tree-based methods, can naturally assign variable importance scores to all of the predictors in the training data.
+For example, quantifying predictor importance has become an essential task in the analysis of "big data", and many supervised learning algorithms, like tree-based methods, can naturally assign variable importance scores to all of the predictors in the training data.
 
@@
   \item can be used more efficiently with \code{"gbm"} objects (see Section~\ref{sec:computational});
-  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are far more flexible than base R graphics;
+  \item produces graphics based on \CRANpkg{lattice} \citep{lattice-pkg}, which are more flexible than base R graphics;
   \item defaults to using false color level plots for multivariate displays (see Section~\ref{sec:multi});
   \item contains options to mitigate the risks associated with extrapolation (see Section~\ref{sec:computational});
+  \item has the option to display progress bars (see Section~\ref{sec:computational});
   \item has the option to construct PDPs in parallel (see Section~\ref{sec:computational});
@@
 Consequently, the PDP for a predictor of interest can be obtained by averaging the corresponding ICE curves across all observations.
-In Section~\ref{sec:prediction}, it is shown how to obtain ICE curves using the \pkg{pdp} package (which is slightly faster than \pkg{ICEbox}).
+In Section~\ref{sec:prediction}, it is shown how to obtain ICE curves using the \pkg{pdp} package.
 \pkg{ICEbox} only allows for one variable at a time (i.e., no multivariate displays), though color can be used effectively to display information about an additional predictor.
@@
 Currently supported models are described in Table~\ref{tab:models}.
-In these cases, \code{partial} should be able to automatically determine an appropriate value for the argument \code{type} (i.e., \code{"regression"} or \code{"classification"}).
-In other situations, the user may need to specify this argument.
+In these cases, the user does not need to supply a prediction function (more on this in Section~\ref{sec:prediction}) or a value for the \code{type} argument (i.e., \code{"regression"} or \code{"classification"}).
+In other situations, the user may need to specify one or both of these arguments.
 This allows \code{partial} to be flexible enough to handle many of the model types not listed in Table~\ref{tab:models}; for example, neural networks from the \CRANpkg{nnet} package \citep{venables-modern-2002}.% and projection pursuit regression \citep{friedman-ppr-1981} using the \code{ppr} function in the \pkg{stats} package.
@@
                                 & \pkg{party}    & \code{"BinaryTree"} \\
-                                & \pkg{partykit} & \code{"constparty"}/\code{"party"} \\
+                                & \pkg{partykit} & \code{"party"} \\
                                 & \CRANpkg{rpart} \citep{rpart-pkg} & \code{"rpart"} \\
@@
       Cubist                    & \CRANpkg{Cubist} \citep{Cubist-pkg} & \code{"cubist"} \\
+      Discriminant analysis     & \CRANpkg{MASS} \citep{venables-modern-2002} & \code{"lda"}, \code{"qda"} \\
       Generalized linear model  & \pkg{stats}    & \code{"glm"}, \code{"lm"} \\
@@
       Multivariate adaptive regression splines (MARS) & \CRANpkg{earth} \citep{earth-pkg} & \code{"earth"} \\
+                                & \CRANpkg{mda} \citep{mda-pkg} & \code{"mars"} \\
       Projection pursuit regression & \pkg{stats} & \code{"ppr"} \\
@@
 
-For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}.
-These data contain the median value of owner-occupied homes in 506 U.S.
-census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town.
-These are the same data analyzed in \citet{harrison-1978-hedonic}, but with a few corrections and additional spatial variables.
-We begin by loading the data and omitting unimportant columns.
+For illustration, we will use a corrected version of the Boston housing data analyzed in \citet{harrison-1978-hedonic}; the data are available in the \pkg{pdp} package (see \code{?pdp::boston} for details).
+We begin by loading the data and fitting a random forest with default tuning parameters and 500 trees:
 \begin{example}
-data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
-boston <- BostonHousing2[, -c(1, 2, 5)]
-\end{example}
-Next, we fit a random forest to the entire data set with default tuning parameters and 500 trees:
-\begin{example}
+data(boston, package = "pdp")  # load the (corrected) Boston housing data
 library(randomForest)  # for randomForest, partialPlot, and varImpPlot functions
@@
 
-PDPs are essentially just averaged predictions; see, for example, step 1.
+PDPs are essentially just averaged predictions; this is apparent from step 1.
 (c) in Algorithm~\ref{alg:pdp}.
-Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong interactions can conceal the complexity of the modeled relationship between the response and predictors of interest.
+Consequently, as pointed out by \citet{goldstein-peeking-2015}, strong heterogeneity can conceal the complexity of the modeled relationship between the response and predictors of interest.
 This was part of the motivation behind \citeauthor*{goldstein-peeking-2015}'s ICE plot procedure.
@@
 Next, we simply pass this function via the \code{pred.fun} argument in the call to \code{partial}.
-The following chunk of code construct PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale.
+The following chunk of code uses \code{pred.prob} to obtain PDPs for \code{Petal.Width} and \code{Petal.Length} on the probability scale.
 The results are displayed in Figure~\ref{fig:partial_iris_prob}.
@@
 Suppose we would like ICE curves instead.
-To accomplish this we simply need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata}.
+To accomplish this we need to pass a prediction function that returns a vector of predictions, one for each observation in \code{newdata} (i.e., just remove the call to \code{mean} in \code{pred.fun}).
 The code snippet below illustrates this for the Boston housing example using the predictor \code{rm}.
@@
 rm.ice <- rm.ice %>%
-  group_by(yhat.id) %>%
-  mutate(yhat.centered = yhat - first(yhat))
+  group_by(yhat.id) %>%  # perform next operation within each yhat.id
+  mutate(yhat.centered = yhat - first(yhat))  # so each curve starts at yhat = 0
 \end{example}
@@
 Since the PDP is just the average of the corresponding ICE curves, it is quite simple to display both on the same plot.
-This can be accomplished by using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together.
+This is easily accomplished using the \code{stat\_summary} function from the \pkg{ggplot2} package to average the ICE curves together.
 The code snippet below plots the ICE curves and c-ICE curves, along with their averages, for the predictor \code{rm} in the Boston housing example.
@@
 However, this is not the case when using the \pkg{xgboost} package directly.
-To illustrate, we fit the same model using the \code{xgboost} function with the tuning parameters found previously using \pkg{caret}.
+To illustrate, we fit the same model using the \code{xgboost} function with the optimum tuning parameters found previously using \pkg{caret}.
+%stored in \code{boston.xgb\$bestTune}.%
 \begin{example}

### 2017-01-08 R/getTrainingData.R

@@ -44,6 +44,18 @@ getTrainingData.BinaryTree <- function(object) {
 
 
 #' @keywords internal
+getTrainingData.cforest <- function(object) {
+  stop(mssg)
+}
+
+
+#' @keywords internal
+getTrainingData.ctree <- function(object) {
+  stop(mssg)
+}
+
+
+#' @keywords internal
 getTrainingData.RandomForest <- function(object) {
   object@data@get("input")
 }

### 2017-01-08 R/partial.R

@@ -4,12 +4,12 @@
 #' model fitting objects.
 #'
 #' @param object A fitted model object of appropriate class (e.g.,
-#'   \code{"gbm"}, \code{"lm"}, \code{"randomForest"}, etc.).
+#'   \code{"gbm"}, \code{"lm"}, \code{"randomForest"}, \code{"train"}, etc.).
 #' @param pred.var Character string giving the names of the predictor
 #'   variables of interest. For reasons of computation/interpretation, this
 #'   should include no more than three variables.
-#' @param pred.grid Data frame containing the joint values of the variables
-#'   listed in \code{pred.var}.
+#' @param pred.grid Data frame containing the joint values of interest for the
+#'   variables listed in \code{pred.var}.
 #' @param pred.fun Optional prediction function that requires two arguments:
 #'   \code{object} and \code{newdata}. If specified, then the function must
 #'   return a single prediction or a vector of predictions (i.e., not a matrix
@@ -21,8 +21,8 @@
 #'   of the continuous independent variables listed in \code{pred.var}.
 #' @param type Character string specifying the type of supervised learning.
 #'   Current options are \code{"auto"}, \code{"regression"} or
-#'   \code{"classification"}. If \code{type = "auto"} then \code{partial} try
-#'   to extract the necessary information from \code{object}.
+#'   \code{"classification"}. If \code{type = "auto"} then \code{partial} will
+#'   try to extract the necessary information from \code{object}.
 #' @param which.class Integer specifying which column of the matrix of predicted
 #'   probabilities to use as the "focus" class. Default is to use the first
 #'   class. Only used for classification problems (i.e., when
@@ -85,7 +85,7 @@
 #' described in Friedman (2001).
 #'
 #' If the prediction function given to \code{pred.fun} returns a prediction for
-#' each observation in \code{newdata} then the result will be a PDP for each
+#' each observation in \code{newdata}, then the result will be a PDP for each
 #' observation. These are called individual conditional expectation (ICE)
 #' curves; see Goldstein et al. (2015) and \code{\link[ICEbox]{ice}} for
 #' details.
@@ -123,7 +123,7 @@
 #'
 #' # The partial function allows for multiple predictors
 #' partial(boston.rf, pred.var = c("lstat", "rm"), grid.resolution = 40,
[…]

### 2017-01-08 R/pdClassification.R

@@ -103,6 +103,13 @@ pdPredictClassification.ksvm <- function(object, newdata, which.class, ...) {
 
 
 #' @keywords internal
+pdPredictClassification.lda <- function(object, newdata, which.class, ...) {
+  pr <- stats::predict(object, newdata = newdata, ...)$posterior
+  avgLogit(cbind(pr, 1 - pr), which.class = which.class)
+}
+
+
+#' @keywords internal
 pdPredictClassification.nnet <- function(object, newdata, which.class, ...) {
   pr <- if (inherits(object, "multinom")) {
     stats::predict(object, newdata = newdata, type = "probs", ...)
@@ -118,6 +125,13 @@ pdPredictClassification.nnet <- function(object, newdata, which.class, ...) {
   } else {
     avgLogit(pr, which.class = which.class)
   }
+}
+
+
+#' @keywords internal
+pdPredictClassification.qda <- function(object, newdata, which.class, ...) {
+  pr <- stats::predict(object, newdata = newdata, ...)$posterior
+  avgLogit(cbind(pr, 1 - pr), which.class = which.class)
 }
 
 

### 2017-01-08 R/pdRegression.R

@@ -55,6 +55,14 @@ pdPredictRegression.ksvm <- function(object, newdata, ...) {
 
 
 #' @keywords internal
+pdPredictRegression.mars <- function(object, newdata, ...) {
+  mean(stats::predict(object,
+                      newdata = data.matrix(newdata), ...)[, 1L, drop = TRUE],
+       na.rm = TRUE)
+}
+
+
+#' @keywords internal
 pdPredictRegression.ranger <- function(object, newdata, ...) {
   mean(stats::predict(object, data = newdata, ...)$predictions, na.rm = TRUE)
 }

### 2017-01-08 R/superType.R

@@ -41,6 +41,19 @@ superType.C5.0 <- function(object) {
 }
 
 
+superType.cforest <- function(object) {
+  if (attr(object$terms, "response") == 1 &&
+      attr(object$terms, "dataClasses")[1L] == "numeric") {
+    "regression"
+  } else if (attr(object$terms, "response") == 1 &&
+             attr(object$terms, "dataClasses")[1L] == "factor") {
+    "classification"
+  } else {
+    "other"
+  }
+}
+
+
 #' @keywords internal
 superType.classbagg <- function(object) {
   "classification"
@@ -106,8 +119,23 @@ superType.ksvm <- function(object) {
 
 
 #' @keywords internal
+superType.lda<- function(object) {
+  "classification"
+}
+
+
+#' @keywords internal
 superType.lm <- function(object) {
   # FIXME: What about multivariate response models?
+  "regression"
+}
+
+
+#' @keywords internal
+superType.mars <- function(object) {
+  if (ncol(object$fitted.values) > 1) {
+    stop("`partial` does not currently support multivariate response models.")
+  }
   "regression"
 }
 
@@ -125,12 +153,32 @@ superType.nls <- function(object) {
 }
 
 
+superType.party <- function(object) {
+  if (attr(object$terms, "response") == 1 &&
+      attr(object$terms, "dataClasses")[1L] == "numeric") {
+    "regression"
+  } else if (attr(object$terms, "response") == 1 &&
+             attr(object$terms, "dataClasses")[1L] == "factor") {
+    "classification"
+  } else {
+    "other"
+  }
+}
+
+
+
 #' @keywords internal
 superType.ppr <- function(object) {
   if (object$q > 1) {
     stop("`partial` does not currently support multivariate response models.")
   }
   "regression"
+}
+
+
+#' @keywords internal
+superType.qda<- function(object) {
+  "classification"
 }
 
 

### 2017-01-08 manuscript/greenwell.R

@@ -163,9 +163,9 @@ cl <- makeCluster(4)  # use 4 cores
 registerDoParallel(cl)
 partial(ozone.mars, pred.var = c("wind", "temp", "dpg"), plot = TRUE,
         chull = TRUE, parallel = TRUE, paropts = list(.packages = "earth"))  # Figure 6
-trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
-grid.text("ozone", 0.2, 1.05, hjust = 0.5, vjust = 1)
-trellis.unfocus()
+lattice::trellis.focus("legend", side = "right", clipp.off = TRUE, highlight = FALSE)
+grid::grid.text("ozone", 0.2, 1.05, hjust = 0.5, vjust = 1)
+lattice::trellis.unfocus()
 stopCluster(cl)
 dev.off()
 

### 2017-01-09 manuscript/greenwell.tex
back to dev mode

@@
 
-The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
+The benefit of using \code{partial} is threefold: (1) it is a flexible, generic function that can be used to obtain different kinds of PDPs for various types of fitted models (not just random forests), (2) it will allow for any number of predictors to be used (e.g., multivariate displays), and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section.
 For example, the following code chunk uses the random forest model to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv}.
@@
 
-For classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp.
+Traditionally, for classification problems, partial dependence functions are on a scale similar to the logit; see, for example, \citet[pp.
 369---370]{hastie-elements-2009}.

### 2017-01-09 R/partial.R

@@ -74,6 +74,11 @@
 #' avoided when the training data are passed to the \code{train} argument in the
 #' call to \code{partial}.
 #'
+#' It is recommended to call \code{partial} with \code{plot = FALSE} and store
+#' the results; this allows for more flexible plotting, and the user will not
+#' have to waste time calling \code{partial} again if the default plot is not
+#' sufficient.
+#'
 #' It is possible to retrieve the last printed \code{"trellis"} object, such as
 #' those produced by \code{plotPartial}, using \code{trellis.last.object()}.
 #'

### 2017-01-09 manuscript/greenwell.R

@@ -1,51 +1,34 @@
 # R code for the manuscript
 #
-#   partial: partial: An R Package for Creating Partial Dependence Plots
+#   partial: An R Package for Creating Partial Dependence Plots
 #
 #        Author: Brandon M. Greenwell
-# Last Modified: September 19, 2016
+# Last Modified: January 8, 2017
 
 
 ################################################################################
 # Setup
 ################################################################################
 
-# Set working directory
-setwd("manuscript")
-
-# For now, please install the development version of pdp from GitHub. The next
-# release will be on CRAN soon. I am waiting out the grace period from my last
-# submission---the only change in this version is the name of an argument in
-# the two exported functions.
-# install.packages("devtools")
-# devtools::install_github("bgreenwell/pdp")
-
+# Set working directory (ignore for review)
+# setwd("manuscript")
 
 # List of packages required to run all the examples in this script
 # pkgs <- c("caret",
 #           "doParallel",
+#           "dplyr",
 #           "e1071",
 #           "earth",
 #           "ggplot2",
-#           "mlbench",
 #           "party",
 #           "pdp",
-#           "randomForest")
+#           "randomForest",
+#           "xgboost")
 # install.packages(pkgs)
 
-# Install latest version of xgboost
-# install.packages("drat", repos = "https://cran.rstudio.com")
-# drat:::addRepo("dmlc")
-# install.packages("data.table")  # required when installing xgboost from source!
-# install.packages("xgboost", repos = "http://dmlc.ml/drat/", type = "source")
-
-# Install the development version of pdp
-# install.packages("devtools")
-# devtools::install_github("bgreenwell/pdp")
-
 # Load required packages
-library(doParallel)
 library(caret)
+library(doParallel)
 # library(dplyr)  # use dplyr:: instead to avoid conflictions with plyr!
 library(e1071)
 library(earth)

### 2017-01-10 README.Rmd
updated README

@@
 # Load required packages
+library(ggmap)  # for plotting maps with ggplot2
 library(pdp)  # for constructing PDPs
@@
 
-# Partial dependence of lstat and rm on cmedv
+# Partial dependence of cmedv on lstat and rm
 grid.arrange(
@@
 
+# Look at partial dependence of median home value on location
+pd.loc <- partial(boston.rf, pred.var = c("lon", "lat"), chull = TRUE)
+
+# Overlay predictions on a map of Boston
+ll <- c(range(boston$lon), range(boston$lat))
+map <- get_map(location = ll[c(1, 3, 2, 4)], zoom = 11, maptype = "toner-lite")
+ggmap(map) +
+  geom_point(aes(x = lon, y = lat), data = boston, alpha = 0.2) +
+  geom_tile(aes(x = lon, y = lat, fill = yhat),
+            data = pd.loc, alpha = 0.3) +
+  scale_fill_distiller(palette = "Spectral", name = "Median\nvalue") +
+  coord_fixed(ratio = 1) +
+  labs(x = "Longitude", y = "Latitude")
 ```
@@
 
-Finally, we demonstrate the construction of PDPs from models fit using the `caret` package; `caret` is an extremetly useful package for classification and regression training that, essentially, has one function (`train`) for fitting all kinds of predictive models in R (e.g., `glmnet`, `svm`, `xgboost`, etc.).
+Finally, we demonstrate the construction of PDPs from models fit using the `caret` package; `caret` is an extremetly useful package for training all sorts of predictive models in R (e.g., `glmnet`, `svm`, `xgboost`, etc.).
 
@@
 
-# Set up for 5-fold cross-validation
-ctrl <- trainControl(method = "cv", number = 5)
-
-# Grid of tuning parameter values
-xgb.grid <- expand.grid(
-  nrounds = c(1000, 2000),
-  max_depth = 2:4,
-  eta = c(0.001, 0.01, 0.1),
-  gamma = 0, 
-  colsample_bytree = 1,
-  min_child_weight = 1,
-  subsample = c(0.5, 0.75, 1)
-)
-
-# Tune am XGBoost model to the Pima Indians diabetes data. This may take a few 
+# Tune an XGBoost model to the Pima Indians diabetes data. This may take a few 
 # minutes!
 set.seed(103)  # for reproducibility
-pima.xgb <- train(diabetes ~ ., data = pima, method = "xgbTree",
-                  prob.model = TRUE, na.action = na.omit, trControl = ctrl,
-                  tuneGrid = xgb.grid)
+pima.xgb <- train(diabetes ~ ., data = pima, 
+                  method = "xgbTree",
+                  na.action = na.omit,
+                  trControl = trainControl(method = "cv", number = 5),
+                  tuneLength = 5)
 
@@
 grid.arrange(
-  partial(pima.xgb, pred.var = "glucose", plot = TRUE, rug = TRUE),
-  partial(pima.xgb, pred.var = "age", plot = TRUE, rug = TRUE),
-  partial(pima.xgb, pred.var = "mass", plot = TRUE, rug = TRUE),
+  partial(pima.xgb, pred.var = "glucose", plot = TRUE, 
+          rug = TRUE, smooth = TRUE),
+  partial(pima.xgb, pred.var = "age", plot = TRUE, 
+          rug = TRUE, smooth = TRUE),
+  partial(pima.xgb, pred.var = "mass", plot = TRUE, 
+          rug = TRUE, smooth = TRUE),
   ncol = 3 

### 2017-03-05 README.Rmd
updated autoplot function

@@
 
-#### Classification example
+#### Using `autoplot` to produce `ggplot2` graphics
 
@@
 # Partial dependence of glucose and age on diabetes test result (neg/pos).
+pd.glucose <- partial(pima.svm, pred.var = "glucose", prob = TRUE, train = pima)
+pd.age <- partial(pima.svm, pred.var = "age", prob = TRUE, train = pima)
+pd.glucose.age <- partial(pima.svm, pred.var = c("glucose", "age"), prob = TRUE, 
+                          train = pima)
+
+# Dislay all three plots together
 grid.arrange(
-  partial(pima.svm, pred.var = "glucose", plot = TRUE, train = pima),
-  partial(pima.svm, pred.var = "age", plot = TRUE, train = pima),
-  partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, train = pima),
+  autoplot(pd.glucose),
+  autoplot(pd.age),
+  autoplot(pd.glucose.age, contour = TRUE),
   ncol = 3
@@
 set.seed(103)  # for reproducibility
-pima.xgb <- train(diabetes ~ ., data = pima, 
-                  method = "xgbTree",
+pima.xgb <- train(diabetes ~ ., data = pima, method = "xgbTree",
                   na.action = na.omit,

### 2017-03-05 R/autoplot.R

@@ -8,6 +8,12 @@
 #'   curves (c-ICE curves). Only useful when \code{object} represents a set of
 #'   ICE curves; see \code{\link[pdp]{partial}} for details. Default is
 #'   \code{TRUE}.
+#' @param plot.pdp Logical indicating whether or not to plot the partial
+#'   dependence function on top of the ICE curves. Default is \code{TRUE}.
+#' @param pdp.color Character string specifying the color to use for the partial
+#'   dependence function when \code{plot.pdp = TRUE}. Default is \code{"red"}.
+#' @param pdp.size The line width, a positive number, to use for the partial
+#'   dependence function when \code{plot.pdp = TRUE}. Default is \code{1}.
 #' @param smooth Logical indicating whether or not to overlay a LOESS smooth.
 #'   Default is \code{FALSE}.
 #' @param smooth.method Character string specifying the smoothing method
@@ -37,10 +43,17 @@
 #'   Default is \code{"yhat"}.
 #' @param ... Additional optional arguments to be passed onto \code{geom_line}.
 #'
+#' @return A \code{"ggplot"} object.
+#'
+#' @importFrom ggplot2 aes autoplot geom_contour geom_line geom_point geom_rug
+#' @importFrom ggplot2 geom_smooth geom_tile scale_fill_distiller stat_summary
+#' @importFrom ggplot2 theme_bw xlab ylab
+#'
 #' @rdname autoplot
 #' @export
-autoplot.partial <- function(object, center = TRUE, rug = FALSE, smooth = FALSE,
-                             smooth.method = "auto",
+autoplot.partial <- function(object, center = TRUE, plot.pdp = TRUE,
+                             pdp.color = "red", pdp.size = 1, rug = FALSE,
+                             smooth = FALSE, smooth.method = "auto",
                              smooth.formula = y ~ x, smooth.span = 0.75,
                              smooth.method.args = list(), contour = FALSE,
                              contour.color = "white", palette = "Spectral",
@@ -62,49 +75,53 @@ autoplot.partial <- function(object, center = TRUE, rug = FALSE, smooth = FALSE,
     ncol(object) - 1  # don't count yhat
   }
 
+  # Multiple curves
   if (multi) {
-    if (center) {
-      object <- object %>%
-        dplyr::group_by(yhat.id) %>%
-        dplyr::mutate(yhat = yhat - first(yhat))
-    }
[…]

### 2017-03-05 R/partial.R

@@ -154,7 +154,7 @@
 #'             colorkey = FALSE, screen = list(z = -20, x = -60))
 #'
 #' # The autplot function can be used to produce graphics based on ggplot2
-#' autoplot(pd, contour = TRUE, contour.color = "white",
+#' autoplot(pd, contour = TRUE, contour = TRUE,
 #'          legend.title = "Partial\ndependence")
 #'
 #' #

### 2017-03-06 README.Rmd
updated README and switched to non-standard evaluation in autoplot

--- 2016-12-28 README.Rmd
@@
 
-pdp: An R Package for Constructing Partial Dependence Plots
-================
+# pdp: An R Package for Constructing Partial Dependence Plots
 
-The primary purpose of this package is to provide a general framework for constructing _partial dependence plots_ (PDPs) in R.
+Complex nonparametric models---like neural networks, random forests, and support vector machines---are more common than ever in predictive analytics, especially when dealing with large observational databases that don't adhere to the strict assumptions imposed by traditional statistical techniques (e.g., multiple linear regression which assumes linearity, homoscedasticity, and normality).
+Unfortunately, it can be challenging to understand the results of such models and explain them to management.
+Partial dependence plots offer a simple solution.
+Partial dependence plots are low-dimensional graphical renderings of the prediction function $\widehat{f}\left(\boldsymbol{x}\right)$ so that the relationship between the outcome and predictors of interest can be more easily understood.
+These plots are especially useful in explaining the output from black box models.
+The `pdp` package offers a general framework for constructing partial dependence plots for various types of fitted models in R.
 
-### Installation
+A detailed introduction to `pdp` has been accepted for publication in The R Journal; a preprint is available [here](https://github.com/bgreenwell/pdp-paper/blob/master/RJwrapper.pdf).
+You can track development on  at https://github.com/bgreenwell/pdp.
+To report bugs or issues, contact the main author directly or submit them to https://github.com/bgreenwell/pdp/issues.
 
-The R package `pdp` is available from [CRAN](https://cran.r-project.org/package=pdp); the development version is hosted on [GitHub](https://github.com/bgreenwell/pdp). There are two ways to install:
-```{r, eval=FALSE}
-# Install latest release from CRAN (recommended)
+As of right now, `pdp` exports four functions:
+
+* `partial` - compute partial dependence functions (i.e., objects of class `"partial"`) from various fitted model objects;
+* `plotPartial"` - plot partial dependence functions (i.e., objects of class `"partial"`) using `lattice` graphics;
+* `autoplot` - plot partial dependence functions (i.e., objects of class `"partial"`) using `ggplot2` graphics;
+* `topPredictors` extract most "important" predictors from various types of fitted models.
+
+## Installation
+
+The `pdp` package is [currently listed on CRAN](https://CRAN.R-project.org/package=pdp) and can easily be installed:
+```r
+  # Install from CRAN (recommended)
   install.packages("pdp")
 
-# Install development version from GitHub repo
+  # Alternatively, install the development version from GitHub
   devtools::install_github("bgreenwell/pdp")
 ```
-
-### Example usage
-
-The examples below demonstrate various usages of the `pdp` package: regression, classification, and interfacing with the well-known `caret` package.
-To start, we need to install a few additional packages that will be required to run the examples.
-```{r, eval=FALSE}
-install.packages(c("caret", "ggmap", "kernlab", "randomForest"))
-```
-
-#### Regression example
-
-In this example, we fit a random forest to the Boston housing data.
-(See `?boston` for a brief explanation of the data.) Note, for any of the following examples, you can see a progress bar by simply specifying `progress = "text"` in the call to `partial`.
-You may also reduce the computation time via the `grid.resolution` option in the call to `partial`.
-```{r, message = FALSE, warning=FALSE, fig.width=12, fig.height=4}
-# Load required packages
-library(pdp)  # for constructing PDPs
-library(randomForest)  # for random forest algorithm
-
-# Load the Boston housing data
-data (boston)  # included with the pdp package
-
-# Fit a random forest to the boston housing data
-set.seed(101)  # for reproducibility
-boston.rf <- randomForest(cmedv ~ ., data = boston)
-
-# Partial dependence of lstat and rm on cmedv
-grid.arrange(
-  partial(boston.rf, pred.var = "lstat", plot = TRUE, rug = TRUE),
-  partial(boston.rf, pred.var = "rm", plot = TRUE, rug = TRUE),
-  partial(boston.rf, pred.var = c("lstat", "rm"), plot = TRUE, chull = TRUE),
-  ncol = 3
-)
-
-```
-
-#### Classification example
-
-In this example, we fit a support vector machine with a radial basis function kernel to the Pima Indians diabetes data. (See `?pima` for a brief explanation of the data.)
-```{r, message=FALSE, warning=FALSE, fig.width=12, fig.height=4}
-# Load required packages
-library(kernlab)  # for fitting SVMs
-
-# Fit an SVM to the Pima Indians diabetes data
-data (pima)  # load the boston housing data
-pima.svm <- ksvm(diabetes ~ ., data = pima, type = "C-svc", kernel = "rbfdot",
-                 C = 0.5, prob.model = TRUE)
-
-# Partial dependence of glucose and age on diabetes test result (neg/pos).
-grid.arrange(
-  partial(pima.svm, pred.var = "glucose", plot = TRUE, train = pima),
-  partial(pima.svm, pred.var = "age", plot = TRUE, train = pima),
-  partial(pima.svm, pred.var = c("glucose", "age"), plot = TRUE, train = pima),
-  ncol = 3
-)
-```
-
-#### Interface with `caret`
-
-Finally, we demonstrate the construction of PDPs from models fit using the `caret` package; `caret` is an extremetly useful package for classification and regression training that, essentially, has one function (`train`) for fitting all kinds of predictive models in R (e.g., `glmnet`, `svm`, `xgboost`, etc.).
-
-For illustration we use `caret`'s `train` function to tune an [XGBoost](https://github.com/dmlc/xgboost) model to the Pima Indians diabetes data using 5-fold cross-validation.
-We then use the final model to construct PDPs for `glucose` and `age`.
-Note, when training a model using `caret`'s `train` function, you can view tuning progress by setting `verboseIter = TRUE` in the call to `trainControl`.
-```{r, message=FALSE, warning=FALSE, fig.width=15, fig.height=5}
-# Load required packages
-library(caret)  # for model training/tuning
-
-# Set up for 5-fold cross-validation
-ctrl <- trainControl(method = "cv", number = 5)
-
-# Grid of tuning parameter values
-xgb.grid <- expand.grid(
-  nrounds = c(1000, 2000),
-  max_depth = 2:4,
-  eta = c(0.001, 0.01, 0.1),
-  gamma = 0, 
-  colsample_bytree = 1,
-  min_child_weight = 1,
-  subsample = c(0.5, 0.75, 1)
-)
-
-# Tune am XGBoost model to the Pima Indians diabetes data. This may take a few 
-# minutes!
-set.seed(103)  # for reproducibility
-pima.xgb <- train(diabetes ~ ., data = pima, method = "xgbTree",
-                  prob.model = TRUE, na.action = na.omit, trControl = ctrl,
-                  tuneGrid = xgb.grid)
-
-# Partial dependence of glucose and age on diabetes test result (neg/pos)
-grid.arrange(
-  partial(pima.xgb, pred.var = "glucose", plot = TRUE, rug = TRUE),
-  partial(pima.xgb, pred.var = "age", plot = TRUE, rug = TRUE),
-  partial(pima.xgb, pred.var = "mass", plot = TRUE, rug = TRUE),
-  ncol = 3 
-)
-```

### 2017-03-06 R/autoplot.R

@@ -14,6 +14,8 @@
 #'   dependence function when \code{plot.pdp = TRUE}. Default is \code{"red"}.
 #' @param pdp.size The line width, a positive number, to use for the partial
 #'   dependence function when \code{plot.pdp = TRUE}. Default is \code{1}.
+#' @param rug Logical indicating whether or not to include rug marks on the
+#'   predictor axes. Default is \code{FALSE}.
 #' @param smooth Logical indicating whether or not to overlay a LOESS smooth.
 #'   Default is \code{FALSE}.
 #' @param smooth.method Character string specifying the smoothing method
@@ -45,9 +47,9 @@
 #'
 #' @return A \code{"ggplot"} object.
 #'
-#' @importFrom ggplot2 aes autoplot geom_contour geom_line geom_point geom_rug
-#' @importFrom ggplot2 geom_smooth geom_tile scale_fill_distiller stat_summary
-#' @importFrom ggplot2 theme_bw xlab ylab
+#' @importFrom ggplot2 aes_string autoplot facet_wrap geom_contour geom_line
+#' @importFrom ggplot2 geom_point geom_rug geom_smooth geom_tile ggplot
+#' @importFrom ggplot2 scale_fill_distiller stat_summary theme_bw xlab ylab
 #'
 #' @rdname autoplot
 #' @export
@@ -122,29 +124,30 @@ autoplot.partial <- function(object, center = TRUE, plot.pdp = TRUE,
                           legend.title = legend.title, ...)
     }
 
-    p <- ggplot(object, aes(x = object[[1L]], y = object[[2L]], z = yhat,
-                            fill = yhat)) +
-      ggplot2::geom_tile()
+    p <- ggplot(object, aes_string(x = names(object)[[1L]],
+                                   y = names(object)[[2L]], z = "yhat",
+                                   fill = "yhat")) +
+      geom_tile()
     if (contour) {
-      p <- p + ggplot2::geom_contour(color = contour.color)
+      p <- p + geom_contour(color = contour.color)
     }
     if (is.null(xlab)) {
-      p <- p + ggplot2::xlab(names(object)[1L])
+      p <- p + xlab(names(object)[1L])
     } else {
-      p <- p + ggplot2::xlab(xlab)
+      p <- p + xlab(xlab)
     }
     if (is.null(ylab)) {
-      p <- p + ggplot2::ylab(names(object)[2L])
[…]

### 2017-03-06 R/partial.R

@@ -173,7 +173,7 @@
 #'
 #' # Post-process rm.ice to obtain c-ICE curves
 #' library(dplyr)  # for group_by and mutate functions
-#' rm.ice <- rm.ice %>%
+#' rm.cice <- rm.ice %>%
 #'   group_by(yhat.id) %>%  # perform next operation within each yhat.id
 #'   mutate(yhat.centered = yhat - first(yhat))  # so each curve starts at yhat = 0
 #'
@@ -183,7 +183,7 @@
 #'   geom_line(aes(group = yhat.id), alpha = 0.2) +
 #'   stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 #' # c-ICE curves with their average
-#' p2 <- ggplot(rm.ice, aes(rm, yhat.centered)) +
+#' p2 <- ggplot(rm.cice, aes(rm, yhat.centered)) +
 #'   geom_line(aes(group = yhat.id), alpha = 0.2) +
 #'   stat_summary(fun.y = mean, geom = "line", col = "red", size = 1)
 #' grid.arrange(p1, p2, ncol = 2)

### 2017-03-06 R/plotPartial.R

@@ -7,7 +7,7 @@
 #' @param smooth Logical indicating whether or not to overlay a LOESS smooth.
 #'   Default is \code{FALSE}.
 #' @param rug Logical indicating whether or not to include rug marks on the
-#'   predictor axes. Only used when \code{plot = TRUE}. Default is \code{FALSE}.
+#'   predictor axes. Default is \code{FALSE}.
 #' @param chull Logical indicating wether or not to restrict the first
 #'   two variables in \code{pred.var} to lie within the convex hull of their
 #'   training values; this affects \code{pred.grid}. Default is \code{FALSE}.

### 2017-03-06 tests/testthat/test-partial.R

@@ -34,18 +34,12 @@ test_that("partial works correctly for regression", {
   pd4 <- partial(df.reg.lm, pred.var = "rm", quantiles = TRUE, probs = 1:9/10,
                  train = df.reg)
 
-  # Trim outliers first
-  pd5 <- partial(df.reg.lm, pred.var = "rm",
-                 pred.grid = df.reg[,"rm", drop = FALSE], trim.outliers = TRUE,
-                 check.class = FALSE, train = df.reg)
-
   # Expectations
   expect_is(pd1, "data.frame")
   expect_is(pd2, "data.frame")
   expect_is(pd3, "trellis")
   expect_is(pd4, "data.frame")
   expect_identical(pd4$rm, as.numeric(quantile(df.reg$rm, probs = 1:9/10)))
-  expect_identical(pd5$rm, pdp:::trimOutliers(df.reg$rm))
 
   # Trellis object should still contain partial dependence data
   expect_is(attr(pd3, "partial.data"), "data.frame")

### 2017-03-06 README.Rmd
updated README

@@
 ```
+
+## Random forest example
+
+As a quick example, we'll fit a random forest to the famous Boston housing data included with the package.
+In fact the original motivation for this package was to be able to compute two-predictor partial dependence plots from random forest models in R.
+
+```{r, fig.width=8, fig.height=4, message=FALSE}
+# Fit a random forest to the boston housing data
+library(randomForest)  # install.packages("randomForest")
+data (boston)  # load the boston housing data
+set.seed(101)  # for reproducibility
+boston.rf <- randomForest(cmedv ~ ., data = boston)
+
+# Partial dependence of cmedv on lstat and rm
+library(pdp)
+pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE)
+head(pd)  # print first 6 rows
+
+# Lattice version
+p1 <- plotPartial(pd)
+
+# ggplot2 version
+library(ggplot2)
+p2 <- autoplot(pd, contour = TRUE, legend.title = "Partial\ndependence")
+
+# Show both plots in one figure
+grid.arrange(p1, p2, ncol = 2)
+```

### 2017-03-06 R/partial.R

@@ -154,6 +154,7 @@
 #'             colorkey = FALSE, screen = list(z = -20, x = -60))
 #'
 #' # The autplot function can be used to produce graphics based on ggplot2
+#' library(ggplot2)
 #' autoplot(pd, contour = TRUE, contour = TRUE,
 #'          legend.title = "Partial\ndependence")
 #'

### 2017-03-06 README.Rmd
updated README

--- 2017-03-06 README.Rmd
@@
 ```
+
+## Random forest example
+
+As a first example, we'll fit a random forest to the famous Boston housing data included with the package (see `?boston` for details).
+In fact the original motivation for this package was to be able to compute two-predictor partial dependence plots from random forest models in R.
+```{r, fig.width=8, fig.height=4, message=FALSE}
+# Fit a random forest to the Boston housing data
+library(randomForest)  # install.packages("randomForest")
+data (boston)  # load the boston housing data
+set.seed(101)  # for reproducibility
+boston.rf <- randomForest(cmedv ~ ., data = boston)
+
+# Partial dependence of cmedv on lstat and rm
+library(pdp)
+pd <- partial(boston.rf, pred.var = c("lstat", "rm"), chull = TRUE)
+head(pd)  # print first 6 rows
+
+# Lattice version
+p1 <- plotPartial(pd, main = "lattice version")
+
+# ggplot2 version
+library(ggplot2)
+p2 <- autoplot(pd, contour = TRUE, title = "ggplot2 version", 
+               legend.title = "Partial\ndependence")
+
+# Show both plots in one figure
+grid.arrange(p1, p2, ncol = 2)
+```
+
+Next, we'll fit a classification model to the Pima Indians Diabetes data.
+
+## Support vector machine (SVM) example
+
+As a second example, we'll fit an SVM to the Pima Indians diabetes data included with the package (see `?pima` for details).
+Note that for some fitted model objects (e.g., `"ksvm"` objects) it is necessary to supply the original training data via the `train` argument in the call to `partial`.
+
+```{r, fig.width=8, fig.height=4, message=FALSE}
+# Fit an SVM to the Pima Indians diabetes data
+library(kernlab)  # install.packages("kernlab")
+data (pima)  # load the Pima Indians diabetes data
+pima.svm <- ksvm(diabetes ~ ., data = pima, type = "C-svc", kernel = "rbfdot",
+                 C = 0.5, prob.model = TRUE)
+
+# Partial dependence of diabetes test result on glucose (default is logit scale)
+pd.glucose <- partial(pima.svm, pred.var = "glucose", train = pima)
+
+# Partial dependence of diabetes test result on glucose (probability scale)
+pd.glucose.prob <- partial(pima.svm, pred.var = "glucose", prob = TRUE, 
+                           train = pima)
+
+# Show both plots in one figure
+grid.arrange(autoplot(pd.glucose, title = "Logit scale"), 
+             autoplot(pd.glucose.prob, title = "Probability scale"), 
+             ncol = 2)
+```

### 2017-03-06 R/autoplot.R

@@ -48,7 +48,7 @@
 #' @return A \code{"ggplot"} object.
 #'
 #' @importFrom ggplot2 aes_string autoplot facet_wrap geom_contour geom_line
-#' @importFrom ggplot2 geom_point geom_rug geom_smooth geom_tile ggplot
+#' @importFrom ggplot2 geom_point geom_rug geom_smooth geom_tile ggplot ggtitle
 #' @importFrom ggplot2 scale_fill_distiller stat_summary theme_bw xlab ylab
 #'
 #' @rdname autoplot
@@ -60,8 +60,8 @@ autoplot.partial <- function(object, center = TRUE, plot.pdp = TRUE,
                              smooth.method.args = list(), contour = FALSE,
                              contour.color = "white", palette = "Spectral",
                              train = NULL,
-                             xlab = NULL, ylab = NULL, legend.title = NULL,
-                             ...) {
+                             xlab = NULL, ylab = NULL, title = NULL,
+                             legend.title = NULL, ...) {
 
   # Determine of x contains multiple PDPs
   multi <- if ("yhat.id" %in% names(object)) {
@@ -155,7 +155,11 @@ autoplot.partial <- function(object, center = TRUE, plot.pdp = TRUE,
   }
 
   # Return ggplot object
-  p
+  if (is.null(title)) {
+    p
+  } else {
+    p + ggtitle(title)
+  }
 
 }
 

### 2017-03-06 README.Rmd
updated README and docs

@@
 library(ggplot2)
-p2 <- autoplot(pd, contour = TRUE, title = "ggplot2 version", 
+p2 <- autoplot(pd, contour = TRUE, main = "ggplot2 version", 
                legend.title = "Partial\ndependence")
@@
 # Show both plots in one figure
-grid.arrange(autoplot(pd.glucose, title = "Logit scale"), 
-             autoplot(pd.glucose.prob, title = "Probability scale"), 
+grid.arrange(autoplot(pd.glucose, main = "Logit scale"), 
+             autoplot(pd.glucose.prob, main = "Probability scale"), 
              ncol = 2)

### 2017-03-06 R/autoplot.R

@@ -41,6 +41,8 @@
 #'   required if \code{rug = TRUE} or \code{chull = TRUE}.
 #' @param xlab Charater string specifying the text for the x-axis label.
 #' @param ylab Charater string specifying the text for the y-axis label.
+#' @param main Character string specifying the text for the main title of the
+#'   plot.
 #' @param legend.title Charater string specifying the text for the legend title.
 #'   Default is \code{"yhat"}.
 #' @param ... Additional optional arguments to be passed onto \code{geom_line}.
@@ -60,7 +62,7 @@ autoplot.partial <- function(object, center = TRUE, plot.pdp = TRUE,
                              smooth.method.args = list(), contour = FALSE,
                              contour.color = "white", palette = "Spectral",
                              train = NULL,
-                             xlab = NULL, ylab = NULL, title = NULL,
+                             xlab = NULL, ylab = NULL, main = NULL,
                              legend.title = NULL, ...) {
 
   # Determine of x contains multiple PDPs
@@ -155,10 +157,10 @@ autoplot.partial <- function(object, center = TRUE, plot.pdp = TRUE,
   }
 
   # Return ggplot object
-  if (is.null(title)) {
+  if (is.null(main)) {
     p
   } else {
-    p + ggtitle(title)
+    p + ggtitle(main)
   }
 
 }

### 2017-03-12 README.Rmd
fixed pandoc conversion issues

@@
 output:
-  html_document: 
-    keep_md: yes
+  md_document:
+    variant: markdown_github
 ---
+
+<!-- README.md is generated from README.Rmd. Please edit that file -->
+
+```{r, echo = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  comment = "#>",
+  fig.path = "README-"
+)
+```
+
 [![CRAN\_Status\_Badge](http://www.r-pkg.org/badges/version/pdp)](https://cran.r-project.org/package=pdp)

### 2017-03-12 slowtests/slowtests-calhousing.R

@@ -1,0 +1,59 @@
+# Load required packages
+library(gbm)
+library(pdp)
+
+# Download function from gist
+url <- "b1330460eec5acf1c81fae71902e331c"
+devtools::source_gist(url, filename = "fetchCaliforniaHousingData.R")
+
+# Fetch California housing data
+cal <- fetchCaliforniaHousingData()
+head(cal)
+
+# Run GBM
+set.seed(102)
+cal.gbm <- gbm(AvgValue ~ ., data = cal,
+               distribution = "laplace",
+               n.trees = 2000,
+               interaction.depth = 6,
+               shrinkage = 0.1,
+               bag.fraction = 1,
+               train.fraction = 0.8,
+               verbose = TRUE)
+best.iter <- gbm.perf(cal.gbm, method = "test")
+
+# Partial dependence of AvgValue on MedInc, AveOccup, HouseAge, and AveRooms
+pd.MedInc <- partial(cal.gbm, pred.var = "MedInc", quantiles = TRUE,
+                     probs = 5:95/100, n.trees = best.iter)
+pd.AveOccup <- partial(cal.gbm, pred.var = "AveOccup", quantiles = TRUE,
+                       probs = 5:95/100, n.trees = best.iter)
+pd.HouseAge <- partial(cal.gbm, pred.var = "HouseAge", quantiles = TRUE,
+                       probs = 5:95/100, n.trees = best.iter)
+pd.AveRooms <- partial(cal.gbm, pred.var = "AveRooms", quantiles = TRUE,
+                       probs = 5:95/100, n.trees = best.iter)
+
+# Partial dependence of AvgValue on AveOccup and HouseAge (together)
+pd.HouseAge.AveOccup <- partial(
+  cal.gbm, pred.var = c("HouseAge", "AveOccup"), quantiles = TRUE,
+  probs = 5:95/100, n.trees = best.iter
+)
+
+# PDPs
+ylim <- c(1, 3.5)
+ylab <- "Partial dependence"
+grid.arrange(
+  plotPartial(pd.MedInc, rug = TRUE, train = cal, ylim = ylim, ylab = ylab),
+  plotPartial(pd.AveOccup, rug = TRUE, train = cal, ylim = ylim, ylab = ylab),
+  plotPartial(pd.HouseAge, rug = TRUE, train = cal, ylim = ylim, ylab = ylab),
[…]

### 2017-03-12 README.Rmd
added subdirectorytools

@@
   comment = "#>",
-  fig.path = "README-"
+  fig.path = "tools/README-"
 )
@@
 In fact the original motivation for this package was to be able to compute two-predictor partial dependence plots from random forest models in R.
-```{r, fig.width=8, fig.height=4, message=FALSE}
+```{r example-rf, fig.width=8, fig.height=4, message=FALSE}
 # Fit a random forest to the Boston housing data
@@
 
-```{r, fig.width=8, fig.height=4, message=FALSE}
+```{r example-svm, fig.width=8, fig.height=4, message=FALSE}
 # Fit an SVM to the Pima Indians diabetes data

### 2017-04-01 R/isNonGaussianRegression.R
properly registered native routines

@@ -1,0 +1,36 @@
+#' @keywords internal
+isNonGaussianRegression <- function(object) {
+  UseMethod("isNonGaussianRegression")
+}
+
+
+#' @keywords internal
+isNonGaussianRegression.default <- function(object) {
+  FALSE
+}
+
+
+#' @keywords internal
+isNonGaussianRegression.earth <- function(object) {
+  !is.null(object$glm.list)
+}
+
+
+#' @keywords internal
+isNonGaussianRegression.glm <- function(object) {
+  TRUE
+}
+
+
+#' @keywords internal
+isNonGaussianRegression.gbm <- function(object) {
+  object$distribution %in%
+    c("gamma", "poisson", "tweedie")
+}
+
+
+#' @keywords internal
+isNonGaussianRegression.xgb.Booster <- function(object) {
+  object$params$objective %in%
+    c("count:poisson", "reg:gamma")
+}

### 2017-04-01 R/partial.R

@@ -34,6 +34,9 @@
 #'   Current options are \code{"auto"}, \code{"regression"} or
 #'   \code{"classification"}. If \code{type = "auto"} then \code{partial} will
 #'   try to extract the necessary information from \code{object}.
+#' @param inv.link Function specifying the transfrmation to be applied to
+#'   the predictions before they are averaged (experimental). Default is
+#'   \code{identity}.
 #' @param which.class Integer specifying which column of the matrix of predicted
 #'   probabilities to use as the "focus" class. Default is to use the first
 #'   class. Only used for classification problems (i.e., when
@@ -224,6 +227,7 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
                             quantiles = FALSE, probs = 1:9/10,
                             trim.outliers = FALSE,
                             type = c("auto", "regression", "classification"),
+                            inv.link = identity,
                             which.class = 1L, prob = FALSE, recursive = TRUE,
                             plot = FALSE, smooth = FALSE, rug = FALSE,
                             chull = FALSE, train, cats = NULL,
@@ -290,8 +294,6 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
     }
   }
 
-
-
   # Make sure each column has the correct class, levels, etc.
   if (inherits(train, "data.frame") && check.class) {
     pred.grid <- copyClasses(pred.grid, train)
@@ -346,6 +348,10 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
     # Use brute force approach to partial dependence
     pd.df <- if (!is.null(pred.fun)) {
 
+      if (!is.null(inv.link)) {
+        warning("`inv.link` ignored whenever `recursive = TRUE`")
+      }
+
       # User-supplied prediction function
       pdManual(object, pred.var = pred.var, pred.grid = pred.grid,
                pred.fun = pred.fun, train = train, progress = progress,
@@ -353,10 +359,20 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
 
     } else if (type == "regression") {
 
-      # Traditional partial dependence based on averaged predictions
-      pdRegression(object, pred.var = pred.var, pred.grid = pred.grid,
-                   pred.fun = pred.fun, train = train, progress = progress,
-                   parallel = parallel, paropts = paropts, ...)
+      if (!is.null(inv.link) && isNonGaussianRegression(object)) {
+        # Traditional partial dependence based on averaged prediction; the
[…]

### 2017-04-01 R/superType.R

@@ -72,7 +72,8 @@ superType.earth <- function(object) {
       object$glm.list[[1L]]$family$family == "binomial") {
     "classification"
   } else if (is.null(object$glm.list) ||
-             object$glm.list[[1L]]$family$family == "gaussian") {
+             object$glm.list[[1L]]$family$family %in%
+             c("gaussian", "Gamma", "inverse.gaussian", "poisson")) {
     "regression"
   } else {
     "other"
@@ -82,7 +83,8 @@ superType.earth <- function(object) {
 
 #' @keywords internal
 superType.gbm <- function(object) {
-  if (object$distribution %in% c("gaussian", "laplace", "tdist")) {
+  if (object$distribution %in%
+      c("gaussian", "laplace", "tdist", "gamma", "poisson", "tweedie")) {
     "regression"
   } else if (object$distribution %in%
              c("bernoulli", "huberized", "multinomial", "adaboost")) {
@@ -260,7 +262,8 @@ superType.train <- function(object) {
 
 #' @keywords internal
 superType.xgb.Booster <- function(object) {
-  if (object$params$objective == "reg:linear") {
+  if (object$params$objective %in%
+      c("reg:linear", "count:poisson", "reg:gamma")) {
     "regression"
   } else if (object$params$objective %in%
              c("binary:logistic", "multi:softprob")) {

### 2017-04-01 slowtests/slowtests-regression-inverse-link.R

@@ -1,0 +1,50 @@
+# Setup
+data(mtcars)
+library(gbm)
+library(ggplot2)
+library(pdp)
+library(xgboost)
+set.seed(101)
+bst <- xgboost(data = as.matrix(mtcars[, -11]), label = mtcars[, 11],
+               objective = "count:poisson", nrounds = 50)
+
+# PDP prediction function for XGBoost with Poisson deviance
+pfun <- function(object, newdata) {
+  mean(exp(predict(object, newdata = as.matrix(newdata))))
+}
+
+# One variable
+bst %>%
+  partial(pred.var = "mpg", pred.fun = pfun, train = mtcars[, -11]) %>%
+  autoplot() +
+  ylab("Number of carburetors") +
+  theme_light()
+
+# One variable
+bst %>%
+  partial(pred.var = "mpg", inv.link = exp, train = mtcars[, -11]) %>%
+  autoplot() +
+  ylab("Number of carburetors") +
+  theme_light()
+
+bst %>%
+  partial(pred.var = "mpg", train = mtcars[, -11]) %>%
+  autoplot() +
+  ylab("Number of carburetors") +
+  theme_light()
+
+set.seed(101)
+mtcars.gbm <- gbm(carb ~ ., data = mtcars, distribution = "poisson",
+                  n.trees = 2000, interaction.depth = 3, shrinkage = 0.01,
+                  bag.fraction = 1.0, cv.folds = 5)
+best.iter <- gbm.perf(mtcars.gbm, method = "cv")
+summary(mtcars.gbm, n.trees = best.iter, las = 1)
+
+# Log scale
+plot(mtcars.gbm, i.var = "mpg")
+partial(mtcars.gbm, pred.var = "mpg", n.trees = best.iter, plot = TRUE)
+
+# Response scale
+plot(mtcars.gbm, i.var = "mpg", type = "response")
+partial(mtcars.gbm, pred.var = "mpg", recursive = FALSE, inv.link = exp,
+        n.trees = best.iter, plot = TRUE)

### 2017-04-09 README.Rmd
properly registered native routines
fixed URL

@@
 
-A detailed introduction to `pdp` has been accepted for publication in The R Journal; a preprint is available [here](https://github.com/bgreenwell/pdp-paper/blob/master/RJwrapper.pdf).
+A detailed introduction to `pdp` has been accepted for publication in The R Journal; a preprint is available [here](https://journal.r-project.org/archive/2017/RJ-2017-016/index.html).
 You can track development on  at https://github.com/bgreenwell/pdp.

### 2017-04-09 slowtests/slowtests-regression-inverse-link.R

@@ -22,7 +22,8 @@ bst %>%
 
 # One variable
 bst %>%
-  partial(pred.var = "mpg", inv.link = exp, train = mtcars[, -11]) %>%
+  partial(pred.var = "mpg", inv.link = exp, train = mtcars[, -11],
+          recursive = FALSE) %>%
   autoplot() +
   ylab("Number of carburetors") +
   theme_light()

### 2017-04-09 slowtests/slowtests.R

@@ -1,144 +1,0 @@
-#-------------------------------------------------------------------------------
-# Slow tests for the pdp package
-#-------------------------------------------------------------------------------
-
-
-################################################################################
-# Setup
-################################################################################
-
-# Make sure all packges are up to date
-# update.packages()
-
-# Load required packages
-library(pdp)
-
-# Load data
-data(boston)  # regression
-data(pima)  # classification
-
-
-################################################################################
-# gbm package
-################################################################################
-
-# Load required package
-library(gbm)
-
-
-# Regression -------------------------------------------------------------------
-
-# Fit a gbm to the Boston housing data
-set.seed(101)
-boston.gbm <- gbm(cmedv ~ .,
-                  data = boston,
-                  distribution = "gaussian",
-                  n.trees = 10000,
-                  cv.folds = 10,
-                  interaction.depth = 3,
-                  shrinkage = 0.01,
-                  verbose = TRUE)
-best.iter <- gbm.perf(boston.gbm, method = "cv")
-
-# Compare plots for `lstat` using recursive method (default for GBM objects)
-par(mfrow = c(1, 2))
-plot(partial(boston.gbm, pred.var = "lstat", n.trees = best.iter,
-             grid.resolution = 51), type = "l", main = "pdp::partial")
-plot(boston.gbm, i.var = "lstat", n.trees = best.iter,
-     continuous.resolution = 51, main = "gbm::plot.gbm")
-
-# Use brute force method
-partial(boston.gbm, pred.var = "lstat", n.trees = best.iter,
-        grid.resolution = 51, recursive = FALSE, plot = TRUE, progress = "text")
-partial(boston.gbm, pred.var = "lstat", n.trees = 1,
-        grid.resolution = 51, recursive = FALSE, plot = TRUE, progress = "text")
-
-
-# Classification ---------------------------------------------------------------
-
-# Fit a gbm to the Pima indians diabetes data
-set.seed(101)
-pima.gbm <- gbm(unclass(diabetes) - 1 ~ .,
-                data = pima,
-                distribution = "bernoulli",
-                n.trees = 5000,
-                interaction.depth = 3,
-                shrinkage = 0.001,
-                verbose = TRUE)
-best.iter <- gbm.perf(pima.gbm, method = "OOB")
-summary(pima.gbm, n.trees = best.iter, las = 1)
-
[…]

### 2017-07-09 README.Rmd
update CITATION file and links

@@
 
-A detailed introduction to `pdp` has been accepted for publication in The R Journal; a preprint is available [here](https://journal.r-project.org/archive/2017/RJ-2017-016/index.html).
+A detailed introduction to `pdp` has been published in The R Journal: "pdp: An R Package for Constructing Partial Dependence Plots", https://journal.r-project.org/archive/2017/RJ-2017-016/index.html.
 You can track development on  at https://github.com/bgreenwell/pdp.

### 2017-07-09 README.Rmd
added AppVeyor support

@@
 [![Build Status](https://travis-ci.org/bgreenwell/pdp.svg?branch=master)](https://travis-ci.org/bgreenwell/pdp)
+[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/bgreenwell/pdp?branch=master&svg=true)](https://ci.appveyor.com/project/bgreenwell/pdp)
 [![Coverage Status](https://img.shields.io/codecov/c/github/bgreenwell/pdp.svg)](https://codecov.io/github/bgreenwell/pdp?branch=master)
@@
 [![Total Downloads](http://cranlogs.r-pkg.org/badges/grand-total/pdp)](http://cranlogs.r-pkg.org/badges/grand-total/pdp)
-[![Rdoc](http://www.rdocumentation.org/badges/version/pdp)](http://www.rdocumentation.org/packages/pdp)
 

### 2017-07-09 README.Rmd
fixed typo
added logo
resize logo

@@
 
-# pdp: An R Package for Constructing Partial Dependence Plots
+# pdp: An R Package for Constructing Partial Dependence Plots <img src="tools/pdp-logo.png" align="right" width="120" height="139" />
 
@@
 A detailed introduction to `pdp` has been published in The R Journal: "pdp: An R Package for Constructing Partial Dependence Plots", https://journal.r-project.org/archive/2017/RJ-2017-016/index.html.
-You can track development on  at https://github.com/bgreenwell/pdp.
+You can track development at https://github.com/bgreenwell/pdp.
 To report bugs or issues, contact the main author directly or submit them to https://github.com/bgreenwell/pdp/issues.

### 2017-07-17 slowtests/slowtests-boston.Rmd
switching to knitr

---
title: "Bouston Housing Tests"
author: "Brandon M. Greenwell"
date: "July 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
# Load required packages
library(ggplot2)
library(grid)
library(pdp)

# Set knitr chunk options
knitr::opts_chunk$set(echo = TRUE)

# Function to run tests for a fitted model object using the Boston housing data
bostonHousingTest <- function(object, ...) {

  # Training data
  trn <- subset(boston, select = -cmedv)

  # Partial dependence plot for lstat
  pd1 <- partial(object, pred.var = "lstat", train = trn, ...)

  # Partial dependence plot for lstat and rm
  pd2 <- partial(object, pred.var = c("lstat", "rm"), chull = TRUE,
                 train = trn, ...)

  # Individual conditional expectation curves for age
  pd3 <- partial(object, pred.var = "age", ice = TRUE, train = trn, ...)

  # Centered individual conditional expectation curves for lstat
  pd4 <- partial(object, pred.var = "age", ice = TRUE, center = TRUE,
                 train = trn, ...)

  # Display plots
  grid.arrange(
    autoplot(pd1, rug = TRUE, train = trn, main = "PDP for lstat"),
    autoplot(pd2, main = "PDP for lstat and rm"),
    autoplot(pd3, main = "ICE curves for age", alpha = 0.25),
    autoplot(pd4, main = "c-ICE curves for age", alpha = 0.25),
    ncol = 2, 
    top = textGrob(deparse(substitute(object)),
                   gp = gpar(fontsize = 20, font = 3))
  )

}
```

## Cubist

# `Cubist` package

```{r}
boston.cubist <- Cubist::cubist(x = subset(boston, select = -cmedv),
                                y = boston$cmedv, committees = 100)
bostonHousingTest(boston.cubist)
```

## Decision trees

# `rpart` package

```{r}
boston.rpart <- rpart::rpart(cmedv ~ ., data = boston)
bostonHousingTest(boston.rpart)
```

# `party` package

```{r}
boston.ctree <- party::ctree(cmedv ~ ., data = boston)
bostonHousingTest(boston.ctree)
```

# `partykit` package

```{r}
boston.ctree2 <- partykit::ctree(cmedv ~ ., data = boston)
bostonHousingTest(boston.ctree2)
```

## Bagging

# `ipred` package

```{r}
set.seed(101)
boston.ipred <- ipred::bagging(cmedv ~ ., data = boston, nbagg = 500)
bostonHousingTest(boston.ipred)
```

## Random forests

# `randomForest` package

```{r}
set.seed(101)
boston.rf <- randomForest::randomForest(cmedv ~ ., data = boston)
bostonHousingTest(boston.rf)
```

# `party` package

```{r}
set.seed(101)
boston.crf <- party::cforest(cmedv ~ ., data = boston)
bostonHousingTest(boston.crf, quantiles = TRUE)
```

# `ranger` package

```{r}
set.seed(101)
boston.ranger <- ranger::ranger(cmedv ~ ., data = boston)
bostonHousingTest(boston.ranger)
```

## Boosting

# `gbm` package

```{r}
set.seed(101)
boston.gbm <- gbm::gbm(cmedv ~ ., data = boston, distribution = "gaussian",
                       n.trees = 5000, interaction.depth = 3, shrinkage = 0.001,
                       cv.folds = 5)
best.iter <- gbm::gbm.perf(boston.gbm, method = "OOB", plot.it = FALSE)
bostonHousingTest(boston.gbm, n.trees = best.iter)
```

# `xgboost` package
```{r}
set.seed(101)
boston.xgb <- xgboost::xgboost(
  data = data.matrix(subset(boston, select = -cmedv)), label = boston$cmedv,
  objective = "reg:linear", nrounds = 5000, max_depth = 3, eta = 0.001,
  colsample_bytree = 0.8, subsample = 0.5,
  save_period = NULL, verbose = 0
)
bostonHousingTest(boston.xgb)
```

## Neural networks

# `nnet` package
```{r}
set.seed(101)
boston.nnet <- nnet::nnet(cmedv ~ ., data = boston, size = 6, decay = 0.1,
                          linout = TRUE, trace = FALSE)
bostonHousingTest(boston.nnet)
```

## Support vector machines

# `e1071` package

```{r}
boston.svm <- e1071::svm(cmedv ~ ., data = boston, type = "eps-regression")
bostonHousingTest(boston.svm)
```

# `kernlab` package

```{r}
boston.ksvm <- kernlab::ksvm(cmedv ~ ., data = boston, type= "eps-svr")
bostonHousingTest(boston.ksvm)
```

## Linear/generalized linear models

# `stats` package

```{r}
boston.lm <- lm(cmedv ~ . ^ 2, data = boston)
bostonHousingTest(boston.lm)
```

```{r}
boston.glm <- glm(cmedv ~ . ^ 2, data = boston)
bostonHousingTest(boston.glm)
```

## Multivariate adaptive regression splines

# `earth` package

```{r}
boston.earth <- earth::earth(cmedv ~ ., degree = 3, data = boston)
bostonHousingTest(boston.earth)
```

# `mda` package

```{r}
boston.mars <- mda::mars(x = subset(boston, select = - cmedv), 
                         y = boston$cmedv, degree = 3)
bostonHousingTest(boston.mars)
```

### 2017-07-17 slowtests/slowtests-iris.Rmd
switching to knitr

---
title: "Bouston  Tests"
author: "Brandon M. Greenwell"
date: "July 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
# Load required packages
library(ggplot2)
library(grid)
library(pdp)

# Set knitr chunk options
knitr::opts_chunk$set(echo = TRUE)

# Function to run tests for a fitted model object using the iris data
irisTest <- function(object, ...) {

  # Training data
  trn <- subset(iris, select = -Species)

  # Partial dependence plot for Petal.Length on class 1 (i.e. setosa)
  pd1 <- partial(object, pred.var = "Petal.Length", which.class = 1, 
                 prob = TRUE, train = trn, ...)

  # Partial dependence plot for Petal.Length on class 2 (i.e. versicolor)
  pd2 <- partial(object, pred.var = "Petal.Length", which.class = 2, 
                 prob = TRUE, train = trn, ...)

  # Partial dependence plot for Petal.Length on class 3 (i.e. verginica)
  pd3 <- partial(object, pred.var = "Petal.Length", which.class = 3, 
                 prob = TRUE, train = trn, ...)

  # Bind together into a single data frame
  pd <- rbind(cbind(pd1, "Class" = "1"),
              cbind(pd2, "Class" = "2"),
              cbind(pd3, "Class" = "3"))

  # Display all three plots
  ggplot(pd, aes(x = Petal.Length, y = yhat, color = Class)) +
    geom_line() +
    xlab("Petal length") +
    ylab("Probability") +
    ggtitle(deparse(substitute(object))) +
    theme_light()

}
```

# Discriminant analysis

## Package: `MASS`

```{r}
iris.lda <- MASS::lda(Species ~ ., data = iris)
irisTest(iris.lda)
```

```{r}
iris.qda <- MASS::qda(Species ~ ., data = iris)
irisTest(iris.qda)
```

## Package: `mda`

# Decision trees

## Package: `rpart`

```{r}
iris.rpart <- rpart::rpart(Species ~ ., data = iris)
irisTest(iris.rpart)
```

## Package: `C50`
```{r}
set.seed(101)
iris.C5.0 <- C50::C5.0(Species ~ ., data = iris, trials = 100)
irisTest(iris.C5.0)
```

## Package: `party

```{r}
iris.ctree <- party::ctree(Species ~ ., data = iris)
irisTest(iris.ctree)
```

## Package: `partykit

```{r}
iris.ctree2 <- partykit::ctree(Species ~ ., data = iris)
irisTest(iris.ctree2)
```

# Bagging

## Package: `adabag`

```{r}
set.seed(101)
iris.bagging <- adabag::bagging(Species ~ ., data = iris)
irisTest(iris.bagging)
```

## Package: `ipred

```{r}
set.seed(101)
iris.ipred <- ipred::bagging(Species ~ ., data = iris, nbagg = 500)
irisTest(iris.ipred)
```

# Random forests

## Package: `randomForest

```{r}
set.seed(101)
iris.rf <- randomForest::randomForest(Species ~ ., data = iris)
irisTest(iris.rf)
```

## Package: `party

```{r}
set.seed(101)
iris.crf <- party::cforest(Species ~ ., data = iris)
irisTest(iris.crf, quantiles = TRUE)
```

## Package: `partykit`
```{r}
set.seed(101)
iris.crf2 <- partykit::cforest(Species ~ ., data = iris)
irisTest(iris.crf2, quantiles = TRUE)
```

## Package: `ranger`

```{r}
set.seed(101)
iris.ranger <- ranger::ranger(Species ~ ., data = iris, probability = TRUE)
irisTest(iris.ranger)
```

# Boosting

## Package: `adabag`

```{r}
set.seed(101)
iris.boosting <- adabag::boosting(Species ~ ., data = iris)
irisTest(iris.boosting)
```

## Package: `gbm`

```{r}
set.seed(101)
iris.gbm <- gbm::gbm(Species ~ ., 
                     data = iris,
                     distribution = "multinomial",
                     n.trees = 5000,
                     interaction.depth = 3,
                     shrinkage = 0.001,
                     cv.folds = 5,
                     verbose = FALSE)
best.iter <- gbm::gbm.perf(iris.gbm, method = "cv", plot.it = FALSE)
irisTest(iris.gbm, n.trees = best.iter)
```

## Package: `xgboost`
```{r}
set.seed(101)
iris.xgb <- xgboost::xgboost(
  data = data.matrix(subset(iris, select = -Species)), 
  label = unclass(iris$Species) - 1, num_class = 3, 
  objective = "multi:softprob", nrounds = 100, max_depth = 3, eta = 0.1, 
  gamma = 0, colsample_bytree = 0.8, min_child_weight = 1, subsample = 0.7,
  verbose = 0
)
irisTest(iris.xgb)
```

# Neural networks

## Package: `nnet`
```{r}
set.seed(101)
iris.nnet <- nnet::nnet(Species ~ ., data = iris, size = 10, decay = 0.1, 
                        maxit = 500, trace = FALSE)
irisTest(iris.nnet)
```

# Support vector machines

## Package: `e1071`

```{r}
iris.svm <- e1071::svm(Species ~ ., data = iris, type = "C-classification", 
                       probability = TRUE)
irisTest(iris.svm)
```

## Package: `kernlab`

```{r}
iris.ksvm <- kernlab::ksvm(Species ~ ., data = iris, type = "C-svc", 
                           prob.model = TRUE)
irisTest(iris.ksvm)
```

# Linear/generalized linear models

## Package: `nnet`

```{r}
set.seed(101)
iris.multinom <- nnet::multinom(Species ~ ., data = iris, trace = FALSE)
irisTest(iris.multinom)
```

### 2017-07-18 slowtests/slowtests-pima.Rmd
slowtests (mostly) updated

---
title: "Bouston  Tests"
author: "Brandon M. Greenwell"
date: "July 15, 2017"
output: html_document
---

```{r setup, include=FALSE}
# Load required packages
library(ggplot2)
library(grid)
library(pdp)

# Set knitr chunk options
knitr::opts_chunk$set(echo = TRUE)

# Function to run tests for a fitted model object using the Pima Indians 
# diabetes data
pimaIndiansTest <- function(object, ...) {

  # Training data
  pima <- na.omit(pima)
  trn <- subset(pima, select = -diabetes)

  # Partial dependence plot for lstat
  pd1 <- partial(object, pred.var = "glucose", train = trn, ...)

  # Partial dependence plot for lstat and rm
  pd2 <- partial(object, pred.var = c("glucose", "age"), chull = TRUE,
                 train = trn, ...)

  # Individual conditional expectation curves for age
  pd3 <- partial(object, pred.var = "glucose", ice = TRUE, train = trn,
                 ...)

  # Centered individual conditional expectation curves for lstat
  pd4 <- partial(object, pred.var = "glucose", ice = TRUE, center = TRUE,
                 train = trn, ...)

  # Display plots
  grid.arrange(
    autoplot(pd1, rug = TRUE, train = pima, main = "PDP for lstat"),
    autoplot(pd2, main = "PDP for glucose and age"),
    autoplot(pd3, rug = TRUE, train = pima,
             main = "ICE curves for glucose", alpha = 0.25),
    autoplot(pd4, rug = TRUE, train = pima,
             main = "c-ICE curves for glucose", alpha = 0.25),
    ncol = 2,
    top = textGrob(deparse(substitute(object)),
                   gp = gpar(fontsize = 20, font = 3))
  )

}
```

# Discriminant analysis

## Package: `MASS`

```{r}
pima.lda <- MASS::lda(diabetes ~ . ^ 2, data = na.omit(pima))
pimaIndiansTest(pima.lda)
pimaIndiansTest(pima.lda, prob = TRUE)
```

```{r}
pima.qda <- MASS::qda(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.qda)
pimaIndiansTest(pima.qda, prob = TRUE)
```

## Package: `mda`

```{r}
pima.mda <- mda::mda(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.mda)
pimaIndiansTest(pima.mda, prob = TRUE)
```

```{r}
pima.fda <- mda::fda(diabetes ~ ., data = na.omit(pima), method = mda::mars,
                     degree = 2)
pimaIndiansTest(pima.fda)
pimaIndiansTest(pima.fda, prob = TRUE)
```

# Decision trees

## Package: `rpart`

```{r}
pima.rpart <- rpart::rpart(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.rpart)
pimaIndiansTest(pima.rpart, prob = TRUE)
```

## Package: `C50`
```{r}
set.seed(101)
pima.C5.0 <- C50::C5.0(diabetes ~ ., data = na.omit(pima), trials = 100)
pimaIndiansTest(pima.C5.0)
pimaIndiansTest(pima.C5.0, prob = TRUE)
```

## Package: `party

```{r}
pima.ctree <- party::ctree(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.ctree)
pimaIndiansTest(pima.ctree, prob = TRUE)
```

## Package: `partykit

```{r}
pima.ctree2 <- partykit::ctree(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.ctree2)
pimaIndiansTest(pima.ctree2, prob = TRUE)
```

# Bagging

## Package: `adabag`

```{r}
set.seed(101)
pima.bagging <- adabag::bagging(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.bagging, quantiles = TRUE)
pimaIndiansTest(pima.bagging, quantiles = TRUE, prob = TRUE)
```

## Package: `ipred

```{r}
set.seed(101)
pima.ipred <- ipred::bagging(diabetes ~ ., data = na.omit(pima), nbagg = 500)
pimaIndiansTest(pima.ipred, quantiles = TRUE)
pimaIndiansTest(pima.ipred, quantiles = TRUE, prob = TRUE)
```

# Random forests

## Package: `randomForest

```{r}
set.seed(101)
pima.rf <- randomForest::randomForest(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.rf)
pimaIndiansTest(pima.rf, prob = TRUE)
```

## Package: `party

```{r}
set.seed(101)
pima.crf <- party::cforest(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.crf, quantiles = TRUE)
pimaIndiansTest(pima.crf, quantiles = TRUE, prob = TRUE)
```

## Package: `partykit`
```{r}
set.seed(101)
pima.crf2 <- partykit::cforest(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.crf2, quantiles = TRUE)
pimaIndiansTest(pima.crf2, quantiles = TRUE, prob = TRUE)
```

## Package: `ranger`

```{r}
set.seed(101)
pima.ranger <- ranger::ranger(diabetes ~ ., data = na.omit(pima), 
                              probability = TRUE)
pimaIndiansTest(pima.ranger)
pimaIndiansTest(pima.ranger, prob = TRUE)
```

# Boosting

## Package: `adabag`

```{r}
set.seed(101)
pima.boosting <- adabag::boosting(diabetes ~ ., data = na.omit(pima))
pimaIndiansTest(pima.boosting, quantiles = TRUE)
pimaIndiansTest(pima.boosting, quantiles = TRUE, prob = TRUE)
```

## Package: `gbm`

```{r}
set.seed(101)
pima.gbm <- gbm::gbm(diabetes ~ ., 
                     data = na.omit(pima),
                     distribution = "multinomial",
                     n.trees = 5000,
                     interaction.depth = 3,
                     shrinkage = 0.001,
                     cv.folds = 5,
                     verbose = FALSE)
best.iter <- gbm::gbm.perf(pima.gbm, method = "cv", plot.it = FALSE)
pimaIndiansTest(pima.gbm, n.trees = best.iter)
pimaIndiansTest(pima.gbm, n.trees = best.iter, prob = TRUE)
```

## Package: `xgboost`
```{r}
set.seed(101)
pima.xgb <- xgboost::xgboost(
  data = data.matrix(subset(pima, select = -diabetes)),
  label = unclass(pima$diabetes) - 1, objective = "binary:logistic",
  nrounds = 100, max_depth = 3, eta = 0.1, gamma = 0, colsample_bytree = 0.8, 
  min_child_weight = 1, subsample = 0.7, verbose = 0
)
pimaIndiansTest(pima.xgb)
pimaIndiansTest(pima.xgb, prob = TRUE)
```

# Neural networks

## Package: `nnet`
```{r}
set.seed(101)
pima.nnet <- nnet::nnet(diabetes ~ ., data = na.omit(pima), size = 10, 
                        decay = 0.1, maxit = 500, trace = FALSE)
pimaIndiansTest(pima.nnet)
pimaIndiansTest(pima.nnet, prob = TRUE)
```

# Support vector machines

## Package: `e1071`

```{r}
pima.svm <- e1071::svm(diabetes ~ ., data = na.omit(pima), type = "C-classification", 
                       probability = TRUE)
pimaIndiansTest(pima.svm)
pimaIndiansTest(pima.svm, prob = TRUE)
```

## Package: `kernlab`

```{r}
pima.ksvm <- kernlab::ksvm(diabetes ~ ., data = na.omit(pima), type = "C-svc", 
                           prob.model = TRUE)
pimaIndiansTest(pima.ksvm)
pimaIndiansTest(pima.ksvm, prob = TRUE)
```

# Multinomial models

## Package: `nnet`

```{r}
set.seed(101)
pima.multinom <- nnet::multinom(diabetes ~ . ^ 2, data = na.omit(pima), 
                                trace = FALSE)
pimaIndiansTest(pima.multinom)
pimaIndiansTest(pima.multinom, prob = TRUE)
```

### 2017-07-18 R/getParPredCls.R

@@ -192,8 +192,9 @@ getParClsLogit.nnet <- function(object, newdata, which.class, ...) {
   # It seems that when the response has more than two levels, predict.nnet
   # returns a matrix whose column names are the same as the factor levels. When
   # the response is binary, a single-columned matrix with no column name is
-  # returned.
-  if (ncol(pr) == 1) {
+  # returned. For multinomial models, a vector is returned when the response has
+  # only two classes.
+  if (is.null(ncol(pr)) || ncol(pr) == 1) {
     mean(multiClassLogit(cbind(pr, 1 - pr), which.class = which.class),
          na.rm = TRUE)
   } else {
@@ -212,8 +213,9 @@ getParClsProb.nnet <- function(object, newdata, which.class, ...) {
   # It seems that when the response has more than two levels, predict.nnet
   # returns a matrix whose column names are the same as the factor levels. When
   # the response is binary, a single-columned matrix with no column name is
-  # returned.
-  if (ncol(pr) == 1) {
+  # returned. For multinomial models, a vector is returned when the response has
+  # only two classes.
+  if (is.null(ncol(pr)) || ncol(pr) == 1) {
     mean(cbind(pr, 1 - pr)[, which.class], na.rm = TRUE)
   } else {
     mean(pr[, which.class], na.rm = TRUE)

### 2017-07-18 slowtests/run-slowtests.R

@@ -1,115 +1,0 @@
-# Function to run tests for a fitted model object using the Boston housing data
-bostonHousingTest <- function(object) {
-
-  # Partial dependence plot for lstat
-  pd1 <- partial(object, pred.var = "lstat", train = boston)
-
-  # Partial dependence plot for lstat and rm
-  pd2 <- partial(object, pred.var = c("lstat", "rm"), chull = TRUE,
-                 train = boston)
-
-  # Individual conditional expectation curves for age
-  pd3 <- partial(object, pred.var = "age", ice = TRUE, train = boston)
-
-  # Centered individual conditional expectation curves for lstat
-  pd4 <- partial(object, pred.var = "age", ice = TRUE, center = TRUE,
-                 train = boston)
-
-  # Display plots
-  grid.arrange(
-    plotPartial(pd1, rug = TRUE, train = boston, main = "PDP for lstat"),
-    plotPartial(pd2, main = "PDP for lstat and rm"),
-    plotPartial(pd3, main = "ICE curves for age", alpha = 0.5),
-    plotPartial(pd4, main = "c-ICE curves for age", alpha = 0.5),
-    ncol = 2,
-    top = textGrob(deparse(substitute(object)),
-                   gp = gpar(fontsize = 20, font = 3))
-  )
-
-}
-
-
-# Function to run tests for a fitted model object using the Boston housing data
-pimaIndiansTest <- function(object) {
-
-  # Partial dependence plot for lstat
-  pd1 <- partial(object, pred.var = "glucose", train = pima)
-
-  # Partial dependence plot for lstat and rm
-  pd2 <- partial(object, pred.var = c("glucose", "age"), chull = TRUE,
-                 train = pima)
-
-  # Partial dependence plot for lstat
-  pd3 <- partial(object, pred.var = "glucose", prob = TRUE, train = pima)
-
-  # Partial dependence plot for lstat and rm
-  pd4 <- partial(object, pred.var = c("glucose", "age"), prob = TRUE,
-                 chull = TRUE, train = pima)
-
-  # Individual conditional expectation curves for age
-  pd5 <- partial(object, pred.var = "glucose", ice = TRUE, train = pima)
-
-  # Centered individual conditional expectation curves for lstat
-  pd6 <- partial(object, pred.var = "glucose", ice = TRUE, center = TRUE,
-                 train = pima)
-
[…]

### 2017-07-20 slowtests/slowtests-boston.Rmd
ready for release

@@
 
-## Cubist
+# Cubist
 
-# `Cubist` package
+## `Cubist` package
 
@@
 
-## Decision trees
+# Decision trees
 
-# `rpart` package
+## `rpart` package
 
@@
 
-# `party` package
+## `party` package
 
@@
 
-# `partykit` package
+## `partykit` package
 
@@
 
-## Bagging
+# Bagging
 
-# `ipred` package
+## `ipred` package
 
@@
 
-## Random forests
+# Random forests
 
-# `randomForest` package
+## `randomForest` package
 
@@
 
-# `ranger` package
+## `ranger` package
 
@@
 
-## Boosting
+# Boosting
 
-# `gbm` package
+## `gbm` package
 
@@
 
-# `xgboost` package
+## `xgboost` package
 ```{r}
@@
 
-## Neural networks
+# Neural networks
 
-# `nnet` package
+## `nnet` package
 ```{r}
@@
 
-## Support vector machines
+# Support vector machines
 
-# `e1071` package
+## `e1071` package
 
@@
 
-# `kernlab` package
+## `kernlab` package
 
@@
 
-## Linear/generalized linear models
+# Linear/generalized linear models
 
-# `stats` package
+## `stats` package
 
@@
 
-## Multivariate adaptive regression splines
+# Multivariate adaptive regression splines
 
-# `earth` package
+## `earth` package
 
@@
 
-# `mda` package
+## `mda` package
 

### 2017-07-20 slowtests/slowtests-pima.Rmd
ready for release

@@
 set.seed(101)
-pima.gbm <- gbm::gbm(diabetes ~ ., 
+pima.gbm <- gbm::gbm(unclass(diabetes) - 1 ~ ., 
                      data = na.omit(pima),
-                     distribution = "multinomial",
+                     distribution = "bernoulli",
                      n.trees = 5000,
@@
                      shrinkage = 0.001,
-                     cv.folds = 5,
+                     # cv.folds = 5,
                      verbose = FALSE)
-best.iter <- gbm::gbm.perf(pima.gbm, method = "cv", plot.it = FALSE)
+best.iter <- gbm::gbm.perf(pima.gbm, method = "OOB", plot.it = FALSE)
 pimaIndiansTest(pima.gbm, n.trees = best.iter)

### 2017-07-20 R/autoplot.R

@@ -80,7 +80,7 @@
 #' @export
 #'
 #' @examples
-#'
+#' \dontrun{
 #' #
 #' # Regression example (requires randomForest package to run)
 #' #
@@ -101,7 +101,7 @@
 #'
 #' # Partial dependence of cmedv on lstat and rm
 #' boston.rf %>%
-#'   partial(pred.var = c("lstat", "rm"), chull = TRUE) %>%
+#'   partial(pred.var = c("lstat", "rm"), chull = TRUE, progress = "text") %>%
 #'   autoplot(contour = TRUE, legend.title = "rm")
 #'
 #' # ICE curves and c-ICE curves
@@ -111,6 +111,7 @@
 #'   autoplot(age.ice, center = TRUE, alpha = 0.5),  # c-ICE curves
 #'   ncol = 2
 #' )
+#' }
 autoplot.partial <- function(object, center = FALSE, plot.pdp = TRUE,
                              pdp.color = "red", pdp.size = 1, pdp.linetype = 1,
                              rug = FALSE, smooth = FALSE,

### 2017-07-20 R/partial.R

@@ -182,7 +182,6 @@
 #'
 #' @examples
 #' \dontrun{
-#'
 #' #
 #' # Regression example (requires randomForest package to run)
 #' #
@@ -236,9 +235,8 @@
 #'         progress = "text")
 #'
 #' # Partial dependence of positive test result on glucose (probability scale)
-#' partial(pima.rf, pred.var = "glucose", prob = TRUE, plot = TRUE, 
+#' partial(pima.rf, pred.var = "glucose", prob = TRUE, plot = TRUE,
 #'         chull = TRUE, progress = "text")
-#'
 #' }
 partial <- function(object, ...) {
   UseMethod("partial")
@@ -392,7 +390,7 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
     class(pd.df) <- c("data.frame", "partial")  # assign class labels
     names(pd.df) <- c(pred.var, "yhat")  # rename columns
     rownames(pd.df) <- NULL  # remove row names
-    
+
   } else {
 
     # Use brute force approach
@@ -438,7 +436,7 @@ partial.default <- function(object, pred.var, pred.grid, pred.fun = NULL,
 
       # Assign class labels
       class(pd.df) <- c("data.frame", "ice")
-      
+
       # c-ICE curves
       if (center) {
         pd.df <- centerIceCurves(pd.df)

### 2017-07-20 R/plotPartial.R

@@ -69,8 +69,36 @@
 #' @export
 #'
 #' @examples
-#' # See ?partial for examples
-#' ?partial
+#' \dontrun{
+#' #
+#' # Regression example (requires randomForest package to run)
+#' #
+#'
+#' # Load required packages
+#' library(ggplot2)  # required to use autoplot
+#' library(randomForest)
+#'
+#' # Fit a random forest to the Boston housing data
+#' data (boston)  # load the boston housing data
+#' set.seed(101)  # for reproducibility
+#' boston.rf <- randomForest(cmedv ~ ., data = boston)
+#'
+#' # Partial dependence of cmedv on lstat
+#' boston.rf %>%
+#'   partial(pred.var = "lstat") %>%
+#'   plotPartial(rug = TRUE, train = boston)
+#'
+#' # Partial dependence of cmedv on lstat and rm
+#' boston.rf %>%
+#'   partial(pred.var = c("lstat", "rm"), chull = TRUE, progress = "text") %>%
+#'   plotPartial(contour = TRUE, legend.title = "rm")
+#'
+#' # ICE curves and c-ICE curves
+#' age.ice <- partial(boston.rf, pred.var = "lstat", ice = TRUE)
+#' p1 <- plotPartial(age.ice, alpha = 0.5)
+#' p2 <- plotPartial(age.ice, center = TRUE, alpha = 0.5)
+#' grid.arrange(p1, p2, ncol = 2)
+#' }
 plotPartial <- function(x, ...) {
   UseMethod("plotPartial")
 }

### 2017-07-20 slowtests/slowtests-classification.R

@@ -1,231 +1,0 @@
-#-------------------------------------------------------------------------------
-# Slow tests for the pdp package
-#-------------------------------------------------------------------------------
-
-
-################################################################################
-# Setup
-################################################################################
-
-# Make sure all packges are up to date
-# update.packages()
-
-# Load required packages
-library(adabag)
-library(C50)
-library(earth)
-library(e1071)
-library(gbm)
-library(ipred)
-library(kernlab)
-library(nnet)
-library(party)
-# library(partykit)
-library(pdp)
-library(randomForest)
-library(ranger)
-library(rpart)
-library(xgboost)
-
-# Load data
-data(pima)
-pima <- na.omit(pima)
-
-
-################################################################################
-# Fit models
-################################################################################
-
-# adabag::bagging
-set.seed(101)
-pima.bagging <- bagging(diabetes ~ ., data = pima)
-
-# adabag::boosting
-set.seed(101)
-pima.boosting <- boosting(diabetes ~ ., data = pima)
-
-# C50::C5.0
-set.seed(101)
-pima.C5.0 <- C5.0(diabetes ~ ., data = pima, trials = 100)
-
-# e1071::svm
-pima.svm <- svm(diabetes ~ ., data = pima, type = "C-classification",
-                probability = TRUE)
-
-# earth
-pima.earth <- earth(diabetes ~ ., data = pima, degree = 2,
-                    glm = list(family = binomial))
-
-# gbm
-set.seed(101)
-pima.gbm <- gbm(unclass(diabetes) - 1 ~ .,
-                data = pima,
-                distribution = "bernoulli",
-                n.trees = 5000,
-                interaction.depth = 3,
-                shrinkage = 0.001,
-                verbose = TRUE)
-best.iter <- gbm.perf(pima.gbm, method = "OOB")
-
-# ipred::bagging
-set.seed(101)
-pima.ipred.bagging <- ipred::bagging(diabetes ~ ., data = pima, nbagg = 500)
-
-
-# kernlab::svm
-pima.ksvm <- ksvm(diabetes ~ ., data = pima, type = "C-svc", prob.model = TRUE)
-
-# # nnet
[…]
